---
title: "Group Breast Cancer EDA + Modelling (Merged Binary GEOs)"
format:
  html:
    embed-resources: true
    code-fold: true
    df-print: paged
    toc: true
editor: visual
---

This notebook builds a **merged binary dataset** from multiple GEO studies and compares multiple class-weighting schemes using leakage-safe leave-one-dataset-out evaluation.

```{r}
#| message: false
#| warning: false
library(GEOquery)
library(limma)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(cvTools)
library(sva)
```

```{r}
#| message: false
#| warning: false
set.seed(3888)
```

## Dataset registry + merge helpers

```{r}
#| message: false
#| warning: false
datasets_multiclass <- c("GSE25066","GSE45827")          # PAM50 multiclass (example; not merged yet)
datasets_binary <- c("GSE15852", "GSE45827","GSE29431","GSE9574", "GSE7904","GSE54002", "GSE25066")  # Binary Normal vs Cancer (merged below)

normal_keywords <- c(
  "normal", "non-tumor", "nontumor", "non tumour", "nontumour",
  "healthy", "control", "adjacent normal", "benign", "none"
)

# Preprocessing knobs
enable_auto_log2 <- TRUE
use_fold_downsampling <- TRUE
downsample_max_ratio <- 3
```

```{r}
#| message: false
#| warning: false
load_geo_eset <- function(gse_id) {
  geo_cache_dir <- file.path("data", "geo-cache")
  if (!dir.exists(geo_cache_dir)) dir.create(geo_cache_dir, recursive = TRUE)

  # Helper: build HTTPS URL for the series matrix
  gse_matrix_url <- function(id, host = "ftp.ncbi.nlm.nih.gov") {
    num <- sub("^GSE", "", id)
    prefix <- substr(num, 1, max(1, nchar(num) - 3))
    series_dir <- paste0("GSE", prefix, "nnn")
    paste0(
      "https://", host, "/geo/series/",
      series_dir, "/", id, "/matrix/",
      id, "_series_matrix.txt.gz"
    )
  }

  # Prefer normal GEOquery path first (uses destdir cache)
  g <- tryCatch(
    GEOquery::getGEO(gse_id, destdir = geo_cache_dir),
    error = function(e) e
  )
  if (!inherits(g, "error")) {
    if (inherits(g, "ExpressionSet")) return(g)
    if (is.list(g) && length(g) >= 1) return(g[[1]])
  }

  # Fallback: explicit HTTPS download with longer timeout + retries + resume, then load from file
  destfile <- file.path(geo_cache_dir, paste0(gse_id, "_series_matrix.txt.gz"))
  urls <- c(
    gse_matrix_url(gse_id, host = "ftp.ncbi.nlm.nih.gov"),
    gse_matrix_url(gse_id, host = "download.ncbi.nlm.nih.gov")
  )

  # R's options(timeout=) does not reliably control libcurl; set curl handle timeout directly.
  h <- curl::new_handle()
  curl::handle_setopt(h,
                      connecttimeout = 60,
                      timeout = 1800,
                      low_speed_time = 300,
                      low_speed_limit = 1000)

  ok <- FALSE
  for (u in urls) {
    for (attempt in 1:6) {
      try({
        size_before <- if (file.exists(destfile)) file.info(destfile)$size else 0
        curl::curl_download(u, destfile, quiet = FALSE, mode = "wb", handle = h, resume = TRUE)
        size_after <- if (file.exists(destfile)) file.info(destfile)$size else 0
        ok <- size_after > 0 && size_after >= size_before
      }, silent = TRUE)
      if (ok) break
      # If we are not making progress, reset the partial file and retry.
      if (file.exists(destfile) && attempt >= 2) file.remove(destfile)
      Sys.sleep(3 * attempt)
    }
    if (ok) break
  }

  if (!ok) {
    stop("Could not download series matrix for ", gse_id, " after retries.\nTried URLs:\n- ",
         paste(urls, collapse = "\n- "))
  }

  g2 <- GEOquery::getGEO(filename = destfile)
  if (inherits(g2, "ExpressionSet")) return(g2)
  if (is.list(g2) && length(g2) >= 1) return(g2[[1]])
  stop("Could not load ExpressionSet for ", gse_id, " from cached file: ", destfile)
}

clean_first_symbol <- function(s) {
  if (is.na(s) || trimws(as.character(s)) == "") return(NA_character_)
  s <- trimws(as.character(s))
  parts <- strsplit(s, " /// ", fixed = TRUE)[[1]]
  trimws(parts[1])
}

needs_log2_transform <- function(e) {
  vals <- as.numeric(e)
  vals <- vals[is.finite(vals)]
  if (length(vals) < 10) return(FALSE)
  qx <- stats::quantile(vals, probs = c(0, 0.25, 0.5, 0.75, 0.99, 1.0), na.rm = TRUE)
  # GEO2R-style heuristic for identifying non-log expression scales.
  (qx[5] > 100) || (qx[6] - qx[1] > 50 && qx[2] > 0) || (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
}

safe_log2_transform <- function(e) {
  pos <- e[e > 0 & is.finite(e)]
  if (!length(pos)) return(e)
  pseudo <- min(pos, na.rm = TRUE) / 2
  e2 <- e
  e2[!is.finite(e2)] <- NA_real_
  e2[e2 <= 0] <- pseudo
  log2(e2)
}

collapse_to_genes <- function(eset) {
  e <- exprs(eset)
  q_raw <- stats::quantile(as.numeric(e), probs = c(0.01, 0.5, 0.99), na.rm = TRUE)
  log2_applied <- isTRUE(enable_auto_log2) && needs_log2_transform(e)
  if (log2_applied) {
    e <- safe_log2_transform(e)
  }
  q_used <- stats::quantile(as.numeric(e), probs = c(0.01, 0.5, 0.99), na.rm = TRUE)
  fd <- fData(eset)
  sym_col <- grep("gene\\s*symbol|symbol", colnames(fd), ignore.case = TRUE, value = TRUE)[1]
  gene_sym_raw <- if (!is.na(sym_col)) as.character(fd[, sym_col]) else rep(NA_character_, nrow(fd))
has_symbol <- !is.na(gene_sym_raw) & trimws(gene_sym_raw) != ""
gene_sym <- rep(NA_character_, length(gene_sym_raw))
gene_sym[has_symbol] <- vapply(gene_sym_raw[has_symbol], clean_first_symbol, character(1))
  gene_sym <- ifelse(is.na(gene_sym) | gene_sym == "", rownames(e), gene_sym)
  e_g <- limma::avereps(e, ID = gene_sym)

  # Feature table: keep one row per gene symbol where possible, otherwise minimal table
  fd$.__gene_sym__ <- gene_sym
  fd_g <- fd %>%
    tibble::as_tibble() %>%
    mutate(.__gene_sym__ = fd$.__gene_sym__) %>%
    group_by(.__gene_sym__) %>%
    slice(1) %>%
    ungroup() %>%
    tibble::column_to_rownames(".__gene_sym__")

  list(
    eMat_g = e_g,
    fMat_g = fd_g,
    pMat = pData(eset),
    log2_applied = log2_applied,
    q01_raw = as.numeric(q_raw[1]),
    q50_raw = as.numeric(q_raw[2]),
    q99_raw = as.numeric(q_raw[3]),
    q01_used = as.numeric(q_used[1]),
    q50_used = as.numeric(q_used[2]),
    q99_used = as.numeric(q_used[3])
  )
}

infer_binary_label <- function(pMat, normal_keywords) {
  # Robust text parsing avoids accidental matches like "abnormal"/"normalized".
  normalize_pheno_txt <- function(x) {
    x <- as.character(x)
    x[is.na(x)] <- ""
    x <- gsub("\u00a0|\u200b", " ", x, fixed = TRUE)
    x <- trimws(gsub("[[:space:]]+", " ", x))
    tolower(x)
  }

  flatten_pheno_col <- function(x) {
    if (is.list(x) && !inherits(x, "data.frame")) {
      vapply(x, function(z) paste(as.character(z), collapse = " "), character(1))
    } else {
      as.character(x)
    }
  }

  rx_escape <- function(s) gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", s, perl = TRUE)

  keyword_hit <- function(xlow, keywords) {
    keywords <- unique(trimws(as.character(keywords)))
    keywords <- keywords[nzchar(keywords)]
    if (!length(keywords)) return(rep(FALSE, length(xlow)))
    parts <- vapply(keywords, function(kw) {
      paste0("(?<![[:alnum:]])", rx_escape(kw), "(?![[:alnum:]])")
    }, character(1))
    grepl(paste0("(?i)(?:", paste(parts, collapse = "|"), ")"), xlow, perl = TRUE)
  }

  explicit_normal_rx <- paste0(
    "(?i)(",
    ",\\s*normal\\b|",
    "\\bbreast\\s+biopsy[, ]+normal\\b|",
    "\\bhuman\\s+breast\\s+biopsy[, ]+normal\\b|",
    "\\bnormal\\s+breast(\\s+tissue)?\\b|",
    "tissue:\\s*normal\\b|",
    "histopathological\\s+exam:\\s*normal\\b",
    ")"
  )

  tumor_hint_rx <- "(?i)(tumor|cancer|carcinoma|malign|invasive|ductal|lobular|metasta)"

  cols_try <- local({
    nm <- colnames(pMat)
    pr <- nm[grepl(
      "source_name|sample_title|\\btitle\\b|histology|pathology|disease|diagnosis|tumor|sample_type|biopsy|specimen",
      nm,
      ignore.case = TRUE
    )]
    ch <- grep("^characteristics_ch", nm, value = TRUE)
    rest <- setdiff(nm, c(pr, ch))
    unique(c(intersect(pr, nm), ch, rest))
  })

  col_has_signal <- function(xlow) {
    nzchar(xlow) &
      (keyword_hit(xlow, normal_keywords) |
         grepl(explicit_normal_rx, xlow, perl = TRUE) |
         grepl(tumor_hint_rx, xlow, perl = TRUE))
  }

  best <- NULL
  best_key <- NULL

  for (idx in seq_along(cols_try)) {
    cn <- cols_try[[idx]]
    x <- flatten_pheno_col(pMat[[cn]])
    xlow <- normalize_pheno_txt(x)
    if (!any(col_has_signal(xlow))) next

    is_normal <- keyword_hit(xlow, normal_keywords) | grepl(explicit_normal_rx, xlow, perl = TRUE)
    label <- ifelse(is_normal, "Normal", "Cancer")
    label[!nzchar(xlow)] <- NA_character_
    label <- factor(label, levels = c("Cancer", "Normal"))

    tab <- table(label, useNA = "no")
    if (!all(c("Cancer", "Normal") %in% names(tab))) next
    if (is.na(tab[["Cancer"]]) || is.na(tab[["Normal"]])) next
    if (tab[["Cancer"]] == 0 || tab[["Normal"]] == 0) next

    min_cnt <- min(as.integer(tab[["Cancer"]]), as.integer(tab[["Normal"]]))
    n_nonempty <- sum(nzchar(xlow) & !is.na(label))
    key <- c(min_cnt, n_nonempty, -idx)
    if (is.null(best_key) || key[1] > best_key[1] ||
        (key[1] == best_key[1] && key[2] > best_key[2]) ||
        (key[1] == best_key[1] && key[2] == best_key[2] && key[3] > best_key[3])) {
      best_key <- key
      best <- list(label = label, label_col = cn)
    }
  }

  if (!is.null(best)) return(best)
  list(label = factor(rep(NA_character_, nrow(pMat)), levels = c("Cancer", "Normal")), label_col = NA_character_)
}

extract_after_scalar <- function(x, prefix) {
  # scalar-safe extractor for strings like "pam50_class: Basal"
  if (length(x) > 1) x <- x[which(!is.na(x))[1]]
  if (length(x) == 0 || is.na(x)) return(NA_character_)
  x <- as.character(x)
  m <- regexpr(prefix, x, ignore.case = TRUE)
  if (m[1] == -1) return(NA_character_)
  out <- substring(x, m[1] + attr(m, "match.length"))
  trimws(out)
}

infer_multiclass_label <- function(pMat) {
  cols_char <- grep("^characteristics_ch", colnames(pMat), value = TRUE)
  cols_other <- setdiff(colnames(pMat), cols_char)
  cols_try <- c(cols_char, cols_other)

  # Prefer columns that look like PAM50
  cols_try <- c(
    intersect(cols_try, grep("pam50", cols_try, ignore.case = TRUE, value = TRUE)),
    setdiff(cols_try, grep("pam50", cols_try, ignore.case = TRUE, value = TRUE))
  )

  # Canonical subtype dictionary (handles common naming variations across datasets)
  # You can extend this list if you see new spellings.
  subtype_variants <- list(
    "Luminal A"    = c("luminal a", "lum a", "luma", "lum-a", "lum_a", "luminala", "lumina"),
    "Luminal B"    = c("luminal b", "lum b", "lumb", "lum-b", "lum_b", "luminalb"),
    "Basal"        = c("basal", "basal-like", "basal like", "basallike", "basal_like", "basal type", "basaltype", "basal/tumor"),
    "HER2"         = c("her2", "her2+", "her-2", "her 2", "her2-enriched", "her2 enriched", "her2enriched", "erb-b2", "erbb2", "erbb2+", "her2e"),
    "Normal-like"  = c("normal-like", "normal like", "normallike", "normal_like", "normal")
  )

  normalize_label_text <- function(x) {
    x <- tolower(as.character(x))
    x[is.na(x)] <- ""
    # Light normalization: trim and collapse non-alphanumeric into single spaces
    x <- trimws(gsub("[^a-z0-9]+", " ", x))
    x
  }

  map_to_subtype <- function(x_raw) {
    x_chr <- as.character(x_raw)
    x_norm <- normalize_label_text(x_chr)
    if (x_norm == "") return(NA_character_)

    # If there is a PAM50 tag embedded in the string, parse after it first
    if (grepl("pam50_class", tolower(x_chr), fixed = TRUE)) {
      x_parsed <- extract_after_scalar(x_chr, prefix = "pam50_class:\\s*")
      x_norm <- normalize_label_text(x_parsed)
    }

    # Match against variants (substring match on normalized text)
    for (canon in names(subtype_variants)) {
      v <- subtype_variants[[canon]]
      if (any(vapply(v, function(kw) grepl(kw, x_norm, fixed = TRUE), logical(1)))) {
        return(canon)
      }
    }
    NA_character_
  }

  for (cn in cols_try) {
    x <- as.character(pMat[[cn]])
    lab <- vapply(x, map_to_subtype, character(1))
    y <- factor(lab)
    y <- droplevels(y[!is.na(y)])
    # Accept if 2+ levels and not degenerate
    if (nlevels(y) >= 2) {
      return(list(label = factor(lab), label_col = cn, parsed_pam50 = any(grepl("pam50_class:", x, ignore.case = TRUE))))
    }
  }

  list(label = factor(rep(NA_character_, nrow(pMat))), label_col = NA_character_, parsed_pam50 = FALSE)
}

rbind_fill <- function(dfs) {
  all_cols <- Reduce(intersect, lapply(dfs, colnames))
  dfs2 <- lapply(dfs, function(d) {
    miss <- setdiff(all_cols, colnames(d))
    for (m in miss) d[[m]] <- NA
    d[, all_cols, drop = FALSE]
  })
  do.call(rbind, dfs2)
}

# Per-fold batch correction with a training-only fit.
combat_fit_apply <- function(expr_train, batch_train, mod_train,
                             expr_test, batch_test) {
  if (length(unique(as.character(batch_train))) < 2) {
    message("Single training batch in this LODO fold; skipping batch correction.")
    return(list(train = expr_train, test = expr_test))
  }

  train_batches <- unique(as.character(batch_train))
  test_batches <- unique(as.character(batch_test))
  common_batches <- intersect(train_batches, test_batches)
  novel_test_idx <- which(!(as.character(batch_test) %in% common_batches))

  corrected_train <- tryCatch(
    sva::ComBat(
      dat = expr_train,
      batch = batch_train,
      mod = mod_train,
      par.prior = TRUE,
      mean.only = FALSE
    ),
    error = function(e) {
      message("ComBat failed on training fold: ", conditionMessage(e),
              "\n  Falling back to limma::removeBatchEffect for this fold.")
      limma::removeBatchEffect(expr_train, batch = batch_train, design = mod_train)
    }
  )

  corrected_test <- tryCatch({
    batch_train_f <- factor(as.character(batch_train))
    batch_test_f <- factor(as.character(batch_test), levels = levels(batch_train_f))

    batch_means_train <- sapply(levels(batch_train_f), function(b) {
      idx <- which(batch_train_f == b)
      if (length(idx) == 0) return(rep(0, nrow(expr_train)))
      rowMeans(expr_train[, idx, drop = FALSE])
    })

    grand_mean_train <- rowMeans(expr_train)
    expr_test_corrected <- expr_test

    for (b in levels(batch_train_f)) {
      test_in_batch <- which(as.character(batch_test_f) == b)
      if (length(test_in_batch) == 0) next
      batch_effect <- batch_means_train[, b] - grand_mean_train
      expr_test_corrected[, test_in_batch] <-
        expr_test[, test_in_batch, drop = FALSE] - batch_effect
    }

    if (length(novel_test_idx) > 0) {
      message("  Test fold contains ", length(novel_test_idx),
              " samples from unseen batches; applying mean-centering only.")
      sample_means <- colMeans(expr_test[, novel_test_idx, drop = FALSE])
      expr_test_corrected[, novel_test_idx] <-
        sweep(expr_test[, novel_test_idx, drop = FALSE], 2, sample_means, "-") +
        grand_mean_train
    }

    expr_test_corrected
  }, error = function(e) {
    message("Test-fold batch correction failed: ", conditionMessage(e),
            "\n  Returning uncorrected test data for this fold.")
    expr_test
  })

  list(train = corrected_train, test = corrected_test)
}

# Missing-value handling for merged expression matrices (use training-fold stats only)
impute_fit <- function(X_train) {
  meds <- apply(X_train, 2, function(v) stats::median(v, na.rm = TRUE))
  meds[is.na(meds)] <- 0
  list(median = meds)
}
impute_apply <- function(X, fit) {
  X2 <- X
  for (j in seq_len(ncol(X2))) {
    idx <- is.na(X2[, j])
    if (any(idx)) X2[idx, j] <- fit$median[j]
  }
  X2
}

# Standardisation for distance-based / margin-based models (fit on training only)
standardize_fit <- function(X_train) {
  mu <- colMeans(X_train)
  sdv <- apply(X_train, 2, sd)
  sdv[sdv == 0] <- 1
  list(mu = mu, sd = sdv)
}
standardize_apply <- function(X, fit) {
  sweep(sweep(X, 2, fit$mu, "-"), 2, fit$sd, "/")
}
```

```{r}
#| message: false
#| warning: false
build_merged_binary_dataset <- function(gse_ids, normal_keywords) {
  pieces <- lapply(gse_ids, function(id) {
    eset <- load_geo_eset(id)
    cc <- collapse_to_genes(eset)
    lab <- infer_binary_label(cc$pMat, normal_keywords)

    p <- cc$pMat
    p$dataset_id <- id
    p$label_col_detected <- lab$label_col
    p$label_binary <- as.character(lab$label)

    list(
      id = id,
      e = cc$eMat_g,
      p = p,
      f = cc$fMat_g,
      log2_applied = cc$log2_applied,
      q01_raw = cc$q01_raw, q50_raw = cc$q50_raw, q99_raw = cc$q99_raw,
      q01_used = cc$q01_used, q50_used = cc$q50_used, q99_used = cc$q99_used
    )
  })

  # Merge expression by INTERSECTION of genes so no dataset-specific padding is needed.
  all_genes <- Reduce(intersect, lapply(pieces, function(x) rownames(x$e)))
  if (length(all_genes) == 0) stop("No genes found across binary datasets.")

  e_big <- do.call(cbind, lapply(pieces, function(x) x$e[all_genes, , drop = FALSE]))

  # Merge phenotype by union of columns
  p_big <- rbind_fill(lapply(pieces, `[[`, "p"))

  # Merge feature data by union of gene rows (keep only genes present in e_big)
  f_big_all <- rbind_fill(lapply(pieces, function(x) {
    ff <- x$f
    ff$dataset_id <- x$id
    tibble::rownames_to_column(ff, var = "gene_id")
  }))
  # Keep one row per gene_id (prefer the first dataset that has it)
  f_big <- f_big_all %>%
    group_by(gene_id) %>%
    slice(1) %>%
    ungroup() %>%
    filter(gene_id %in% rownames(e_big)) %>%
    tibble::column_to_rownames("gene_id")

  list(eMat = e_big, pMat = p_big, fMat = f_big, all_genes = all_genes, pieces = pieces)
}

build_merged_multiclass_dataset <- function(gse_ids) {
  pieces <- lapply(gse_ids, function(id) {
    eset <- load_geo_eset(id)
    cc <- collapse_to_genes(eset)
    lab <- infer_multiclass_label(cc$pMat)

    p <- cc$pMat
    p$dataset_id <- id
    p$label_col_detected <- lab$label_col
    p$label_multiclass <- as.character(lab$label)

    list(id = id, e = cc$eMat_g, p = p, f = cc$fMat_g)
  })

  all_genes <- Reduce(intersect, lapply(pieces, function(x) rownames(x$e)))
  if (length(all_genes) == 0) stop("No genes found across multiclass datasets.")

  e_big <- do.call(cbind, lapply(pieces, function(x) x$e[all_genes, , drop = FALSE]))
  p_big <- rbind_fill(lapply(pieces, `[[`, "p"))

  f_big_all <- rbind_fill(lapply(pieces, function(x) {
    ff <- x$f
    ff$dataset_id <- x$id
    tibble::rownames_to_column(ff, var = "gene_id")
  }))
  f_big <- f_big_all %>%
    group_by(gene_id) %>%
    slice(1) %>%
    ungroup() %>%
    filter(gene_id %in% rownames(e_big)) %>%
    tibble::column_to_rownames("gene_id")

  list(eMat = e_big, pMat = p_big, fMat = f_big, all_genes = all_genes, pieces = pieces)
}

merged_bin <- build_merged_binary_dataset(datasets_binary, normal_keywords)
eMat_big_bin <- merged_bin$eMat
pMat_big_bin <- merged_bin$pMat
fMat_big_bin <- merged_bin$fMat

na_fraction_bin <- rowMeans(is.na(eMat_big_bin))
eMat_big_bin <- eMat_big_bin[na_fraction_bin < 0.3, , drop = FALSE]

cat("Merged binary expression dims (genes x samples):", paste(dim(eMat_big_bin), collapse=" x "), "\n")
print(table(pMat_big_bin$dataset_id))
print(table(pMat_big_bin$label_binary, useNA = "ifany"))
```

```{r}
#| message: false
#| warning: false
merged_mc <- build_merged_multiclass_dataset(datasets_multiclass)
eMat_big_mc <- merged_mc$eMat
pMat_big_mc <- merged_mc$pMat
fMat_big_mc <- merged_mc$fMat

na_fraction_mc <- rowMeans(is.na(eMat_big_mc))
eMat_big_mc <- eMat_big_mc[na_fraction_mc < 0.3, , drop = FALSE]

cat("Merged multiclass expression dims (genes x samples):", paste(dim(eMat_big_mc), collapse=" x "), "\n")
print(table(pMat_big_mc$dataset_id))
print(table(pMat_big_mc$label_multiclass, useNA = "ifany"))
```

## Multiclass modelling (merged dataset)

```{r}
#| message: false
#| warning: false
macro_f1 <- function(truth, pred) {
  truth <- factor(truth)
  pred <- factor(pred, levels = levels(truth))
  lvls <- levels(truth)
  f1s <- c()
  for (k in lvls) {
    tp <- sum(pred == k & truth == k, na.rm = TRUE)
    fp <- sum(pred == k & truth != k, na.rm = TRUE)
    fn <- sum(pred != k & truth == k, na.rm = TRUE)
    prec <- if ((tp + fp) == 0) NA_real_ else tp / (tp + fp)
    rec  <- if ((tp + fn) == 0) NA_real_ else tp / (tp + fn)
    f1   <- if (is.na(prec) || is.na(rec) || (prec + rec) == 0) NA_real_ else 2 * prec * rec / (prec + rec)
    f1s <- c(f1s, f1)
  }
  mean(f1s, na.rm = TRUE)
}

select_genes_one_vs_rest <- function(eMat_g, y, n_per_class = 25) {
  ok <- !is.na(y) & y != ""
  e <- eMat_g[, ok, drop = FALSE]
  y <- factor(y[ok])

  lvls_orig <- levels(y)
  lvls_safe <- make.names(lvls_orig)
  names(lvls_safe) <- lvls_orig

  design <- model.matrix(~ 0 + y)
  colnames(design) <- lvls_safe[lvls_orig]
  fit <- lmFit(e, design)

  genes <- character(0)
  for (lvl in lvls_orig) {
    others <- setdiff(lvls_orig, lvl)
    if (length(others) == 0) next
    lvl_s <- unname(lvls_safe[lvl])
    others_s <- unname(lvls_safe[others])
    contrast_str <- paste0(lvl_s, " - (", paste0(others_s, collapse = " + "), ")/", length(others_s))
    cont <- makeContrasts(contrasts = contrast_str, levels = design)
    fit2 <- contrasts.fit(fit, cont)
    fit2 <- eBayes(fit2)
    tt <- topTable(fit2, n = n_per_class, sort.by = "P")
    genes <- union(genes, rownames(tt))
  }
  genes
}

ok_mc <- !is.na(pMat_big_mc$label_multiclass) & pMat_big_mc$label_multiclass != ""
y_mc <- factor(pMat_big_mc$label_multiclass[ok_mc])
e_mc <- eMat_big_mc[, ok_mc, drop = FALSE]
batch_mc <- pMat_big_mc$dataset_id[ok_mc]

cat("Multiclass class counts (merged):\n")
print(table(y_mc))

genes_mc_vis <- select_genes_one_vs_rest(e_mc, y = y_mc, n_per_class = 25)
cat("Selected genes for multiclass visualization:", length(genes_mc_vis), "\n")

X_mc <- t(e_mc)
mc_vis_genes <- intersect(genes_mc_vis, rownames(e_mc))
X_mc_vis <- if (length(mc_vis_genes) > 0) {
  t(e_mc[mc_vis_genes, , drop = FALSE])
} else {
  X_mc
}
```

## PCA (multiclass, merged)

```{r, fig.cap="PCA of merged multiclass expression (after imputation + scaling). Points coloured by PAM50 subtype."}
#| message: false
#| warning: false
# PCA is for visualisation only; we impute/scale using the exploratory gene subset when available.
imp_mc_all <- impute_fit(X_mc_vis)
X_mc_i_all <- impute_apply(X_mc_vis, imp_mc_all)
sc_mc_all <- standardize_fit(X_mc_i_all)
X_mc_s_all <- standardize_apply(X_mc_i_all, sc_mc_all)

pca_mc <- prcomp(X_mc_s_all, center = FALSE, scale. = FALSE)
pca_mc_df <- data.frame(
  PC1 = pca_mc$x[, 1],
  PC2 = pca_mc$x[, 2],
  truth = y_mc,
  dataset_id = pMat_big_mc$dataset_id[ok_mc]
)

ggplot(pca_mc_df, aes(x = PC1, y = PC2, color = truth)) +
  geom_point(alpha = 0.75, size = 2) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Subtype", title = "Multiclass PCA (merged)") +
  theme(legend.position = "right")
```

```{r}
#| message: false
#| warning: false
run_lodo_cv_multiclass <- function(X, y, batch) {
  y <- factor(y)
  batch <- as.character(batch)
  perf <- data.frame()
  all_preds <- data.frame()
  has_multinom <- requireNamespace("nnet", quietly = TRUE)

  for (ds in unique(batch)) {
    test_id <- which(batch == ds)
    train_id <- setdiff(seq_along(batch), test_id)

    y_train <- droplevels(y[train_id])
    y_test <- droplevels(y[test_id])
    if (nlevels(y_train) < 2) next
    if (length(unique(y_test)) < 2) next

    if (!all(levels(y_test) %in% levels(y_train))) {
      y_test <- factor(as.character(y_test), levels = levels(y_train))
    }

    expr_train_raw <- t(X[train_id, , drop = FALSE])
    expr_test_raw <- t(X[test_id, , drop = FALSE])
    batch_train <- batch[train_id]
    batch_test <- batch[test_id]
    mod_train <- model.matrix(~ y_train)

    cb <- combat_fit_apply(
      expr_train = expr_train_raw,
      batch_train = batch_train,
      mod_train = mod_train,
      expr_test = expr_test_raw,
      batch_test = batch_test
    )

    X_train_full <- t(cb$train)
    X_test_full <- t(cb$test)

    genes_sel <- select_genes_one_vs_rest(t(X_train_full), y = y_train, n_per_class = 25)
    sel_genes <- intersect(genes_sel, colnames(X_train_full))
    if (length(sel_genes) < 10) next

    X_train <- X_train_full[, sel_genes, drop = FALSE]
    X_test <- X_test_full[, sel_genes, drop = FALSE]

    imp <- impute_fit(X_train)
    X_train_i <- impute_apply(X_train, imp)
    X_test_i <- impute_apply(X_test, imp)

    sc <- standardize_fit(X_train_i)
    X_train_s <- standardize_apply(X_train_i, sc)
    X_test_s <- standardize_apply(X_test_i, sc)

    y_test_chr <- as.character(y_test)

    pred_knn <- class::knn(train = X_train_s, test = X_test_s, cl = y_train, k = 7)
    pred_knn_chr <- as.character(pred_knn)
    perf <- rbind(perf, data.frame(task = "multiclass", model = "kNN", fold = ds,
                                   accuracy = mean(pred_knn_chr == y_test_chr),
                                   macro_f1 = macro_f1(y_test_chr, pred_knn_chr)))
    all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "kNN", fold = ds,
                                             truth = y_test_chr, pred = pred_knn_chr))

    svm_fit <- e1071::svm(x = X_train_s, y = y_train, kernel = "radial", cost = 1, gamma = 1 / ncol(X_train_s))
    pred_svm <- predict(svm_fit, X_test_s)
    pred_svm_chr <- as.character(pred_svm)
    perf <- rbind(perf, data.frame(task = "multiclass", model = "SVM_RBF", fold = ds,
                                   accuracy = mean(pred_svm_chr == y_test_chr),
                                   macro_f1 = macro_f1(y_test_chr, pred_svm_chr)))
    all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "SVM_RBF", fold = ds,
                                             truth = y_test_chr, pred = pred_svm_chr))

    rf_fit <- randomForest::randomForest(x = X_train_i, y = y_train)
    pred_rf <- predict(rf_fit, X_test_i)
    pred_rf_chr <- as.character(pred_rf)
    perf <- rbind(perf, data.frame(task = "multiclass", model = "RF", fold = ds,
                                   accuracy = mean(pred_rf_chr == y_test_chr),
                                   macro_f1 = macro_f1(y_test_chr, pred_rf_chr)))
    all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "RF", fold = ds,
                                             truth = y_test_chr, pred = pred_rf_chr))

    if (has_multinom) {
      mn_fit <- nnet::multinom(y_train ~ ., data = data.frame(y_train = y_train, X_train_s),
                               trace = FALSE, MaxNWts = 50000)
      pred_mn <- predict(mn_fit, newdata = data.frame(X_test_s))
      pred_mn_chr <- as.character(pred_mn)
      perf <- rbind(perf, data.frame(task = "multiclass", model = "Multinom", fold = ds,
                                     accuracy = mean(pred_mn_chr == y_test_chr),
                                     macro_f1 = macro_f1(y_test_chr, pred_mn_chr)))
      all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "Multinom", fold = ds,
                                               truth = y_test_chr, pred = pred_mn_chr))
    }
  }

  list(perf = perf, preds = all_preds)
}

res_mc <- run_lodo_cv_multiclass(X_mc, y_mc, batch_mc)
```

```{r, fig.cap="Multiclass (merged): accuracy and macro-F1 across models (leave-one-dataset-out CV)."}
#| message: false
#| warning: false
perf_mc <- res_mc$perf %>%
  pivot_longer(cols = c(accuracy, macro_f1), names_to = "metric", values_to = "value")

ggplot(perf_mc, aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.6) +
  facet_wrap(~metric, scales = "free_y") +
  theme_bw() +
  labs(x = "", y = "", title = "Multiclass (merged) model comparison") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r, fig.cap="Multiclass (merged): confusion matrix for best model (by mean macro-F1)."}
#| message: false
#| warning: false
if (!is.data.frame(perf_mc) || nrow(perf_mc) == 0 || !all(c("model", "metric", "value") %in% names(perf_mc))) {
  best_model_mc <- NA_character_
} else {
  best_mc_tbl <- perf_mc %>%
    filter(metric == "macro_f1") %>%
    group_by(model) %>%
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean))

  best_model_mc <- if (nrow(best_mc_tbl) == 0) NA_character_ else as.character(best_mc_tbl$model[[1]])
}

cat("Best multiclass model by mean macro-F1:", best_model_mc, "\n")

preds_best_mc <- res_mc$preds %>% filter(model == best_model_mc)
tab_mc <- table(truth = preds_best_mc$truth, pred = preds_best_mc$pred)
tab_mc_df <- as.data.frame(tab_mc)

ggplot(tab_mc_df, aes(x = pred, y = truth, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#2C7FB8") +
  theme_bw() +
  labs(x = "Predicted", y = "True", title = paste0("Multiclass confusion matrix: ", best_model_mc)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Best model separation (multiclass)

```{r, fig.cap="Best multiclass model: PCA view with refit predictions on the full merged dataset (visualisation only)."}
#| message: false
#| warning: false
fit_predict_best_multiclass <- function(best_model, X, y) {
  y <- factor(y)
  imp <- impute_fit(X)
  X_i <- impute_apply(X, imp)

  if (best_model %in% c("kNN", "SVM_RBF", "Multinom")) {
    sc <- standardize_fit(X_i)
    X_s <- standardize_apply(X_i, sc)
  } else {
    X_s <- X_i
  }

  if (best_model == "kNN") {
    pred <- class::knn(train = X_s, test = X_s, cl = y, k = 7)
    return(list(pred = pred))
  }
  if (best_model == "SVM_RBF") {
    fit <- e1071::svm(x = X_s, y = y, kernel = "radial", cost = 1, gamma = 1 / ncol(X_s))
    pred <- predict(fit, X_s)
    return(list(pred = pred))
  }
  if (best_model == "RF") {
    fit <- randomForest::randomForest(x = X_s, y = y)
    pred <- predict(fit, X_s)
    return(list(pred = pred))
  }
  if (best_model == "Multinom") {
    if (!requireNamespace("nnet", quietly = TRUE)) stop("Package 'nnet' not installed; can't fit Multinom.")
    fit <- nnet::multinom(y ~ ., data = data.frame(y = y, X_s), trace = FALSE, MaxNWts = 50000)
    pred <- predict(fit, newdata = data.frame(X_s))
    return(list(pred = pred))
  }
  stop("Unknown multiclass model name: ", best_model)
}

pred_full_mc <- fit_predict_best_multiclass(best_model_mc, X_mc, y_mc)$pred

pca_mc_pred_df <- pca_mc_df %>%
  mutate(pred = pred_full_mc)

ggplot(pca_mc_pred_df, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = truth, shape = pred), alpha = 0.75, size = 2) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "True", shape = "Pred (refit)", title = paste0("Best multiclass model separation: ", best_model_mc)) +
  theme(legend.position = "right")
```

## Binary DE gene selection (merged dataset)

```{r}
#| message: false
#| warning: false
# Binary DE gene selection on the MERGED binary dataset
ok_bin_de <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""
y_bin_de <- factor(pMat_big_bin$label_binary[ok_bin_de], levels = c("Cancer", "Normal"))
e_bin_de <- eMat_big_bin[, ok_bin_de, drop = FALSE]

tab_bin_de <- table(y_bin_de)
cat("Binary class table (merged):\n"); print(tab_bin_de)
if (nlevels(y_bin_de) < 2 || any(tab_bin_de == 0)) {
  stop("Merged binary labels must contain at least two classes with non-zero counts. Check label inference per dataset.")
}

if (nrow(e_bin_de) == 0 || ncol(e_bin_de) == 0 || all(is.na(e_bin_de))) {
  stop("Merged binary expression matrix is empty or all NA after filtering; cannot run DE selection.")
}

design_bin <- model.matrix(~ y_bin_de)
fit_bin <- tryCatch(
  eBayes(lmFit(e_bin_de, design_bin)),
  error = function(e) {
    stop("Binary DE selection failed: ", conditionMessage(e))
  }
)
top_bin <- topTable(fit_bin, n = 50, sort.by = "P")
genes_bin <- rownames(top_bin)
cat("Selected genes for binary (merged):", length(genes_bin), "\n")
top_bin
```

## Model zoo + LODO CV (binary: merged datasets)

```{r}
#| message: false
#| warning: false
auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)
  pos <- truth_pos == 1
  neg <- truth_pos == 0
  n_pos <- sum(pos)
  n_neg <- sum(neg)
  if (n_pos == 0 || n_neg == 0) return(NA_real_)
  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

pr_curve_df <- function(truth_pos, score_pos) {
  ord <- order(score_pos, decreasing = TRUE)
  y <- as.integer(truth_pos[ord])
  s <- score_pos[ord]
  tp <- cumsum(y == 1)
  fp <- cumsum(y == 0)
  fn_total <- sum(y == 1)
  precision <- tp / pmax(tp + fp, 1)
  recall <- tp / pmax(fn_total, 1)
  tibble(threshold = s, precision = precision, recall = recall)
}

auc_pr <- function(truth_pos, score_pos) {
  df <- pr_curve_df(truth_pos, score_pos)
  if (nrow(df) < 2) return(NA_real_)
  o <- order(df$recall, df$precision)
  x <- df$recall[o]
  y <- df$precision[o]
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2, na.rm = TRUE)
}

binary_metrics <- function(truth, pred, positive = "Normal") {
  keep <- !(is.na(truth) | is.na(pred))
  truth <- factor(as.character(truth[keep]))
  pred <- factor(as.character(pred[keep]), levels = levels(truth))
  if (length(truth) == 0) {
    return(tibble(
      accuracy = NA_real_,
      balanced_accuracy = NA_real_,
      recall_normal = NA_real_,
      precision_normal = NA_real_,
      f1_normal = NA_real_
    ))
  }
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)

  recall_pos <- if ((tp + fn) == 0) NA_real_ else tp / (tp + fn)
  precision_pos <- if ((tp + fp) == 0) NA_real_ else tp / (tp + fp)
  f1_pos <- if (is.na(recall_pos) || is.na(precision_pos) || (recall_pos + precision_pos) == 0) NA_real_ else {
    2 * recall_pos * precision_pos / (recall_pos + precision_pos)
  }
  specificity <- if ((tn + fp) == 0) NA_real_ else tn / (tn + fp)
  bal_acc <- mean(c(recall_pos, specificity), na.rm = TRUE)
  acc <- mean(truth == pred)

  tibble(
    accuracy = acc,
    balanced_accuracy = bal_acc,
    recall_normal = recall_pos,
    precision_normal = precision_pos,
    f1_normal = f1_pos
  )
}

best_threshold_f1 <- function(truth, score_pos, positive = "Normal") {
  truth <- as.character(truth)
  neg <- setdiff(unique(na.omit(truth)), positive)[1]
  if (is.na(neg) || length(neg) == 0) return(0.5)
  cand <- sort(unique(score_pos))
  if (length(cand) > 200) cand <- quantile(cand, probs = seq(0, 1, length.out = 200), na.rm = TRUE)
  f1 <- sapply(cand, function(t) {
    pred <- ifelse(score_pos >= t, positive, neg)
    m <- binary_metrics(truth, pred, positive = positive)
    m$f1_normal
  })
  cand[which.max(f1)]
}
```

```{r}
#| message: false
#| warning: false
# Use merged binary dataset (across `datasets_binary`)
ok_bin <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""
y_bin2 <- factor(pMat_big_bin$label_binary[ok_bin], levels = c("Cancer", "Normal"))
batch_bin <- pMat_big_bin$dataset_id[ok_bin]

X_bin <- t(eMat_big_bin[, ok_bin, drop = FALSE])
bin_vis_genes <- intersect(genes_bin, rownames(eMat_big_bin))
X_bin_vis <- if (length(bin_vis_genes) > 0) {
  t(eMat_big_bin[bin_vis_genes, ok_bin, drop = FALSE])
} else {
  X_bin
}

cat("Binary X dims (samples x genes):", paste(dim(X_bin), collapse=" x "), "\n")
cat("Binary class counts:\n")
print(table(y_bin2))

run_lodo_cv_binary <- function(X, y, batch, positive = "Normal", use_weights = TRUE,
                               apply_batch_correction = TRUE,
                               weight_scheme = c("ratio", "balanced", "balanced_capped"),
                               weight_cap = 4,
                               use_downsampling = TRUE,
                               downsample_ratio_cap = 3) {
  y <- factor(y)
  batch <- as.character(batch)
  weight_scheme <- match.arg(weight_scheme)
  setting_label <- if (!isTRUE(use_weights)) {
    "unweighted"
  } else if (weight_scheme == "ratio") {
    "weighted_ratio"
  } else if (weight_scheme == "balanced") {
    "weighted_balanced"
  } else {
    paste0("weighted_balanced_cap", format(weight_cap, trim = TRUE, scientific = FALSE))
  }
  perf <- data.frame()
  all_preds <- data.frame()
  all_scores <- data.frame()
  fold_diag <- data.frame()

  for (ds in unique(batch)) {
    test_id <- which(batch == ds)
    train_id <- setdiff(seq_along(batch), test_id)

    y_train <- droplevels(y[train_id])
    y_test <- droplevels(y[test_id])
    if (nlevels(y_train) < 2) {
      message("Skipping binary fold ", ds, ": training has < 2 classes.")
      next
    }
    if (!all(levels(y_test) %in% levels(y_train))) {
      message("Skipping binary fold ", ds, ": test contains class(es) not present in training.")
      next
    }
    y_test <- factor(as.character(y_test), levels = levels(y_train))
    if (length(unique(y_test)) < 2) next

    expr_train_raw <- t(X[train_id, , drop = FALSE])
    expr_test_raw <- t(X[test_id, , drop = FALSE])
    batch_train <- batch[train_id]
    batch_test <- batch[test_id]

    n_pos_before <- sum(y_train == positive)
    n_neg_before <- sum(y_train != positive)

    downsample_applied <- FALSE
    if (isTRUE(use_downsampling)) {
      idx_pos <- which(y_train == positive)
      idx_neg <- which(y_train != positive)
      if (length(idx_pos) > 0 && length(idx_neg) > 0) {
        max_n <- function(n_min) max(1L, as.integer(downsample_ratio_cap * n_min))
        keep_idx <- seq_along(y_train)
        if (length(idx_neg) > max_n(length(idx_pos))) {
          keep_idx <- sort(c(idx_pos, sample(idx_neg, max_n(length(idx_pos)))))
        } else if (length(idx_pos) > max_n(length(idx_neg))) {
          keep_idx <- sort(c(sample(idx_pos, max_n(length(idx_neg))), idx_neg))
        }
        if (length(keep_idx) < length(y_train)) {
          downsample_applied <- TRUE
          y_train <- droplevels(y_train[keep_idx])
          expr_train_raw <- expr_train_raw[, keep_idx, drop = FALSE]
          batch_train <- batch_train[keep_idx]
        }
      }
    }
    mod_train <- model.matrix(~ y_train)

    cb <- if (isTRUE(apply_batch_correction)) {
      combat_fit_apply(
        expr_train = expr_train_raw,
        batch_train = batch_train,
        mod_train = mod_train,
        expr_test = expr_test_raw,
        batch_test = batch_test
      )
    } else {
      list(train = expr_train_raw, test = expr_test_raw)
    }

    X_train_full <- t(cb$train)
    X_test_full <- t(cb$test)

    if (nrow(X_train_full) == 0 || ncol(X_train_full) == 0 || all(is.na(X_train_full))) {
      message("Skipping binary fold ", ds, ": corrected training matrix is empty or all NA.")
      next
    }

    design_train <- model.matrix(~ y_train)
    fit_lm <- tryCatch(
      eBayes(lmFit(t(X_train_full), design_train)),
      error = function(e) {
        message(
          "Skipping binary fold ", ds,
          ": lmFit failed (samples=", nrow(X_train_full),
          ", genes=", ncol(X_train_full),
          ", NA_fraction=", round(mean(is.na(X_train_full)), 4),
          ") -> ", conditionMessage(e)
        )
        NULL
      }
    )
    if (is.null(fit_lm)) next
    top_genes <- topTable(fit_lm, n = 50, sort.by = "P")
    sel_genes <- intersect(rownames(top_genes), colnames(X_train_full))
    if (length(sel_genes) < 10) next

    X_train <- X_train_full[, sel_genes, drop = FALSE]
    X_test <- X_test_full[, sel_genes, drop = FALSE]

    imp <- impute_fit(X_train)
    X_train_i <- impute_apply(X_train, imp)
    X_test_i <- impute_apply(X_test, imp)

    sc <- standardize_fit(X_train_i)
    X_train_s <- standardize_apply(X_train_i, sc)
    X_test_s <- standardize_apply(X_test_i, sc)

    n_pos <- sum(y_train == positive)
    n_neg <- sum(y_train != positive)

    if (isTRUE(use_weights)) {
      if (weight_scheme == "ratio") {
        pos_w <- if (n_pos > 0) n_neg / n_pos else 1
        neg_w <- 1
      } else {
        n_total <- n_pos + n_neg
        pos_w <- if (n_pos > 0) n_total / (2 * n_pos) else 1
        neg_w <- if (n_neg > 0) n_total / (2 * n_neg) else 1
        if (weight_scheme == "balanced_capped") {
          pos_w <- min(pos_w, weight_cap)
          neg_w <- min(neg_w, weight_cap)
        }
      }
    } else {
      pos_w <- 1
      neg_w <- 1
    }

    n_pos_after <- sum(y_train == positive)
    n_neg_after <- sum(y_train != positive)
    fold_diag <- rbind(fold_diag, data.frame(
      setting = setting_label,
      fold = ds,
      positive_class = positive,
      n_train_before = as.integer(n_pos_before + n_neg_before),
      n_pos_before = as.integer(n_pos_before),
      n_neg_before = as.integer(n_neg_before),
      n_train_after = as.integer(n_pos_after + n_neg_after),
      n_pos_after = as.integer(n_pos_after),
      n_neg_after = as.integer(n_neg_after),
      downsample_used = isTRUE(use_downsampling),
      downsample_applied = downsample_applied,
      downsample_ratio_cap = downsample_ratio_cap,
      pos_weight = as.numeric(pos_w),
      neg_weight = as.numeric(neg_w),
      stringsAsFactors = FALSE
    ))

    neg_label <- setdiff(levels(y_train), positive)[1]
    cw <- c()
    cw[positive] <- pos_w
    cw[neg_label] <- neg_w

    pred_knn <- class::knn(train = X_train_s, test = X_test_s, cl = y_train, k = 7, prob = TRUE)
    win_prob <- attr(pred_knn, "prob")
    score_knn <- ifelse(pred_knn == positive, win_prob, 1 - win_prob)
    m_knn <- binary_metrics(y_test, pred_knn, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = "kNN", fold = ds, m_knn))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = "kNN", fold = ds, truth = y_test, pred = pred_knn))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = "kNN", fold = ds, truth = y_test, score_normal = score_knn))

    svm_fit <- e1071::svm(
      x = X_train_s, y = y_train, kernel = "radial",
      cost = 1, gamma = 1 / ncol(X_train_s),
      probability = TRUE,
      class.weights = if (use_weights) cw else NULL
    )
    pred_svm <- predict(svm_fit, X_test_s, probability = TRUE)
    prob_svm <- attr(pred_svm, "probabilities")
    score_svm <- if ("Normal" %in% colnames(prob_svm)) prob_svm[, "Normal"] else prob_svm[, 1]
    m_svm <- binary_metrics(y_test, pred_svm, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = ifelse(use_weights, "SVM_RBF_w", "SVM_RBF"),
                                   fold = ds, m_svm))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = ifelse(use_weights, "SVM_RBF_w", "SVM_RBF"),
                                             fold = ds, truth = y_test, pred = pred_svm))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = ifelse(use_weights, "SVM_RBF_w", "SVM_RBF"),
                                               fold = ds, truth = y_test, score_normal = score_svm))

    rf_fit <- randomForest::randomForest(x = X_train_i, y = y_train, classwt = if (use_weights) cw else NULL)
    pred_rf <- predict(rf_fit, X_test_i)
    prob_rf <- predict(rf_fit, X_test_i, type = "prob")
    score_rf <- if ("Normal" %in% colnames(prob_rf)) prob_rf[, "Normal"] else prob_rf[, 1]
    m_rf <- binary_metrics(y_test, pred_rf, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = ifelse(use_weights, "RF_w", "RF"),
                                   fold = ds, m_rf))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = ifelse(use_weights, "RF_w", "RF"),
                                             fold = ds, truth = y_test, pred = pred_rf))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = ifelse(use_weights, "RF_w", "RF"),
                                               fold = ds, truth = y_test, score_normal = score_rf))

    w <- if (use_weights) ifelse(y_train == positive, pos_w, neg_w) else rep(1, length(y_train))
    glm_fit <- glm(y_train ~ ., data = data.frame(y_train = y_train, X_train_s),
                   family = binomial(), weights = w)
    p_hat <- predict(glm_fit, newdata = data.frame(X_test_s), type = "response")
    pred_glm <- ifelse(p_hat >= 0.5, positive, setdiff(levels(y_train), positive)[1])
    pred_glm <- factor(pred_glm, levels = levels(y_train))
    m_glm <- binary_metrics(y_test, pred_glm, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = ifelse(use_weights, "Logistic_w", "Logistic"),
                                   fold = ds, m_glm))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = ifelse(use_weights, "Logistic_w", "Logistic"),
                                             fold = ds, truth = y_test, pred = pred_glm))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = ifelse(use_weights, "Logistic_w", "Logistic"),
                                               fold = ds, truth = y_test, score_normal = p_hat))

    if (requireNamespace("glmnet", quietly = TRUE)) {
      x_train_mat <- as.matrix(X_train_s)
      x_test_mat <- as.matrix(X_test_s)
      y_train_num <- as.numeric(y_train == positive)
      lasso_fit <- tryCatch(
        glmnet::cv.glmnet(
          x = x_train_mat,
          y = y_train_num,
          family = "binomial",
          alpha = 1,
          weights = w,
          standardize = FALSE
        ),
        error = function(e) {
          message("Skipping Lasso fold ", ds, ": ", conditionMessage(e))
          NULL
        }
      )
      if (!is.null(lasso_fit)) {
        p_lasso <- as.numeric(predict(lasso_fit, newx = x_test_mat, type = "response", s = "lambda.min"))
        pred_lasso <- ifelse(p_lasso >= 0.5, positive, setdiff(levels(y_train), positive)[1])
        pred_lasso <- factor(pred_lasso, levels = levels(y_train))
        m_lasso <- binary_metrics(y_test, pred_lasso, positive = positive)
        perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                       model = ifelse(use_weights, "Lasso_w", "Lasso"),
                                       fold = ds, m_lasso))
        all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                                 model = ifelse(use_weights, "Lasso_w", "Lasso"),
                                                 fold = ds, truth = y_test, pred = pred_lasso))
        all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                                   model = ifelse(use_weights, "Lasso_w", "Lasso"),
                                                   fold = ds, truth = y_test, score_normal = p_lasso))
      }
    }
  }

  list(perf = perf, preds = all_preds, scores = all_scores, diag = fold_diag)
}

weight_cfg <- tibble::tribble(
  ~use_weights, ~weight_scheme, ~weight_cap,
  FALSE, "ratio", Inf,
  TRUE, "ratio", Inf,
  TRUE, "balanced", Inf,
  TRUE, "balanced_capped", 4
)

bin_runs <- lapply(seq_len(nrow(weight_cfg)), function(i) {
  cfg <- weight_cfg[i, ]
  run_lodo_cv_binary(
    X = X_bin,
    y = y_bin2,
    batch = batch_bin,
    positive = "Normal",
    use_weights = cfg$use_weights,
    apply_batch_correction = TRUE,
    weight_scheme = cfg$weight_scheme,
    weight_cap = cfg$weight_cap,
    use_downsampling = use_fold_downsampling,
    downsample_ratio_cap = downsample_max_ratio
  )
})

perf_bin <- bind_rows(lapply(bin_runs, `[[`, "perf"))
preds_bin <- bind_rows(lapply(bin_runs, `[[`, "preds"))
scores_bin <- bind_rows(lapply(bin_runs, `[[`, "scores"))
diag_bin <- bind_rows(lapply(bin_runs, `[[`, "diag"))
```

## Preprocessing + imbalance diagnostics

```{r}
#| message: false
#| warning: false
prep_diag <- bind_rows(lapply(merged_bin$pieces, function(pc) {
  tibble::tibble(
    dataset_id = pc$id,
    log2_applied = isTRUE(pc$log2_applied),
    q01_raw = round(pc$q01_raw, 3),
    q50_raw = round(pc$q50_raw, 3),
    q99_raw = round(pc$q99_raw, 3),
    q01_used = round(pc$q01_used, 3),
    q50_used = round(pc$q50_used, 3),
    q99_used = round(pc$q99_used, 3)
  )
}))

knitr::kable(prep_diag, caption = "Per-dataset preprocessing diagnostics (raw scale vs post-log scale)")

diag_fold_summary <- diag_bin %>%
  mutate(
    imbalance_before = round(n_neg_before / pmax(n_pos_before, 1), 2),
    imbalance_after = round(n_neg_after / pmax(n_pos_after, 1), 2)
  ) %>%
  select(setting, fold, n_train_before, n_pos_before, n_neg_before, imbalance_before,
         n_train_after, n_pos_after, n_neg_after, imbalance_after,
         downsample_used, downsample_applied, pos_weight, neg_weight)

knitr::kable(diag_fold_summary, caption = "Per-fold class-balance diagnostics (before/after downsampling and effective weights)")
```

## PCA (binary, merged)

```{r, fig.cap="PCA of merged binary expression (after imputation + scaling). Points coloured by class."}
#| message: false
#| warning: false
# PCA is for visualisation only; we impute/scale using the exploratory gene subset when available.
imp_bin_all <- impute_fit(X_bin_vis)
X_bin_i_all <- impute_apply(X_bin_vis, imp_bin_all)
sc_bin_all <- standardize_fit(X_bin_i_all)
X_bin_s_all <- standardize_apply(X_bin_i_all, sc_bin_all)

pca_bin <- prcomp(X_bin_s_all, center = FALSE, scale. = FALSE)
pca_bin_df <- data.frame(
  PC1 = pca_bin$x[, 1],
  PC2 = pca_bin$x[, 2],
  truth = y_bin2,
  dataset_id = pMat_big_bin$dataset_id[ok_bin]
)

ggplot(pca_bin_df, aes(x = PC1, y = PC2, color = truth)) +
  geom_point(alpha = 0.75, size = 2) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Class", title = "Binary PCA (merged)") +
  theme(legend.position = "right")
```

```{r, fig.cap="Binary performance across weight settings: unweighted vs weighted_ratio vs weighted_balanced vs weighted_balanced_cap4 (LODO CV)."}
#| message: false
#| warning: false
perf_bin_long <- perf_bin %>%
  pivot_longer(
    cols = c(accuracy, balanced_accuracy, recall_normal, precision_normal, f1_normal),
    names_to = "metric",
    values_to = "value"
  )

ggplot(perf_bin_long, aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.6) +
  facet_grid(metric ~ setting, scales = "free_y") +
  theme_bw() +
  labs(x = "", y = "", title = "Binary model comparison by weighting scheme") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r, fig.cap="Binary: confusion matrices for best model (default threshold 0.5 vs tuned threshold maximizing pooled F1 for Normal)."}
#| message: false
#| warning: false
best_run_bin <- perf_bin %>%
  group_by(setting, model) %>%
  summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    mean_recall = mean(recall_normal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    scores_bin %>%
      group_by(setting, model) %>%
      summarise(
        mean_roc_auc = auc_roc(truth == "Normal", score_normal),
        mean_pr_auc = auc_pr(truth == "Normal", score_normal),
        .groups = "drop"
      ),
    by = c("setting", "model")
  ) %>%
  arrange(desc(mean_roc_auc), desc(mean_pr_auc), desc(mean_recall), desc(mean_accuracy)) %>%
  slice(1)

best_setting_bin <- best_run_bin$setting[1]
best_model_bin <- best_run_bin$model[1]
cat("Best binary run by ROC AUC then PR AUC then recall:", best_setting_bin, "/", best_model_bin,
  "(ROC AUC=", round(best_run_bin$mean_roc_auc[1], 3),
  ", PR AUC=", round(best_run_bin$mean_pr_auc[1], 3),
  ", Recall=", round(best_run_bin$mean_recall[1], 3),
  ", Accuracy=", round(best_run_bin$mean_accuracy[1], 3), ")\n", sep = "")

score_best <- scores_bin %>% filter(setting == best_setting_bin, model == best_model_bin)
thr_best <- best_threshold_f1(score_best$truth, score_best$score_normal, positive = "Normal")
cat("Chosen threshold (max pooled F1 for Normal):", round(thr_best, 4), "\n")

neg_label_bin <- setdiff(levels(y_bin2), "Normal")[1]
pred_default <- ifelse(score_best$score_normal >= 0.5, "Normal", neg_label_bin)
pred_tuned <- ifelse(score_best$score_normal >= thr_best, "Normal", neg_label_bin)
tab_default <- table(truth = score_best$truth, pred = pred_default)
tab_tuned <- table(truth = score_best$truth, pred = pred_tuned)

acc_check <- tibble::tibble(
  metric = c(
    "Mean fold accuracy (from perf table)",
    "Pooled accuracy @ threshold 0.5",
    "Pooled accuracy @ tuned threshold"
  ),
  value = c(
    perf_bin %>% filter(setting == best_setting_bin, model == best_model_bin) %>% summarise(v = mean(accuracy, na.rm = TRUE)) %>% pull(v),
    mean(score_best$truth == pred_default, na.rm = TRUE),
    mean(score_best$truth == pred_tuned, na.rm = TRUE)
  )
)
knitr::kable(acc_check %>% mutate(value = round(value, 3)),
             caption = paste0("Accuracy consistency check for ", best_setting_bin, " / ", best_model_bin))

tab_default_df <- as.data.frame(tab_default)
tab_tuned_df <- as.data.frame(tab_tuned)

p_cm_default <- ggplot(tab_default_df, aes(x = pred, y = truth, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradient(low = "white", high = "#41AB5D") +
  theme_bw() +
  labs(x = "Predicted", y = "True", title = paste0("Pooled confusion matrix @ 0.5: ", best_setting_bin, " / ", best_model_bin)) +
  theme(legend.position = "right")

p_cm_tuned <- ggplot(tab_tuned_df, aes(x = pred, y = truth, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradient(low = "white", high = "#41AB5D") +
  theme_bw() +
  labs(x = "Predicted", y = "True", title = paste0("Pooled confusion matrix @ tuned threshold: ", best_setting_bin, " / ", best_model_bin)) +
  theme(legend.position = "right")

print(p_cm_default)
print(p_cm_tuned)
```

```{r, fig.cap="Binary: pooled ROC and PR curves for the best model."}
#| message: false
#| warning: false
curve_best <- scores_bin %>% filter(setting == best_setting_bin, model == best_model_bin)
truth01 <- as.integer(curve_best$truth == "Normal")
score <- curve_best$score_normal

# ROC points
roc_thresh <- sort(unique(score), decreasing = TRUE)
roc_df <- lapply(roc_thresh, function(t) {
  pred <- score >= t
  tp <- sum(pred & truth01 == 1)
  fn <- sum(!pred & truth01 == 1)
  fp <- sum(pred & truth01 == 0)
  tn <- sum(!pred & truth01 == 0)
  tibble(
    threshold = t,
    tpr = if ((tp + fn) == 0) NA_real_ else tp / (tp + fn),
    fpr = if ((fp + tn) == 0) NA_real_ else fp / (fp + tn)
  )
}) %>% bind_rows()

pr_df <- pr_curve_df(truth01, score)

p1 <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "#2C7FB8", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  theme_bw() +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    title = paste0("ROC curve (AUC = ", round(auc_roc(truth01, score), 3), ")")
  )

p2 <- ggplot(pr_df, aes(x = recall, y = precision)) +
  geom_line(color = "#41AB5D", linewidth = 1) +
  theme_bw() +
  labs(
    x = "Recall (Normal)",
    y = "Precision (Normal)",
    title = paste0("PR curve (AUC = ", round(auc_pr(truth01, score), 3), ")")
  )

print(p1)
print(p2)
```

```{r}
#| message: false
#| warning: false
# Binary: aggregate metrics table (mean \u00B1 SD) plus ROC/PR AUC across all weight settings

model_auc <- scores_bin %>%
  group_by(setting, model) %>%
  summarise(
    roc_auc = auc_roc(truth == "Normal", score_normal),
    pr_auc = auc_pr(truth == "Normal", score_normal),
    .groups = "drop"
  )

model_summary <- perf_bin %>%
  group_by(setting, model) %>%
  summarise(
    accuracy_mean = mean(accuracy, na.rm = TRUE),
    accuracy_sd = sd(accuracy, na.rm = TRUE),
    balanced_accuracy_mean = mean(balanced_accuracy, na.rm = TRUE),
    balanced_accuracy_sd = sd(balanced_accuracy, na.rm = TRUE),
    sensitivity_mean = mean(recall_normal, na.rm = TRUE),
    sensitivity_sd = sd(recall_normal, na.rm = TRUE),
    precision_mean = mean(precision_normal, na.rm = TRUE),
    precision_sd = sd(precision_normal, na.rm = TRUE),
    f1_mean = mean(f1_normal, na.rm = TRUE),
    f1_sd = sd(f1_normal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(model_auc, by = c("setting", "model")) %>%
  mutate(
    rank_roc = dense_rank(desc(roc_auc)),
    rank_pr = dense_rank(desc(pr_auc)),
    rank_recall = dense_rank(desc(sensitivity_mean))
  ) %>%
  arrange(desc(roc_auc), desc(pr_auc), desc(sensitivity_mean), desc(accuracy_mean)) %>%
  mutate(overall_rank = row_number()) %>%
  select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall,
         accuracy_mean, accuracy_sd, balanced_accuracy_mean, balanced_accuracy_sd,
         sensitivity_mean, sensitivity_sd, precision_mean, precision_sd, f1_mean, f1_sd,
         roc_auc, pr_auc) %>%
  mutate(
    accuracy = sprintf("%.3f \u00B1 %.3f", accuracy_mean, accuracy_sd),
    balanced_accuracy = sprintf("%.3f \u00B1 %.3f", balanced_accuracy_mean, balanced_accuracy_sd),
    sensitivity = sprintf("%.3f \u00B1 %.3f", sensitivity_mean, sensitivity_sd),
    precision = sprintf("%.3f \u00B1 %.3f", precision_mean, precision_sd),
    f1 = sprintf("%.3f \u00B1 %.3f", f1_mean, f1_sd),
    roc_auc = round(roc_auc, 3),
    pr_auc = round(pr_auc, 3)
  ) %>%
  select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall,
         accuracy, balanced_accuracy, sensitivity, precision, f1, roc_auc, pr_auc)

knitr::kable(model_summary, caption = "Binary: model scorecard ranked by ROC AUC, then PR AUC, then recall")
```

## How many DE genes do we need? (binary)

```{r, fig.cap="Binary: best balanced accuracy vs number of DE genes (weighted models only)."}
#| message: false
#| warning: false
# Goal: find the smallest top-N DE gene set that achieves strong performance.
# We evaluate a grid of N, run the weighted leave-one-dataset-out CV once per N, and take the best mean balanced accuracy across models.

de_gene_sweep <- function(e_bin_de, y_bin_de, n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
                          K = 4, repeats = 10, positive = "Normal") {
  if (nrow(e_bin_de) == 0 || ncol(e_bin_de) == 0 || all(is.na(e_bin_de))) {
    message("Skipping DE sweep: expression matrix is empty or all NA.")
    return(tibble::tibble(
      n_genes = integer(0),
      best_model = character(0),
      best_mean_balanced_accuracy = numeric(0)
    ))
  }

  design_bin <- model.matrix(~ y_bin_de)
  fit_bin <- tryCatch(
    eBayes(lmFit(e_bin_de, design_bin)),
    error = function(e) {
      message("Skipping DE sweep: ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(fit_bin)) {
    return(tibble::tibble(
      n_genes = integer(0),
      best_model = character(0),
      best_mean_balanced_accuracy = numeric(0)
    ))
  }

  out <- list()
  for (n_genes in n_grid) {
    tt <- topTable(fit_bin, n = n_genes, sort.by = "P")
    genes <- rownames(tt)
    X <- t(e_bin_de[intersect(genes, rownames(e_bin_de)), , drop = FALSE])

    # If something went wrong (e.g., no overlap), skip safely
    if (ncol(X) < 2) next

    batch_de <- pMat_big_bin$dataset_id[ok_bin_de]
    res_w <- run_lodo_cv_binary(X, y_bin_de, batch_de, positive = positive, use_weights = TRUE)
    perf_w <- res_w$perf

    if (nrow(perf_w) == 0 || !"model" %in% names(perf_w)) {
      out[[as.character(n_genes)]] <- data.frame(
        n_genes = n_genes,
        best_model = NA_character_,
        best_mean_balanced_accuracy = NA_real_
      )
      next
    }

    best <- perf_w %>%
      group_by(model) %>%
      summarise(mean_bal_acc = mean(balanced_accuracy, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(mean_bal_acc)) %>%
      slice(1)

    out[[as.character(n_genes)]] <- data.frame(
      n_genes = n_genes,
      best_model = best$model[1],
      best_mean_balanced_accuracy = best$mean_bal_acc[1]
    )
  }
  bind_rows(out)
}

sweep_res <- de_gene_sweep(e_bin_de, y_bin_de,
                          n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
                          K = 4, repeats = 10, positive = "Normal")

knitr::kable(sweep_res, caption = "Binary DE gene sweep (weighted CV): best mean balanced accuracy by top-N DE genes")

ggplot(sweep_res, aes(x = n_genes, y = best_mean_balanced_accuracy)) +
  geom_line(color = "#2C7FB8", linewidth = 1) +
  geom_point(size = 2) +
  theme_bw() +
  scale_x_continuous(breaks = sweep_res$n_genes) +
  labs(x = "Top-N DE genes", y = "Best mean balanced accuracy (weighted CV)", title = "How many DE genes do we need?") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## QC inclusion criteria overview (per dataset)

```{r}
#| message: false
#| warning: false
qc_summary_from_piece <- function(piece, task = c("binary", "multiclass")) {
  task <- match.arg(task)
  p <- piece$p
  e <- piece$e
  f <- piece$f

  label_col <- unique(as.character(p$label_col_detected))
  label_col <- label_col[!is.na(label_col) & label_col != ""]
  label_col <- if (length(label_col) > 0) label_col[1] else NA_character_

  label_vec <- if (task == "binary") as.character(p$label_binary) else as.character(p$label_multiclass)
  label_vec[is.na(label_vec) | trimws(label_vec) == ""] <- NA_character_
  n_labeled <- sum(!is.na(label_vec))
  n_classes <- dplyr::n_distinct(label_vec[!is.na(label_vec)])

  # Heuristic: if metadata says normalized / log transformed, mark yes.
  p_text <- tolower(apply(as.data.frame(p), 1, function(r) paste(r, collapse = " | ")))
  has_norm_meta <- any(grepl("normaliz|log2|rma|mas5|quantile", p_text))

  # Gene symbol coverage proxy: proportion of feature rownames that look like symbols.
  gene_ids <- rownames(f)
  gene_symbol_like <- mean(grepl("^[A-Za-z0-9][A-Za-z0-9\\-\\.]{1,}$", gene_ids))

  tibble::tibble(
    dataset_id = piece$id,
    task = task,
    n_samples = ncol(e),
    n_genes_after_collapse = nrow(e),
    normalized_data = ifelse(has_norm_meta, "Likely (metadata indicates transformed/normalized)", "Likely (series matrix); explicit flag not found"),
    outcome_variable_clear = ifelse(!is.na(label_col) && n_classes >= 2, "Yes", "No / unclear"),
    detected_outcome_column = ifelse(is.na(label_col), "NA", label_col),
    labeled_samples = n_labeled,
    classes_detected = n_classes,
    gene_symbols_available = sprintf("%.1f%% symbol-like rownames", 100 * gene_symbol_like),
    include_in_merge = ifelse(n_labeled >= 10 && n_classes >= 2, "Include", "Exclude (insufficient/unclear labels)")
  )
}

qc_bin <- bind_rows(lapply(merged_bin$pieces, qc_summary_from_piece, task = "binary"))
qc_mc  <- bind_rows(lapply(merged_mc$pieces,  qc_summary_from_piece, task = "multiclass"))
qc_table <- bind_rows(qc_bin, qc_mc) %>% arrange(task, dataset_id)

knitr::kable(qc_table, caption = "QC inclusion criteria overview by dataset")
```

## Dataset and Report Flow Statistics

```{r}
#| message: false
#| warning: false
# Detailed flow statistics mappings inspired by PRISMA 2020

flow_stats_df <- tibble::tibble(
  Metric = c(
    "Databases",
    "Registers",
    "Reports sought for retrieval",
    "Reports assessed for eligibility",
    "Duplicate records removed",
    "Records marked as ineligible by automation tools",
    "Records removed for other reasons",
    "Records screened",
    "Records excluded",
    "Reports excluded"
  ),
  Count = c(
    1, # GEO database
    0, 
    length(c(datasets_binary, datasets_multiclass)), # Total datasets attempted for retrieval
    sum(qc_table$n_samples > 0), # Datasets successfully matched/loaded limits
    0, # No explicit duplication check across raw samples here
    sum(qc_table$n_samples) - sum(qc_table$labeled_samples), # Samples lacking valid inferred labels
    0, 
    sum(qc_table$n_samples), # Total raw samples extracted prior to labelling filtering
    sum(qc_table$n_samples) - sum(qc_table$labeled_samples), # Same as ineligible drop exclusion
    sum(qc_table$include_in_merge == "Exclude (insufficient/unclear labels)")
  )
)

knitr::kable(flow_stats_df, caption = "Data Flow and Exclusion Statistics")
```

## Does adding more datasets help? (AUC vs number of datasets)

```{r}
#| message: false
#| warning: false
dataset_count_auc_sweep <- function(dataset_ids, normal_keywords,
                                    top_n_genes = 50, K = 4, repeats = 8,
                                    positive = "Normal") {
  out <- list()

  for (k in seq_along(dataset_ids)) {
    ids_k <- dataset_ids[seq_len(k)]
    merged_k <- build_merged_binary_dataset(ids_k, normal_keywords)

    ok_k <- !is.na(merged_k$pMat$label_binary) & merged_k$pMat$label_binary != ""
    y_k <- factor(merged_k$pMat$label_binary[ok_k], levels = c("Cancer", "Normal"))
    e_k <- merged_k$eMat[, ok_k, drop = FALSE]
    if (nlevels(droplevels(y_k)) < 2) next
    if (nrow(e_k) == 0 || ncol(e_k) == 0 || all(is.na(e_k))) next

    design_k <- model.matrix(~ y_k)
    fit_k <- tryCatch(
      eBayes(lmFit(e_k, design_k)),
      error = function(e) {
        message("Skipping dataset prefix ", k, " in dataset sweep: ", conditionMessage(e))
        NULL
      }
    )
    if (is.null(fit_k)) next
    top_k <- topTable(fit_k, n = top_n_genes, sort.by = "P")
    genes_k <- rownames(top_k)
    X_k <- t(e_k[intersect(genes_k, rownames(e_k)), , drop = FALSE])
    if (ncol(X_k) < 2 || nrow(X_k) < (K + 2)) next

    batch_k <- merged_k$pMat$dataset_id[ok_k]
    res_k <- run_lodo_cv_binary(X_k, y_k, batch_k, positive = positive, use_weights = TRUE)

    req_cols <- c("model", "truth", "score_normal")
    if (!is.data.frame(res_k$scores) || nrow(res_k$scores) == 0 ||
        !all(req_cols %in% names(res_k$scores))) {
      message("Skipping dataset prefix ", k, ": score output is empty or missing required columns.")
      next
    }

    auc_k <- tryCatch({
      res_k$scores %>%
        group_by(model) %>%
        summarise(
          roc_auc = auc_roc(truth == positive, score_normal),
          pr_auc = auc_pr(truth == positive, score_normal),
          .groups = "drop"
        ) %>%
        arrange(desc(roc_auc))
    }, error = function(e) {
      message("Skipping dataset prefix ", k, " during AUC summary: ", conditionMessage(e))
      tibble::tibble()
    })

    if (nrow(auc_k) == 0) next

    out[[length(out) + 1]] <- tibble::tibble(
      n_datasets = k,
      datasets_used = paste(ids_k, collapse = ","),
      n_samples = nrow(X_k),
      best_model = auc_k$model[1],
      best_roc_auc = auc_k$roc_auc[1],
      best_pr_auc = auc_k$pr_auc[1]
    )
  }

  bind_rows(out)
}

dataset_auc_res <- dataset_count_auc_sweep(
  dataset_ids = datasets_binary,
  normal_keywords = normal_keywords,
  top_n_genes = 50, K = 4, repeats = 8, positive = "Normal"
)

knitr::kable(dataset_auc_res, caption = "AUC vs number of merged binary datasets")

ggplot(dataset_auc_res, aes(x = n_datasets, y = best_roc_auc)) +
  geom_line(color = "#2C7FB8", linewidth = 1) +
  geom_point(size = 2) +
  theme_bw() +
  scale_x_continuous(breaks = dataset_auc_res$n_datasets) +
  labs(x = "Number of merged datasets", y = "Best ROC AUC", title = "Does adding datasets improve ROC AUC?")
```

## Cross-Dataset Generalization Heatmap

```{r, fig.cap="Heatmap of external validation ROC AUC across all pairwise combinations of binary datasets (columns = test set, rows = training set)."}
#| message: false
#| warning: false
compute_cross_dataset_matrix <- function(dataset_ids, eMat_big, pMat_big, target_genes, positive = "Normal") {
  n <- length(dataset_ids)
  auc_mat <- matrix(NA, n, n, dimnames = list(Train = dataset_ids, Test = dataset_ids))
  
  # For safety, pre-extract the unified datasets
  ok_bin <- !is.na(pMat_big$label_binary) & pMat_big$label_binary != ""
  p_valid <- pMat_big[ok_bin, ]
  e_valid <- eMat_big[intersect(target_genes, rownames(eMat_big)), ok_bin, drop = FALSE]
  
  for (i in seq_along(dataset_ids)) {
    id_train <- dataset_ids[i]
    idx_train <- which(p_valid$dataset_id == id_train)
    if (length(idx_train) == 0) next
    
    y_train <- factor(p_valid$label_binary[idx_train], levels = c("Cancer", "Normal"))
    X_train <- t(e_valid[, idx_train, drop = FALSE])
    
    if (nlevels(droplevels(y_train)) < 2) next
    
    imp <- impute_fit(X_train)
    X_train_i <- impute_apply(X_train, imp)
    sc <- standardize_fit(X_train_i)
    X_train_s <- standardize_apply(X_train_i, sc)
    
    tab <- table(y_train)
    w_train <- ifelse(y_train == positive, as.numeric(tab["Cancer"] / max(1, tab["Normal"])), 1)
    
    fit <- glm(y_train ~ ., data = data.frame(y_train = y_train, X_train_s),
               family = binomial(), weights = w_train)
    
    for (j in seq_along(dataset_ids)) {
      id_test <- dataset_ids[j]
      idx_test <- which(p_valid$dataset_id == id_test)
      if (length(idx_test) == 0) next
      
      y_test <- factor(p_valid$label_binary[idx_test], levels = c("Cancer", "Normal"))
      X_test <- t(e_valid[, idx_test, drop = FALSE])
      
      if (nlevels(droplevels(y_test)) < 2) next
      
      X_test_s <- standardize_apply(impute_apply(X_test, imp), sc)
      
      # The score returned is P(Y="Normal")
      p_hat <- predict(fit, newdata = data.frame(X_test_s), type = "response")
      auc_mat[i, j] <- auc_roc(y_test == positive, p_hat)
    }
  }
  auc_mat
}

cross_auc_mat <- compute_cross_dataset_matrix(
  dataset_ids = datasets_binary,
  eMat_big = eMat_big_bin, 
  pMat_big = pMat_big_bin, 
  target_genes = genes_bin, 
  positive = "Normal"
)

# Convert to long format for ggplot heatmap
cross_auc_df <- as.data.frame(as.table(cross_auc_mat))
colnames(cross_auc_df) <- c("Train", "Test", "AUC")

ggplot(cross_auc_df, aes(x = Test, y = Train, fill = AUC)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", AUC)), color = ifelse(is.na(cross_auc_df$AUC) | cross_auc_df$AUC > 0.75, "white", "black"), size = 4) +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B", na.value = "grey90") +
  theme_bw() +
  labs(
    x = "Test Dataset (Evaluation)", 
    y = "Train Dataset (Fit)", 
    title = "Cross-Dataset Model Generalization (Logistic Regression ROC AUC)",
    fill = "ROC AUC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Feature count vs performance (AUC extension)

```{r}
#| message: false
#| warning: false
de_gene_sweep_auc <- function(e_bin_de, y_bin_de, n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
                              K = 4, repeats = 8, positive = "Normal") {
  if (nrow(e_bin_de) == 0 || ncol(e_bin_de) == 0 || all(is.na(e_bin_de))) {
    message("Skipping DE sweep AUC: expression matrix is empty or all NA.")
    return(tibble::tibble(
      n_genes = integer(0),
      best_model_bal_acc = character(0),
      best_mean_balanced_accuracy = numeric(0),
      best_model_auc = character(0),
      best_roc_auc = numeric(0),
      best_pr_auc = numeric(0)
    ))
  }

  design_bin <- model.matrix(~ y_bin_de)
  fit_bin <- tryCatch(
    eBayes(lmFit(e_bin_de, design_bin)),
    error = function(e) {
      message("Skipping DE sweep AUC: ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(fit_bin)) {
    return(tibble::tibble(
      n_genes = integer(0),
      best_model_bal_acc = character(0),
      best_mean_balanced_accuracy = numeric(0),
      best_model_auc = character(0),
      best_roc_auc = numeric(0),
      best_pr_auc = numeric(0)
    ))
  }
  out <- list()

  for (n_genes in n_grid) {
    tt <- topTable(fit_bin, n = n_genes, sort.by = "P")
    genes <- rownames(tt)
    X <- t(e_bin_de[intersect(genes, rownames(e_bin_de)), , drop = FALSE])
    if (ncol(X) < 2) next

    batch_de <- pMat_big_bin$dataset_id[ok_bin_de]
    res_w <- run_lodo_cv_binary(X, y_bin_de, batch_de, positive = positive, use_weights = TRUE)

    if (nrow(res_w$perf) == 0 || !"model" %in% names(res_w$perf) ||
        nrow(res_w$scores) == 0 || !"model" %in% names(res_w$scores)) {
      out[[as.character(n_genes)]] <- tibble::tibble(
        n_genes = n_genes,
        best_model_bal_acc = NA_character_,
        best_mean_balanced_accuracy = NA_real_,
        best_model_auc = NA_character_,
        best_roc_auc = NA_real_,
        best_pr_auc = NA_real_
      )
      next
    }

    perf_best <- res_w$perf %>%
      group_by(model) %>%
      summarise(mean_bal_acc = mean(balanced_accuracy, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(mean_bal_acc)) %>%
      slice(1)

    auc_by_model <- res_w$scores %>%
      group_by(model) %>%
      summarise(
        roc_auc = auc_roc(truth == positive, score_normal),
        pr_auc = auc_pr(truth == positive, score_normal),
        .groups = "drop"
      ) %>%
      arrange(desc(roc_auc))

    out[[as.character(n_genes)]] <- tibble::tibble(
      n_genes = n_genes,
      best_model_bal_acc = perf_best$model[1],
      best_mean_balanced_accuracy = perf_best$mean_bal_acc[1],
      best_model_auc = auc_by_model$model[1],
      best_roc_auc = auc_by_model$roc_auc[1],
      best_pr_auc = auc_by_model$pr_auc[1]
    )
  }

  bind_rows(out)
}

sweep_auc_res <- de_gene_sweep_auc(
  e_bin_de, y_bin_de,
  n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
  K = 4, repeats = 8, positive = "Normal"
)

knitr::kable(sweep_auc_res, caption = "Feature-count sweep: balanced accuracy + ROC/PR AUC")

ggplot(sweep_auc_res, aes(x = n_genes)) +
  geom_line(aes(y = best_roc_auc, color = "ROC AUC"), linewidth = 1) +
  geom_point(aes(y = best_roc_auc, color = "ROC AUC"), size = 2) +
  geom_line(aes(y = best_pr_auc, color = "PR AUC"), linewidth = 1) +
  geom_point(aes(y = best_pr_auc, color = "PR AUC"), size = 2) +
  theme_bw() +
  scale_x_continuous(breaks = sweep_auc_res$n_genes) +
  labs(x = "Top-N DE genes", y = "AUC", color = "", title = "AUC vs number of features")
```

## Risk calculator (binary prototype)

```{r}
#| message: false
#| warning: false
# Build a simple risk calculator from merged binary data:
# output = estimated probability of Cancer for each sample.

# Pick feature count by best ROC AUC from sweep (fallback to 50 if unavailable)
n_risk <- if (nrow(sweep_auc_res) > 0) sweep_auc_res$n_genes[which.max(sweep_auc_res$best_roc_auc)] else 50

design_risk <- model.matrix(~ y_bin_de)
fit_risk_de <- eBayes(lmFit(e_bin_de, design_risk))
top_risk <- topTable(fit_risk_de, n = n_risk, sort.by = "P")
genes_risk <- rownames(top_risk)

X_risk_raw <- t(e_bin_de[intersect(genes_risk, rownames(e_bin_de)), , drop = FALSE])

imp_risk <- impute_fit(X_risk_raw)
X_risk_i <- impute_apply(X_risk_raw, imp_risk)
sc_risk <- standardize_fit(X_risk_i)
X_risk_s <- standardize_apply(X_risk_i, sc_risk)

# Class weighting to reduce bias under imbalance
tab_risk <- table(y_bin_de)
w_risk <- ifelse(y_bin_de == "Normal", as.numeric(tab_risk["Cancer"] / tab_risk["Normal"]), 1)

# Extract Clinical Metadata (Age, Gender, Location)
p_risk <- pMat_big_bin[ok_bin_de, ]
extract_metadata_cols <- function(p) {
  df <- data.frame(sample = rownames(p), age = NA_real_, gender = NA_character_, location = NA_character_)
  chars <- p[, grep("^characteristics_ch", colnames(p)), drop = FALSE]
  
  if (ncol(chars) == 0) return(df)
  
  for (i in seq_len(nrow(p))) {
    row_text <- tolower(paste(chars[i, ], collapse = " "))
    
    # Age extraction
    m_age <- regexpr("age[^a-z0-9]*([0-9]{1,3})", row_text)
    if (m_age > 0) df$age[i] <- as.numeric(gsub("[^0-9]", "", regmatches(row_text, m_age)))
    
    # Gender extraction
    if (grepl("female", row_text)) {
       df$gender[i] <- "Female"
    } else if (grepl("male[\\s,;]?", row_text)) {
       df$gender[i] <- "Male"
    }
    
    # Location / Country
    m_loc <- regexpr("country[:\\s]*([a-z]+)", row_text)
    if (m_loc > 0) df$location[i] <- gsub("country[:\\s]*", "", regmatches(row_text, m_loc))
  }
  df
}

meta_df <- extract_metadata_cols(p_risk)

cat("\n--- Metadata Extraction Summary ---\n")
if (all(is.na(meta_df$age))) {
  cat("No age metadata found across datasets. Proceeding with genetic markers only.\n")
} else {
  cat("Age successfully extracted for", sum(!is.na(meta_df$age)), "samples.\n")
}
cat("\nGender distribution:\n")
print(table(meta_df$gender, useNA = "always"))
cat("\nLocation/Country availability:\n")
print(table(meta_df$location, useNA = "always"))

# Append available metadata features to X_risk_s for modelling
X_risk_meta <- data.frame(X_risk_s)

# Impute and add Age if reasonably present
if (!all(is.na(meta_df$age))) {
  med_age <- median(meta_df$age, na.rm = TRUE)
  filled_age <- meta_df$age
  filled_age[is.na(filled_age)] <- med_age
  X_risk_meta$meta_Age <- scale(filled_age)
}

# Add Gender if variability exists
if (length(na.omit(unique(meta_df$gender))) > 1) {
  is_female <- as.numeric(meta_df$gender == "Female")
  is_female[is.na(is_female)] <- 1 # Base rate breast cancer assumption
  X_risk_meta$meta_IsFemale <- is_female
}
risk_band <- function(p) {
  ifelse(p < 0.20, "Low",
         ifelse(p < 0.50, "Intermediate", "High"))
}

# Risk calculator reports that mirror the classification pipeline.
summarise_risk_report <- function(perf_df, scores_df, report_label) {
  risk_scores_df <- scores_df %>%
    mutate(
      p_normal = score_normal,
      p_cancer = 1 - score_normal,
      risk_score = qlogis(pmin(pmax(p_normal, 1e-6), 1 - 1e-6)),
      risk_band = risk_band(p_cancer),
      sample_id = paste0("fold_", fold, "_", row_number())
    )

  risk_metrics_df <- risk_scores_df %>%
    group_by(setting, model) %>%
    summarise(
      roc_auc = auc_roc(truth == "Normal", score_normal),
      pr_auc = auc_pr(truth == "Normal", score_normal),
      .groups = "drop"
    ) %>%
    left_join(
      perf_df %>%
        group_by(setting, model) %>%
        summarise(
          accuracy = mean(accuracy, na.rm = TRUE),
          recall = mean(recall_normal, na.rm = TRUE),
          .groups = "drop"
        ),
      by = c("setting", "model")
    ) %>%
    mutate(
      rank_roc = dense_rank(desc(roc_auc)),
      rank_pr = dense_rank(desc(pr_auc)),
      rank_recall = dense_rank(desc(recall))
    ) %>%
    arrange(desc(roc_auc), desc(pr_auc), desc(recall), desc(accuracy)) %>%
    mutate(overall_rank = row_number())

  best_setting <- risk_metrics_df$setting[1]
  best_model <- risk_metrics_df$model[1]
  risk_df <- risk_scores_df %>% filter(setting == best_setting, model == best_model)
  plot_df <- risk_df %>% mutate(fold = factor(fold, levels = unique(fold)))

  list(
    report_label = report_label,
    risk_scores_df = risk_scores_df,
    risk_metrics_df = risk_metrics_df,
    best_setting = best_setting,
    best_model = best_model,
    risk_df = risk_df,
    plot_df = plot_df
  )
}

risk_runs_nb <- lapply(seq_len(nrow(weight_cfg)), function(i) {
  cfg <- weight_cfg[i, ]
  run_lodo_cv_binary(
    X = X_bin,
    y = y_bin2,
    batch = batch_bin,
    positive = "Normal",
    use_weights = cfg$use_weights,
    apply_batch_correction = FALSE,
    weight_scheme = cfg$weight_scheme,
    weight_cap = cfg$weight_cap
  )
})

risk_perf_nb <- bind_rows(lapply(risk_runs_nb, `[[`, "perf"))
risk_scores_nb <- bind_rows(lapply(risk_runs_nb, `[[`, "scores"))
risk_report_nb <- summarise_risk_report(risk_perf_nb, risk_scores_nb, "No batch correction")

risk_report_bc <- summarise_risk_report(perf_bin, scores_bin, "Batch corrected")

cat("Risk model feature count (Target):", n_risk, "\n")

cat("\n--- Risk Calculator: No Batch Correction ---\n")
knitr::kable(
  risk_report_nb$risk_metrics_df %>%
    mutate(across(c(roc_auc, pr_auc, accuracy, recall), ~ round(.x, 3))) %>%
    select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall, accuracy, recall, roc_auc, pr_auc),
  caption = "Risk calculator (no batch correction): model scorecard ranked by ROC AUC, then PR AUC, then recall"
)
cat("Best risk model (no batch correction):", risk_report_nb$best_setting, "/", risk_report_nb$best_model, "\n")
knitr::kable(
  head(risk_report_nb$risk_df %>% arrange(desc(p_cancer)), 15),
  caption = paste0("Risk output (no batch correction) using ", risk_report_nb$best_model, ", top 15 highest predicted cancer risk")
)

p_risk_nb_folds <- ggplot(risk_report_nb$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ fold, ncol = 3) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("LODO risk separation by fold (no batch): ", risk_report_nb$best_model))

p_risk_nb_all <- ggplot(risk_report_nb$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("Selected model: ", risk_report_nb$best_model, " (no batch)"))

print(p_risk_nb_folds)
print(p_risk_nb_all)

cat("\n--- Risk Calculator: Batch Corrected ---\n")
knitr::kable(
  risk_report_bc$risk_metrics_df %>%
    mutate(across(c(roc_auc, pr_auc, accuracy, recall), ~ round(.x, 3))) %>%
    select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall, accuracy, recall, roc_auc, pr_auc),
  caption = "Risk calculator (batch corrected): model scorecard ranked by ROC AUC, then PR AUC, then recall"
)
cat("Best risk model (batch corrected):", risk_report_bc$best_setting, "/", risk_report_bc$best_model, "\n")
knitr::kable(
  head(risk_report_bc$risk_df %>% arrange(desc(p_cancer)), 15),
  caption = paste0("Risk output (batch corrected) using ", risk_report_bc$best_model, ", top 15 highest predicted cancer risk")
)

p_risk_bc_folds <- ggplot(risk_report_bc$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ fold, ncol = 3) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("LODO risk separation by fold (batch corrected): ", risk_report_bc$best_model))

p_risk_bc_all <- ggplot(risk_report_bc$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("Selected model: ", risk_report_bc$best_model, " (batch corrected)"))

print(p_risk_bc_folds)
print(p_risk_bc_all)
```

# Workflow Description

For each dataset, samples were first split in a stratified manner by class (Cancer vs Normal), where:

75% of samples were assigned to the development set 25% of samples were reserved as an independent hold-out test set

This split was performed within each dataset and within each class, ensuring that both dataset structure and class distribution were keep.

## Development Set (75%)

The development set was used for model selection and ensemble design:

Leave-One-Dataset-Out (LODO) validation was performed to evaluate model performance The mean LODO AUC was computed for each candidate model Models were ranked based on their mean LODO performance Fixed ensemble groups were constructed using a Top-K selection strategy (e.g., Top 4, Top 5, etc.) Multiple weighting strategies were defined to combine model outputs

This stage was used exclusively for designing the ensemble, without accessing the hold-out test set.

## Hold-out Test Set (25%)

The hold-out test set was used for external evaluation:

All predefined ensemble configurations (Top-K × weighting strategies) were applied Performance was evaluated on unseen samples using: AUC Sensitivity Specificity Balanced Accuracy Additionally, density overlap between Cancer and Normal risk score distributions was computed The results were used to compare how well different ensemble configurations generalize to new data

Importantly, the hold-out test set was not used for model selection or tuning, but only for final evaluation. 
# No batch ensemble risk

```{r}
# ============================================================
# NOT BATCH CORRECTION Ensemble Risk Calculator
# Top-K fixed ensemble version
# 75% development + 25% hold-out test
# Development: LODO mean AUC for model ranking
# Hold-out: evaluate Top-K fixed ensemble combinations
# ============================================================

library(dplyr)
library(tidyr)
library(tibble)
library(limma)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(glmnet)

set.seed(3888)

# -----------------------------
# 0. Settings
# -----------------------------
positive_class <- "Cancer"
negative_class <- "Normal"

top_n_genes <- 50

top_k_vec <- c(4, 5, 6, 7)

weight_methods <- c(
  "rank_auc",
  "softmax_auc",
  "auc_x_balacc",
  "auc_minus_spec_penalty",
  "equal_weight"
)

T <- 0.20
lambda <- 0.50

candidate_models <- c(
  "RF",
  "SVM_RBF",
  "SVM_LINEAR",
  "LASSO",
  "Ridge",
  "ElasticNet",
  "kNN",
  "Logistic",
  "NaiveBayes"
)

# -----------------------------
# 1. Helper functions
# -----------------------------
clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)

  pos <- truth_pos == 1
  neg <- truth_pos == 0

  n_pos <- sum(pos)
  n_neg <- sum(neg)

  if (n_pos == 0 || n_neg == 0) return(NA_real_)

  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

binary_metrics_cancer <- function(truth, score_cancer, threshold = 0.5) {
  truth <- as.character(truth)
  pred <- ifelse(score_cancer >= threshold, positive_class, negative_class)

  tp <- sum(pred == positive_class & truth == positive_class, na.rm = TRUE)
  fn <- sum(pred != positive_class & truth == positive_class, na.rm = TRUE)
  tn <- sum(pred == negative_class & truth == negative_class, na.rm = TRUE)
  fp <- sum(pred != negative_class & truth == negative_class, na.rm = TRUE)

  accuracy <- mean(pred == truth, na.rm = TRUE)
  sensitivity <- ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn))
  specificity <- ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp))
  balanced_accuracy <- mean(c(sensitivity, specificity), na.rm = TRUE)
  auc <- auc_roc(truth == positive_class, score_cancer)

  tibble(
    AUC = auc,
    Accuracy = accuracy,
    Sensitivity = sensitivity,
    Specificity = specificity,
    BalancedAccuracy = balanced_accuracy
  )
}

impute_fit <- function(X_train) {
  meds <- apply(X_train, 2, function(v) median(v, na.rm = TRUE))
  meds[is.na(meds)] <- 0
  list(median = meds)
}

impute_apply <- function(X, fit) {
  X2 <- X
  for (j in seq_len(ncol(X2))) {
    idx <- is.na(X2[, j])
    if (any(idx)) X2[idx, j] <- fit$median[j]
  }
  X2
}

standardize_fit <- function(X_train) {
  mu <- colMeans(X_train)
  sdv <- apply(X_train, 2, sd)
  sdv[sdv == 0 | is.na(sdv)] <- 1
  list(mu = mu, sd = sdv)
}

standardize_apply <- function(X, fit) {
  sweep(sweep(X, 2, fit$mu, "-"), 2, fit$sd, "/")
}

select_de_genes <- function(X_train_full, y_train, top_n = 50) {
  y_train <- factor(y_train, levels = c(negative_class, positive_class))
  design <- model.matrix(~ y_train)

  fit <- eBayes(lmFit(t(X_train_full), design))
  tt <- topTable(fit, n = top_n, sort.by = "P")

  intersect(rownames(tt), colnames(X_train_full))
}

calc_density_overlap <- function(df, score_col = "ensemble_risk_score") {
  x_cancer <- df %>% filter(truth == positive_class) %>% pull(.data[[score_col]])
  x_normal <- df %>% filter(truth == negative_class) %>% pull(.data[[score_col]])

  x_cancer <- x_cancer[is.finite(x_cancer)]
  x_normal <- x_normal[is.finite(x_normal)]

  if (length(x_cancer) < 2 || length(x_normal) < 2) return(NA_real_)

  x_min <- min(c(x_cancer, x_normal))
  x_max <- max(c(x_cancer, x_normal))

  d_cancer <- density(x_cancer, from = x_min, to = x_max, n = 1000, adjust = 1.2)
  d_normal <- density(x_normal, from = x_min, to = x_max, n = 1000, adjust = 1.2)

  overlap_y <- pmin(d_cancer$y, d_normal$y)
  grid_x <- seq(x_min, x_max, length.out = 1000)

  sum(diff(grid_x) * head(overlap_y, -1))
}

# -----------------------------
# 2. Prepare binary data
# -----------------------------
ok_bin <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""

X_all <- t(eMat_big_bin[, ok_bin, drop = FALSE])
y_all <- factor(pMat_big_bin$label_binary[ok_bin], levels = c(negative_class, positive_class))
dataset_all <- as.character(pMat_big_bin$dataset_id[ok_bin])

valid <- !is.na(y_all)
X_all <- X_all[valid, , drop = FALSE]
y_all <- droplevels(y_all[valid])
dataset_all <- dataset_all[valid]

cat("Total samples:", nrow(X_all), "\n")
cat("Total genes:", ncol(X_all), "\n")
cat("Class distribution:\n")
print(table(y_all))
cat("Dataset distribution:\n")
print(table(dataset_all))

# -----------------------------
# 3. Stratified 75/25 split within each dataset and class
# -----------------------------
split_stratified_by_dataset <- function(y, dataset, train_prop = 0.75) {
  dev_idx <- c()
  test_idx <- c()

  for (ds in unique(dataset)) {
    for (cl in unique(as.character(y))) {
      idx <- which(dataset == ds & as.character(y) == cl)
      if (length(idx) == 0) next

      idx <- sample(idx)

      if (length(idx) == 1) {
        dev_idx <- c(dev_idx, idx)
      } else {
        n_dev <- floor(length(idx) * train_prop)
        n_dev <- max(1, min(n_dev, length(idx) - 1))

        dev_idx <- c(dev_idx, idx[seq_len(n_dev)])
        test_idx <- c(test_idx, idx[(n_dev + 1):length(idx)])
      }
    }
  }

  list(
    dev_idx = sort(dev_idx),
    test_idx = sort(test_idx)
  )
}

split_obj <- split_stratified_by_dataset(y_all, dataset_all, train_prop = 0.75)

dev_idx <- split_obj$dev_idx
test_idx <- split_obj$test_idx

X_dev <- X_all[dev_idx, , drop = FALSE]
y_dev <- y_all[dev_idx]
dataset_dev <- dataset_all[dev_idx]

X_test <- X_all[test_idx, , drop = FALSE]
y_test <- y_all[test_idx]
dataset_test <- dataset_all[test_idx]

cat("\nDevelopment set:\n")
print(table(y_dev))
print(table(dataset_dev))

cat("\nHold-out test set:\n")
print(table(y_test))
print(table(dataset_test))

# -----------------------------
# 4. Fit models and predict Cancer probability
# -----------------------------
fit_predict_models <- function(X_train_full, y_train, X_test_full, model_set = candidate_models) {

  sel_genes <- select_de_genes(X_train_full, y_train, top_n = top_n_genes)

  if (length(sel_genes) < 5) {
    stop("Too few selected genes.")
  }

  X_train <- X_train_full[, sel_genes, drop = FALSE]
  X_test <- X_test_full[, sel_genes, drop = FALSE]

  imp <- impute_fit(X_train)
  X_train_i <- impute_apply(X_train, imp)
  X_test_i <- impute_apply(X_test, imp)

  sc <- standardize_fit(X_train_i)
  X_train_s <- standardize_apply(X_train_i, sc)
  X_test_s <- standardize_apply(X_test_i, sc)

  y_train <- factor(y_train, levels = c(negative_class, positive_class))

  out <- list()

  if ("RF" %in% model_set) {
    rf_fit <- randomForest(
      x = X_train_i,
      y = y_train,
      ntree = 500
    )
    prob_rf <- predict(rf_fit, X_test_i, type = "prob")
    out[["RF"]] <- prob_rf[, positive_class]
  }

  if ("SVM_RBF" %in% model_set) {
    svm_rbf_fit <- e1071::svm(
      x = X_train_s,
      y = y_train,
      kernel = "radial",
      cost = 1,
      gamma = 1 / ncol(X_train_s),
      probability = TRUE
    )
    pred_svm_rbf <- predict(svm_rbf_fit, X_test_s, probability = TRUE)
    prob_svm_rbf <- attr(pred_svm_rbf, "probabilities")
    out[["SVM_RBF"]] <- prob_svm_rbf[, positive_class]
  }

  if ("SVM_LINEAR" %in% model_set) {
    svm_linear_fit <- e1071::svm(
      x = X_train_s,
      y = y_train,
      kernel = "linear",
      cost = 1,
      probability = TRUE
    )
    pred_svm_linear <- predict(svm_linear_fit, X_test_s, probability = TRUE)
    prob_svm_linear <- attr(pred_svm_linear, "probabilities")
    out[["SVM_LINEAR"]] <- prob_svm_linear[, positive_class]
  }

  if ("LASSO" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    lasso_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 1,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(lasso_fit)) {
      out[["LASSO"]] <- as.numeric(
        predict(lasso_fit, newx = as.matrix(X_test_s), type = "response", s = "lambda.min")
      )
    }
  }

  if ("Ridge" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    ridge_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 0,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(ridge_fit)) {
      out[["Ridge"]] <- as.numeric(
        predict(ridge_fit, newx = as.matrix(X_test_s), type = "response", s = "lambda.min")
      )
    }
  }

  if ("ElasticNet" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    enet_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 0.5,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(enet_fit)) {
      out[["ElasticNet"]] <- as.numeric(
        predict(enet_fit, newx = as.matrix(X_test_s), type = "response", s = "lambda.min")
      )
    }
  }

  if ("kNN" %in% model_set) {
    pred_knn <- class::knn(
      train = X_train_s,
      test = X_test_s,
      cl = y_train,
      k = 7,
      prob = TRUE
    )

    win_prob <- attr(pred_knn, "prob")
    score_knn <- ifelse(pred_knn == positive_class, win_prob, 1 - win_prob)
    out[["kNN"]] <- score_knn
  }

  if ("Logistic" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    df_train <- data.frame(y_num = y_num, X_train_s, check.names = TRUE)
    df_test <- data.frame(X_test_s, check.names = TRUE)

    logistic_fit <- tryCatch(
      glm(y_num ~ ., data = df_train, family = binomial()),
      error = function(e) NULL
    )

    if (!is.null(logistic_fit)) {
      score_logistic <- tryCatch(
        as.numeric(predict(logistic_fit, newdata = df_test, type = "response")),
        error = function(e) rep(NA_real_, nrow(X_test_s))
      )
      out[["Logistic"]] <- score_logistic
    }
  }

  if ("NaiveBayes" %in% model_set) {
    nb_fit <- tryCatch(
      e1071::naiveBayes(x = X_train_s, y = y_train),
      error = function(e) NULL
    )

    if (!is.null(nb_fit)) {
      prob_nb <- tryCatch(
        predict(nb_fit, X_test_s, type = "raw"),
        error = function(e) NULL
      )

      if (!is.null(prob_nb) && positive_class %in% colnames(prob_nb)) {
        out[["NaiveBayes"]] <- prob_nb[, positive_class]
      }
    }
  }

  out <- lapply(out, clip_prob)
  as_tibble(out)
}

# -----------------------------
# 5. Development LODO
# -----------------------------
lodo_scores_list <- list()
lodo_metrics_list <- list()

for (ds in unique(dataset_dev)) {
  cat("Running development LODO fold:", ds, "\n")

  test_id <- which(dataset_dev == ds)
  train_id <- setdiff(seq_along(y_dev), test_id)

  y_train <- y_dev[train_id]
  y_fold <- y_dev[test_id]

  if (length(unique(y_train)) < 2 || length(unique(y_fold)) < 2) {
    cat("Skipping fold", ds, "because train/test has only one class.\n")
    next
  }

  pred_scores <- fit_predict_models(
    X_train_full = X_dev[train_id, , drop = FALSE],
    y_train = y_train,
    X_test_full = X_dev[test_id, , drop = FALSE],
    model_set = candidate_models
  )

  available_models <- intersect(candidate_models, colnames(pred_scores))

  fold_scores <- pred_scores %>%
    mutate(
      fold = ds,
      sample_index = seq_len(n()),
      truth = as.character(y_fold)
    ) %>%
    pivot_longer(
      cols = all_of(available_models),
      names_to = "model",
      values_to = "score_cancer"
    )

  lodo_scores_list[[length(lodo_scores_list) + 1]] <- fold_scores

  fold_metrics <- fold_scores %>%
    group_by(model) %>%
    summarise(
      AUC = binary_metrics_cancer(truth, score_cancer)$AUC,
      Accuracy = binary_metrics_cancer(truth, score_cancer)$Accuracy,
      Sensitivity = binary_metrics_cancer(truth, score_cancer)$Sensitivity,
      Specificity = binary_metrics_cancer(truth, score_cancer)$Specificity,
      BalancedAccuracy = binary_metrics_cancer(truth, score_cancer)$BalancedAccuracy,
      .groups = "drop"
    ) %>%
    mutate(fold = ds)

  lodo_metrics_list[[length(lodo_metrics_list) + 1]] <- fold_metrics
}

lodo_scores_dev <- bind_rows(lodo_scores_list)
lodo_metrics_dev <- bind_rows(lodo_metrics_list)

cat("\n===== Development LODO metrics =====\n")
print(lodo_metrics_dev)

# -----------------------------
# 6. Model ranking by mean LODO AUC
# -----------------------------
model_rank_dev <- lodo_metrics_dev %>%
  group_by(model) %>%
  summarise(
    mean_LODO_AUC = mean(AUC, na.rm = TRUE),
    sd_LODO_AUC = sd(AUC, na.rm = TRUE),
    mean_BalancedAccuracy = mean(BalancedAccuracy, na.rm = TRUE),
    mean_Sensitivity = mean(Sensitivity, na.rm = TRUE),
    mean_Specificity = mean(Specificity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy))

cat("\n===== Model ranking by mean development LODO AUC =====\n")
print(model_rank_dev)

# -----------------------------
# 7. Fixed groups by Top-K
# -----------------------------
fixed_groups <- list()

for (k in top_k_vec) {
  selected <- model_rank_dev %>%
    arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy)) %>%
    slice_head(n = k)

  if (nrow(selected) == 0) {
    cat("No model selected for Top-K =", k, "\n")
    next
  }

  fixed_groups[[as.character(k)]] <- selected
}

cat("\n===== Fixed model groups by Top-K =====\n")
print(fixed_groups)

# -----------------------------
# 8. Weighting strategy function
# -----------------------------
make_weights <- function(selected_df, weight_method) {
  df <- selected_df %>%
    arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy), desc(mean_Sensitivity)) %>%
    mutate(
      rank = row_number(),
      rank_score = rev(seq_len(n()))
    )

  if (weight_method == "rank_auc") {
    df <- df %>%
      mutate(raw_weight = mean_LODO_AUC * rank_score)

  } else if (weight_method == "softmax_auc") {
    df <- df %>%
      mutate(raw_weight = exp(mean_LODO_AUC / T))

  } else if (weight_method == "auc_x_balacc") {
    df <- df %>%
      mutate(raw_weight = exp((mean_LODO_AUC * mean_BalancedAccuracy) / T))

  } else if (weight_method == "auc_minus_spec_penalty") {
    df <- df %>%
      mutate(raw_weight = exp((mean_LODO_AUC - lambda * (1 - mean_Specificity)) / T))

  } else if (weight_method == "equal_weight") {
    df <- df %>%
      mutate(raw_weight = 1)

  } else {
    stop("Unknown weight method: ", weight_method)
  }

  df %>%
    mutate(
      raw_weight = ifelse(is.na(raw_weight) | raw_weight <= 0, 1e-8, raw_weight),
      final_weight = raw_weight / sum(raw_weight)
    )
}

# -----------------------------
# 9. Hold-out test evaluation
# -----------------------------
all_holdout_scores <- list()
all_holdout_metrics <- list()
all_holdout_weights <- list()

for (k in names(fixed_groups)) {
  selected_df <- fixed_groups[[k]]
  selected_models <- selected_df$model

  cat("\nTraining fixed models for Top-K:", k, "\n")
  cat("Selected models:", paste(selected_models, collapse = ", "), "\n")

  pred_scores_test <- fit_predict_models(
    X_train_full = X_dev,
    y_train = y_dev,
    X_test_full = X_test,
    model_set = selected_models
  )

  pred_scores_test <- pred_scores_test %>%
    mutate(
      sample_index = seq_len(n()),
      truth = as.character(y_test)
    )

  for (wm in weight_methods) {
    weight_df <- make_weights(selected_df, wm)

    score_cols <- intersect(weight_df$model, colnames(pred_scores_test))

    if (length(score_cols) == 0) next

    weights_use <- weight_df$final_weight[match(score_cols, weight_df$model)]
    weights_use <- weights_use / sum(weights_use)

    score_mat <- as.matrix(pred_scores_test[, score_cols, drop = FALSE])
    ensemble_score <- as.numeric(score_mat %*% weights_use)
    ensemble_score <- clip_prob(ensemble_score)

    score_df <- pred_scores_test %>%
      select(sample_index, truth, all_of(score_cols)) %>%
      mutate(
        top_k = as.numeric(k),
        weight_method = wm,
        combo = paste0("top", k, " | ", wm),
        ensemble_score_cancer = ensemble_score,
        ensemble_risk_score = qlogis(ensemble_score)
      )

    metric_df <- binary_metrics_cancer(
      truth = score_df$truth,
      score_cancer = score_df$ensemble_score_cancer,
      threshold = 0.5
    ) %>%
      mutate(
        top_k = as.numeric(k),
        weight_method = wm,
        combo = paste0("top", k, " | ", wm),
        n_models = length(score_cols),
        selected_models = paste(score_cols, collapse = " + ")
      )

    weight_out <- weight_df %>%
      select(model, mean_LODO_AUC, mean_BalancedAccuracy, mean_Sensitivity, mean_Specificity, final_weight) %>%
      mutate(
        top_k = as.numeric(k),
        weight_method = wm,
        combo = paste0("top", k, " | ", wm)
      )

    all_holdout_scores[[length(all_holdout_scores) + 1]] <- score_df
    all_holdout_metrics[[length(all_holdout_metrics) + 1]] <- metric_df
    all_holdout_weights[[length(all_holdout_weights) + 1]] <- weight_out
  }
}

holdout_scores_df <- bind_rows(all_holdout_scores)
holdout_metrics_df <- bind_rows(all_holdout_metrics)
holdout_weights_df <- bind_rows(all_holdout_weights)

if (nrow(holdout_scores_df) == 0) {
  stop("No hold-out ensemble result generated. Check Top-K settings.")
}

holdout_metrics_df <- holdout_metrics_df %>%
  arrange(desc(BalancedAccuracy), desc(AUC), desc(Sensitivity))

cat("\n===== Hold-out test performance =====\n")
print(holdout_metrics_df)

cat("\n===== Hold-out ensemble weights =====\n")
print(holdout_weights_df)

# -----------------------------
# 10. Density overlap ranking
# -----------------------------
overlap_df <- holdout_scores_df %>%
  group_by(combo, top_k, weight_method) %>%
  summarise(
    overlap_area = calc_density_overlap(cur_data(), score_col = "ensemble_risk_score"),
    cancer_median = median(ensemble_risk_score[truth == positive_class], na.rm = TRUE),
    normal_median = median(ensemble_risk_score[truth == negative_class], na.rm = TRUE),
    cancer_right_rate = mean(ensemble_risk_score[truth == positive_class] > 0, na.rm = TRUE),
    normal_right_rate = mean(ensemble_risk_score[truth == negative_class] > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(cancer_median > normal_median) %>%
  arrange(overlap_area, normal_right_rate)

cat("\n===== Density overlap ranking =====\n")
print(overlap_df)

# -----------------------------
# 11. Plots
# -----------------------------

ggplot(model_rank_dev, aes(x = reorder(model, mean_LODO_AUC), y = mean_LODO_AUC)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  theme_bw() +
  labs(
    x = "Model",
    y = "Mean development LODO AUC",
    title = "Model ranking based on 75% development LODO"
  )

holdout_metrics_long <- holdout_metrics_df %>%
  select(combo, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy) %>%
  pivot_longer(
    cols = c(AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy),
    names_to = "metric",
    values_to = "value"
  )

ggplot(holdout_metrics_long, aes(x = reorder(combo, value), y = value)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  facet_wrap(~ metric, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Combination",
    y = "Value",
    title = "Hold-out test performance of Top-K fixed ensemble combinations"
  )

ggplot(holdout_scores_df, aes(x = ensemble_risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.2, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_grid(weight_method ~ top_k, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Ensemble risk score: log-odds of Cancer",
    y = "Density",
    title = "NOT BATCH: Top-K Ensemble Risk Calculator Density Comparison"
  )

if (nrow(overlap_df) > 0) {
  best_overlap_combo <- overlap_df$combo[1]

  df_best_overlap <- holdout_scores_df %>%
    filter(combo == best_overlap_combo)

  ggplot(df_best_overlap, aes(x = ensemble_risk_score, fill = truth)) +
    geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
    theme_bw(base_size = 15) +
    labs(
      title = paste0("Minimum Density Overlap on Hold-out Test: ", best_overlap_combo),
      x = "Ensemble risk score: log-odds of Cancer",
      y = "Density"
    )
}

# -----------------------------
# 12. Final summaries
# -----------------------------
best_by_balacc <- holdout_metrics_df %>% slice(1)

cat("\n===== Best hold-out combination by Balanced Accuracy =====\n")
print(best_by_balacc)

if (nrow(overlap_df) > 0) {
  cat("\n===== Best hold-out combination by minimum density overlap =====\n")
  print(overlap_df %>% slice(1))
}

cat("\n===== Top 10 hold-out combinations =====\n")
print(holdout_metrics_df %>% head(10))

# -----------------------------
# 13. Print final score formulas
# -----------------------------
make_formula_text <- function(weight_table, combo_name) {
  df <- weight_table %>%
    filter(combo == combo_name) %>%
    arrange(desc(final_weight))

  if (nrow(df) == 0) return("No formula available.")

  parts <- paste0(
    round(df$final_weight, 4),
    " * ",
    df$model
  )

  paste0(
    "FINAL SCORE = ",
    paste(parts, collapse = " + ")
  )
}

best_ba_combo <- best_by_balacc$combo[1]

formula_best_ba <- make_formula_text(
  weight_table = holdout_weights_df,
  combo_name = best_ba_combo
)

cat("\n===== Final score formula: Best by Balanced Accuracy =====\n")
cat("Combination:", best_ba_combo, "\n")
cat(formula_best_ba, "\n")

if (nrow(overlap_df) > 0) {
  best_overlap_combo <- overlap_df$combo[1]

  formula_best_overlap <- make_formula_text(
    weight_table = holdout_weights_df,
    combo_name = best_overlap_combo
  )

  cat("\n===== Final score formula: Best by Minimum Density Overlap =====\n")
  cat("Combination:", best_overlap_combo, "\n")
  cat(formula_best_overlap, "\n")
}
# -----------------------------
# 14. Single best model hold-out performance
# -----------------------------

single_best_model <- model_rank_dev %>%
  arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy)) %>%
  slice(1) %>%
  pull(model)

cat("\n===== Single Best Model selected by development LODO mean AUC =====\n")
cat("Single best model:", single_best_model, "\n")

single_pred_test <- fit_predict_models(
  X_train_full = X_dev,
  y_train = y_dev,
  X_test_full = X_test,
  model_set = single_best_model
)

single_score <- single_pred_test[[single_best_model]]
single_score <- clip_prob(single_score)

single_metrics <- binary_metrics_cancer(
  truth = as.character(y_test),
  score_cancer = single_score,
  threshold = 0.5
) %>%
  mutate(
    model = single_best_model,
    selected_by = "highest mean development LODO AUC"
  )

cat("\n===== Single Best Model Hold-out Test Performance =====\n")
print(single_metrics)

single_score_df <- tibble(
  sample_index = seq_along(single_score),
  truth = as.character(y_test),
  model = single_best_model,
  score_cancer = single_score,
  risk_score = qlogis(single_score)
)

ggplot(single_score_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  theme_bw(base_size = 15) +
  labs(
    title = paste0("Single Best Model Density on Hold-out Test: ", single_best_model),
    x = "Risk score: log-odds of Cancer",
    y = "Density"
  )

# -----------------------------
# 15. Compare single best vs best ensemble
# -----------------------------

best_ensemble_metrics <- best_by_balacc %>%
  mutate(method = "Best Ensemble by Balanced Accuracy") %>%
  select(method, combo, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy)

single_compare_metrics <- single_metrics %>%
  mutate(
    method = "Single Best Model",
    combo = single_best_model
  ) %>%
  select(method, combo, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy)

comparison_table <- bind_rows(
  single_compare_metrics,
  best_ensemble_metrics
)

cat("\n===== Single Best Model vs Best Ensemble Comparison =====\n")
print(comparison_table)
```

Not Batch ===== Final score formula: Best by Balanced Accuracy ===== Combination: top6 | softmax_auc FINAL SCORE = 0.1898 * RF + 0.1782 * Ridge + 0.1662 * LASSO + 0.1655 * ElasticNet + 0.1529 * SVM_LINEAR + 0.1473 * NaiveBayes 
===== Final score formula: Best by Minimum Density Overlap ===== Combination: top4 | softmax_auc FINAL SCORE = 0.2713 * RF + 0.2547 * Ridge + 0.2375 * LASSO + 0.2365 * ElasticNet

# batched ensemble risk calculator

```{r}
# ============================================================
# BATCH CORRECTION Ensemble Risk Calculator
# Top-K fixed ensemble version
# 75% development + 25% hold-out test
# Development: LODO mean AUC for model ranking
# Hold-out: evaluate Top-K fixed ensemble combinations
# ============================================================

library(dplyr)
library(tidyr)
library(tibble)
library(limma)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(glmnet)
library(sva)

set.seed(3888)

# -----------------------------
# 0. Settings
# -----------------------------
positive_class <- "Cancer"
negative_class <- "Normal"

top_n_genes <- 50
top_k_vec <- c(4, 5, 6, 7)

weight_methods <- c(
  "rank_auc",
  "softmax_auc",
  "auc_x_balacc",
  "auc_minus_spec_penalty",
  "equal_weight"
)

T <- 0.20
lambda <- 0.50

candidate_models <- c(
  "RF", "SVM_RBF", "SVM_LINEAR",
  "LASSO", "Ridge", "ElasticNet",
  "kNN", "Logistic", "NaiveBayes"
)

# -----------------------------
# 1. Helper functions
# -----------------------------
clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)

  pos <- truth_pos == 1
  neg <- truth_pos == 0

  n_pos <- sum(pos)
  n_neg <- sum(neg)

  if (n_pos == 0 || n_neg == 0) return(NA_real_)

  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

binary_metrics_cancer <- function(truth, score_cancer, threshold = 0.5) {
  truth <- as.character(truth)
  pred <- ifelse(score_cancer >= threshold, positive_class, negative_class)

  tp <- sum(pred == positive_class & truth == positive_class, na.rm = TRUE)
  fn <- sum(pred != positive_class & truth == positive_class, na.rm = TRUE)
  tn <- sum(pred == negative_class & truth == negative_class, na.rm = TRUE)
  fp <- sum(pred != negative_class & truth == negative_class, na.rm = TRUE)

  accuracy <- mean(pred == truth, na.rm = TRUE)
  sensitivity <- ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn))
  specificity <- ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp))
  balanced_accuracy <- mean(c(sensitivity, specificity), na.rm = TRUE)
  auc <- auc_roc(truth == positive_class, score_cancer)

  tibble(
    AUC = auc,
    Accuracy = accuracy,
    Sensitivity = sensitivity,
    Specificity = specificity,
    BalancedAccuracy = balanced_accuracy
  )
}

impute_fit <- function(X_train) {
  meds <- apply(X_train, 2, function(v) median(v, na.rm = TRUE))
  meds[is.na(meds)] <- 0
  list(median = meds)
}

impute_apply <- function(X, fit) {
  X2 <- X
  for (j in seq_len(ncol(X2))) {
    idx <- is.na(X2[, j])
    if (any(idx)) X2[idx, j] <- fit$median[j]
  }
  X2
}

standardize_fit <- function(X_train) {
  mu <- colMeans(X_train)
  sdv <- apply(X_train, 2, sd)
  sdv[sdv == 0 | is.na(sdv)] <- 1
  list(mu = mu, sd = sdv)
}

standardize_apply <- function(X, fit) {
  sweep(sweep(X, 2, fit$mu, "-"), 2, fit$sd, "/")
}

# Training-only batch correction
combat_fit_apply <- function(expr_train, batch_train, y_train, expr_test, batch_test) {
  batch_train <- as.character(batch_train)
  batch_test <- as.character(batch_test)

  if (length(unique(batch_train)) < 2) {
    message("Only one training batch; skipping batch correction.")
    return(list(train = expr_train, test = expr_test))
  }

  mod_train <- model.matrix(~ y_train)

  corrected_train <- tryCatch(
    sva::ComBat(
      dat = expr_train,
      batch = batch_train,
      mod = mod_train,
      par.prior = TRUE,
      mean.only = FALSE
    ),
    error = function(e) {
      message("ComBat failed; using limma::removeBatchEffect for train. Error: ", conditionMessage(e))
      limma::removeBatchEffect(expr_train, batch = batch_train, design = mod_train)
    }
  )

  train_batches <- unique(batch_train)
  batch_train_f <- factor(batch_train, levels = train_batches)

  batch_means_train <- sapply(train_batches, function(b) {
    idx <- which(batch_train_f == b)
    rowMeans(expr_train[, idx, drop = FALSE], na.rm = TRUE)
  })

  grand_mean_train <- rowMeans(expr_train, na.rm = TRUE)
  expr_test_corrected <- expr_test

  for (b in train_batches) {
    test_idx <- which(batch_test == b)
    if (length(test_idx) == 0) next

    batch_effect <- batch_means_train[, b] - grand_mean_train

    expr_test_corrected[, test_idx] <-
      sweep(expr_test[, test_idx, drop = FALSE], 1, batch_effect, "-")
  }

  novel_test_idx <- which(!(batch_test %in% train_batches))

  if (length(novel_test_idx) > 0) {
    message("Test contains unseen batch(es); applying mean-centering fallback.")

    sample_means <- colMeans(expr_test[, novel_test_idx, drop = FALSE], na.rm = TRUE)

    expr_test_corrected[, novel_test_idx] <-
      sweep(expr_test[, novel_test_idx, drop = FALSE], 2, sample_means, "-") +
      mean(grand_mean_train, na.rm = TRUE)
  }

  list(
    train = corrected_train,
    test = expr_test_corrected
  )
}

select_de_genes <- function(X_train_full, y_train, top_n = 50) {
  y_train <- factor(y_train, levels = c(negative_class, positive_class))
  design <- model.matrix(~ y_train)

  fit <- eBayes(lmFit(t(X_train_full), design))
  tt <- topTable(fit, n = top_n, sort.by = "P")

  intersect(rownames(tt), colnames(X_train_full))
}

calc_density_overlap <- function(df, score_col = "ensemble_risk_score") {
  x_cancer <- df %>% filter(truth == positive_class) %>% pull(.data[[score_col]])
  x_normal <- df %>% filter(truth == negative_class) %>% pull(.data[[score_col]])

  x_cancer <- x_cancer[is.finite(x_cancer)]
  x_normal <- x_normal[is.finite(x_normal)]

  if (length(x_cancer) < 2 || length(x_normal) < 2) return(NA_real_)

  x_min <- min(c(x_cancer, x_normal))
  x_max <- max(c(x_cancer, x_normal))

  d_cancer <- density(x_cancer, from = x_min, to = x_max, n = 1000, adjust = 1.2)
  d_normal <- density(x_normal, from = x_min, to = x_max, n = 1000, adjust = 1.2)

  overlap_y <- pmin(d_cancer$y, d_normal$y)
  grid_x <- seq(x_min, x_max, length.out = 1000)

  sum(diff(grid_x) * head(overlap_y, -1))
}

# -----------------------------
# 2. Prepare binary data
# -----------------------------
ok_bin <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""

X_all <- t(eMat_big_bin[, ok_bin, drop = FALSE])
y_all <- factor(pMat_big_bin$label_binary[ok_bin], levels = c(negative_class, positive_class))
dataset_all <- as.character(pMat_big_bin$dataset_id[ok_bin])

valid <- !is.na(y_all)
X_all <- X_all[valid, , drop = FALSE]
y_all <- droplevels(y_all[valid])
dataset_all <- dataset_all[valid]

cat("Total samples:", nrow(X_all), "\n")
cat("Total genes:", ncol(X_all), "\n")
cat("Class distribution:\n")
print(table(y_all))
cat("Dataset distribution:\n")
print(table(dataset_all))

# -----------------------------
# 3. Stratified 75/25 split
# -----------------------------
split_stratified_by_dataset <- function(y, dataset, train_prop = 0.75) {
  dev_idx <- c()
  test_idx <- c()

  for (ds in unique(dataset)) {
    for (cl in unique(as.character(y))) {
      idx <- which(dataset == ds & as.character(y) == cl)
      if (length(idx) == 0) next

      idx <- sample(idx)

      if (length(idx) == 1) {
        dev_idx <- c(dev_idx, idx)
      } else {
        n_dev <- floor(length(idx) * train_prop)
        n_dev <- max(1, min(n_dev, length(idx) - 1))

        dev_idx <- c(dev_idx, idx[seq_len(n_dev)])
        test_idx <- c(test_idx, idx[(n_dev + 1):length(idx)])
      }
    }
  }

  list(dev_idx = sort(dev_idx), test_idx = sort(test_idx))
}

split_obj <- split_stratified_by_dataset(y_all, dataset_all, train_prop = 0.75)

dev_idx <- split_obj$dev_idx
test_idx <- split_obj$test_idx

X_dev <- X_all[dev_idx, , drop = FALSE]
y_dev <- y_all[dev_idx]
dataset_dev <- dataset_all[dev_idx]

X_test <- X_all[test_idx, , drop = FALSE]
y_test <- y_all[test_idx]
dataset_test <- dataset_all[test_idx]

cat("\nDevelopment set:\n")
print(table(y_dev))
print(table(dataset_dev))

cat("\nHold-out test set:\n")
print(table(y_test))
print(table(dataset_test))

# -----------------------------
# 4. Fit models and predict Cancer probability WITH batch correction
# -----------------------------
fit_predict_models <- function(
    X_train_full,
    y_train,
    batch_train,
    X_test_full,
    batch_test,
    model_set = candidate_models
) {

  y_train <- factor(y_train, levels = c(negative_class, positive_class))

  # batch correction before DE gene selection
  expr_train_raw <- t(X_train_full)
  expr_test_raw <- t(X_test_full)

  cb <- combat_fit_apply(
    expr_train = expr_train_raw,
    batch_train = batch_train,
    y_train = y_train,
    expr_test = expr_test_raw,
    batch_test = batch_test
  )

  X_train_bc_full <- t(cb$train)
  X_test_bc_full <- t(cb$test)

  sel_genes <- select_de_genes(X_train_bc_full, y_train, top_n = top_n_genes)

  if (length(sel_genes) < 5) {
    stop("Too few selected genes.")
  }

  X_train <- X_train_bc_full[, sel_genes, drop = FALSE]
  X_test <- X_test_bc_full[, sel_genes, drop = FALSE]

  imp <- impute_fit(X_train)
  X_train_i <- impute_apply(X_train, imp)
  X_test_i <- impute_apply(X_test, imp)

  sc <- standardize_fit(X_train_i)
  X_train_s <- standardize_apply(X_train_i, sc)
  X_test_s <- standardize_apply(X_test_i, sc)

  out <- list()

  if ("RF" %in% model_set) {
    rf_fit <- randomForest(x = X_train_i, y = y_train, ntree = 500)
    prob_rf <- predict(rf_fit, X_test_i, type = "prob")
    out[["RF"]] <- prob_rf[, positive_class]
  }

  if ("SVM_RBF" %in% model_set) {
    svm_rbf_fit <- e1071::svm(
      x = X_train_s,
      y = y_train,
      kernel = "radial",
      cost = 1,
      gamma = 1 / ncol(X_train_s),
      probability = TRUE
    )
    pred_svm_rbf <- predict(svm_rbf_fit, X_test_s, probability = TRUE)
    prob_svm_rbf <- attr(pred_svm_rbf, "probabilities")
    out[["SVM_RBF"]] <- prob_svm_rbf[, positive_class]
  }

  if ("SVM_LINEAR" %in% model_set) {
    svm_linear_fit <- e1071::svm(
      x = X_train_s,
      y = y_train,
      kernel = "linear",
      cost = 1,
      probability = TRUE
    )
    pred_svm_linear <- predict(svm_linear_fit, X_test_s, probability = TRUE)
    prob_svm_linear <- attr(pred_svm_linear, "probabilities")
    out[["SVM_LINEAR"]] <- prob_svm_linear[, positive_class]
  }

  if ("LASSO" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    lasso_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 1,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(lasso_fit)) {
      out[["LASSO"]] <- as.numeric(
        predict(lasso_fit, newx = as.matrix(X_test_s), type = "response", s = "lambda.min")
      )
    }
  }

  if ("Ridge" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    ridge_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 0,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(ridge_fit)) {
      out[["Ridge"]] <- as.numeric(
        predict(ridge_fit, newx = as.matrix(X_test_s), type = "response", s = "lambda.min")
      )
    }
  }

  if ("ElasticNet" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    enet_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 0.5,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(enet_fit)) {
      out[["ElasticNet"]] <- as.numeric(
        predict(enet_fit, newx = as.matrix(X_test_s), type = "response", s = "lambda.min")
      )
    }
  }

  if ("kNN" %in% model_set) {
    pred_knn <- class::knn(
      train = X_train_s,
      test = X_test_s,
      cl = y_train,
      k = 7,
      prob = TRUE
    )

    win_prob <- attr(pred_knn, "prob")
    score_knn <- ifelse(pred_knn == positive_class, win_prob, 1 - win_prob)
    out[["kNN"]] <- score_knn
  }

  if ("Logistic" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    df_train <- data.frame(y_num = y_num, X_train_s, check.names = TRUE)
    df_test <- data.frame(X_test_s, check.names = TRUE)

    logistic_fit <- tryCatch(
      glm(y_num ~ ., data = df_train, family = binomial()),
      error = function(e) NULL
    )

    if (!is.null(logistic_fit)) {
      score_logistic <- tryCatch(
        as.numeric(predict(logistic_fit, newdata = df_test, type = "response")),
        error = function(e) rep(NA_real_, nrow(X_test_s))
      )
      out[["Logistic"]] <- score_logistic
    }
  }

  if ("NaiveBayes" %in% model_set) {
    nb_fit <- tryCatch(
      e1071::naiveBayes(x = X_train_s, y = y_train),
      error = function(e) NULL
    )

    if (!is.null(nb_fit)) {
      prob_nb <- tryCatch(
        predict(nb_fit, X_test_s, type = "raw"),
        error = function(e) NULL
      )

      if (!is.null(prob_nb) && positive_class %in% colnames(prob_nb)) {
        out[["NaiveBayes"]] <- prob_nb[, positive_class]
      }
    }
  }

  out <- lapply(out, clip_prob)
  as_tibble(out)
}

# -----------------------------
# 5. Development LODO
# -----------------------------
lodo_scores_list <- list()
lodo_metrics_list <- list()

for (ds in unique(dataset_dev)) {
  cat("Running batch-corrected development LODO fold:", ds, "\n")

  test_id <- which(dataset_dev == ds)
  train_id <- setdiff(seq_along(y_dev), test_id)

  y_train <- y_dev[train_id]
  y_fold <- y_dev[test_id]

  batch_train <- dataset_dev[train_id]
  batch_fold <- dataset_dev[test_id]

  if (length(unique(y_train)) < 2 || length(unique(y_fold)) < 2) {
    cat("Skipping fold", ds, "because train/test has only one class.\n")
    next
  }

  pred_scores <- fit_predict_models(
    X_train_full = X_dev[train_id, , drop = FALSE],
    y_train = y_train,
    batch_train = batch_train,
    X_test_full = X_dev[test_id, , drop = FALSE],
    batch_test = batch_fold,
    model_set = candidate_models
  )

  available_models <- intersect(candidate_models, colnames(pred_scores))

  fold_scores <- pred_scores %>%
    mutate(
      fold = ds,
      sample_index = seq_len(n()),
      truth = as.character(y_fold)
    ) %>%
    pivot_longer(
      cols = all_of(available_models),
      names_to = "model",
      values_to = "score_cancer"
    )

  lodo_scores_list[[length(lodo_scores_list) + 1]] <- fold_scores

  fold_metrics <- fold_scores %>%
    group_by(model) %>%
    summarise(
      AUC = binary_metrics_cancer(truth, score_cancer)$AUC,
      Accuracy = binary_metrics_cancer(truth, score_cancer)$Accuracy,
      Sensitivity = binary_metrics_cancer(truth, score_cancer)$Sensitivity,
      Specificity = binary_metrics_cancer(truth, score_cancer)$Specificity,
      BalancedAccuracy = binary_metrics_cancer(truth, score_cancer)$BalancedAccuracy,
      .groups = "drop"
    ) %>%
    mutate(fold = ds)

  lodo_metrics_list[[length(lodo_metrics_list) + 1]] <- fold_metrics
}

lodo_scores_dev <- bind_rows(lodo_scores_list)
lodo_metrics_dev <- bind_rows(lodo_metrics_list)

cat("\n===== Batch-corrected development LODO metrics =====\n")
print(lodo_metrics_dev)

# -----------------------------
# 6. Model ranking
# -----------------------------
model_rank_dev <- lodo_metrics_dev %>%
  group_by(model) %>%
  summarise(
    mean_LODO_AUC = mean(AUC, na.rm = TRUE),
    sd_LODO_AUC = sd(AUC, na.rm = TRUE),
    mean_BalancedAccuracy = mean(BalancedAccuracy, na.rm = TRUE),
    mean_Sensitivity = mean(Sensitivity, na.rm = TRUE),
    mean_Specificity = mean(Specificity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy))

cat("\n===== Batch-corrected model ranking by mean development LODO AUC =====\n")
print(model_rank_dev)

# -----------------------------
# 7. Fixed groups by Top-K
# -----------------------------
fixed_groups <- list()

for (k in top_k_vec) {
  selected <- model_rank_dev %>%
    arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy)) %>%
    slice_head(n = k)

  if (nrow(selected) == 0) next

  fixed_groups[[as.character(k)]] <- selected
}

cat("\n===== Batch-corrected fixed model groups by Top-K =====\n")
print(fixed_groups)

# -----------------------------
# 8. Weighting strategy
# -----------------------------
make_weights <- function(selected_df, weight_method) {
  df <- selected_df %>%
    arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy), desc(mean_Sensitivity)) %>%
    mutate(
      rank = row_number(),
      rank_score = rev(seq_len(n()))
    )

  if (weight_method == "rank_auc") {
    df <- df %>% mutate(raw_weight = mean_LODO_AUC * rank_score)

  } else if (weight_method == "softmax_auc") {
    df <- df %>% mutate(raw_weight = exp(mean_LODO_AUC / T))

  } else if (weight_method == "auc_x_balacc") {
    df <- df %>% mutate(raw_weight = exp((mean_LODO_AUC * mean_BalancedAccuracy) / T))

  } else if (weight_method == "auc_minus_spec_penalty") {
    df <- df %>% mutate(raw_weight = exp((mean_LODO_AUC - lambda * (1 - mean_Specificity)) / T))

  } else if (weight_method == "equal_weight") {
    df <- df %>% mutate(raw_weight = 1)

  } else {
    stop("Unknown weight method: ", weight_method)
  }

  df %>%
    mutate(
      raw_weight = ifelse(is.na(raw_weight) | raw_weight <= 0, 1e-8, raw_weight),
      final_weight = raw_weight / sum(raw_weight)
    )
}

# -----------------------------
# 9. Hold-out evaluation
# -----------------------------
all_holdout_scores <- list()
all_holdout_metrics <- list()
all_holdout_weights <- list()

for (k in names(fixed_groups)) {
  selected_df <- fixed_groups[[k]]
  selected_models <- selected_df$model

  cat("\nTraining batch-corrected fixed models for Top-K:", k, "\n")
  cat("Selected models:", paste(selected_models, collapse = ", "), "\n")

  pred_scores_test <- fit_predict_models(
    X_train_full = X_dev,
    y_train = y_dev,
    batch_train = dataset_dev,
    X_test_full = X_test,
    batch_test = dataset_test,
    model_set = selected_models
  )

  pred_scores_test <- pred_scores_test %>%
    mutate(
      sample_index = seq_len(n()),
      truth = as.character(y_test)
    )

  for (wm in weight_methods) {
    weight_df <- make_weights(selected_df, wm)

    score_cols <- intersect(weight_df$model, colnames(pred_scores_test))
    if (length(score_cols) == 0) next

    weights_use <- weight_df$final_weight[match(score_cols, weight_df$model)]
    weights_use <- weights_use / sum(weights_use)

    score_mat <- as.matrix(pred_scores_test[, score_cols, drop = FALSE])
    ensemble_score <- clip_prob(as.numeric(score_mat %*% weights_use))

    score_df <- pred_scores_test %>%
      select(sample_index, truth, all_of(score_cols)) %>%
      mutate(
        top_k = as.numeric(k),
        weight_method = wm,
        combo = paste0("top", k, " | ", wm),
        ensemble_score_cancer = ensemble_score,
        ensemble_risk_score = qlogis(ensemble_score)
      )

    metric_df <- binary_metrics_cancer(
      truth = score_df$truth,
      score_cancer = score_df$ensemble_score_cancer,
      threshold = 0.5
    ) %>%
      mutate(
        top_k = as.numeric(k),
        weight_method = wm,
        combo = paste0("top", k, " | ", wm),
        n_models = length(score_cols),
        selected_models = paste(score_cols, collapse = " + ")
      )

    weight_out <- weight_df %>%
      select(model, mean_LODO_AUC, mean_BalancedAccuracy, mean_Sensitivity, mean_Specificity, final_weight) %>%
      mutate(
        top_k = as.numeric(k),
        weight_method = wm,
        combo = paste0("top", k, " | ", wm)
      )

    all_holdout_scores[[length(all_holdout_scores) + 1]] <- score_df
    all_holdout_metrics[[length(all_holdout_metrics) + 1]] <- metric_df
    all_holdout_weights[[length(all_holdout_weights) + 1]] <- weight_out
  }
}

holdout_scores_df <- bind_rows(all_holdout_scores)
holdout_metrics_df <- bind_rows(all_holdout_metrics)
holdout_weights_df <- bind_rows(all_holdout_weights)

if (nrow(holdout_scores_df) == 0) {
  stop("No batch-corrected hold-out ensemble result generated.")
}

holdout_metrics_df <- holdout_metrics_df %>%
  arrange(desc(BalancedAccuracy), desc(AUC), desc(Sensitivity))

cat("\n===== Batch-corrected hold-out test performance =====\n")
print(holdout_metrics_df)

cat("\n===== Batch-corrected hold-out ensemble weights =====\n")
print(holdout_weights_df)

# -----------------------------
# 10. Density overlap
# -----------------------------
overlap_df <- holdout_scores_df %>%
  group_by(combo, top_k, weight_method) %>%
  summarise(
    overlap_area = calc_density_overlap(cur_data(), score_col = "ensemble_risk_score"),
    cancer_median = median(ensemble_risk_score[truth == positive_class], na.rm = TRUE),
    normal_median = median(ensemble_risk_score[truth == negative_class], na.rm = TRUE),
    cancer_right_rate = mean(ensemble_risk_score[truth == positive_class] > 0, na.rm = TRUE),
    normal_right_rate = mean(ensemble_risk_score[truth == negative_class] > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(cancer_median > normal_median) %>%
  arrange(overlap_area, normal_right_rate)

cat("\n===== Batch-corrected density overlap ranking =====\n")
print(overlap_df)

# -----------------------------
# 11. Plots
# -----------------------------
ggplot(model_rank_dev, aes(x = reorder(model, mean_LODO_AUC), y = mean_LODO_AUC)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  theme_bw() +
  labs(
    x = "Model",
    y = "Mean development LODO AUC",
    title = "BATCH: Model ranking based on 75% development LODO"
  )

holdout_metrics_long <- holdout_metrics_df %>%
  select(combo, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy) %>%
  pivot_longer(
    cols = c(AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy),
    names_to = "metric",
    values_to = "value"
  )

ggplot(holdout_metrics_long, aes(x = reorder(combo, value), y = value)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  facet_wrap(~ metric, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Combination",
    y = "Value",
    title = "BATCH: Hold-out test performance of Top-K fixed ensemble combinations"
  )

ggplot(holdout_scores_df, aes(x = ensemble_risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.2, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_grid(weight_method ~ top_k, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Ensemble risk score: log-odds of Cancer",
    y = "Density",
    title = "BATCH: Top-K Ensemble Risk Calculator Density Comparison"
  )

if (nrow(overlap_df) > 0) {
  best_overlap_combo <- overlap_df$combo[1]

  df_best_overlap <- holdout_scores_df %>%
    filter(combo == best_overlap_combo)

  ggplot(df_best_overlap, aes(x = ensemble_risk_score, fill = truth)) +
    geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
    theme_bw(base_size = 15) +
    labs(
      title = paste0("BATCH: Minimum Density Overlap on Hold-out Test: ", best_overlap_combo),
      x = "Ensemble risk score: log-odds of Cancer",
      y = "Density"
    )
}

# -----------------------------
# 12. Final summaries
# -----------------------------
best_by_balacc <- holdout_metrics_df %>% slice(1)

cat("\n===== BATCH: Best hold-out combination by Balanced Accuracy =====\n")
print(best_by_balacc)

if (nrow(overlap_df) > 0) {
  cat("\n===== BATCH: Best hold-out combination by minimum density overlap =====\n")
  print(overlap_df %>% slice(1))
}

cat("\n===== BATCH: Top 10 hold-out combinations =====\n")
print(holdout_metrics_df %>% head(10))

# -----------------------------
# 13. Print final score formulas
# -----------------------------
make_formula_text <- function(weight_table, combo_name) {
  df <- weight_table %>%
    filter(combo == combo_name) %>%
    arrange(desc(final_weight))

  if (nrow(df) == 0) return("No formula available.")

  parts <- paste0(round(df$final_weight, 4), " * ", df$model)

  paste0("FINAL SCORE = ", paste(parts, collapse = " + "))
}

best_ba_combo <- best_by_balacc$combo[1]

formula_best_ba <- make_formula_text(
  weight_table = holdout_weights_df,
  combo_name = best_ba_combo
)

cat("\n===== BATCH: Final score formula: Best by Balanced Accuracy =====\n")
cat("Combination:", best_ba_combo, "\n")
cat(formula_best_ba, "\n")

if (nrow(overlap_df) > 0) {
  best_overlap_combo <- overlap_df$combo[1]

  formula_best_overlap <- make_formula_text(
    weight_table = holdout_weights_df,
    combo_name = best_overlap_combo
  )

  cat("\n===== BATCH: Final score formula: Best by Minimum Density Overlap =====\n")
  cat("Combination:", best_overlap_combo, "\n")
  cat(formula_best_overlap, "\n")
}

# -----------------------------
# 14. Single best model hold-out performance
# -----------------------------
single_best_model <- model_rank_dev %>%
  arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy)) %>%
  slice(1) %>%
  pull(model)

cat("\n===== BATCH: Single Best Model selected by development LODO mean AUC =====\n")
cat("Single best model:", single_best_model, "\n")

single_pred_test <- fit_predict_models(
  X_train_full = X_dev,
  y_train = y_dev,
  batch_train = dataset_dev,
  X_test_full = X_test,
  batch_test = dataset_test,
  model_set = single_best_model
)

single_score <- single_pred_test[[single_best_model]]
single_score <- clip_prob(single_score)

single_metrics <- binary_metrics_cancer(
  truth = as.character(y_test),
  score_cancer = single_score,
  threshold = 0.5
) %>%
  mutate(
    model = single_best_model,
    selected_by = "highest mean development LODO AUC"
  )

cat("\n===== BATCH: Single Best Model Hold-out Test Performance =====\n")
print(single_metrics)

single_score_df <- tibble(
  sample_index = seq_along(single_score),
  truth = as.character(y_test),
  model = single_best_model,
  score_cancer = single_score,
  risk_score = qlogis(single_score)
)

ggplot(single_score_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  theme_bw(base_size = 15) +
  labs(
    title = paste0("BATCH: Single Best Model Density on Hold-out Test: ", single_best_model),
    x = "Risk score: log-odds of Cancer",
    y = "Density"
  )

# -----------------------------
# 15. Compare single best vs best ensemble
# -----------------------------
best_ensemble_metrics <- best_by_balacc %>%
  mutate(method = "Best Ensemble by Balanced Accuracy") %>%
  select(method, combo, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy)

single_compare_metrics <- single_metrics %>%
  mutate(
    method = "Single Best Model",
    combo = single_best_model
  ) %>%
  select(method, combo, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy)

comparison_table <- bind_rows(
  single_compare_metrics,
  best_ensemble_metrics
)

cat("\n===== BATCH: Single Best Model vs Best Ensemble Comparison =====\n")
print(comparison_table)
```

===== BATCH: Final score formula: Best by Balanced Accuracy ===== Combination: top7 | softmax_auc FINAL SCORE = 0.1601 * Ridge + 0.1564 * RF + 0.1493 * ElasticNet + 0.1475 * LASSO + 0.1351 * SVM_LINEAR + 0.1301 * kNN + 0.1214 * SVM_RBF

===== BATCH: Final score formula: Best by Minimum Density Overlap ===== Combination: top7 | equal_weight FINAL SCORE = 0.1429 * Ridge + 0.1429 * RF + 0.1429 * ElasticNet + 0.1429 * LASSO + 0.1429 * SVM_LINEAR + 0.1429 * kNN + 0.1429 * SVM_RBF

final step: Use all samples to train the models. Use final fomular(best combination) to calculate final score.
