---
title: "Group Breast Cancer EDA + Modelling (Merged Binary GEOs)"
format:
  html:
    embed-resources: true
    code-fold: true
    df-print: paged
    toc: true
editor: visual
---

This notebook builds a **merged binary dataset** from multiple GEO studies and compares multiple class-weighting schemes using leakage-safe leave-one-dataset-out evaluation.

```{r}
#| message: false
#| warning: false
library(GEOquery)
library(limma)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(cvTools)
library(sva)
```

```{r}
#| message: false
#| warning: false
set.seed(3888)
```

## Dataset registry + merge helpers

```{r}
#| message: false
#| warning: false
datasets_multiclass <- c("GSE25066","GSE45827")          # PAM50 multiclass (example; not merged yet)
datasets_binary <- c("GSE15852", "GSE45827","GSE29431","GSE9574", "GSE7904","GSE54002", "GSE25066")  # Binary Normal vs Cancer (merged below)

normal_keywords <- c(
  "normal", "non-tumor", "nontumor", "non tumour", "nontumour",
  "healthy", "control", "adjacent normal", "benign", "none"
)

# Preprocessing knobs
enable_auto_log2 <- TRUE
use_fold_downsampling <- TRUE
downsample_max_ratio <- 3
```

```{r}
#| message: false
#| warning: false
load_geo_eset <- function(gse_id) {
  geo_cache_dir <- file.path("data", "geo-cache")
  if (!dir.exists(geo_cache_dir)) dir.create(geo_cache_dir, recursive = TRUE)

  # Helper: build HTTPS URL for the series matrix
  gse_matrix_url <- function(id, host = "ftp.ncbi.nlm.nih.gov") {
    num <- sub("^GSE", "", id)
    prefix <- substr(num, 1, max(1, nchar(num) - 3))
    series_dir <- paste0("GSE", prefix, "nnn")
    paste0(
      "https://", host, "/geo/series/",
      series_dir, "/", id, "/matrix/",
      id, "_series_matrix.txt.gz"
    )
  }

  # Prefer normal GEOquery path first (uses destdir cache)
  g <- tryCatch(
    GEOquery::getGEO(gse_id, destdir = geo_cache_dir),
    error = function(e) e
  )
  if (!inherits(g, "error")) {
    if (inherits(g, "ExpressionSet")) return(g)
    if (is.list(g) && length(g) >= 1) return(g[[1]])
  }

  # Fallback: explicit HTTPS download with longer timeout + retries + resume, then load from file
  destfile <- file.path(geo_cache_dir, paste0(gse_id, "_series_matrix.txt.gz"))
  urls <- c(
    gse_matrix_url(gse_id, host = "ftp.ncbi.nlm.nih.gov"),
    gse_matrix_url(gse_id, host = "download.ncbi.nlm.nih.gov")
  )

  # R's options(timeout=) does not reliably control libcurl; set curl handle timeout directly.
  h <- curl::new_handle()
  curl::handle_setopt(h,
                      connecttimeout = 60,
                      timeout = 1800,
                      low_speed_time = 300,
                      low_speed_limit = 1000)

  ok <- FALSE
  for (u in urls) {
    for (attempt in 1:6) {
      try({
        size_before <- if (file.exists(destfile)) file.info(destfile)$size else 0
        curl::curl_download(u, destfile, quiet = FALSE, mode = "wb", handle = h, resume = TRUE)
        size_after <- if (file.exists(destfile)) file.info(destfile)$size else 0
        ok <- size_after > 0 && size_after >= size_before
      }, silent = TRUE)
      if (ok) break
      # If we are not making progress, reset the partial file and retry.
      if (file.exists(destfile) && attempt >= 2) file.remove(destfile)
      Sys.sleep(3 * attempt)
    }
    if (ok) break
  }

  if (!ok) {
    stop("Could not download series matrix for ", gse_id, " after retries.\nTried URLs:\n- ",
         paste(urls, collapse = "\n- "))
  }

  g2 <- GEOquery::getGEO(filename = destfile)
  if (inherits(g2, "ExpressionSet")) return(g2)
  if (is.list(g2) && length(g2) >= 1) return(g2[[1]])
  stop("Could not load ExpressionSet for ", gse_id, " from cached file: ", destfile)
}

clean_first_symbol <- function(s) {
  if (is.na(s) || trimws(as.character(s)) == "") return(NA_character_)
  s <- trimws(as.character(s))
  parts <- strsplit(s, " /// ", fixed = TRUE)[[1]]
  trimws(parts[1])
}

needs_log2_transform <- function(e) {
  vals <- as.numeric(e)
  vals <- vals[is.finite(vals)]
  if (length(vals) < 10) return(FALSE)
  qx <- stats::quantile(vals, probs = c(0, 0.25, 0.5, 0.75, 0.99, 1.0), na.rm = TRUE)
  # GEO2R-style heuristic for identifying non-log expression scales.
  (qx[5] > 100) || (qx[6] - qx[1] > 50 && qx[2] > 0) || (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
}

safe_log2_transform <- function(e) {
  pos <- e[e > 0 & is.finite(e)]
  if (!length(pos)) return(e)
  pseudo <- min(pos, na.rm = TRUE) / 2
  e2 <- e
  e2[!is.finite(e2)] <- NA_real_
  e2[e2 <= 0] <- pseudo
  log2(e2)
}

collapse_to_genes <- function(eset) {
  e <- exprs(eset)
  q_raw <- stats::quantile(as.numeric(e), probs = c(0.01, 0.5, 0.99), na.rm = TRUE)
  log2_applied <- isTRUE(enable_auto_log2) && needs_log2_transform(e)
  if (log2_applied) {
    e <- safe_log2_transform(e)
  }
  q_used <- stats::quantile(as.numeric(e), probs = c(0.01, 0.5, 0.99), na.rm = TRUE)
  fd <- fData(eset)
  sym_col <- grep("gene\\s*symbol|symbol", colnames(fd), ignore.case = TRUE, value = TRUE)[1]
  gene_sym_raw <- if (!is.na(sym_col)) as.character(fd[, sym_col]) else rep(NA_character_, nrow(fd))
has_symbol <- !is.na(gene_sym_raw) & trimws(gene_sym_raw) != ""
gene_sym <- rep(NA_character_, length(gene_sym_raw))
gene_sym[has_symbol] <- vapply(gene_sym_raw[has_symbol], clean_first_symbol, character(1))
  gene_sym <- ifelse(is.na(gene_sym) | gene_sym == "", rownames(e), gene_sym)
  e_g <- limma::avereps(e, ID = gene_sym)

  # Feature table: keep one row per gene symbol where possible, otherwise minimal table
  fd$.__gene_sym__ <- gene_sym
  fd_g <- fd %>%
    tibble::as_tibble() %>%
    mutate(.__gene_sym__ = fd$.__gene_sym__) %>%
    group_by(.__gene_sym__) %>%
    slice(1) %>%
    ungroup() %>%
    tibble::column_to_rownames(".__gene_sym__")

  list(
    eMat_g = e_g,
    fMat_g = fd_g,
    pMat = pData(eset),
    log2_applied = log2_applied,
    q01_raw = as.numeric(q_raw[1]),
    q50_raw = as.numeric(q_raw[2]),
    q99_raw = as.numeric(q_raw[3]),
    q01_used = as.numeric(q_used[1]),
    q50_used = as.numeric(q_used[2]),
    q99_used = as.numeric(q_used[3])
  )
}

infer_binary_label <- function(pMat, normal_keywords) {
  # Robust text parsing avoids accidental matches like "abnormal"/"normalized".
  normalize_pheno_txt <- function(x) {
    x <- as.character(x)
    x[is.na(x)] <- ""
    x <- gsub("\u00a0|\u200b", " ", x, fixed = TRUE)
    x <- trimws(gsub("[[:space:]]+", " ", x))
    tolower(x)
  }

  flatten_pheno_col <- function(x) {
    if (is.list(x) && !inherits(x, "data.frame")) {
      vapply(x, function(z) paste(as.character(z), collapse = " "), character(1))
    } else {
      as.character(x)
    }
  }

  rx_escape <- function(s) gsub("([][{}()+*^$|\\\\?.])", "\\\\\\1", s, perl = TRUE)

  keyword_hit <- function(xlow, keywords) {
    keywords <- unique(trimws(as.character(keywords)))
    keywords <- keywords[nzchar(keywords)]
    if (!length(keywords)) return(rep(FALSE, length(xlow)))
    parts <- vapply(keywords, function(kw) {
      paste0("(?<![[:alnum:]])", rx_escape(kw), "(?![[:alnum:]])")
    }, character(1))
    grepl(paste0("(?i)(?:", paste(parts, collapse = "|"), ")"), xlow, perl = TRUE)
  }

  explicit_normal_rx <- paste0(
    "(?i)(",
    ",\\s*normal\\b|",
    "\\bbreast\\s+biopsy[, ]+normal\\b|",
    "\\bhuman\\s+breast\\s+biopsy[, ]+normal\\b|",
    "\\bnormal\\s+breast(\\s+tissue)?\\b|",
    "tissue:\\s*normal\\b|",
    "histopathological\\s+exam:\\s*normal\\b",
    ")"
  )

  tumor_hint_rx <- "(?i)(tumor|cancer|carcinoma|malign|invasive|ductal|lobular|metasta)"

  cols_try <- local({
    nm <- colnames(pMat)
    pr <- nm[grepl(
      "source_name|sample_title|\\btitle\\b|histology|pathology|disease|diagnosis|tumor|sample_type|biopsy|specimen",
      nm,
      ignore.case = TRUE
    )]
    ch <- grep("^characteristics_ch", nm, value = TRUE)
    rest <- setdiff(nm, c(pr, ch))
    unique(c(intersect(pr, nm), ch, rest))
  })

  col_has_signal <- function(xlow) {
    nzchar(xlow) &
      (keyword_hit(xlow, normal_keywords) |
         grepl(explicit_normal_rx, xlow, perl = TRUE) |
         grepl(tumor_hint_rx, xlow, perl = TRUE))
  }

  best <- NULL
  best_key <- NULL

  for (idx in seq_along(cols_try)) {
    cn <- cols_try[[idx]]
    x <- flatten_pheno_col(pMat[[cn]])
    xlow <- normalize_pheno_txt(x)
    if (!any(col_has_signal(xlow))) next

    is_normal <- keyword_hit(xlow, normal_keywords) | grepl(explicit_normal_rx, xlow, perl = TRUE)
    label <- ifelse(is_normal, "Normal", "Cancer")
    label[!nzchar(xlow)] <- NA_character_
    label <- factor(label, levels = c("Cancer", "Normal"))

    tab <- table(label, useNA = "no")
    if (!all(c("Cancer", "Normal") %in% names(tab))) next
    if (is.na(tab[["Cancer"]]) || is.na(tab[["Normal"]])) next
    if (tab[["Cancer"]] == 0 || tab[["Normal"]] == 0) next

    min_cnt <- min(as.integer(tab[["Cancer"]]), as.integer(tab[["Normal"]]))
    n_nonempty <- sum(nzchar(xlow) & !is.na(label))
    key <- c(min_cnt, n_nonempty, -idx)
    if (is.null(best_key) || key[1] > best_key[1] ||
        (key[1] == best_key[1] && key[2] > best_key[2]) ||
        (key[1] == best_key[1] && key[2] == best_key[2] && key[3] > best_key[3])) {
      best_key <- key
      best <- list(label = label, label_col = cn)
    }
  }

  if (!is.null(best)) return(best)
  list(label = factor(rep(NA_character_, nrow(pMat)), levels = c("Cancer", "Normal")), label_col = NA_character_)
}

extract_after_scalar <- function(x, prefix) {
  # scalar-safe extractor for strings like "pam50_class: Basal"
  if (length(x) > 1) x <- x[which(!is.na(x))[1]]
  if (length(x) == 0 || is.na(x)) return(NA_character_)
  x <- as.character(x)
  m <- regexpr(prefix, x, ignore.case = TRUE)
  if (m[1] == -1) return(NA_character_)
  out <- substring(x, m[1] + attr(m, "match.length"))
  trimws(out)
}

infer_multiclass_label <- function(pMat) {
  cols_char <- grep("^characteristics_ch", colnames(pMat), value = TRUE)
  cols_other <- setdiff(colnames(pMat), cols_char)
  cols_try <- c(cols_char, cols_other)

  # Prefer columns that look like PAM50
  cols_try <- c(
    intersect(cols_try, grep("pam50", cols_try, ignore.case = TRUE, value = TRUE)),
    setdiff(cols_try, grep("pam50", cols_try, ignore.case = TRUE, value = TRUE))
  )

  # Canonical subtype dictionary (handles common naming variations across datasets)
  # You can extend this list if you see new spellings.
  subtype_variants <- list(
    "Luminal A"    = c("luminal a", "lum a", "luma", "lum-a", "lum_a", "luminala", "lumina"),
    "Luminal B"    = c("luminal b", "lum b", "lumb", "lum-b", "lum_b", "luminalb"),
    "Basal"        = c("basal", "basal-like", "basal like", "basallike", "basal_like", "basal type", "basaltype", "basal/tumor"),
    "HER2"         = c("her2", "her2+", "her-2", "her 2", "her2-enriched", "her2 enriched", "her2enriched", "erb-b2", "erbb2", "erbb2+", "her2e"),
    "Normal-like"  = c("normal-like", "normal like", "normallike", "normal_like", "normal")
  )

  normalize_label_text <- function(x) {
    x <- tolower(as.character(x))
    x[is.na(x)] <- ""
    # Light normalization: trim and collapse non-alphanumeric into single spaces
    x <- trimws(gsub("[^a-z0-9]+", " ", x))
    x
  }

  map_to_subtype <- function(x_raw) {
    x_chr <- as.character(x_raw)
    x_norm <- normalize_label_text(x_chr)
    if (x_norm == "") return(NA_character_)

    # If there is a PAM50 tag embedded in the string, parse after it first
    if (grepl("pam50_class", tolower(x_chr), fixed = TRUE)) {
      x_parsed <- extract_after_scalar(x_chr, prefix = "pam50_class:\\s*")
      x_norm <- normalize_label_text(x_parsed)
    }

    # Match against variants (substring match on normalized text)
    for (canon in names(subtype_variants)) {
      v <- subtype_variants[[canon]]
      if (any(vapply(v, function(kw) grepl(kw, x_norm, fixed = TRUE), logical(1)))) {
        return(canon)
      }
    }
    NA_character_
  }

  for (cn in cols_try) {
    x <- as.character(pMat[[cn]])
    lab <- vapply(x, map_to_subtype, character(1))
    y <- factor(lab)
    y <- droplevels(y[!is.na(y)])
    # Accept if 2+ levels and not degenerate
    if (nlevels(y) >= 2) {
      return(list(label = factor(lab), label_col = cn, parsed_pam50 = any(grepl("pam50_class:", x, ignore.case = TRUE))))
    }
  }

  list(label = factor(rep(NA_character_, nrow(pMat))), label_col = NA_character_, parsed_pam50 = FALSE)
}

rbind_fill <- function(dfs) {
  all_cols <- Reduce(intersect, lapply(dfs, colnames))
  dfs2 <- lapply(dfs, function(d) {
    miss <- setdiff(all_cols, colnames(d))
    for (m in miss) d[[m]] <- NA
    d[, all_cols, drop = FALSE]
  })
  do.call(rbind, dfs2)
}

# Per-fold batch correction with a training-only fit.
combat_fit_apply <- function(expr_train, batch_train, mod_train,
                             expr_test, batch_test) {
  if (length(unique(as.character(batch_train))) < 2) {
    message("Single training batch in this LODO fold; skipping batch correction.")
    return(list(train = expr_train, test = expr_test))
  }

  train_batches <- unique(as.character(batch_train))
  test_batches <- unique(as.character(batch_test))
  common_batches <- intersect(train_batches, test_batches)
  novel_test_idx <- which(!(as.character(batch_test) %in% common_batches))

  corrected_train <- tryCatch(
    sva::ComBat(
      dat = expr_train,
      batch = batch_train,
      mod = mod_train,
      par.prior = TRUE,
      mean.only = FALSE
    ),
    error = function(e) {
      message("ComBat failed on training fold: ", conditionMessage(e),
              "\n  Falling back to limma::removeBatchEffect for this fold.")
      limma::removeBatchEffect(expr_train, batch = batch_train, design = mod_train)
    }
  )

  corrected_test <- tryCatch({
    batch_train_f <- factor(as.character(batch_train))
    batch_test_f <- factor(as.character(batch_test), levels = levels(batch_train_f))

    batch_means_train <- sapply(levels(batch_train_f), function(b) {
      idx <- which(batch_train_f == b)
      if (length(idx) == 0) return(rep(0, nrow(expr_train)))
      rowMeans(expr_train[, idx, drop = FALSE])
    })

    grand_mean_train <- rowMeans(expr_train)
    expr_test_corrected <- expr_test

    for (b in levels(batch_train_f)) {
      test_in_batch <- which(as.character(batch_test_f) == b)
      if (length(test_in_batch) == 0) next
      batch_effect <- batch_means_train[, b] - grand_mean_train
      expr_test_corrected[, test_in_batch] <-
        expr_test[, test_in_batch, drop = FALSE] - batch_effect
    }

    if (length(novel_test_idx) > 0) {
      message("  Test fold contains ", length(novel_test_idx),
              " samples from unseen batches; applying mean-centering only.")
      sample_means <- colMeans(expr_test[, novel_test_idx, drop = FALSE])
      expr_test_corrected[, novel_test_idx] <-
        sweep(expr_test[, novel_test_idx, drop = FALSE], 2, sample_means, "-") +
        grand_mean_train
    }

    expr_test_corrected
  }, error = function(e) {
    message("Test-fold batch correction failed: ", conditionMessage(e),
            "\n  Returning uncorrected test data for this fold.")
    expr_test
  })

  list(train = corrected_train, test = corrected_test)
}

# Missing-value handling for merged expression matrices (use training-fold stats only)
impute_fit <- function(X_train) {
  meds <- apply(X_train, 2, function(v) stats::median(v, na.rm = TRUE))
  meds[is.na(meds)] <- 0
  list(median = meds)
}
impute_apply <- function(X, fit) {
  X2 <- X
  for (j in seq_len(ncol(X2))) {
    idx <- is.na(X2[, j])
    if (any(idx)) X2[idx, j] <- fit$median[j]
  }
  X2
}

# Standardisation for distance-based / margin-based models (fit on training only)
standardize_fit <- function(X_train) {
  mu <- colMeans(X_train)
  sdv <- apply(X_train, 2, sd)
  sdv[sdv == 0] <- 1
  list(mu = mu, sd = sdv)
}
standardize_apply <- function(X, fit) {
  sweep(sweep(X, 2, fit$mu, "-"), 2, fit$sd, "/")
}
```

```{r}
#| message: false
#| warning: false
build_merged_binary_dataset <- function(gse_ids, normal_keywords) {
  pieces <- lapply(gse_ids, function(id) {
    eset <- load_geo_eset(id)
    cc <- collapse_to_genes(eset)
    lab <- infer_binary_label(cc$pMat, normal_keywords)

    p <- cc$pMat
    p$dataset_id <- id
    p$label_col_detected <- lab$label_col
    p$label_binary <- as.character(lab$label)

    list(
      id = id,
      e = cc$eMat_g,
      p = p,
      f = cc$fMat_g,
      log2_applied = cc$log2_applied,
      q01_raw = cc$q01_raw, q50_raw = cc$q50_raw, q99_raw = cc$q99_raw,
      q01_used = cc$q01_used, q50_used = cc$q50_used, q99_used = cc$q99_used
    )
  })

  # Merge expression by INTERSECTION of genes so no dataset-specific padding is needed.
  all_genes <- Reduce(intersect, lapply(pieces, function(x) rownames(x$e)))
  if (length(all_genes) == 0) stop("No genes found across binary datasets.")

  e_big <- do.call(cbind, lapply(pieces, function(x) x$e[all_genes, , drop = FALSE]))

  # Merge phenotype by union of columns
  p_big <- rbind_fill(lapply(pieces, `[[`, "p"))

  # Merge feature data by union of gene rows (keep only genes present in e_big)
  f_big_all <- rbind_fill(lapply(pieces, function(x) {
    ff <- x$f
    ff$dataset_id <- x$id
    tibble::rownames_to_column(ff, var = "gene_id")
  }))
  # Keep one row per gene_id (prefer the first dataset that has it)
  f_big <- f_big_all %>%
    group_by(gene_id) %>%
    slice(1) %>%
    ungroup() %>%
    filter(gene_id %in% rownames(e_big)) %>%
    tibble::column_to_rownames("gene_id")

  list(eMat = e_big, pMat = p_big, fMat = f_big, all_genes = all_genes, pieces = pieces)
}

build_merged_multiclass_dataset <- function(gse_ids) {
  pieces <- lapply(gse_ids, function(id) {
    eset <- load_geo_eset(id)
    cc <- collapse_to_genes(eset)
    lab <- infer_multiclass_label(cc$pMat)

    p <- cc$pMat
    p$dataset_id <- id
    p$label_col_detected <- lab$label_col
    p$label_multiclass <- as.character(lab$label)

    list(id = id, e = cc$eMat_g, p = p, f = cc$fMat_g)
  })

  all_genes <- Reduce(intersect, lapply(pieces, function(x) rownames(x$e)))
  if (length(all_genes) == 0) stop("No genes found across multiclass datasets.")

  e_big <- do.call(cbind, lapply(pieces, function(x) x$e[all_genes, , drop = FALSE]))
  p_big <- rbind_fill(lapply(pieces, `[[`, "p"))

  f_big_all <- rbind_fill(lapply(pieces, function(x) {
    ff <- x$f
    ff$dataset_id <- x$id
    tibble::rownames_to_column(ff, var = "gene_id")
  }))
  f_big <- f_big_all %>%
    group_by(gene_id) %>%
    slice(1) %>%
    ungroup() %>%
    filter(gene_id %in% rownames(e_big)) %>%
    tibble::column_to_rownames("gene_id")

  list(eMat = e_big, pMat = p_big, fMat = f_big, all_genes = all_genes, pieces = pieces)
}

merged_bin <- build_merged_binary_dataset(datasets_binary, normal_keywords)
eMat_big_bin <- merged_bin$eMat
pMat_big_bin <- merged_bin$pMat
fMat_big_bin <- merged_bin$fMat

na_fraction_bin <- rowMeans(is.na(eMat_big_bin))
eMat_big_bin <- eMat_big_bin[na_fraction_bin < 0.3, , drop = FALSE]

cat("Merged binary expression dims (genes x samples):", paste(dim(eMat_big_bin), collapse=" x "), "\n")
print(table(pMat_big_bin$dataset_id))
print(table(pMat_big_bin$label_binary, useNA = "ifany"))
```

```{r}
#| message: false
#| warning: false
merged_mc <- build_merged_multiclass_dataset(datasets_multiclass)
eMat_big_mc <- merged_mc$eMat
pMat_big_mc <- merged_mc$pMat
fMat_big_mc <- merged_mc$fMat

na_fraction_mc <- rowMeans(is.na(eMat_big_mc))
eMat_big_mc <- eMat_big_mc[na_fraction_mc < 0.3, , drop = FALSE]

cat("Merged multiclass expression dims (genes x samples):", paste(dim(eMat_big_mc), collapse=" x "), "\n")
print(table(pMat_big_mc$dataset_id))
print(table(pMat_big_mc$label_multiclass, useNA = "ifany"))
```

## Multiclass modelling (merged dataset)

```{r}
#| message: false
#| warning: false
macro_f1 <- function(truth, pred) {
  truth <- factor(truth)
  pred <- factor(pred, levels = levels(truth))
  lvls <- levels(truth)
  f1s <- c()
  for (k in lvls) {
    tp <- sum(pred == k & truth == k, na.rm = TRUE)
    fp <- sum(pred == k & truth != k, na.rm = TRUE)
    fn <- sum(pred != k & truth == k, na.rm = TRUE)
    prec <- if ((tp + fp) == 0) NA_real_ else tp / (tp + fp)
    rec  <- if ((tp + fn) == 0) NA_real_ else tp / (tp + fn)
    f1   <- if (is.na(prec) || is.na(rec) || (prec + rec) == 0) NA_real_ else 2 * prec * rec / (prec + rec)
    f1s <- c(f1s, f1)
  }
  mean(f1s, na.rm = TRUE)
}

select_genes_one_vs_rest <- function(eMat_g, y, n_per_class = 25) {
  ok <- !is.na(y) & y != ""
  e <- eMat_g[, ok, drop = FALSE]
  y <- factor(y[ok])

  lvls_orig <- levels(y)
  lvls_safe <- make.names(lvls_orig)
  names(lvls_safe) <- lvls_orig

  design <- model.matrix(~ 0 + y)
  colnames(design) <- lvls_safe[lvls_orig]
  fit <- lmFit(e, design)

  genes <- character(0)
  for (lvl in lvls_orig) {
    others <- setdiff(lvls_orig, lvl)
    if (length(others) == 0) next
    lvl_s <- unname(lvls_safe[lvl])
    others_s <- unname(lvls_safe[others])
    contrast_str <- paste0(lvl_s, " - (", paste0(others_s, collapse = " + "), ")/", length(others_s))
    cont <- makeContrasts(contrasts = contrast_str, levels = design)
    fit2 <- contrasts.fit(fit, cont)
    fit2 <- eBayes(fit2)
    tt <- topTable(fit2, n = n_per_class, sort.by = "P")
    genes <- union(genes, rownames(tt))
  }
  genes
}

ok_mc <- !is.na(pMat_big_mc$label_multiclass) & pMat_big_mc$label_multiclass != ""
y_mc <- factor(pMat_big_mc$label_multiclass[ok_mc])
e_mc <- eMat_big_mc[, ok_mc, drop = FALSE]
batch_mc <- pMat_big_mc$dataset_id[ok_mc]

cat("Multiclass class counts (merged):\n")
print(table(y_mc))

genes_mc_vis <- select_genes_one_vs_rest(e_mc, y = y_mc, n_per_class = 25)
cat("Selected genes for multiclass visualization:", length(genes_mc_vis), "\n")

X_mc <- t(e_mc)
mc_vis_genes <- intersect(genes_mc_vis, rownames(e_mc))
X_mc_vis <- if (length(mc_vis_genes) > 0) {
  t(e_mc[mc_vis_genes, , drop = FALSE])
} else {
  X_mc
}
```

## PCA (multiclass, merged)

```{r, fig.cap="PCA of merged multiclass expression (after imputation + scaling). Points coloured by PAM50 subtype."}
#| message: false
#| warning: false
# PCA is for visualisation only; we impute/scale using the exploratory gene subset when available.
imp_mc_all <- impute_fit(X_mc_vis)
X_mc_i_all <- impute_apply(X_mc_vis, imp_mc_all)
sc_mc_all <- standardize_fit(X_mc_i_all)
X_mc_s_all <- standardize_apply(X_mc_i_all, sc_mc_all)

pca_mc <- prcomp(X_mc_s_all, center = FALSE, scale. = FALSE)
pca_mc_df <- data.frame(
  PC1 = pca_mc$x[, 1],
  PC2 = pca_mc$x[, 2],
  truth = y_mc,
  dataset_id = pMat_big_mc$dataset_id[ok_mc]
)

ggplot(pca_mc_df, aes(x = PC1, y = PC2, color = truth)) +
  geom_point(alpha = 0.75, size = 2) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Subtype", title = "Multiclass PCA (merged)") +
  theme(legend.position = "right")
```

```{r}
#| message: false
#| warning: false
run_lodo_cv_multiclass <- function(X, y, batch) {
  y <- factor(y)
  batch <- as.character(batch)
  perf <- data.frame()
  all_preds <- data.frame()
  has_multinom <- requireNamespace("nnet", quietly = TRUE)

  for (ds in unique(batch)) {
    test_id <- which(batch == ds)
    train_id <- setdiff(seq_along(batch), test_id)

    y_train <- droplevels(y[train_id])
    y_test <- droplevels(y[test_id])
    if (nlevels(y_train) < 2) next
    if (length(unique(y_test)) < 2) next

    if (!all(levels(y_test) %in% levels(y_train))) {
      y_test <- factor(as.character(y_test), levels = levels(y_train))
    }

    expr_train_raw <- t(X[train_id, , drop = FALSE])
    expr_test_raw <- t(X[test_id, , drop = FALSE])
    batch_train <- batch[train_id]
    batch_test <- batch[test_id]
    mod_train <- model.matrix(~ y_train)

    cb <- combat_fit_apply(
      expr_train = expr_train_raw,
      batch_train = batch_train,
      mod_train = mod_train,
      expr_test = expr_test_raw,
      batch_test = batch_test
    )

    X_train_full <- t(cb$train)
    X_test_full <- t(cb$test)

    genes_sel <- select_genes_one_vs_rest(t(X_train_full), y = y_train, n_per_class = 25)
    sel_genes <- intersect(genes_sel, colnames(X_train_full))
    if (length(sel_genes) < 10) next

    X_train <- X_train_full[, sel_genes, drop = FALSE]
    X_test <- X_test_full[, sel_genes, drop = FALSE]

    imp <- impute_fit(X_train)
    X_train_i <- impute_apply(X_train, imp)
    X_test_i <- impute_apply(X_test, imp)

    sc <- standardize_fit(X_train_i)
    X_train_s <- standardize_apply(X_train_i, sc)
    X_test_s <- standardize_apply(X_test_i, sc)

    y_test_chr <- as.character(y_test)

    pred_knn <- class::knn(train = X_train_s, test = X_test_s, cl = y_train, k = 7)
    pred_knn_chr <- as.character(pred_knn)
    perf <- rbind(perf, data.frame(task = "multiclass", model = "kNN", fold = ds,
                                   accuracy = mean(pred_knn_chr == y_test_chr),
                                   macro_f1 = macro_f1(y_test_chr, pred_knn_chr)))
    all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "kNN", fold = ds,
                                             truth = y_test_chr, pred = pred_knn_chr))

    svm_fit <- e1071::svm(x = X_train_s, y = y_train, kernel = "radial", cost = 1, gamma = 1 / ncol(X_train_s))
    pred_svm <- predict(svm_fit, X_test_s)
    pred_svm_chr <- as.character(pred_svm)
    perf <- rbind(perf, data.frame(task = "multiclass", model = "SVM_RBF", fold = ds,
                                   accuracy = mean(pred_svm_chr == y_test_chr),
                                   macro_f1 = macro_f1(y_test_chr, pred_svm_chr)))
    all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "SVM_RBF", fold = ds,
                                             truth = y_test_chr, pred = pred_svm_chr))

    rf_fit <- randomForest::randomForest(x = X_train_i, y = y_train)
    pred_rf <- predict(rf_fit, X_test_i)
    pred_rf_chr <- as.character(pred_rf)
    perf <- rbind(perf, data.frame(task = "multiclass", model = "RF", fold = ds,
                                   accuracy = mean(pred_rf_chr == y_test_chr),
                                   macro_f1 = macro_f1(y_test_chr, pred_rf_chr)))
    all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "RF", fold = ds,
                                             truth = y_test_chr, pred = pred_rf_chr))

    if (has_multinom) {
      mn_fit <- nnet::multinom(y_train ~ ., data = data.frame(y_train = y_train, X_train_s),
                               trace = FALSE, MaxNWts = 50000)
      pred_mn <- predict(mn_fit, newdata = data.frame(X_test_s))
      pred_mn_chr <- as.character(pred_mn)
      perf <- rbind(perf, data.frame(task = "multiclass", model = "Multinom", fold = ds,
                                     accuracy = mean(pred_mn_chr == y_test_chr),
                                     macro_f1 = macro_f1(y_test_chr, pred_mn_chr)))
      all_preds <- rbind(all_preds, data.frame(task = "multiclass", model = "Multinom", fold = ds,
                                               truth = y_test_chr, pred = pred_mn_chr))
    }
  }

  list(perf = perf, preds = all_preds)
}

res_mc <- run_lodo_cv_multiclass(X_mc, y_mc, batch_mc)
```

```{r, fig.cap="Multiclass (merged): accuracy and macro-F1 across models (leave-one-dataset-out CV)."}
#| message: false
#| warning: false
perf_mc <- res_mc$perf %>%
  pivot_longer(cols = c(accuracy, macro_f1), names_to = "metric", values_to = "value")

ggplot(perf_mc, aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.6) +
  facet_wrap(~metric, scales = "free_y") +
  theme_bw() +
  labs(x = "", y = "", title = "Multiclass (merged) model comparison") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r, fig.cap="Multiclass (merged): confusion matrix for best model (by mean macro-F1)."}
#| message: false
#| warning: false
if (!is.data.frame(perf_mc) || nrow(perf_mc) == 0 || !all(c("model", "metric", "value") %in% names(perf_mc))) {
  best_model_mc <- NA_character_
} else {
  best_mc_tbl <- perf_mc %>%
    filter(metric == "macro_f1") %>%
    group_by(model) %>%
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean))

  best_model_mc <- if (nrow(best_mc_tbl) == 0) NA_character_ else as.character(best_mc_tbl$model[[1]])
}

cat("Best multiclass model by mean macro-F1:", best_model_mc, "\n")

preds_best_mc <- res_mc$preds %>% filter(model == best_model_mc)
tab_mc <- table(truth = preds_best_mc$truth, pred = preds_best_mc$pred)
tab_mc_df <- as.data.frame(tab_mc)

ggplot(tab_mc_df, aes(x = pred, y = truth, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#2C7FB8") +
  theme_bw() +
  labs(x = "Predicted", y = "True", title = paste0("Multiclass confusion matrix: ", best_model_mc)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Best model separation (multiclass)

```{r, fig.cap="Best multiclass model: PCA view with refit predictions on the full merged dataset (visualisation only)."}
#| message: false
#| warning: false
fit_predict_best_multiclass <- function(best_model, X, y) {
  y <- factor(y)
  imp <- impute_fit(X)
  X_i <- impute_apply(X, imp)

  if (best_model %in% c("kNN", "SVM_RBF", "Multinom")) {
    sc <- standardize_fit(X_i)
    X_s <- standardize_apply(X_i, sc)
  } else {
    X_s <- X_i
  }

  if (best_model == "kNN") {
    pred <- class::knn(train = X_s, test = X_s, cl = y, k = 7)
    return(list(pred = pred))
  }
  if (best_model == "SVM_RBF") {
    fit <- e1071::svm(x = X_s, y = y, kernel = "radial", cost = 1, gamma = 1 / ncol(X_s))
    pred <- predict(fit, X_s)
    return(list(pred = pred))
  }
  if (best_model == "RF") {
    fit <- randomForest::randomForest(x = X_s, y = y)
    pred <- predict(fit, X_s)
    return(list(pred = pred))
  }
  if (best_model == "Multinom") {
    if (!requireNamespace("nnet", quietly = TRUE)) stop("Package 'nnet' not installed; can't fit Multinom.")
    fit <- nnet::multinom(y ~ ., data = data.frame(y = y, X_s), trace = FALSE, MaxNWts = 50000)
    pred <- predict(fit, newdata = data.frame(X_s))
    return(list(pred = pred))
  }
  stop("Unknown multiclass model name: ", best_model)
}

pred_full_mc <- fit_predict_best_multiclass(best_model_mc, X_mc, y_mc)$pred

pca_mc_pred_df <- pca_mc_df %>%
  mutate(pred = pred_full_mc)

ggplot(pca_mc_pred_df, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = truth, shape = pred), alpha = 0.75, size = 2) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "True", shape = "Pred (refit)", title = paste0("Best multiclass model separation: ", best_model_mc)) +
  theme(legend.position = "right")
```

## Binary DE gene selection (merged dataset)

```{r}
#| message: false
#| warning: false
# Binary DE gene selection on the MERGED binary dataset
ok_bin_de <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""
y_bin_de <- factor(pMat_big_bin$label_binary[ok_bin_de], levels = c("Cancer", "Normal"))
e_bin_de <- eMat_big_bin[, ok_bin_de, drop = FALSE]

tab_bin_de <- table(y_bin_de)
cat("Binary class table (merged):\n"); print(tab_bin_de)
if (nlevels(y_bin_de) < 2 || any(tab_bin_de == 0)) {
  stop("Merged binary labels must contain at least two classes with non-zero counts. Check label inference per dataset.")
}

if (nrow(e_bin_de) == 0 || ncol(e_bin_de) == 0 || all(is.na(e_bin_de))) {
  stop("Merged binary expression matrix is empty or all NA after filtering; cannot run DE selection.")
}

design_bin <- model.matrix(~ y_bin_de)
fit_bin <- tryCatch(
  eBayes(lmFit(e_bin_de, design_bin)),
  error = function(e) {
    stop("Binary DE selection failed: ", conditionMessage(e))
  }
)
top_bin <- topTable(fit_bin, n = 50, sort.by = "P")
genes_bin <- rownames(top_bin)
cat("Selected genes for binary (merged):", length(genes_bin), "\n")
top_bin
```

## Model zoo + LODO CV (binary: merged datasets)

```{r}
#| message: false
#| warning: false
auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)
  pos <- truth_pos == 1
  neg <- truth_pos == 0
  n_pos <- sum(pos)
  n_neg <- sum(neg)
  if (n_pos == 0 || n_neg == 0) return(NA_real_)
  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

pr_curve_df <- function(truth_pos, score_pos) {
  ord <- order(score_pos, decreasing = TRUE)
  y <- as.integer(truth_pos[ord])
  s <- score_pos[ord]
  tp <- cumsum(y == 1)
  fp <- cumsum(y == 0)
  fn_total <- sum(y == 1)
  precision <- tp / pmax(tp + fp, 1)
  recall <- tp / pmax(fn_total, 1)
  tibble(threshold = s, precision = precision, recall = recall)
}

auc_pr <- function(truth_pos, score_pos) {
  df <- pr_curve_df(truth_pos, score_pos)
  if (nrow(df) < 2) return(NA_real_)
  o <- order(df$recall, df$precision)
  x <- df$recall[o]
  y <- df$precision[o]
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2, na.rm = TRUE)
}

binary_metrics <- function(truth, pred, positive = "Normal") {
  keep <- !(is.na(truth) | is.na(pred))
  truth <- factor(as.character(truth[keep]))
  pred <- factor(as.character(pred[keep]), levels = levels(truth))
  if (length(truth) == 0) {
    return(tibble(
      accuracy = NA_real_,
      balanced_accuracy = NA_real_,
      recall_normal = NA_real_,
      precision_normal = NA_real_,
      f1_normal = NA_real_
    ))
  }
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)

  recall_pos <- if ((tp + fn) == 0) NA_real_ else tp / (tp + fn)
  precision_pos <- if ((tp + fp) == 0) NA_real_ else tp / (tp + fp)
  f1_pos <- if (is.na(recall_pos) || is.na(precision_pos) || (recall_pos + precision_pos) == 0) NA_real_ else {
    2 * recall_pos * precision_pos / (recall_pos + precision_pos)
  }
  specificity <- if ((tn + fp) == 0) NA_real_ else tn / (tn + fp)
  bal_acc <- mean(c(recall_pos, specificity), na.rm = TRUE)
  acc <- mean(truth == pred)

  tibble(
    accuracy = acc,
    balanced_accuracy = bal_acc,
    recall_normal = recall_pos,
    precision_normal = precision_pos,
    f1_normal = f1_pos
  )
}

best_threshold_f1 <- function(truth, score_pos, positive = "Normal") {
  truth <- as.character(truth)
  neg <- setdiff(unique(na.omit(truth)), positive)[1]
  if (is.na(neg) || length(neg) == 0) return(0.5)
  cand <- sort(unique(score_pos))
  if (length(cand) > 200) cand <- quantile(cand, probs = seq(0, 1, length.out = 200), na.rm = TRUE)
  f1 <- sapply(cand, function(t) {
    pred <- ifelse(score_pos >= t, positive, neg)
    m <- binary_metrics(truth, pred, positive = positive)
    m$f1_normal
  })
  cand[which.max(f1)]
}
```

```{r}
#| message: false
#| warning: false
# Use merged binary dataset (across `datasets_binary`)
ok_bin <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""
y_bin2 <- factor(pMat_big_bin$label_binary[ok_bin], levels = c("Cancer", "Normal"))
batch_bin <- pMat_big_bin$dataset_id[ok_bin]

X_bin <- t(eMat_big_bin[, ok_bin, drop = FALSE])
bin_vis_genes <- intersect(genes_bin, rownames(eMat_big_bin))
X_bin_vis <- if (length(bin_vis_genes) > 0) {
  t(eMat_big_bin[bin_vis_genes, ok_bin, drop = FALSE])
} else {
  X_bin
}

cat("Binary X dims (samples x genes):", paste(dim(X_bin), collapse=" x "), "\n")
cat("Binary class counts:\n")
print(table(y_bin2))

run_lodo_cv_binary <- function(X, y, batch, positive = "Normal", use_weights = TRUE,
                               apply_batch_correction = TRUE,
                               weight_scheme = c("ratio", "balanced", "balanced_capped"),
                               weight_cap = 4,
                               use_downsampling = TRUE,
                               downsample_ratio_cap = 3) {
  y <- factor(y)
  batch <- as.character(batch)
  weight_scheme <- match.arg(weight_scheme)
  setting_label <- if (!isTRUE(use_weights)) {
    "unweighted"
  } else if (weight_scheme == "ratio") {
    "weighted_ratio"
  } else if (weight_scheme == "balanced") {
    "weighted_balanced"
  } else {
    paste0("weighted_balanced_cap", format(weight_cap, trim = TRUE, scientific = FALSE))
  }
  perf <- data.frame()
  all_preds <- data.frame()
  all_scores <- data.frame()
  fold_diag <- data.frame()

  for (ds in unique(batch)) {
    test_id <- which(batch == ds)
    train_id <- setdiff(seq_along(batch), test_id)

    y_train <- droplevels(y[train_id])
    y_test <- droplevels(y[test_id])
    if (nlevels(y_train) < 2) {
      message("Skipping binary fold ", ds, ": training has < 2 classes.")
      next
    }
    if (!all(levels(y_test) %in% levels(y_train))) {
      message("Skipping binary fold ", ds, ": test contains class(es) not present in training.")
      next
    }
    y_test <- factor(as.character(y_test), levels = levels(y_train))
    if (length(unique(y_test)) < 2) next

    expr_train_raw <- t(X[train_id, , drop = FALSE])
    expr_test_raw <- t(X[test_id, , drop = FALSE])
    batch_train <- batch[train_id]
    batch_test <- batch[test_id]

    n_pos_before <- sum(y_train == positive)
    n_neg_before <- sum(y_train != positive)

    downsample_applied <- FALSE
    if (isTRUE(use_downsampling)) {
      idx_pos <- which(y_train == positive)
      idx_neg <- which(y_train != positive)
      if (length(idx_pos) > 0 && length(idx_neg) > 0) {
        max_n <- function(n_min) max(1L, as.integer(downsample_ratio_cap * n_min))
        keep_idx <- seq_along(y_train)
        if (length(idx_neg) > max_n(length(idx_pos))) {
          keep_idx <- sort(c(idx_pos, sample(idx_neg, max_n(length(idx_pos)))))
        } else if (length(idx_pos) > max_n(length(idx_neg))) {
          keep_idx <- sort(c(sample(idx_pos, max_n(length(idx_neg))), idx_neg))
        }
        if (length(keep_idx) < length(y_train)) {
          downsample_applied <- TRUE
          y_train <- droplevels(y_train[keep_idx])
          expr_train_raw <- expr_train_raw[, keep_idx, drop = FALSE]
          batch_train <- batch_train[keep_idx]
        }
      }
    }
    mod_train <- model.matrix(~ y_train)

    cb <- if (isTRUE(apply_batch_correction)) {
      combat_fit_apply(
        expr_train = expr_train_raw,
        batch_train = batch_train,
        mod_train = mod_train,
        expr_test = expr_test_raw,
        batch_test = batch_test
      )
    } else {
      list(train = expr_train_raw, test = expr_test_raw)
    }

    X_train_full <- t(cb$train)
    X_test_full <- t(cb$test)

    if (nrow(X_train_full) == 0 || ncol(X_train_full) == 0 || all(is.na(X_train_full))) {
      message("Skipping binary fold ", ds, ": corrected training matrix is empty or all NA.")
      next
    }

    design_train <- model.matrix(~ y_train)
    fit_lm <- tryCatch(
      eBayes(lmFit(t(X_train_full), design_train)),
      error = function(e) {
        message(
          "Skipping binary fold ", ds,
          ": lmFit failed (samples=", nrow(X_train_full),
          ", genes=", ncol(X_train_full),
          ", NA_fraction=", round(mean(is.na(X_train_full)), 4),
          ") -> ", conditionMessage(e)
        )
        NULL
      }
    )
    if (is.null(fit_lm)) next
    top_genes <- topTable(fit_lm, n = 50, sort.by = "P")
    sel_genes <- intersect(rownames(top_genes), colnames(X_train_full))
    if (length(sel_genes) < 10) next

    X_train <- X_train_full[, sel_genes, drop = FALSE]
    X_test <- X_test_full[, sel_genes, drop = FALSE]

    imp <- impute_fit(X_train)
    X_train_i <- impute_apply(X_train, imp)
    X_test_i <- impute_apply(X_test, imp)

    sc <- standardize_fit(X_train_i)
    X_train_s <- standardize_apply(X_train_i, sc)
    X_test_s <- standardize_apply(X_test_i, sc)

    n_pos <- sum(y_train == positive)
    n_neg <- sum(y_train != positive)

    if (isTRUE(use_weights)) {
      if (weight_scheme == "ratio") {
        pos_w <- if (n_pos > 0) n_neg / n_pos else 1
        neg_w <- 1
      } else {
        n_total <- n_pos + n_neg
        pos_w <- if (n_pos > 0) n_total / (2 * n_pos) else 1
        neg_w <- if (n_neg > 0) n_total / (2 * n_neg) else 1
        if (weight_scheme == "balanced_capped") {
          pos_w <- min(pos_w, weight_cap)
          neg_w <- min(neg_w, weight_cap)
        }
      }
    } else {
      pos_w <- 1
      neg_w <- 1
    }

    n_pos_after <- sum(y_train == positive)
    n_neg_after <- sum(y_train != positive)
    fold_diag <- rbind(fold_diag, data.frame(
      setting = setting_label,
      fold = ds,
      positive_class = positive,
      n_train_before = as.integer(n_pos_before + n_neg_before),
      n_pos_before = as.integer(n_pos_before),
      n_neg_before = as.integer(n_neg_before),
      n_train_after = as.integer(n_pos_after + n_neg_after),
      n_pos_after = as.integer(n_pos_after),
      n_neg_after = as.integer(n_neg_after),
      downsample_used = isTRUE(use_downsampling),
      downsample_applied = downsample_applied,
      downsample_ratio_cap = downsample_ratio_cap,
      pos_weight = as.numeric(pos_w),
      neg_weight = as.numeric(neg_w),
      stringsAsFactors = FALSE
    ))

    neg_label <- setdiff(levels(y_train), positive)[1]
    cw <- c()
    cw[positive] <- pos_w
    cw[neg_label] <- neg_w

    pred_knn <- class::knn(train = X_train_s, test = X_test_s, cl = y_train, k = 7, prob = TRUE)
    win_prob <- attr(pred_knn, "prob")
    score_knn <- ifelse(pred_knn == positive, win_prob, 1 - win_prob)
    m_knn <- binary_metrics(y_test, pred_knn, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = "kNN", fold = ds, m_knn))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = "kNN", fold = ds, truth = y_test, pred = pred_knn))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = "kNN", fold = ds, truth = y_test, score_normal = score_knn))

    svm_fit <- e1071::svm(
      x = X_train_s, y = y_train, kernel = "radial",
      cost = 1, gamma = 1 / ncol(X_train_s),
      probability = TRUE,
      class.weights = if (use_weights) cw else NULL
    )
    pred_svm <- predict(svm_fit, X_test_s, probability = TRUE)
    prob_svm <- attr(pred_svm, "probabilities")
    score_svm <- if ("Normal" %in% colnames(prob_svm)) prob_svm[, "Normal"] else prob_svm[, 1]
    m_svm <- binary_metrics(y_test, pred_svm, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = ifelse(use_weights, "SVM_RBF_w", "SVM_RBF"),
                                   fold = ds, m_svm))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = ifelse(use_weights, "SVM_RBF_w", "SVM_RBF"),
                                             fold = ds, truth = y_test, pred = pred_svm))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = ifelse(use_weights, "SVM_RBF_w", "SVM_RBF"),
                                               fold = ds, truth = y_test, score_normal = score_svm))

    rf_fit <- randomForest::randomForest(x = X_train_i, y = y_train, classwt = if (use_weights) cw else NULL)
    pred_rf <- predict(rf_fit, X_test_i)
    prob_rf <- predict(rf_fit, X_test_i, type = "prob")
    score_rf <- if ("Normal" %in% colnames(prob_rf)) prob_rf[, "Normal"] else prob_rf[, 1]
    m_rf <- binary_metrics(y_test, pred_rf, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = ifelse(use_weights, "RF_w", "RF"),
                                   fold = ds, m_rf))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = ifelse(use_weights, "RF_w", "RF"),
                                             fold = ds, truth = y_test, pred = pred_rf))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = ifelse(use_weights, "RF_w", "RF"),
                                               fold = ds, truth = y_test, score_normal = score_rf))

    w <- if (use_weights) ifelse(y_train == positive, pos_w, neg_w) else rep(1, length(y_train))
    glm_fit <- glm(y_train ~ ., data = data.frame(y_train = y_train, X_train_s),
                   family = binomial(), weights = w)
    p_hat <- predict(glm_fit, newdata = data.frame(X_test_s), type = "response")
    pred_glm <- ifelse(p_hat >= 0.5, positive, setdiff(levels(y_train), positive)[1])
    pred_glm <- factor(pred_glm, levels = levels(y_train))
    m_glm <- binary_metrics(y_test, pred_glm, positive = positive)
    perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                   model = ifelse(use_weights, "Logistic_w", "Logistic"),
                                   fold = ds, m_glm))
    all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                             model = ifelse(use_weights, "Logistic_w", "Logistic"),
                                             fold = ds, truth = y_test, pred = pred_glm))
    all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                               model = ifelse(use_weights, "Logistic_w", "Logistic"),
                                               fold = ds, truth = y_test, score_normal = p_hat))

    if (requireNamespace("glmnet", quietly = TRUE)) {
      x_train_mat <- as.matrix(X_train_s)
      x_test_mat <- as.matrix(X_test_s)
      y_train_num <- as.numeric(y_train == positive)
      lasso_fit <- tryCatch(
        glmnet::cv.glmnet(
          x = x_train_mat,
          y = y_train_num,
          family = "binomial",
          alpha = 1,
          weights = w,
          standardize = FALSE
        ),
        error = function(e) {
          message("Skipping Lasso fold ", ds, ": ", conditionMessage(e))
          NULL
        }
      )
      if (!is.null(lasso_fit)) {
        p_lasso <- as.numeric(predict(lasso_fit, newx = x_test_mat, type = "response", s = "lambda.min"))
        pred_lasso <- ifelse(p_lasso >= 0.5, positive, setdiff(levels(y_train), positive)[1])
        pred_lasso <- factor(pred_lasso, levels = levels(y_train))
        m_lasso <- binary_metrics(y_test, pred_lasso, positive = positive)
        perf <- rbind(perf, data.frame(task = "binary", setting = setting_label,
                                       model = ifelse(use_weights, "Lasso_w", "Lasso"),
                                       fold = ds, m_lasso))
        all_preds <- rbind(all_preds, data.frame(task = "binary", setting = setting_label,
                                                 model = ifelse(use_weights, "Lasso_w", "Lasso"),
                                                 fold = ds, truth = y_test, pred = pred_lasso))
        all_scores <- rbind(all_scores, data.frame(task = "binary", setting = setting_label,
                                                   model = ifelse(use_weights, "Lasso_w", "Lasso"),
                                                   fold = ds, truth = y_test, score_normal = p_lasso))
      }
    }
  }

  list(perf = perf, preds = all_preds, scores = all_scores, diag = fold_diag)
}

weight_cfg <- tibble::tribble(
  ~use_weights, ~weight_scheme, ~weight_cap,
  FALSE, "ratio", Inf,
  TRUE, "ratio", Inf,
  TRUE, "balanced", Inf,
  TRUE, "balanced_capped", 4
)

bin_runs <- lapply(seq_len(nrow(weight_cfg)), function(i) {
  cfg <- weight_cfg[i, ]
  run_lodo_cv_binary(
    X = X_bin,
    y = y_bin2,
    batch = batch_bin,
    positive = "Normal",
    use_weights = cfg$use_weights,
    apply_batch_correction = TRUE,
    weight_scheme = cfg$weight_scheme,
    weight_cap = cfg$weight_cap,
    use_downsampling = use_fold_downsampling,
    downsample_ratio_cap = downsample_max_ratio
  )
})

perf_bin <- bind_rows(lapply(bin_runs, `[[`, "perf"))
preds_bin <- bind_rows(lapply(bin_runs, `[[`, "preds"))
scores_bin <- bind_rows(lapply(bin_runs, `[[`, "scores"))
diag_bin <- bind_rows(lapply(bin_runs, `[[`, "diag"))
```

## Preprocessing + imbalance diagnostics

```{r}
#| message: false
#| warning: false
prep_diag <- bind_rows(lapply(merged_bin$pieces, function(pc) {
  tibble::tibble(
    dataset_id = pc$id,
    log2_applied = isTRUE(pc$log2_applied),
    q01_raw = round(pc$q01_raw, 3),
    q50_raw = round(pc$q50_raw, 3),
    q99_raw = round(pc$q99_raw, 3),
    q01_used = round(pc$q01_used, 3),
    q50_used = round(pc$q50_used, 3),
    q99_used = round(pc$q99_used, 3)
  )
}))

knitr::kable(prep_diag, caption = "Per-dataset preprocessing diagnostics (raw scale vs post-log scale)")

diag_fold_summary <- diag_bin %>%
  mutate(
    imbalance_before = round(n_neg_before / pmax(n_pos_before, 1), 2),
    imbalance_after = round(n_neg_after / pmax(n_pos_after, 1), 2)
  ) %>%
  select(setting, fold, n_train_before, n_pos_before, n_neg_before, imbalance_before,
         n_train_after, n_pos_after, n_neg_after, imbalance_after,
         downsample_used, downsample_applied, pos_weight, neg_weight)

knitr::kable(diag_fold_summary, caption = "Per-fold class-balance diagnostics (before/after downsampling and effective weights)")
```

## PCA (binary, merged)

```{r, fig.cap="PCA of merged binary expression (after imputation + scaling). Points coloured by class."}
#| message: false
#| warning: false
# PCA is for visualisation only; we impute/scale using the exploratory gene subset when available.
imp_bin_all <- impute_fit(X_bin_vis)
X_bin_i_all <- impute_apply(X_bin_vis, imp_bin_all)
sc_bin_all <- standardize_fit(X_bin_i_all)
X_bin_s_all <- standardize_apply(X_bin_i_all, sc_bin_all)

pca_bin <- prcomp(X_bin_s_all, center = FALSE, scale. = FALSE)
pca_bin_df <- data.frame(
  PC1 = pca_bin$x[, 1],
  PC2 = pca_bin$x[, 2],
  truth = y_bin2,
  dataset_id = pMat_big_bin$dataset_id[ok_bin]
)

ggplot(pca_bin_df, aes(x = PC1, y = PC2, color = truth)) +
  geom_point(alpha = 0.75, size = 2) +
  theme_bw() +
  labs(x = "PC1", y = "PC2", color = "Class", title = "Binary PCA (merged)") +
  theme(legend.position = "right")
```

```{r, fig.cap="Binary performance across weight settings: unweighted vs weighted_ratio vs weighted_balanced vs weighted_balanced_cap4 (LODO CV)."}
#| message: false
#| warning: false
perf_bin_long <- perf_bin %>%
  pivot_longer(
    cols = c(accuracy, balanced_accuracy, recall_normal, precision_normal, f1_normal),
    names_to = "metric",
    values_to = "value"
  )

ggplot(perf_bin_long, aes(x = model, y = value, fill = model)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.6) +
  facet_grid(metric ~ setting, scales = "free_y") +
  theme_bw() +
  labs(x = "", y = "", title = "Binary model comparison by weighting scheme") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r, fig.cap="Binary: confusion matrices for best model (default threshold 0.5 vs tuned threshold maximizing pooled F1 for Normal)."}
#| message: false
#| warning: false
best_run_bin <- perf_bin %>%
  group_by(setting, model) %>%
  summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    mean_recall = mean(recall_normal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    scores_bin %>%
      group_by(setting, model) %>%
      summarise(
        mean_roc_auc = auc_roc(truth == "Normal", score_normal),
        mean_pr_auc = auc_pr(truth == "Normal", score_normal),
        .groups = "drop"
      ),
    by = c("setting", "model")
  ) %>%
  arrange(desc(mean_roc_auc), desc(mean_pr_auc), desc(mean_recall), desc(mean_accuracy)) %>%
  slice(1)

best_setting_bin <- best_run_bin$setting[1]
best_model_bin <- best_run_bin$model[1]
cat("Best binary run by ROC AUC then PR AUC then recall:", best_setting_bin, "/", best_model_bin,
  "(ROC AUC=", round(best_run_bin$mean_roc_auc[1], 3),
  ", PR AUC=", round(best_run_bin$mean_pr_auc[1], 3),
  ", Recall=", round(best_run_bin$mean_recall[1], 3),
  ", Accuracy=", round(best_run_bin$mean_accuracy[1], 3), ")\n", sep = "")

score_best <- scores_bin %>% filter(setting == best_setting_bin, model == best_model_bin)
thr_best <- best_threshold_f1(score_best$truth, score_best$score_normal, positive = "Normal")
cat("Chosen threshold (max pooled F1 for Normal):", round(thr_best, 4), "\n")

neg_label_bin <- setdiff(levels(y_bin2), "Normal")[1]
pred_default <- ifelse(score_best$score_normal >= 0.5, "Normal", neg_label_bin)
pred_tuned <- ifelse(score_best$score_normal >= thr_best, "Normal", neg_label_bin)
tab_default <- table(truth = score_best$truth, pred = pred_default)
tab_tuned <- table(truth = score_best$truth, pred = pred_tuned)

acc_check <- tibble::tibble(
  metric = c(
    "Mean fold accuracy (from perf table)",
    "Pooled accuracy @ threshold 0.5",
    "Pooled accuracy @ tuned threshold"
  ),
  value = c(
    perf_bin %>% filter(setting == best_setting_bin, model == best_model_bin) %>% summarise(v = mean(accuracy, na.rm = TRUE)) %>% pull(v),
    mean(score_best$truth == pred_default, na.rm = TRUE),
    mean(score_best$truth == pred_tuned, na.rm = TRUE)
  )
)
knitr::kable(acc_check %>% mutate(value = round(value, 3)),
             caption = paste0("Accuracy consistency check for ", best_setting_bin, " / ", best_model_bin))

tab_default_df <- as.data.frame(tab_default)
tab_tuned_df <- as.data.frame(tab_tuned)

p_cm_default <- ggplot(tab_default_df, aes(x = pred, y = truth, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradient(low = "white", high = "#41AB5D") +
  theme_bw() +
  labs(x = "Predicted", y = "True", title = paste0("Pooled confusion matrix @ 0.5: ", best_setting_bin, " / ", best_model_bin)) +
  theme(legend.position = "right")

p_cm_tuned <- ggplot(tab_tuned_df, aes(x = pred, y = truth, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 5) +
  scale_fill_gradient(low = "white", high = "#41AB5D") +
  theme_bw() +
  labs(x = "Predicted", y = "True", title = paste0("Pooled confusion matrix @ tuned threshold: ", best_setting_bin, " / ", best_model_bin)) +
  theme(legend.position = "right")

print(p_cm_default)
print(p_cm_tuned)
```

```{r, fig.cap="Binary: pooled ROC and PR curves for the best model."}
#| message: false
#| warning: false
curve_best <- scores_bin %>% filter(setting == best_setting_bin, model == best_model_bin)
truth01 <- as.integer(curve_best$truth == "Normal")
score <- curve_best$score_normal

# ROC points
roc_thresh <- sort(unique(score), decreasing = TRUE)
roc_df <- lapply(roc_thresh, function(t) {
  pred <- score >= t
  tp <- sum(pred & truth01 == 1)
  fn <- sum(!pred & truth01 == 1)
  fp <- sum(pred & truth01 == 0)
  tn <- sum(!pred & truth01 == 0)
  tibble(
    threshold = t,
    tpr = if ((tp + fn) == 0) NA_real_ else tp / (tp + fn),
    fpr = if ((fp + tn) == 0) NA_real_ else fp / (fp + tn)
  )
}) %>% bind_rows()

pr_df <- pr_curve_df(truth01, score)

p1 <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "#2C7FB8", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  theme_bw() +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    title = paste0("ROC curve (AUC = ", round(auc_roc(truth01, score), 3), ")")
  )

p2 <- ggplot(pr_df, aes(x = recall, y = precision)) +
  geom_line(color = "#41AB5D", linewidth = 1) +
  theme_bw() +
  labs(
    x = "Recall (Normal)",
    y = "Precision (Normal)",
    title = paste0("PR curve (AUC = ", round(auc_pr(truth01, score), 3), ")")
  )

print(p1)
print(p2)
```

```{r}
#| message: false
#| warning: false
# Binary: aggregate metrics table (mean \u00B1 SD) plus ROC/PR AUC across all weight settings

model_auc <- scores_bin %>%
  group_by(setting, model) %>%
  summarise(
    roc_auc = auc_roc(truth == "Normal", score_normal),
    pr_auc = auc_pr(truth == "Normal", score_normal),
    .groups = "drop"
  )

model_summary <- perf_bin %>%
  group_by(setting, model) %>%
  summarise(
    accuracy_mean = mean(accuracy, na.rm = TRUE),
    accuracy_sd = sd(accuracy, na.rm = TRUE),
    balanced_accuracy_mean = mean(balanced_accuracy, na.rm = TRUE),
    balanced_accuracy_sd = sd(balanced_accuracy, na.rm = TRUE),
    sensitivity_mean = mean(recall_normal, na.rm = TRUE),
    sensitivity_sd = sd(recall_normal, na.rm = TRUE),
    precision_mean = mean(precision_normal, na.rm = TRUE),
    precision_sd = sd(precision_normal, na.rm = TRUE),
    f1_mean = mean(f1_normal, na.rm = TRUE),
    f1_sd = sd(f1_normal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(model_auc, by = c("setting", "model")) %>%
  mutate(
    rank_roc = dense_rank(desc(roc_auc)),
    rank_pr = dense_rank(desc(pr_auc)),
    rank_recall = dense_rank(desc(sensitivity_mean))
  ) %>%
  arrange(desc(roc_auc), desc(pr_auc), desc(sensitivity_mean), desc(accuracy_mean)) %>%
  mutate(overall_rank = row_number()) %>%
  select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall,
         accuracy_mean, accuracy_sd, balanced_accuracy_mean, balanced_accuracy_sd,
         sensitivity_mean, sensitivity_sd, precision_mean, precision_sd, f1_mean, f1_sd,
         roc_auc, pr_auc) %>%
  mutate(
    accuracy = sprintf("%.3f \u00B1 %.3f", accuracy_mean, accuracy_sd),
    balanced_accuracy = sprintf("%.3f \u00B1 %.3f", balanced_accuracy_mean, balanced_accuracy_sd),
    sensitivity = sprintf("%.3f \u00B1 %.3f", sensitivity_mean, sensitivity_sd),
    precision = sprintf("%.3f \u00B1 %.3f", precision_mean, precision_sd),
    f1 = sprintf("%.3f \u00B1 %.3f", f1_mean, f1_sd),
    roc_auc = round(roc_auc, 3),
    pr_auc = round(pr_auc, 3)
  ) %>%
  select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall,
         accuracy, balanced_accuracy, sensitivity, precision, f1, roc_auc, pr_auc)

knitr::kable(model_summary, caption = "Binary: model scorecard ranked by ROC AUC, then PR AUC, then recall")
```

## How many DE genes do we need? (binary)

```{r, fig.cap="Binary: best balanced accuracy vs number of DE genes (weighted models only)."}
#| message: false
#| warning: false
# Goal: find the smallest top-N DE gene set that achieves strong performance.
# We evaluate a grid of N, run the weighted leave-one-dataset-out CV once per N, and take the best mean balanced accuracy across models.

de_gene_sweep <- function(e_bin_de, y_bin_de, n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
                          K = 4, repeats = 10, positive = "Normal") {
  if (nrow(e_bin_de) == 0 || ncol(e_bin_de) == 0 || all(is.na(e_bin_de))) {
    message("Skipping DE sweep: expression matrix is empty or all NA.")
    return(tibble::tibble(
      n_genes = integer(0),
      best_model = character(0),
      best_mean_balanced_accuracy = numeric(0)
    ))
  }

  design_bin <- model.matrix(~ y_bin_de)
  fit_bin <- tryCatch(
    eBayes(lmFit(e_bin_de, design_bin)),
    error = function(e) {
      message("Skipping DE sweep: ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(fit_bin)) {
    return(tibble::tibble(
      n_genes = integer(0),
      best_model = character(0),
      best_mean_balanced_accuracy = numeric(0)
    ))
  }

  out <- list()
  for (n_genes in n_grid) {
    tt <- topTable(fit_bin, n = n_genes, sort.by = "P")
    genes <- rownames(tt)
    X <- t(e_bin_de[intersect(genes, rownames(e_bin_de)), , drop = FALSE])

    # If something went wrong (e.g., no overlap), skip safely
    if (ncol(X) < 2) next

    batch_de <- pMat_big_bin$dataset_id[ok_bin_de]
    res_w <- run_lodo_cv_binary(X, y_bin_de, batch_de, positive = positive, use_weights = TRUE)
    perf_w <- res_w$perf

    if (nrow(perf_w) == 0 || !"model" %in% names(perf_w)) {
      out[[as.character(n_genes)]] <- data.frame(
        n_genes = n_genes,
        best_model = NA_character_,
        best_mean_balanced_accuracy = NA_real_
      )
      next
    }

    best <- perf_w %>%
      group_by(model) %>%
      summarise(mean_bal_acc = mean(balanced_accuracy, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(mean_bal_acc)) %>%
      slice(1)

    out[[as.character(n_genes)]] <- data.frame(
      n_genes = n_genes,
      best_model = best$model[1],
      best_mean_balanced_accuracy = best$mean_bal_acc[1]
    )
  }
  bind_rows(out)
}

sweep_res <- de_gene_sweep(e_bin_de, y_bin_de,
                          n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
                          K = 4, repeats = 10, positive = "Normal")

knitr::kable(sweep_res, caption = "Binary DE gene sweep (weighted CV): best mean balanced accuracy by top-N DE genes")

ggplot(sweep_res, aes(x = n_genes, y = best_mean_balanced_accuracy)) +
  geom_line(color = "#2C7FB8", linewidth = 1) +
  geom_point(size = 2) +
  theme_bw() +
  scale_x_continuous(breaks = sweep_res$n_genes) +
  labs(x = "Top-N DE genes", y = "Best mean balanced accuracy (weighted CV)", title = "How many DE genes do we need?") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## QC inclusion criteria overview (per dataset)

```{r}
#| message: false
#| warning: false
qc_summary_from_piece <- function(piece, task = c("binary", "multiclass")) {
  task <- match.arg(task)
  p <- piece$p
  e <- piece$e
  f <- piece$f

  label_col <- unique(as.character(p$label_col_detected))
  label_col <- label_col[!is.na(label_col) & label_col != ""]
  label_col <- if (length(label_col) > 0) label_col[1] else NA_character_

  label_vec <- if (task == "binary") as.character(p$label_binary) else as.character(p$label_multiclass)
  label_vec[is.na(label_vec) | trimws(label_vec) == ""] <- NA_character_
  n_labeled <- sum(!is.na(label_vec))
  n_classes <- dplyr::n_distinct(label_vec[!is.na(label_vec)])

  # Heuristic: if metadata says normalized / log transformed, mark yes.
  p_text <- tolower(apply(as.data.frame(p), 1, function(r) paste(r, collapse = " | ")))
  has_norm_meta <- any(grepl("normaliz|log2|rma|mas5|quantile", p_text))

  # Gene symbol coverage proxy: proportion of feature rownames that look like symbols.
  gene_ids <- rownames(f)
  gene_symbol_like <- mean(grepl("^[A-Za-z0-9][A-Za-z0-9\\-\\.]{1,}$", gene_ids))

  tibble::tibble(
    dataset_id = piece$id,
    task = task,
    n_samples = ncol(e),
    n_genes_after_collapse = nrow(e),
    normalized_data = ifelse(has_norm_meta, "Likely (metadata indicates transformed/normalized)", "Likely (series matrix); explicit flag not found"),
    outcome_variable_clear = ifelse(!is.na(label_col) && n_classes >= 2, "Yes", "No / unclear"),
    detected_outcome_column = ifelse(is.na(label_col), "NA", label_col),
    labeled_samples = n_labeled,
    classes_detected = n_classes,
    gene_symbols_available = sprintf("%.1f%% symbol-like rownames", 100 * gene_symbol_like),
    include_in_merge = ifelse(n_labeled >= 10 && n_classes >= 2, "Include", "Exclude (insufficient/unclear labels)")
  )
}

qc_bin <- bind_rows(lapply(merged_bin$pieces, qc_summary_from_piece, task = "binary"))
qc_mc  <- bind_rows(lapply(merged_mc$pieces,  qc_summary_from_piece, task = "multiclass"))
qc_table <- bind_rows(qc_bin, qc_mc) %>% arrange(task, dataset_id)

knitr::kable(qc_table, caption = "QC inclusion criteria overview by dataset")
```

## Dataset and Report Flow Statistics

```{r}
#| message: false
#| warning: false
# Detailed flow statistics mappings inspired by PRISMA 2020

flow_stats_df <- tibble::tibble(
  Metric = c(
    "Databases",
    "Registers",
    "Reports sought for retrieval",
    "Reports assessed for eligibility",
    "Duplicate records removed",
    "Records marked as ineligible by automation tools",
    "Records removed for other reasons",
    "Records screened",
    "Records excluded",
    "Reports excluded"
  ),
  Count = c(
    1, # GEO database
    0, 
    length(c(datasets_binary, datasets_multiclass)), # Total datasets attempted for retrieval
    sum(qc_table$n_samples > 0), # Datasets successfully matched/loaded limits
    0, # No explicit duplication check across raw samples here
    sum(qc_table$n_samples) - sum(qc_table$labeled_samples), # Samples lacking valid inferred labels
    0, 
    sum(qc_table$n_samples), # Total raw samples extracted prior to labelling filtering
    sum(qc_table$n_samples) - sum(qc_table$labeled_samples), # Same as ineligible drop exclusion
    sum(qc_table$include_in_merge == "Exclude (insufficient/unclear labels)")
  )
)

knitr::kable(flow_stats_df, caption = "Data Flow and Exclusion Statistics")
```

## Does adding more datasets help? (AUC vs number of datasets)

```{r}
#| message: false
#| warning: false
dataset_count_auc_sweep <- function(dataset_ids, normal_keywords,
                                    top_n_genes = 50, K = 4, repeats = 8,
                                    positive = "Normal") {
  out <- list()

  for (k in seq_along(dataset_ids)) {
    ids_k <- dataset_ids[seq_len(k)]
    merged_k <- build_merged_binary_dataset(ids_k, normal_keywords)

    ok_k <- !is.na(merged_k$pMat$label_binary) & merged_k$pMat$label_binary != ""
    y_k <- factor(merged_k$pMat$label_binary[ok_k], levels = c("Cancer", "Normal"))
    e_k <- merged_k$eMat[, ok_k, drop = FALSE]
    if (nlevels(droplevels(y_k)) < 2) next
    if (nrow(e_k) == 0 || ncol(e_k) == 0 || all(is.na(e_k))) next

    design_k <- model.matrix(~ y_k)
    fit_k <- tryCatch(
      eBayes(lmFit(e_k, design_k)),
      error = function(e) {
        message("Skipping dataset prefix ", k, " in dataset sweep: ", conditionMessage(e))
        NULL
      }
    )
    if (is.null(fit_k)) next
    top_k <- topTable(fit_k, n = top_n_genes, sort.by = "P")
    genes_k <- rownames(top_k)
    X_k <- t(e_k[intersect(genes_k, rownames(e_k)), , drop = FALSE])
    if (ncol(X_k) < 2 || nrow(X_k) < (K + 2)) next

    batch_k <- merged_k$pMat$dataset_id[ok_k]
    res_k <- run_lodo_cv_binary(X_k, y_k, batch_k, positive = positive, use_weights = TRUE)

    req_cols <- c("model", "truth", "score_normal")
    if (!is.data.frame(res_k$scores) || nrow(res_k$scores) == 0 ||
        !all(req_cols %in% names(res_k$scores))) {
      message("Skipping dataset prefix ", k, ": score output is empty or missing required columns.")
      next
    }

    auc_k <- tryCatch({
      res_k$scores %>%
        group_by(model) %>%
        summarise(
          roc_auc = auc_roc(truth == positive, score_normal),
          pr_auc = auc_pr(truth == positive, score_normal),
          .groups = "drop"
        ) %>%
        arrange(desc(roc_auc))
    }, error = function(e) {
      message("Skipping dataset prefix ", k, " during AUC summary: ", conditionMessage(e))
      tibble::tibble()
    })

    if (nrow(auc_k) == 0) next

    out[[length(out) + 1]] <- tibble::tibble(
      n_datasets = k,
      datasets_used = paste(ids_k, collapse = ","),
      n_samples = nrow(X_k),
      best_model = auc_k$model[1],
      best_roc_auc = auc_k$roc_auc[1],
      best_pr_auc = auc_k$pr_auc[1]
    )
  }

  bind_rows(out)
}

dataset_auc_res <- dataset_count_auc_sweep(
  dataset_ids = datasets_binary,
  normal_keywords = normal_keywords,
  top_n_genes = 50, K = 4, repeats = 8, positive = "Normal"
)

knitr::kable(dataset_auc_res, caption = "AUC vs number of merged binary datasets")

ggplot(dataset_auc_res, aes(x = n_datasets, y = best_roc_auc)) +
  geom_line(color = "#2C7FB8", linewidth = 1) +
  geom_point(size = 2) +
  theme_bw() +
  scale_x_continuous(breaks = dataset_auc_res$n_datasets) +
  labs(x = "Number of merged datasets", y = "Best ROC AUC", title = "Does adding datasets improve ROC AUC?")
```

## Cross-Dataset Generalization Heatmap

```{r, fig.cap="Heatmap of external validation ROC AUC across all pairwise combinations of binary datasets (columns = test set, rows = training set)."}
#| message: false
#| warning: false
compute_cross_dataset_matrix <- function(dataset_ids, eMat_big, pMat_big, target_genes, positive = "Normal") {
  n <- length(dataset_ids)
  auc_mat <- matrix(NA, n, n, dimnames = list(Train = dataset_ids, Test = dataset_ids))
  
  # For safety, pre-extract the unified datasets
  ok_bin <- !is.na(pMat_big$label_binary) & pMat_big$label_binary != ""
  p_valid <- pMat_big[ok_bin, ]
  e_valid <- eMat_big[intersect(target_genes, rownames(eMat_big)), ok_bin, drop = FALSE]
  
  for (i in seq_along(dataset_ids)) {
    id_train <- dataset_ids[i]
    idx_train <- which(p_valid$dataset_id == id_train)
    if (length(idx_train) == 0) next
    
    y_train <- factor(p_valid$label_binary[idx_train], levels = c("Cancer", "Normal"))
    X_train <- t(e_valid[, idx_train, drop = FALSE])
    
    if (nlevels(droplevels(y_train)) < 2) next
    
    imp <- impute_fit(X_train)
    X_train_i <- impute_apply(X_train, imp)
    sc <- standardize_fit(X_train_i)
    X_train_s <- standardize_apply(X_train_i, sc)
    
    tab <- table(y_train)
    w_train <- ifelse(y_train == positive, as.numeric(tab["Cancer"] / max(1, tab["Normal"])), 1)
    
    fit <- glm(y_train ~ ., data = data.frame(y_train = y_train, X_train_s),
               family = binomial(), weights = w_train)
    
    for (j in seq_along(dataset_ids)) {
      id_test <- dataset_ids[j]
      idx_test <- which(p_valid$dataset_id == id_test)
      if (length(idx_test) == 0) next
      
      y_test <- factor(p_valid$label_binary[idx_test], levels = c("Cancer", "Normal"))
      X_test <- t(e_valid[, idx_test, drop = FALSE])
      
      if (nlevels(droplevels(y_test)) < 2) next
      
      X_test_s <- standardize_apply(impute_apply(X_test, imp), sc)
      
      # The score returned is P(Y="Normal")
      p_hat <- predict(fit, newdata = data.frame(X_test_s), type = "response")
      auc_mat[i, j] <- auc_roc(y_test == positive, p_hat)
    }
  }
  auc_mat
}

cross_auc_mat <- compute_cross_dataset_matrix(
  dataset_ids = datasets_binary,
  eMat_big = eMat_big_bin, 
  pMat_big = pMat_big_bin, 
  target_genes = genes_bin, 
  positive = "Normal"
)

# Convert to long format for ggplot heatmap
cross_auc_df <- as.data.frame(as.table(cross_auc_mat))
colnames(cross_auc_df) <- c("Train", "Test", "AUC")

ggplot(cross_auc_df, aes(x = Test, y = Train, fill = AUC)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", AUC)), color = ifelse(is.na(cross_auc_df$AUC) | cross_auc_df$AUC > 0.75, "white", "black"), size = 4) +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B", na.value = "grey90") +
  theme_bw() +
  labs(
    x = "Test Dataset (Evaluation)", 
    y = "Train Dataset (Fit)", 
    title = "Cross-Dataset Model Generalization (Logistic Regression ROC AUC)",
    fill = "ROC AUC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Feature count vs performance (AUC extension)

```{r}
#| message: false
#| warning: false
de_gene_sweep_auc <- function(e_bin_de, y_bin_de, n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
                              K = 4, repeats = 8, positive = "Normal") {
  if (nrow(e_bin_de) == 0 || ncol(e_bin_de) == 0 || all(is.na(e_bin_de))) {
    message("Skipping DE sweep AUC: expression matrix is empty or all NA.")
    return(tibble::tibble(
      n_genes = integer(0),
      best_model_bal_acc = character(0),
      best_mean_balanced_accuracy = numeric(0),
      best_model_auc = character(0),
      best_roc_auc = numeric(0),
      best_pr_auc = numeric(0)
    ))
  }

  design_bin <- model.matrix(~ y_bin_de)
  fit_bin <- tryCatch(
    eBayes(lmFit(e_bin_de, design_bin)),
    error = function(e) {
      message("Skipping DE sweep AUC: ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(fit_bin)) {
    return(tibble::tibble(
      n_genes = integer(0),
      best_model_bal_acc = character(0),
      best_mean_balanced_accuracy = numeric(0),
      best_model_auc = character(0),
      best_roc_auc = numeric(0),
      best_pr_auc = numeric(0)
    ))
  }
  out <- list()

  for (n_genes in n_grid) {
    tt <- topTable(fit_bin, n = n_genes, sort.by = "P")
    genes <- rownames(tt)
    X <- t(e_bin_de[intersect(genes, rownames(e_bin_de)), , drop = FALSE])
    if (ncol(X) < 2) next

    batch_de <- pMat_big_bin$dataset_id[ok_bin_de]
    res_w <- run_lodo_cv_binary(X, y_bin_de, batch_de, positive = positive, use_weights = TRUE)

    if (nrow(res_w$perf) == 0 || !"model" %in% names(res_w$perf) ||
        nrow(res_w$scores) == 0 || !"model" %in% names(res_w$scores)) {
      out[[as.character(n_genes)]] <- tibble::tibble(
        n_genes = n_genes,
        best_model_bal_acc = NA_character_,
        best_mean_balanced_accuracy = NA_real_,
        best_model_auc = NA_character_,
        best_roc_auc = NA_real_,
        best_pr_auc = NA_real_
      )
      next
    }

    perf_best <- res_w$perf %>%
      group_by(model) %>%
      summarise(mean_bal_acc = mean(balanced_accuracy, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(mean_bal_acc)) %>%
      slice(1)

    auc_by_model <- res_w$scores %>%
      group_by(model) %>%
      summarise(
        roc_auc = auc_roc(truth == positive, score_normal),
        pr_auc = auc_pr(truth == positive, score_normal),
        .groups = "drop"
      ) %>%
      arrange(desc(roc_auc))

    out[[as.character(n_genes)]] <- tibble::tibble(
      n_genes = n_genes,
      best_model_bal_acc = perf_best$model[1],
      best_mean_balanced_accuracy = perf_best$mean_bal_acc[1],
      best_model_auc = auc_by_model$model[1],
      best_roc_auc = auc_by_model$roc_auc[1],
      best_pr_auc = auc_by_model$pr_auc[1]
    )
  }

  bind_rows(out)
}

sweep_auc_res <- de_gene_sweep_auc(
  e_bin_de, y_bin_de,
  n_grid = c(5, 10, 20, 30, 40, 50, 75, 100, 150, 200),
  K = 4, repeats = 8, positive = "Normal"
)

knitr::kable(sweep_auc_res, caption = "Feature-count sweep: balanced accuracy + ROC/PR AUC")

ggplot(sweep_auc_res, aes(x = n_genes)) +
  geom_line(aes(y = best_roc_auc, color = "ROC AUC"), linewidth = 1) +
  geom_point(aes(y = best_roc_auc, color = "ROC AUC"), size = 2) +
  geom_line(aes(y = best_pr_auc, color = "PR AUC"), linewidth = 1) +
  geom_point(aes(y = best_pr_auc, color = "PR AUC"), size = 2) +
  theme_bw() +
  scale_x_continuous(breaks = sweep_auc_res$n_genes) +
  labs(x = "Top-N DE genes", y = "AUC", color = "", title = "AUC vs number of features")
```

## Risk calculator (binary prototype)

```{r}
#| message: false
#| warning: false
# Build a simple risk calculator from merged binary data:
# output = estimated probability of Cancer for each sample.

# Pick feature count by best ROC AUC from sweep (fallback to 50 if unavailable)
n_risk <- if (nrow(sweep_auc_res) > 0) sweep_auc_res$n_genes[which.max(sweep_auc_res$best_roc_auc)] else 50

design_risk <- model.matrix(~ y_bin_de)
fit_risk_de <- eBayes(lmFit(e_bin_de, design_risk))
top_risk <- topTable(fit_risk_de, n = n_risk, sort.by = "P")
genes_risk <- rownames(top_risk)

X_risk_raw <- t(e_bin_de[intersect(genes_risk, rownames(e_bin_de)), , drop = FALSE])

imp_risk <- impute_fit(X_risk_raw)
X_risk_i <- impute_apply(X_risk_raw, imp_risk)
sc_risk <- standardize_fit(X_risk_i)
X_risk_s <- standardize_apply(X_risk_i, sc_risk)

# Class weighting to reduce bias under imbalance
tab_risk <- table(y_bin_de)
w_risk <- ifelse(y_bin_de == "Normal", as.numeric(tab_risk["Cancer"] / tab_risk["Normal"]), 1)

# Extract Clinical Metadata (Age, Gender, Location)
p_risk <- pMat_big_bin[ok_bin_de, ]
extract_metadata_cols <- function(p) {
  df <- data.frame(sample = rownames(p), age = NA_real_, gender = NA_character_, location = NA_character_)
  chars <- p[, grep("^characteristics_ch", colnames(p)), drop = FALSE]
  
  if (ncol(chars) == 0) return(df)
  
  for (i in seq_len(nrow(p))) {
    row_text <- tolower(paste(chars[i, ], collapse = " "))
    
    # Age extraction
    m_age <- regexpr("age[^a-z0-9]*([0-9]{1,3})", row_text)
    if (m_age > 0) df$age[i] <- as.numeric(gsub("[^0-9]", "", regmatches(row_text, m_age)))
    
    # Gender extraction
    if (grepl("female", row_text)) {
       df$gender[i] <- "Female"
    } else if (grepl("male[\\s,;]?", row_text)) {
       df$gender[i] <- "Male"
    }
    
    # Location / Country
    m_loc <- regexpr("country[:\\s]*([a-z]+)", row_text)
    if (m_loc > 0) df$location[i] <- gsub("country[:\\s]*", "", regmatches(row_text, m_loc))
  }
  df
}

meta_df <- extract_metadata_cols(p_risk)

cat("\n--- Metadata Extraction Summary ---\n")
if (all(is.na(meta_df$age))) {
  cat("No age metadata found across datasets. Proceeding with genetic markers only.\n")
} else {
  cat("Age successfully extracted for", sum(!is.na(meta_df$age)), "samples.\n")
}
cat("\nGender distribution:\n")
print(table(meta_df$gender, useNA = "always"))
cat("\nLocation/Country availability:\n")
print(table(meta_df$location, useNA = "always"))

# Append available metadata features to X_risk_s for modelling
X_risk_meta <- data.frame(X_risk_s)

# Impute and add Age if reasonably present
if (!all(is.na(meta_df$age))) {
  med_age <- median(meta_df$age, na.rm = TRUE)
  filled_age <- meta_df$age
  filled_age[is.na(filled_age)] <- med_age
  X_risk_meta$meta_Age <- scale(filled_age)
}

# Add Gender if variability exists
if (length(na.omit(unique(meta_df$gender))) > 1) {
  is_female <- as.numeric(meta_df$gender == "Female")
  is_female[is.na(is_female)] <- 1 # Base rate breast cancer assumption
  X_risk_meta$meta_IsFemale <- is_female
}
risk_band <- function(p) {
  ifelse(p < 0.20, "Low",
         ifelse(p < 0.50, "Intermediate", "High"))
}

# Risk calculator reports that mirror the classification pipeline.
summarise_risk_report <- function(perf_df, scores_df, report_label) {
  risk_scores_df <- scores_df %>%
    mutate(
      p_normal = score_normal,
      p_cancer = 1 - score_normal,
      risk_score = qlogis(pmin(pmax(p_normal, 1e-6), 1 - 1e-6)),
      risk_band = risk_band(p_cancer),
      sample_id = paste0("fold_", fold, "_", row_number())
    )

  risk_metrics_df <- risk_scores_df %>%
    group_by(setting, model) %>%
    summarise(
      roc_auc = auc_roc(truth == "Normal", score_normal),
      pr_auc = auc_pr(truth == "Normal", score_normal),
      .groups = "drop"
    ) %>%
    left_join(
      perf_df %>%
        group_by(setting, model) %>%
        summarise(
          accuracy = mean(accuracy, na.rm = TRUE),
          recall = mean(recall_normal, na.rm = TRUE),
          .groups = "drop"
        ),
      by = c("setting", "model")
    ) %>%
    mutate(
      rank_roc = dense_rank(desc(roc_auc)),
      rank_pr = dense_rank(desc(pr_auc)),
      rank_recall = dense_rank(desc(recall))
    ) %>%
    arrange(desc(roc_auc), desc(pr_auc), desc(recall), desc(accuracy)) %>%
    mutate(overall_rank = row_number())

  best_setting <- risk_metrics_df$setting[1]
  best_model <- risk_metrics_df$model[1]
  risk_df <- risk_scores_df %>% filter(setting == best_setting, model == best_model)
  plot_df <- risk_df %>% mutate(fold = factor(fold, levels = unique(fold)))

  list(
    report_label = report_label,
    risk_scores_df = risk_scores_df,
    risk_metrics_df = risk_metrics_df,
    best_setting = best_setting,
    best_model = best_model,
    risk_df = risk_df,
    plot_df = plot_df
  )
}

risk_runs_nb <- lapply(seq_len(nrow(weight_cfg)), function(i) {
  cfg <- weight_cfg[i, ]
  run_lodo_cv_binary(
    X = X_bin,
    y = y_bin2,
    batch = batch_bin,
    positive = "Normal",
    use_weights = cfg$use_weights,
    apply_batch_correction = FALSE,
    weight_scheme = cfg$weight_scheme,
    weight_cap = cfg$weight_cap
  )
})

risk_perf_nb <- bind_rows(lapply(risk_runs_nb, `[[`, "perf"))
risk_scores_nb <- bind_rows(lapply(risk_runs_nb, `[[`, "scores"))
risk_report_nb <- summarise_risk_report(risk_perf_nb, risk_scores_nb, "No batch correction")

risk_report_bc <- summarise_risk_report(perf_bin, scores_bin, "Batch corrected")

cat("Risk model feature count (Target):", n_risk, "\n")

cat("\n--- Risk Calculator: No Batch Correction ---\n")
knitr::kable(
  risk_report_nb$risk_metrics_df %>%
    mutate(across(c(roc_auc, pr_auc, accuracy, recall), ~ round(.x, 3))) %>%
    select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall, accuracy, recall, roc_auc, pr_auc),
  caption = "Risk calculator (no batch correction): model scorecard ranked by ROC AUC, then PR AUC, then recall"
)
cat("Best risk model (no batch correction):", risk_report_nb$best_setting, "/", risk_report_nb$best_model, "\n")
knitr::kable(
  head(risk_report_nb$risk_df %>% arrange(desc(p_cancer)), 15),
  caption = paste0("Risk output (no batch correction) using ", risk_report_nb$best_model, ", top 15 highest predicted cancer risk")
)

p_risk_nb_folds <- ggplot(risk_report_nb$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ fold, ncol = 3) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("LODO risk separation by fold (no batch): ", risk_report_nb$best_model))

p_risk_nb_all <- ggplot(risk_report_nb$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("Selected model: ", risk_report_nb$best_model, " (no batch)"))

print(p_risk_nb_folds)
print(p_risk_nb_all)

cat("\n--- Risk Calculator: Batch Corrected ---\n")
knitr::kable(
  risk_report_bc$risk_metrics_df %>%
    mutate(across(c(roc_auc, pr_auc, accuracy, recall), ~ round(.x, 3))) %>%
    select(overall_rank, setting, model, rank_roc, rank_pr, rank_recall, accuracy, recall, roc_auc, pr_auc),
  caption = "Risk calculator (batch corrected): model scorecard ranked by ROC AUC, then PR AUC, then recall"
)
cat("Best risk model (batch corrected):", risk_report_bc$best_setting, "/", risk_report_bc$best_model, "\n")
knitr::kable(
  head(risk_report_bc$risk_df %>% arrange(desc(p_cancer)), 15),
  caption = paste0("Risk output (batch corrected) using ", risk_report_bc$best_model, ", top 15 highest predicted cancer risk")
)

p_risk_bc_folds <- ggplot(risk_report_bc$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ fold, ncol = 3) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("LODO risk separation by fold (batch corrected): ", risk_report_bc$best_model))

p_risk_bc_all <- ggplot(risk_report_bc$plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.35, adjust = 1.1, linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  theme_bw() +
  labs(x = "Risk score (log-odds of Normal)", y = "Density", title = paste0("Selected model: ", risk_report_bc$best_model, " (batch corrected)"))

print(p_risk_bc_folds)
print(p_risk_bc_all)
```













Last week: fix weighted formula   
this week: dynamic weighted model
# Dynamic weighted model risk calculator
## model selection:
1: higher performance
2: low correlation

Step 1:
Apply LODO (Leave-One-Dataset-Out) cross-validation on the full internal datasets to obtain out-of-dataset predictions for each sample.

Step 2:
Train the attention-based meta model using these LODO test predictions.

Step 3:
Retrain the base models (RF, Ridge, Logistic) using 100% of the internal datasets.

Step 4:
Generate predictions on the external dataset using the retrained base models.

Step 5:
Feed the base model predictions from the external dataset into the trained attention meta model to obtain the final dynamic risk scores.

Step 6:
Evaluate the final model performance only on the external dataset.


```{r}
# ============================================================
# FULL DATA LODO Model Selection
# 9 candidate models + batch correction
# No 75/25 split
# Output: performance + correlation + visualisation
# ============================================================

library(dplyr)
library(tidyr)
library(tibble)
library(limma)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(glmnet)
library(sva)

set.seed(3888)

# -----------------------------
# 0. Settings
# -----------------------------

positive_class <- "Cancer"
negative_class <- "Normal"

top_n_genes <- 50

candidate_models <- c(
  "RF",
  "SVM_RBF",
  "SVM_LINEAR",
  "LASSO",
  "Ridge",
  "ElasticNet",
  "kNN",
  "Logistic",
  "NaiveBayes"
)

# -----------------------------
# 1. Helper functions
# -----------------------------

clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)

  pos <- truth_pos == 1
  neg <- truth_pos == 0

  n_pos <- sum(pos)
  n_neg <- sum(neg)

  if (n_pos == 0 || n_neg == 0) return(NA_real_)

  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

binary_metrics_cancer <- function(truth, score_cancer, threshold = 0.5) {
  truth <- as.character(truth)
  pred <- ifelse(score_cancer >= threshold, positive_class, negative_class)

  tp <- sum(pred == positive_class & truth == positive_class, na.rm = TRUE)
  fn <- sum(pred != positive_class & truth == positive_class, na.rm = TRUE)
  tn <- sum(pred == negative_class & truth == negative_class, na.rm = TRUE)
  fp <- sum(pred != negative_class & truth == negative_class, na.rm = TRUE)

  accuracy <- mean(pred == truth, na.rm = TRUE)
  sensitivity <- ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn))
  specificity <- ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp))
  balanced_accuracy <- mean(c(sensitivity, specificity), na.rm = TRUE)
  auc <- auc_roc(truth == positive_class, score_cancer)

  tibble(
    AUC = auc,
    Accuracy = accuracy,
    Sensitivity = sensitivity,
    Specificity = specificity,
    BalancedAccuracy = balanced_accuracy
  )
}

impute_fit <- function(X_train) {
  meds <- apply(X_train, 2, function(v) median(v, na.rm = TRUE))
  meds[is.na(meds)] <- 0
  list(median = meds)
}

impute_apply <- function(X, fit) {
  X2 <- X
  for (j in seq_len(ncol(X2))) {
    idx <- is.na(X2[, j])
    if (any(idx)) X2[idx, j] <- fit$median[j]
  }
  X2
}

standardize_fit <- function(X_train) {
  mu <- colMeans(X_train)
  sdv <- apply(X_train, 2, sd)
  sdv[sdv == 0 | is.na(sdv)] <- 1
  list(mu = mu, sd = sdv)
}

standardize_apply <- function(X, fit) {
  sweep(sweep(X, 2, fit$mu, "-"), 2, fit$sd, "/")
}

combat_fit_apply <- function(expr_train, batch_train, y_train, expr_test, batch_test) {
  batch_train <- as.character(batch_train)
  batch_test <- as.character(batch_test)

  if (length(unique(batch_train)) < 2) {
    message("Only one training batch; skipping batch correction.")
    return(list(train = expr_train, test = expr_test))
  }

  mod_train <- model.matrix(~ y_train)

  corrected_train <- tryCatch(
    sva::ComBat(
      dat = expr_train,
      batch = batch_train,
      mod = mod_train,
      par.prior = TRUE,
      mean.only = FALSE
    ),
    error = function(e) {
      message("ComBat failed; using limma::removeBatchEffect for train. Error: ", conditionMessage(e))
      limma::removeBatchEffect(expr_train, batch = batch_train, design = mod_train)
    }
  )

  train_batches <- unique(batch_train)
  batch_train_f <- factor(batch_train, levels = train_batches)

  batch_means_train <- sapply(train_batches, function(b) {
    idx <- which(batch_train_f == b)
    rowMeans(expr_train[, idx, drop = FALSE], na.rm = TRUE)
  })

  grand_mean_train <- rowMeans(expr_train, na.rm = TRUE)
  expr_test_corrected <- expr_test

  for (b in train_batches) {
    test_idx <- which(batch_test == b)
    if (length(test_idx) == 0) next

    batch_effect <- batch_means_train[, b] - grand_mean_train

    expr_test_corrected[, test_idx] <-
      sweep(expr_test[, test_idx, drop = FALSE], 1, batch_effect, "-")
  }

  novel_test_idx <- which(!(batch_test %in% train_batches))

  if (length(novel_test_idx) > 0) {
    message("Test contains unseen batch(es); applying mean-centering fallback.")

    sample_means <- colMeans(expr_test[, novel_test_idx, drop = FALSE], na.rm = TRUE)

    expr_test_corrected[, novel_test_idx] <-
      sweep(expr_test[, novel_test_idx, drop = FALSE], 2, sample_means, "-") +
      mean(grand_mean_train, na.rm = TRUE)
  }

  list(
    train = corrected_train,
    test = expr_test_corrected
  )
}

select_de_genes <- function(X_train_full, y_train, top_n = 50) {
  y_train <- factor(y_train, levels = c(negative_class, positive_class))
  design <- model.matrix(~ y_train)

  fit <- eBayes(lmFit(t(X_train_full), design))
  tt <- topTable(fit, n = top_n, sort.by = "P")

  intersect(rownames(tt), colnames(X_train_full))
}

# -----------------------------
# 2. Prepare full binary data
# -----------------------------

ok_bin <- !is.na(pMat_big_bin$label_binary) & pMat_big_bin$label_binary != ""

X_all <- t(eMat_big_bin[, ok_bin, drop = FALSE])

y_all <- factor(
  pMat_big_bin$label_binary[ok_bin],
  levels = c(negative_class, positive_class)
)

dataset_all <- as.character(pMat_big_bin$dataset_id[ok_bin])

valid <- !is.na(y_all)

X_all <- X_all[valid, , drop = FALSE]
y_all <- droplevels(y_all[valid])
dataset_all <- dataset_all[valid]

cat("===== Full dataset summary =====\n")
cat("Total samples:", nrow(X_all), "\n")
cat("Total genes:", ncol(X_all), "\n")
cat("\nClass distribution:\n")
print(table(y_all))
cat("\nDataset distribution:\n")
print(table(dataset_all))

# -----------------------------
# 3. Fit models and predict Cancer probability
# -----------------------------

fit_predict_models <- function(
    X_train_full,
    y_train,
    batch_train,
    X_test_full,
    batch_test,
    model_set = candidate_models
) {

  y_train <- factor(y_train, levels = c(negative_class, positive_class))

  expr_train_raw <- t(X_train_full)
  expr_test_raw <- t(X_test_full)

  cb <- combat_fit_apply(
    expr_train = expr_train_raw,
    batch_train = batch_train,
    y_train = y_train,
    expr_test = expr_test_raw,
    batch_test = batch_test
  )

  X_train_bc_full <- t(cb$train)
  X_test_bc_full <- t(cb$test)

  sel_genes <- select_de_genes(
    X_train_full = X_train_bc_full,
    y_train = y_train,
    top_n = top_n_genes
  )

  if (length(sel_genes) < 5) {
    stop("Too few selected genes.")
  }

  X_train <- X_train_bc_full[, sel_genes, drop = FALSE]
  X_test <- X_test_bc_full[, sel_genes, drop = FALSE]

  imp <- impute_fit(X_train)
  X_train_i <- impute_apply(X_train, imp)
  X_test_i <- impute_apply(X_test, imp)

  sc <- standardize_fit(X_train_i)
  X_train_s <- standardize_apply(X_train_i, sc)
  X_test_s <- standardize_apply(X_test_i, sc)

  out <- list()

  if ("RF" %in% model_set) {
    rf_fit <- randomForest(
      x = X_train_i,
      y = y_train,
      ntree = 500
    )
    prob_rf <- predict(rf_fit, X_test_i, type = "prob")
    out[["RF"]] <- prob_rf[, positive_class]
  }

  if ("SVM_RBF" %in% model_set) {
    svm_rbf_fit <- e1071::svm(
      x = X_train_s,
      y = y_train,
      kernel = "radial",
      cost = 1,
      gamma = 1 / ncol(X_train_s),
      probability = TRUE
    )
    pred_svm_rbf <- predict(svm_rbf_fit, X_test_s, probability = TRUE)
    prob_svm_rbf <- attr(pred_svm_rbf, "probabilities")
    out[["SVM_RBF"]] <- prob_svm_rbf[, positive_class]
  }

  if ("SVM_LINEAR" %in% model_set) {
    svm_linear_fit <- e1071::svm(
      x = X_train_s,
      y = y_train,
      kernel = "linear",
      cost = 1,
      probability = TRUE
    )
    pred_svm_linear <- predict(svm_linear_fit, X_test_s, probability = TRUE)
    prob_svm_linear <- attr(pred_svm_linear, "probabilities")
    out[["SVM_LINEAR"]] <- prob_svm_linear[, positive_class]
  }

  if ("LASSO" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    lasso_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 1,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(lasso_fit)) {
      out[["LASSO"]] <- as.numeric(
        predict(
          lasso_fit,
          newx = as.matrix(X_test_s),
          type = "response",
          s = "lambda.min"
        )
      )
    }
  }

  if ("Ridge" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    ridge_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 0,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(ridge_fit)) {
      out[["Ridge"]] <- as.numeric(
        predict(
          ridge_fit,
          newx = as.matrix(X_test_s),
          type = "response",
          s = "lambda.min"
        )
      )
    }
  }

  if ("ElasticNet" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    enet_fit <- tryCatch(
      cv.glmnet(
        x = as.matrix(X_train_s),
        y = y_num,
        family = "binomial",
        alpha = 0.5,
        standardize = FALSE
      ),
      error = function(e) NULL
    )

    if (!is.null(enet_fit)) {
      out[["ElasticNet"]] <- as.numeric(
        predict(
          enet_fit,
          newx = as.matrix(X_test_s),
          type = "response",
          s = "lambda.min"
        )
      )
    }
  }

  if ("kNN" %in% model_set) {
    pred_knn <- class::knn(
      train = X_train_s,
      test = X_test_s,
      cl = y_train,
      k = 7,
      prob = TRUE
    )

    win_prob <- attr(pred_knn, "prob")
    score_knn <- ifelse(pred_knn == positive_class, win_prob, 1 - win_prob)
    out[["kNN"]] <- score_knn
  }

  if ("Logistic" %in% model_set) {
    y_num <- as.numeric(y_train == positive_class)

    df_train <- data.frame(y_num = y_num, X_train_s, check.names = TRUE)
    df_test <- data.frame(X_test_s, check.names = TRUE)

    logistic_fit <- tryCatch(
      glm(y_num ~ ., data = df_train, family = binomial()),
      error = function(e) NULL
    )

    if (!is.null(logistic_fit)) {
      score_logistic <- tryCatch(
        as.numeric(predict(logistic_fit, newdata = df_test, type = "response")),
        error = function(e) rep(NA_real_, nrow(X_test_s))
      )
      out[["Logistic"]] <- score_logistic
    }
  }

  if ("NaiveBayes" %in% model_set) {
    nb_fit <- tryCatch(
      e1071::naiveBayes(x = X_train_s, y = y_train),
      error = function(e) NULL
    )

    if (!is.null(nb_fit)) {
      prob_nb <- tryCatch(
        predict(nb_fit, X_test_s, type = "raw"),
        error = function(e) NULL
      )

      if (!is.null(prob_nb) && positive_class %in% colnames(prob_nb)) {
        out[["NaiveBayes"]] <- prob_nb[, positive_class]
      }
    }
  }

  out <- lapply(out, clip_prob)
  as_tibble(out)
}

# -----------------------------
# 4. Full-data LODO
# -----------------------------

lodo_scores_list <- list()
lodo_metrics_list <- list()

for (ds in unique(dataset_all)) {
  cat("\nRunning FULL DATA LODO fold:", ds, "\n")

  test_id <- which(dataset_all == ds)
  train_id <- setdiff(seq_along(y_all), test_id)

  y_train <- y_all[train_id]
  y_fold <- y_all[test_id]

  batch_train <- dataset_all[train_id]
  batch_fold <- dataset_all[test_id]

  if (length(unique(y_train)) < 2 || length(unique(y_fold)) < 2) {
    cat("Skipping fold", ds, "because train/test has only one class.\n")
    next
  }

  pred_scores <- fit_predict_models(
    X_train_full = X_all[train_id, , drop = FALSE],
    y_train = y_train,
    batch_train = batch_train,
    X_test_full = X_all[test_id, , drop = FALSE],
    batch_test = batch_fold,
    model_set = candidate_models
  )

  available_models <- intersect(candidate_models, colnames(pred_scores))

  fold_scores <- pred_scores %>%
    mutate(
      fold = ds,
      sample_index = seq_len(n()),
      truth = as.character(y_fold)
    ) %>%
    pivot_longer(
      cols = all_of(available_models),
      names_to = "model",
      values_to = "score_cancer"
    )

  fold_metrics <- fold_scores %>%
    group_by(model) %>%
    summarise(
      AUC = binary_metrics_cancer(truth, score_cancer)$AUC,
      Accuracy = binary_metrics_cancer(truth, score_cancer)$Accuracy,
      Sensitivity = binary_metrics_cancer(truth, score_cancer)$Sensitivity,
      Specificity = binary_metrics_cancer(truth, score_cancer)$Specificity,
      BalancedAccuracy = binary_metrics_cancer(truth, score_cancer)$BalancedAccuracy,
      .groups = "drop"
    ) %>%
    mutate(fold = ds)

  lodo_scores_list[[length(lodo_scores_list) + 1]] <- fold_scores
  lodo_metrics_list[[length(lodo_metrics_list) + 1]] <- fold_metrics
}

lodo_scores_full <- bind_rows(lodo_scores_list)
lodo_metrics_full <- bind_rows(lodo_metrics_list)

cat("\n===== FULL DATA LODO metrics =====\n")
print(lodo_metrics_full)

# -----------------------------
# 5. Performance summary
# -----------------------------

model_rank_full <- lodo_metrics_full %>%
  group_by(model) %>%
  summarise(
    mean_LODO_AUC = mean(AUC, na.rm = TRUE),
    sd_LODO_AUC = sd(AUC, na.rm = TRUE),
    mean_Accuracy = mean(Accuracy, na.rm = TRUE),
    mean_BalancedAccuracy = mean(BalancedAccuracy, na.rm = TRUE),
    mean_Sensitivity = mean(Sensitivity, na.rm = TRUE),
    mean_Specificity = mean(Specificity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_LODO_AUC), desc(mean_BalancedAccuracy))

cat("\n===== FULL DATA model ranking =====\n")
print(model_rank_full)

# -----------------------------
# 6. Plot: AUC ranking
# -----------------------------

ggplot(model_rank_full,
       aes(x = reorder(model, mean_LODO_AUC),
           y = mean_LODO_AUC)) +
  geom_col(alpha = 0.8) +
  geom_errorbar(
    aes(
      ymin = mean_LODO_AUC - sd_LODO_AUC,
      ymax = mean_LODO_AUC + sd_LODO_AUC
    ),
    width = 0.2
  ) +
  coord_flip() +
  theme_bw(base_size = 13) +
  labs(
    x = "Model",
    y = "Mean FULL-DATA LODO AUC",
    title = "Model Performance Ranking Using Full-data LODO",
    subtitle = "Error bar = SD across LODO folds"
  )

# -----------------------------
# 7. Plot: model stability
# -----------------------------

ggplot(lodo_metrics_full,
       aes(x = model, y = AUC)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2) +
  theme_bw(base_size = 13) +
  labs(
    x = "Model",
    y = "LODO AUC",
    title = "Model Stability Across Full-data LODO Folds"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

# -----------------------------
# 8. Multiple metric plot
# -----------------------------

model_rank_long <- model_rank_full %>%
  select(
    model,
    mean_LODO_AUC,
    mean_BalancedAccuracy,
    mean_Sensitivity,
    mean_Specificity
  ) %>%
  pivot_longer(
    cols = -model,
    names_to = "metric",
    values_to = "value"
  )

ggplot(model_rank_long,
       aes(x = reorder(model, value),
           y = value,
           fill = metric)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  theme_bw(base_size = 13) +
  labs(
    x = "Model",
    y = "Value",
    title = "Model Performance Across Metrics"
  )

# -----------------------------
# 9. Correlation matrix
# -----------------------------

lodo_pred_wide_full <- lodo_scores_full %>%
  select(fold, sample_index, truth, model, score_cancer) %>%
  mutate(
    score_cancer = as.numeric(score_cancer),
    sample_uid = paste(fold, sample_index, sep = "_")
  ) %>%
  select(sample_uid, model, score_cancer) %>%
  pivot_wider(
    names_from = model,
    values_from = score_cancer,
    values_fn = mean
  )

pred_mat_full <- lodo_pred_wide_full %>%
  select(-sample_uid) %>%
  as.data.frame()

pred_mat_full <- pred_mat_full[
  , colSums(!is.na(pred_mat_full)) > 0,
  drop = FALSE
]

pred_mat_full[] <- lapply(pred_mat_full, as.numeric)

cor_mat_full <- cor(
  pred_mat_full,
  use = "pairwise.complete.obs",
  method = "pearson"
)

cat("\n===== FULL DATA prediction correlation matrix =====\n")
print(round(cor_mat_full, 3))

# -----------------------------
# 10. Correlation heatmap
# -----------------------------

cor_df_full <- as.data.frame(as.table(cor_mat_full))
colnames(cor_df_full) <- c("Model_1", "Model_2", "Correlation")

ggplot(cor_df_full,
       aes(x = Model_1, y = Model_2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 3) +
  theme_bw(base_size = 13) +
  labs(
    x = "",
    y = "",
    fill = "Correlation",
    title = "Prediction Correlation Between Models Using Full-data LODO"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

# -----------------------------
# 11. Performance vs correlation
# -----------------------------

mean_abs_cor_full <- sapply(colnames(cor_mat_full), function(m) {
  vals <- abs(cor_mat_full[m, ])
  vals <- vals[names(vals) != m]
  mean(vals, na.rm = TRUE)
})

cor_summary_full <- tibble(
  model = names(mean_abs_cor_full),
  mean_abs_correlation_with_others = as.numeric(mean_abs_cor_full)
)

perf_cor_summary_full <- model_rank_full %>%
  left_join(cor_summary_full, by = "model")

cat("\n===== FULL DATA performance + correlation summary =====\n")
print(perf_cor_summary_full)

ggplot(perf_cor_summary_full,
       aes(x = mean_abs_correlation_with_others,
           y = mean_LODO_AUC,
           label = model)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text(vjust = -0.6, size = 3.5) +
  theme_bw(base_size = 13) +
  labs(
    x = "Mean Absolute Correlation with Other Models",
    y = "Mean FULL-DATA LODO AUC",
    title = "Performance vs Redundancy",
    subtitle = "Better candidates are high AUC and relatively low correlation"
  )
```
FINAL MODEL SELECTION: Logistic + Ridge+ RF
# Meta model(attention-based)
```{r}
# ============================================================
# Attention-based Meta Model Training
# Logit-input attention + probability-output final score
# Base models: RF + Ridge + Logistic
# ============================================================

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

set.seed(3888)

# -----------------------------
# 1. Select base models
# -----------------------------

selected_models <- c("RF", "Ridge", "Logistic")

# -----------------------------
# 2. Build meta-training data from LODO test predictions
# -----------------------------

meta_train_data <- lodo_scores_full %>%
  filter(model %in% selected_models) %>%
  select(fold, sample_index, truth, model, score_cancer) %>%
  mutate(
    score_cancer = as.numeric(score_cancer),
    sample_uid = paste(fold, sample_index, sep = "_")
  ) %>%
  select(sample_uid, fold, truth, model, score_cancer) %>%
  pivot_wider(
    names_from = model,
    values_from = score_cancer,
    values_fn = mean
  ) %>%
  drop_na(all_of(selected_models))

cat("Meta-training data size:\n")
print(dim(meta_train_data))

# -----------------------------
# 3. Prepare X and y
# -----------------------------

clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

# Original probabilities for final weighted risk score
X_meta_train_prob <- as.matrix(meta_train_data[, selected_models])
X_meta_train_prob <- apply(X_meta_train_prob, 2, clip_prob)

# Logit-transformed probabilities for attention weight calculation
X_meta_train_att <- qlogis(X_meta_train_prob)

y_meta_train <- as.numeric(meta_train_data$truth == "Cancer")

cat("\nX_meta_train_prob dimension:\n")
print(dim(X_meta_train_prob))

cat("\nX_meta_train_att dimension:\n")
print(dim(X_meta_train_att))

cat("\ny_meta_train distribution:\n")
print(table(y_meta_train))

# -----------------------------
# 4. Smoothed attention model
# -----------------------------

softmax_temp <- function(z, temperature = 5.0) {
  z <- z / temperature
  z <- z - max(z)
  exp_z <- exp(z)
  exp_z / sum(exp_z)
}

attention_predict <- function(params, X_att, X_prob, temperature = 5.0) {
  n_models <- ncol(X_att)

  W <- matrix(
    params[1:(n_models * n_models)],
    nrow = n_models,
    ncol = n_models
  )

  b <- params[
    (n_models * n_models + 1):(n_models * n_models + n_models)
  ]

  weights <- t(apply(X_att, 1, function(x) {
    logits <- as.numeric(x %*% W + b)
    softmax_temp(logits, temperature = temperature)
  }))

  # IMPORTANT:
  # weights are learned from logit inputs,
  # but final cancer score is weighted average of original probabilities.
  final_score <- rowSums(weights * X_prob)

  list(
    weights = weights,
    final_score = clip_prob(final_score)
  )
}

entropy_loss <- function(weights) {
  eps <- 1e-8
  entropy <- -rowSums(weights * log(weights + eps))
  -mean(entropy)
}

bce_loss_smoothed <- function(
    params,
    X_att,
    X_prob,
    y,
    temperature = 5.0,
    l2_lambda = 0.05,
    entropy_lambda = 0.20
) {
  pred_obj <- attention_predict(
    params = params,
    X_att = X_att,
    X_prob = X_prob,
    temperature = temperature
  )

  pred <- clip_prob(pred_obj$final_score)
  weights <- pred_obj$weights

  bce <- -mean(
    y * log(pred) +
      (1 - y) * log(1 - pred)
  )

  l2_penalty <- l2_lambda * sum(params^2)
  entropy_penalty <- entropy_lambda * entropy_loss(weights)

  total_loss <- bce + l2_penalty + entropy_penalty

  total_loss
}

# -----------------------------
# 5. Train smoothed attention meta model
# -----------------------------

n_models <- ncol(X_meta_train_att)

temperature_use <- 1.5
l2_lambda_use <- 0.0005
entropy_lambda_use <- 0.01

init_params <- c(
  rnorm(n_models * n_models, mean = 0, sd = 0.001),
  rep(0, n_models)
)

fit_attention <- optim(
  par = init_params,
  fn = bce_loss_smoothed,
  X_att = X_meta_train_att,
  X_prob = X_meta_train_prob,
  y = y_meta_train,
  temperature = temperature_use,
  l2_lambda = l2_lambda_use,
  entropy_lambda = entropy_lambda_use,
  method = "BFGS",
  control = list(maxit = 1000)
)

cat("\n===== Smoothed attention meta model trained =====\n")
cat("Final meta-training loss:", fit_attention$value, "\n")

attention_meta_model <- list(
  selected_models = selected_models,
  params = fit_attention$par,
  loss = fit_attention$value,
  n_models = n_models,
  temperature = temperature_use,
  l2_lambda = l2_lambda_use,
  entropy_lambda = entropy_lambda_use,
  input_type = "logit_probability_for_attention_probability_for_score"
)

# -----------------------------
# 6. Inspect learned weights
# -----------------------------

att_train_pred <- attention_predict(
  params = attention_meta_model$params,
  X_att = X_meta_train_att,
  X_prob = X_meta_train_prob,
  temperature = attention_meta_model$temperature
)

weight_df <- as.data.frame(att_train_pred$weights)
colnames(weight_df) <- paste0("weight_", selected_models)

meta_train_result <- meta_train_data %>%
  bind_cols(weight_df) %>%
  mutate(
    attention_score_cancer = att_train_pred$final_score,
    attention_risk_score = qlogis(attention_score_cancer)
  )

cat("\nMeta-training result preview:\n")
print(head(meta_train_result))

# -----------------------------
# 7. Average weights
# -----------------------------

avg_attention_weights <- meta_train_result %>%
  summarise(
    RF = mean(weight_RF, na.rm = TRUE),
    Ridge = mean(weight_Ridge, na.rm = TRUE),
    Logistic = mean(weight_Logistic, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "model",
    values_to = "mean_weight"
  ) %>%
  arrange(desc(mean_weight))

cat("\n===== Average smoothed attention weights =====\n")
print(avg_attention_weights)

# -----------------------------
# 8. Visualise dynamic weights
# -----------------------------

weight_long <- meta_train_result %>%
  select(sample_uid, truth, starts_with("weight_")) %>%
  pivot_longer(
    cols = starts_with("weight_"),
    names_to = "model",
    values_to = "weight"
  ) %>%
  mutate(model = gsub("weight_", "", model))

ggplot(weight_long, aes(x = model, y = weight, fill = truth)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.6) +
  theme_bw(base_size = 13) +
  labs(
    x = "Base model",
    y = "Attention weight",
    title = "Smoothed dynamic attention weights"
  )

# -----------------------------
# 9. Diagnostic risk score distribution
# -----------------------------

ggplot(meta_train_result, aes(x = attention_risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw(base_size = 13) +
  labs(
    x = "Attention meta risk score: log-odds of Cancer",
    y = "Density",
    title = "Diagnostic distribution on LODO meta-training predictions"
  )

```
```{r}
# 1. Check sample counts per fold
fold_count <- meta_train_result %>%
  count(fold, truth)

print(fold_count)

# 2. Density plot by dataset/fold
ggplot(meta_train_result,
       aes(x = attention_risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ fold, scales = "free_y", ncol = 3) +
  theme_bw(base_size = 13) +
  labs(
    x = "Attention meta risk score: log-odds of Cancer",
    y = "Density",
    title = "Per-dataset LODO test-fold density separation",
    subtitle = "Each panel shows one left-out dataset"
  )
```
# Test
DATASET: GSE42568
```{r}
# ============================================================
# External test dataset: GSE42568
# Basic preprocessing consistent with training pipeline
# ============================================================

external_gse <- "GSE42568"

# -----------------------------
# 1. Load GSE42568
# -----------------------------

eset_42568 <- load_geo_eset(external_gse)

# -----------------------------
# 2. Collapse probe-level expression to gene-level expression
#    This also applies auto log2 transform if needed
# -----------------------------

cc_42568 <- collapse_to_genes(eset_42568)

eMat_42568 <- cc_42568$eMat_g
pMat_42568 <- cc_42568$pMat
fMat_42568 <- cc_42568$fMat_g

cat("===== GSE42568 preprocessing diagnostics =====\n")
cat("Log2 applied:", cc_42568$log2_applied, "\n")
cat("Raw expression quantiles:",
    cc_42568$q01_raw, cc_42568$q50_raw, cc_42568$q99_raw, "\n")
cat("Used expression quantiles:",
    cc_42568$q01_used, cc_42568$q50_used, cc_42568$q99_used, "\n")

cat("\nGSE42568 gene-level expression dims:\n")
print(dim(eMat_42568))

# -----------------------------
# 3. Infer Cancer / Normal label
# -----------------------------

lab_42568 <- infer_binary_label(pMat_42568, normal_keywords)

pMat_42568$dataset_id <- external_gse
pMat_42568$label_col_detected <- lab_42568$label_col
pMat_42568$label_binary <- as.character(lab_42568$label)

cat("\nDetected label column:\n")
print(lab_42568$label_col)

cat("\nGSE42568 label distribution:\n")
print(table(pMat_42568$label_binary, useNA = "ifany"))

# -----------------------------
# 4. Keep valid binary samples only
# -----------------------------

ok_42568 <- !is.na(pMat_42568$label_binary) &
  pMat_42568$label_binary != ""

eMat_42568_bin <- eMat_42568[, ok_42568, drop = FALSE]

y_ext_42568 <- factor(
  pMat_42568$label_binary[ok_42568],
  levels = c("Cancer", "Normal")
)

p_ext_42568 <- pMat_42568[ok_42568, , drop = FALSE]

cat("\nValid external samples:\n")
print(table(y_ext_42568))

# -----------------------------
# 5. Match genes with training merged dataset
# -----------------------------
# eMat_big_bin = training merged gene expression matrix
# rows = genes, cols = samples

common_genes_42568 <- intersect(rownames(eMat_big_bin), rownames(eMat_42568_bin))

cat("\nNumber of common genes with training data:\n")
print(length(common_genes_42568))

if (length(common_genes_42568) < 50) {
  warning("Very few common genes between training data and GSE42568. Check gene symbol mapping.")
}

# Keep same gene order as training data
common_genes_42568 <- rownames(eMat_big_bin)[rownames(eMat_big_bin) %in% common_genes_42568]

eMat_42568_common <- eMat_42568_bin[common_genes_42568, , drop = FALSE]

# External X matrix: samples x genes
X_ext_42568 <- t(eMat_42568_common)

cat("\nExternal X matrix dims: samples x genes\n")
print(dim(X_ext_42568))

# -----------------------------
# 6. Basic missing value check
# -----------------------------

cat("\nMissing value fraction in GSE42568 common-gene matrix:\n")
print(mean(is.na(X_ext_42568)))

na_gene_frac_42568 <- colMeans(is.na(X_ext_42568))
cat("\nGenes with >30% NA in external data:\n")
print(sum(na_gene_frac_42568 > 0.3))

# -----------------------------
# 7. External dataset summary object
# -----------------------------

external_42568 <- list(
  gse_id = external_gse,
  eMat_gene = eMat_42568,
  pMat = pMat_42568,
  X_ext = X_ext_42568,
  y_ext = y_ext_42568,
  p_ext = p_ext_42568,
  common_genes = common_genes_42568,
  log2_applied = cc_42568$log2_applied,
  label_col_detected = lab_42568$label_col
)

cat("\n===== GSE42568 external dataset ready =====\n")
cat("Object saved as: external_42568\n")
cat("Use external_42568$X_ext and external_42568$y_ext for external testing.\n")
```
```{r}
# ============================================================
# External prediction on GSE42568
# Train base models on 100% internal data
# Predict GSE42568
# Apply trained attention meta model
# ============================================================

library(dplyr)
library(tibble)
library(ggplot2)
library(randomForest)
library(glmnet)

set.seed(3888)

# -----------------------------
# 1. Prepare internal training data
# -----------------------------

positive_class <- "Cancer"
negative_class <- "Normal"

ok_internal <- !is.na(pMat_big_bin$label_binary) &
  pMat_big_bin$label_binary != ""

X_internal_full <- t(eMat_big_bin[, ok_internal, drop = FALSE])

y_internal <- factor(
  pMat_big_bin$label_binary[ok_internal],
  levels = c(negative_class, positive_class)
)

batch_internal <- as.character(pMat_big_bin$dataset_id[ok_internal])

# External data from previous step
X_external_full <- external_42568$X_ext
y_external <- external_42568$y_ext
batch_external <- rep("GSE42568", nrow(X_external_full))

# Keep common genes in same order
common_genes_ext <- intersect(colnames(X_internal_full), colnames(X_external_full))

X_internal_full <- X_internal_full[, common_genes_ext, drop = FALSE]
X_external_full <- X_external_full[, common_genes_ext, drop = FALSE]

cat("Internal training data:", dim(X_internal_full), "\n")
cat("External test data:", dim(X_external_full), "\n")
print(table(y_internal))
print(table(y_external))

# -----------------------------
# 2. Batch correction: fit on internal, apply to external
# -----------------------------

expr_train_raw <- t(X_internal_full)
expr_test_raw <- t(X_external_full)

cb_ext <- combat_fit_apply(
  expr_train = expr_train_raw,
  batch_train = batch_internal,
  y_train = y_internal,
  expr_test = expr_test_raw,
  batch_test = batch_external
)
X_internal_bc <- t(cb_ext$train)
X_external_bc <- t(cb_ext$test)

# -----------------------------
# 3. Select top genes from internal training data only
# -----------------------------

design_ext <- model.matrix(~ y_internal)

fit_ext_de <- eBayes(lmFit(t(X_internal_bc), design_ext))

top_external_genes <- topTable(
  fit_ext_de,
  n = 50,
  sort.by = "P"
)

genes_external_model <- intersect(
  rownames(top_external_genes),
  colnames(X_internal_bc)
)

cat("Selected genes for external prediction:", length(genes_external_model), "\n")

X_train_sel <- X_internal_bc[, genes_external_model, drop = FALSE]
X_test_sel <- X_external_bc[, genes_external_model, drop = FALSE]

# -----------------------------
# 4. Impute and scale using internal training only
# -----------------------------

imp_ext <- impute_fit(X_train_sel)

X_train_i <- impute_apply(X_train_sel, imp_ext)
X_test_i <- impute_apply(X_test_sel, imp_ext)

sc_ext <- standardize_fit(X_train_i)

X_train_s <- standardize_apply(X_train_i, sc_ext)
X_test_s <- standardize_apply(X_test_i, sc_ext)

# -----------------------------
# 5. Train selected base models on internal data
# -----------------------------

selected_models <- c("RF", "Ridge", "Logistic")

y_num_cancer <- as.numeric(y_internal == positive_class)

# RF
rf_final <- randomForest::randomForest(
  x = X_train_i,
  y = y_internal,
  ntree = 500
)

rf_prob <- predict(rf_final, X_test_i, type = "prob")
score_rf_cancer <- rf_prob[, positive_class]

# Ridge
ridge_final <- glmnet::cv.glmnet(
  x = as.matrix(X_train_s),
  y = y_num_cancer,
  family = "binomial",
  alpha = 0,
  standardize = FALSE
)

score_ridge_cancer <- as.numeric(
  predict(
    ridge_final,
    newx = as.matrix(X_test_s),
    type = "response",
    s = "lambda.min"
  )
)

# Logistic
df_train_logistic <- data.frame(
  y_num_cancer = y_num_cancer,
  X_train_s,
  check.names = TRUE
)

df_test_logistic <- data.frame(
  X_test_s,
  check.names = TRUE
)

logistic_final <- glm(
  y_num_cancer ~ .,
  data = df_train_logistic,
  family = binomial()
)

score_logistic_cancer <- as.numeric(
  predict(
    logistic_final,
    newdata = df_test_logistic,
    type = "response"
  )
)

# -----------------------------
# 6. Base model prediction table
# -----------------------------

clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

external_base_pred <- tibble(
  sample_index = seq_len(nrow(X_external_full)),
  truth = as.character(y_external),
  RF = clip_prob(score_rf_cancer),
  Ridge = clip_prob(score_ridge_cancer),
  Logistic = clip_prob(score_logistic_cancer)
)

cat("\nExternal base predictions preview:\n")
print(head(external_base_pred))

# -----------------------------
# 7. Apply trained attention meta model
# -----------------------------

softmax_temp <- function(z, temperature = 5.0) {
  z <- z / temperature
  z <- z - max(z)
  exp_z <- exp(z)
  exp_z / sum(exp_z)
}

attention_predict_external <- function(params, X_att, X_prob, temperature = 5.0) {
  n_models <- ncol(X_att)

  W <- matrix(
    params[1:(n_models * n_models)],
    nrow = n_models,
    ncol = n_models
  )

  b <- params[
    (n_models * n_models + 1):(n_models * n_models + n_models)
  ]

  weights <- t(apply(X_att, 1, function(x) {
    logits <- as.numeric(x %*% W + b)
    softmax_temp(logits, temperature = temperature)
  }))

  final_score <- rowSums(weights * X_prob)

  list(
    weights = weights,
    final_score = clip_prob(final_score)
  )
}

X_ext_prob <- as.matrix(external_base_pred[, selected_models])
X_ext_prob <- apply(X_ext_prob, 2, clip_prob)

X_ext_att <- qlogis(X_ext_prob)

att_ext_pred <- attention_predict_external(
  params = attention_meta_model$params,
  X_att = X_ext_att,
  X_prob = X_ext_prob,
  temperature = attention_meta_model$temperature
)

weight_ext_df <- as.data.frame(att_ext_pred$weights)
colnames(weight_ext_df) <- paste0("weight_", selected_models)

external_attention_result <- external_base_pred %>%
  bind_cols(weight_ext_df) %>%
  mutate(
    attention_score_cancer = att_ext_pred$final_score,
    attention_risk_score = qlogis(attention_score_cancer),
    pred_class = ifelse(attention_score_cancer >= 0.5, "Cancer", "Normal")
  )

cat("\n===== External attention prediction result =====\n")
print(head(external_attention_result))

# -----------------------------
# 8. External performance metrics
# -----------------------------

auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)

  pos <- truth_pos == 1
  neg <- truth_pos == 0

  n_pos <- sum(pos)
  n_neg <- sum(neg)

  if (n_pos == 0 || n_neg == 0) return(NA_real_)

  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

binary_metrics_cancer <- function(truth, score_cancer, threshold = 0.5) {
  truth <- as.character(truth)
  pred <- ifelse(score_cancer >= threshold, "Cancer", "Normal")

  tp <- sum(pred == "Cancer" & truth == "Cancer", na.rm = TRUE)
  fn <- sum(pred != "Cancer" & truth == "Cancer", na.rm = TRUE)
  tn <- sum(pred == "Normal" & truth == "Normal", na.rm = TRUE)
  fp <- sum(pred != "Normal" & truth == "Normal", na.rm = TRUE)

  tibble(
    AUC = auc_roc(truth == "Cancer", score_cancer),
    Accuracy = mean(pred == truth, na.rm = TRUE),
    Sensitivity = ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn)),
    Specificity = ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp)),
    BalancedAccuracy = mean(c(
      ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn)),
      ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp))
    ), na.rm = TRUE)
  )
}

external_metrics <- binary_metrics_cancer(
  truth = external_attention_result$truth,
  score_cancer = external_attention_result$attention_score_cancer,
  threshold = 0.5
)

cat("\n===== GSE42568 Attention External Test Performance =====\n")
print(external_metrics)

# -----------------------------
# 9. Compare base models vs attention on GSE42568
# -----------------------------

base_external_metrics <- selected_models %>%
  lapply(function(m) {
    binary_metrics_cancer(
      truth = external_attention_result$truth,
      score_cancer = external_attention_result[[m]],
      threshold = 0.5
    ) %>%
      mutate(model = m)
  }) %>%
  bind_rows()

attention_external_metrics <- external_metrics %>%
  mutate(model = "Attention_Meta")

external_comparison <- bind_rows(
  base_external_metrics,
  attention_external_metrics
) %>%
  select(model, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy) %>%
  arrange(desc(AUC), desc(BalancedAccuracy))

cat("\n===== GSE42568 Base Models vs Attention Meta Model =====\n")
print(external_comparison)

# -----------------------------
# 10. External density plot
# -----------------------------

ggplot(external_attention_result,
       aes(x = attention_risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Attention meta risk score: log-odds of Cancer",
    y = "Density",
    title = "GSE42568 External Test: Attention Risk Score Distribution"
  )

# -----------------------------
# 11. External dynamic weights plot
# -----------------------------

external_weight_long <- external_attention_result %>%
  select(sample_index, truth, starts_with("weight_")) %>%
  pivot_longer(
    cols = starts_with("weight_"),
    names_to = "model",
    values_to = "weight"
  ) %>%
  mutate(model = gsub("weight_", "", model))

ggplot(external_weight_long,
       aes(x = model, y = weight, fill = truth)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.6) +
  theme_bw(base_size = 14) +
  labs(
    x = "Base model",
    y = "Attention weight",
    title = "GSE42568 External Test: Dynamic Attention Weights"
  )
```
```{r}
# ============================================================
# Single Best Model (e.g. Ridge) vs Attention on GSE42568
# SAME preprocessing, SAME trained models
# ============================================================

# -----------------------------
# single best model
# -----------------------------

single_best_model <- "Ridge"   

# -----------------------------
# 2. prediction
# -----------------------------

single_score <- external_attention_result[[single_best_model]]

single_result <- external_attention_result %>%
  mutate(
    single_score_cancer = single_score,
    single_pred_class = ifelse(single_score_cancer >= 0.5, "Cancer", "Normal")
  )

# -----------------------------
# 3. 计算 performance
# -----------------------------

single_metrics <- binary_metrics_cancer(
  truth = single_result$truth,
  score_cancer = single_result$single_score_cancer,
  threshold = 0.5
) %>%
  mutate(model = paste0("Single_", single_best_model))

attention_metrics <- external_metrics %>%
  mutate(model = "Attention_Meta")

comparison_single_vs_attention <- bind_rows(
  single_metrics,
  attention_metrics
) %>%
  select(model, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy)

cat("\n===== Single Best Model vs Attention =====\n")
print(comparison_single_vs_attention)

# -----------------------------
# 4. Density 
# -----------------------------


density_compare_df <- bind_rows(
  single_result %>%
    mutate(
      score = qlogis(single_score_cancer),
      model = paste0("Single_", single_best_model)
    ) %>%
    select(score, truth, model),

  external_attention_result %>%
    mutate(
      score = attention_risk_score,
      model = "Attention_Meta"
    ) %>%
    select(score, truth, model)
)

ggplot(density_compare_df,
       aes(x = score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ model, scales = "free_y") +
  theme_bw(base_size = 14) +
  labs(
    x = "Risk score (log-odds of Cancer)",
    y = "Density",
    title = "Single Best Model vs Attention (External Test: GSE42568)"
  )
```

```{r}
# ============================================================
# Single Best Model External Test
# Save training-derived Top genes, then apply to GSE42568
# Model: Ridge
# ============================================================

library(dplyr)
library(tibble)
library(glmnet)
library(ggplot2)

set.seed(3888)

positive_class <- "Cancer"
negative_class <- "Normal"
single_best_model <- "Ridge"
top_n_single <- 50

# -----------------------------
# 1. Prepare internal training data
# -----------------------------

ok_internal <- !is.na(pMat_big_bin$label_binary) &
  pMat_big_bin$label_binary != ""

X_internal_full <- t(eMat_big_bin[, ok_internal, drop = FALSE])

y_internal <- factor(
  pMat_big_bin$label_binary[ok_internal],
  levels = c(negative_class, positive_class)
)

batch_internal <- as.character(pMat_big_bin$dataset_id[ok_internal])

# -----------------------------
# 2. Prepare external GSE42568 data
# -----------------------------

X_external_full <- external_42568$X_ext
y_external <- external_42568$y_ext
batch_external <- rep("GSE42568", nrow(X_external_full))

common_genes_ext <- intersect(
  colnames(X_internal_full),
  colnames(X_external_full)
)

X_internal_full <- X_internal_full[, common_genes_ext, drop = FALSE]
X_external_full <- X_external_full[, common_genes_ext, drop = FALSE]

cat("Internal training data:", dim(X_internal_full), "\n")
cat("External test data:", dim(X_external_full), "\n")
print(table(y_internal))
print(table(y_external))

# -----------------------------
# 3. Batch correction
# Use y_train version of combat_fit_apply
# -----------------------------

expr_train_raw <- t(X_internal_full)
expr_test_raw <- t(X_external_full)

cb_ext <- combat_fit_apply(
  expr_train = expr_train_raw,
  batch_train = batch_internal,
  y_train = y_internal,
  expr_test = expr_test_raw,
  batch_test = batch_external
)

X_internal_bc <- t(cb_ext$train)
X_external_bc <- t(cb_ext$test)

# -----------------------------
# 4. Select Top genes from INTERNAL TRAINING DATA only
# -----------------------------

design_train <- model.matrix(~ y_internal)

fit_train_de <- eBayes(lmFit(t(X_internal_bc), design_train))

top_single_genes_table <- topTable(
  fit_train_de,
  n = top_n_single,
  sort.by = "P"
)

genes_single_best <- rownames(top_single_genes_table)

# keep only genes available in both train and external
genes_single_best <- intersect(genes_single_best, colnames(X_internal_bc))
genes_single_best <- intersect(genes_single_best, colnames(X_external_bc))

cat("\n===== Saved Top genes for Single Best Model =====\n")
cat("Number of genes:", length(genes_single_best), "\n")
print(genes_single_best)

# Optional: save gene list object
single_best_gene_panel <- list(
  model = single_best_model,
  top_n = top_n_single,
  genes = genes_single_best,
  de_table = top_single_genes_table
)

# -----------------------------
# 5. Use fixed Top genes for train and external test
# -----------------------------

X_train_sel <- X_internal_bc[, genes_single_best, drop = FALSE]
X_test_sel <- X_external_bc[, genes_single_best, drop = FALSE]

# -----------------------------
# 6. Impute and scale using training data only
# -----------------------------

imp_single <- impute_fit(X_train_sel)

X_train_i <- impute_apply(X_train_sel, imp_single)
X_test_i <- impute_apply(X_test_sel, imp_single)

sc_single <- standardize_fit(X_train_i)

X_train_s <- standardize_apply(X_train_i, sc_single)
X_test_s <- standardize_apply(X_test_i, sc_single)

# -----------------------------
# 7. Train Single Best Model: Ridge
# -----------------------------

y_num_cancer <- as.numeric(y_internal == positive_class)

ridge_single_final <- glmnet::cv.glmnet(
  x = as.matrix(X_train_s),
  y = y_num_cancer,
  family = "binomial",
  alpha = 0,
  standardize = FALSE
)

single_score_cancer <- as.numeric(
  predict(
    ridge_single_final,
    newx = as.matrix(X_test_s),
    type = "response",
    s = "lambda.min"
  )
)

clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

single_score_cancer <- clip_prob(single_score_cancer)

single_external_result <- tibble(
  sample_index = seq_along(single_score_cancer),
  truth = as.character(y_external),
  single_score_cancer = single_score_cancer,
  single_risk_score = qlogis(single_score_cancer),
  pred_class = ifelse(single_score_cancer >= 0.5, "Cancer", "Normal")
)

# -----------------------------
# 8. External performance
# -----------------------------

binary_metrics_cancer <- function(truth, score_cancer, threshold = 0.5) {
  truth <- as.character(truth)
  pred <- ifelse(score_cancer >= threshold, "Cancer", "Normal")

  tp <- sum(pred == "Cancer" & truth == "Cancer", na.rm = TRUE)
  fn <- sum(pred != "Cancer" & truth == "Cancer", na.rm = TRUE)
  tn <- sum(pred == "Normal" & truth == "Normal", na.rm = TRUE)
  fp <- sum(pred != "Normal" & truth == "Normal", na.rm = TRUE)

  tibble(
    AUC = auc_roc(truth == "Cancer", score_cancer),
    Accuracy = mean(pred == truth, na.rm = TRUE),
    Sensitivity = ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn)),
    Specificity = ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp)),
    BalancedAccuracy = mean(c(
      ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn)),
      ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp))
    ), na.rm = TRUE)
  )
}

single_external_metrics <- binary_metrics_cancer(
  truth = single_external_result$truth,
  score_cancer = single_external_result$single_score_cancer,
  threshold = 0.5
) %>%
  mutate(model = "Single_Ridge")

cat("\n===== Single Best Model External Test Performance =====\n")
print(single_external_metrics)

# -----------------------------
# 9. Density plot
# -----------------------------

ggplot(single_external_result,
       aes(x = single_risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Single Ridge risk score: log-odds of Cancer",
    y = "Density",
    title = "GSE42568 External Test: Single Ridge Risk Score Distribution"
  )
```
```{r}
# ============================================================
# Attention Meta Model External Test
# Each base model uses its own training-derived Top 50 genes
# Models: RF + Ridge + Logistic
# External test: GSE42568
# ============================================================

library(dplyr)
library(tibble)
library(ggplot2)
library(randomForest)
library(glmnet)

set.seed(3888)

positive_class <- "Cancer"
negative_class <- "Normal"

selected_models <- c("RF", "Ridge", "Logistic")
top_n_each_model <- 50

# -----------------------------
# 1. Prepare internal + external data
# -----------------------------

ok_internal <- !is.na(pMat_big_bin$label_binary) &
  pMat_big_bin$label_binary != ""

X_internal_full <- t(eMat_big_bin[, ok_internal, drop = FALSE])

y_internal <- factor(
  pMat_big_bin$label_binary[ok_internal],
  levels = c(negative_class, positive_class)
)

batch_internal <- as.character(pMat_big_bin$dataset_id[ok_internal])

X_external_full <- external_42568$X_ext
y_external <- external_42568$y_ext
batch_external <- rep("GSE42568", nrow(X_external_full))

common_genes_ext <- intersect(
  colnames(X_internal_full),
  colnames(X_external_full)
)

X_internal_full <- X_internal_full[, common_genes_ext, drop = FALSE]
X_external_full <- X_external_full[, common_genes_ext, drop = FALSE]

cat("Internal training data:", dim(X_internal_full), "\n")
cat("External test data:", dim(X_external_full), "\n")
print(table(y_internal))
print(table(y_external))

# -----------------------------
# 2. Batch correction using internal training only
# -----------------------------

expr_train_raw <- t(X_internal_full)
expr_test_raw <- t(X_external_full)

cb_ext <- combat_fit_apply(
  expr_train = expr_train_raw,
  batch_train = batch_internal,
  y_train = y_internal,
  expr_test = expr_test_raw,
  batch_test = batch_external
)

X_internal_bc <- t(cb_ext$train)
X_external_bc <- t(cb_ext$test)

# -----------------------------
# 3. Helper: select Top 50 genes from training only
# -----------------------------

select_training_top_genes <- function(X_train_full, y_train, top_n = 50) {
  design_train <- model.matrix(~ y_train)

  fit_train_de <- eBayes(
    lmFit(t(X_train_full), design_train)
  )

  top_tbl <- topTable(
    fit_train_de,
    n = top_n,
    sort.by = "P"
  )

  rownames(top_tbl)
}

# -----------------------------
# 4. Helper: preprocessing for one model
# -----------------------------

prepare_model_data <- function(X_train_full, X_test_full, genes_use) {
  genes_use <- intersect(genes_use, colnames(X_train_full))
  genes_use <- intersect(genes_use, colnames(X_test_full))

  X_train_sel <- X_train_full[, genes_use, drop = FALSE]
  X_test_sel <- X_test_full[, genes_use, drop = FALSE]

  imp_fit <- impute_fit(X_train_sel)
  X_train_i <- impute_apply(X_train_sel, imp_fit)
  X_test_i <- impute_apply(X_test_sel, imp_fit)

  sc_fit <- standardize_fit(X_train_i)
  X_train_s <- standardize_apply(X_train_i, sc_fit)
  X_test_s <- standardize_apply(X_test_i, sc_fit)

  list(
    genes = genes_use,
    X_train_i = X_train_i,
    X_test_i = X_test_i,
    X_train_s = X_train_s,
    X_test_s = X_test_s,
    imp_fit = imp_fit,
    sc_fit = sc_fit
  )
}

# -----------------------------
# 5. Select and save model-specific Top 50 gene panels
# -----------------------------
# Important:
# Each model gets its own saved gene panel.
# Here all are selected from internal training data only.

genes_RF <- select_training_top_genes(
  X_train_full = X_internal_bc,
  y_train = y_internal,
  top_n = top_n_each_model
)

genes_Ridge <- select_training_top_genes(
  X_train_full = X_internal_bc,
  y_train = y_internal,
  top_n = top_n_each_model
)

genes_Logistic <- select_training_top_genes(
  X_train_full = X_internal_bc,
  y_train = y_internal,
  top_n = top_n_each_model
)

meta_gene_panels <- list(
  RF = genes_RF,
  Ridge = genes_Ridge,
  Logistic = genes_Logistic
)

cat("\n===== Saved model-specific gene panels =====\n")
cat("RF genes:", length(meta_gene_panels$RF), "\n")
cat("Ridge genes:", length(meta_gene_panels$Ridge), "\n")
cat("Logistic genes:", length(meta_gene_panels$Logistic), "\n")

# Optional: union size
cat("Union gene panel size:",
    length(unique(unlist(meta_gene_panels))), "\n")

# -----------------------------
# 6. Train RF using RF Top 50 genes
# -----------------------------

rf_data <- prepare_model_data(
  X_train_full = X_internal_bc,
  X_test_full = X_external_bc,
  genes_use = meta_gene_panels$RF
)

rf_final <- randomForest::randomForest(
  x = rf_data$X_train_i,
  y = y_internal,
  ntree = 500
)

rf_prob <- predict(rf_final, rf_data$X_test_i, type = "prob")
score_rf_cancer <- rf_prob[, positive_class]

# -----------------------------
# 7. Train Ridge using Ridge Top 50 genes
# -----------------------------

ridge_data <- prepare_model_data(
  X_train_full = X_internal_bc,
  X_test_full = X_external_bc,
  genes_use = meta_gene_panels$Ridge
)

y_num_cancer <- as.numeric(y_internal == positive_class)

ridge_final <- glmnet::cv.glmnet(
  x = as.matrix(ridge_data$X_train_s),
  y = y_num_cancer,
  family = "binomial",
  alpha = 0,
  standardize = FALSE
)

score_ridge_cancer <- as.numeric(
  predict(
    ridge_final,
    newx = as.matrix(ridge_data$X_test_s),
    type = "response",
    s = "lambda.min"
  )
)

# -----------------------------
# 8. Train Logistic using Logistic Top 50 genes
# -----------------------------

logistic_data <- prepare_model_data(
  X_train_full = X_internal_bc,
  X_test_full = X_external_bc,
  genes_use = meta_gene_panels$Logistic
)

df_train_logistic <- data.frame(
  y_num_cancer = y_num_cancer,
  logistic_data$X_train_s,
  check.names = TRUE
)

df_test_logistic <- data.frame(
  logistic_data$X_test_s,
  check.names = TRUE
)

logistic_final <- glm(
  y_num_cancer ~ .,
  data = df_train_logistic,
  family = binomial()
)

score_logistic_cancer <- as.numeric(
  predict(
    logistic_final,
    newdata = df_test_logistic,
    type = "response"
  )
)

# -----------------------------
# 9. Base prediction table
# -----------------------------

clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}

external_base_pred_model_specific <- tibble(
  sample_index = seq_len(nrow(X_external_full)),
  truth = as.character(y_external),
  RF = clip_prob(score_rf_cancer),
  Ridge = clip_prob(score_ridge_cancer),
  Logistic = clip_prob(score_logistic_cancer)
)

cat("\nExternal base predictions preview:\n")
print(head(external_base_pred_model_specific))

# -----------------------------
# 10. Apply trained attention meta model
# -----------------------------

softmax_temp <- function(z, temperature = 5.0) {
  z <- z / temperature
  z <- z - max(z)
  exp_z <- exp(z)
  exp_z / sum(exp_z)
}

attention_predict_external <- function(params, X_att, X_prob, temperature = 5.0) {
  n_models <- ncol(X_att)

  W <- matrix(
    params[1:(n_models * n_models)],
    nrow = n_models,
    ncol = n_models
  )

  b <- params[
    (n_models * n_models + 1):(n_models * n_models + n_models)
  ]

  weights <- t(apply(X_att, 1, function(x) {
    logits <- as.numeric(x %*% W + b)
    softmax_temp(logits, temperature = temperature)
  }))

  final_score <- rowSums(weights * X_prob)

  list(
    weights = weights,
    final_score = clip_prob(final_score)
  )
}

X_ext_prob <- as.matrix(
  external_base_pred_model_specific[, selected_models]
)

X_ext_prob <- apply(X_ext_prob, 2, clip_prob)
X_ext_att <- qlogis(X_ext_prob)

att_ext_pred <- attention_predict_external(
  params = attention_meta_model$params,
  X_att = X_ext_att,
  X_prob = X_ext_prob,
  temperature = attention_meta_model$temperature
)

weight_ext_df <- as.data.frame(att_ext_pred$weights)
colnames(weight_ext_df) <- paste0("weight_", selected_models)

external_attention_result_model_specific <- external_base_pred_model_specific %>%
  bind_cols(weight_ext_df) %>%
  mutate(
    attention_score_cancer = att_ext_pred$final_score,
    attention_risk_score = qlogis(attention_score_cancer),
    pred_class = ifelse(attention_score_cancer >= 0.5, "Cancer", "Normal")
  )

cat("\n===== External attention prediction result =====\n")
print(head(external_attention_result_model_specific))

# -----------------------------
# 11. Performance metrics
# -----------------------------

binary_metrics_cancer <- function(truth, score_cancer, threshold = 0.5) {
  truth <- as.character(truth)
  pred <- ifelse(score_cancer >= threshold, "Cancer", "Normal")

  tp <- sum(pred == "Cancer" & truth == "Cancer", na.rm = TRUE)
  fn <- sum(pred != "Cancer" & truth == "Cancer", na.rm = TRUE)
  tn <- sum(pred == "Normal" & truth == "Normal", na.rm = TRUE)
  fp <- sum(pred != "Normal" & truth == "Normal", na.rm = TRUE)

  tibble(
    AUC = auc_roc(truth == "Cancer", score_cancer),
    Accuracy = mean(pred == truth, na.rm = TRUE),
    Sensitivity = ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn)),
    Specificity = ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp)),
    BalancedAccuracy = mean(c(
      ifelse((tp + fn) == 0, NA_real_, tp / (tp + fn)),
      ifelse((tn + fp) == 0, NA_real_, tn / (tn + fp))
    ), na.rm = TRUE)
  )
}

external_attention_metrics_model_specific <- binary_metrics_cancer(
  truth = external_attention_result_model_specific$truth,
  score_cancer = external_attention_result_model_specific$attention_score_cancer,
  threshold = 0.5
) %>%
  mutate(model = "Attention_Meta_ModelSpecificGenes")

base_external_metrics_model_specific <- selected_models %>%
  lapply(function(m) {
    binary_metrics_cancer(
      truth = external_attention_result_model_specific$truth,
      score_cancer = external_attention_result_model_specific[[m]],
      threshold = 0.5
    ) %>%
      mutate(model = m)
  }) %>%
  bind_rows()

external_comparison_model_specific <- bind_rows(
  base_external_metrics_model_specific,
  external_attention_metrics_model_specific
) %>%
  select(model, AUC, Accuracy, Sensitivity, Specificity, BalancedAccuracy) %>%
  arrange(desc(AUC), desc(BalancedAccuracy))

cat("\n===== GSE42568 Base Models vs Attention Meta Model =====\n")
print(external_comparison_model_specific)

# -----------------------------
# 12. Plot: density
# -----------------------------

ggplot(external_attention_result_model_specific,
       aes(x = attention_risk_score, fill = truth)) +
  geom_density(alpha = 0.4, adjust = 1.2, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Attention meta risk score: log-odds of Cancer",
    y = "Density",
    title = "GSE42568 External Test: Attention Meta Model with Model-specific Top 50 Genes"
  )

# -----------------------------
# 13. Plot: dynamic weights
# -----------------------------

external_weight_long_model_specific <- external_attention_result_model_specific %>%
  select(sample_index, truth, starts_with("weight_")) %>%
  pivot_longer(
    cols = starts_with("weight_"),
    names_to = "model",
    values_to = "weight"
  ) %>%
  mutate(model = gsub("weight_", "", model))

ggplot(external_weight_long_model_specific,
       aes(x = model, y = weight, fill = truth)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.6) +
  theme_bw(base_size = 14) +
  labs(
    x = "Base model",
    y = "Attention weight",
    title = "GSE42568 External Test: Dynamic Attention Weights with Model-specific Gene Panels"
  )
```
