---
title: "Breast cancer"
format: html
editor: visual
---

```{r}
library(GEOquery)
library(randomForest)
library(caret)
library(e1071)
library(class)
```

## 1. Load GSE15852

```{r}
gse15852 <- getGEO(filename = "GSE15852_series_matrix.txt.gz")
```

```{r}
expr1 <- exprs(gse15852)
pheno1 <- pData(gse15852)

dim(expr1)
head(pheno1)
colnames(pheno1)
```

```{r}
table(pheno1$`characteristics_ch1.1`)
```

```{r}
labels1 <- ifelse(
  pheno1$characteristics_ch1.1 == "histopathological exam: normal",
  "normal",
  "tumor"
)

table(labels1)
```

```{r}
expr1 <- exprs(gse15852)
dim(expr1)
table(labels1)
```

## 2. Load GSE9574

```{r}
gse9574 <- getGEO(filename = "GSE9574_series_matrix.txt.gz")

expr2 <- exprs(gse9574)
pheno2 <- pData(gse9574)
```

```{r}
labels2 <- ifelse(
  grepl("tumor|cancer", pheno2$source_name_ch1, ignore.case = TRUE),
  "tumor",
  "normal"
)

table(labels2)
```

```{r}
colnames(pheno2)
head(pheno2)
```

```{r}
table(pheno2$`source_name_ch1`)
```

## 3. Load GSE25066

```{r}
gse25066 <- getGEO(filename = "GSE25066_series_matrix.txt.gz")

expr3 <- exprs(gse25066)
pheno3 <- pData(gse25066)
```

```{r}
dim(expr3)
colnames(pheno3)
head(pheno3)
```

```{r}
table(pheno3$`characteristics_ch1.20`)
```

```{r}
labels3 <- ifelse(
  grepl("Normal", pheno3$characteristics_ch1.20, ignore.case = TRUE),
  "normal",
  "tumor"
)

table(labels3)
```

## 4. Merging 3 datasets

```{r}
dim(expr1)
dim(expr2)
dim(expr3)

table(labels1)
table(labels2)
table(labels3)
```

```{r}
common_genes <- Reduce(intersect, list(
  rownames(expr1),
  rownames(expr2),
  rownames(expr3)
))

length(common_genes)
```

```{r}
expr1 <- expr1[common_genes, ]
expr2 <- expr2[common_genes, ]
expr3 <- expr3[common_genes, ]
```

```{r}
merged_expr <- cbind(expr1, expr2, expr3)

dim(merged_expr)
```

```{r}
labels <- c(labels1, labels2, labels3)

batch <- c(
  rep("GSE15852", length(labels1)),
  rep("GSE9574", length(labels2)),
  rep("GSE25066", length(labels3))
)

table(labels)
table(batch)
```

```{r}
# Normalization
library(limma)

merged_expr <- normalizeBetweenArrays(merged_expr)
```

```{r}
# Batch Effect Correction using ComBat

library(sva)

combat_expr <- ComBat(
  dat = merged_expr,
  batch = batch,
  mod = model.matrix(~labels)
)
```

```{r}
# check whether merge successfully
pca <- prcomp(t(combat_expr))

plot(
  pca$x[,1],
  pca$x[,2],
  col = as.factor(labels),
  pch = 16,
  main = "PCA after ComBat",
  xlab = "PC1 (Principal Component 1)",
  ylab = "PC2 (Principal Component 2)"
)
```

```{r}
# Handling imbalance - Downsampling tumor samples

set.seed(1)

tumor_idx <- which(labels == "tumor")
normal_idx <- which(labels == "normal")

tumor_sampled <- sample(tumor_idx, length(normal_idx))

final_idx <- c(tumor_sampled, normal_idx)

final_expr <- combat_expr[, final_idx]
final_labels <- labels[final_idx]

table(final_labels)
```

## 5.1 EDA of merged dataset (after downsampling) 

```{r}
# balanced dataset
table(final_labels)
prop.table(table(final_labels))
```

```{r}
boxplot(
  final_expr,
  outline = FALSE,
  las = 2,
  main = "Boxplot after Downsampling",
  ylab = "Expression"
)
```

The boxplot shows comparable expression distributions across samples after normalization and batch correction.

```{r}
# PCA

pca_final <- prcomp(t(final_expr))

plot(
  pca_final$x[,1],
  pca_final$x[,2],
  col = as.factor(final_labels),
  pch = 16,
  main = "PCA after Downsampling",
  xlab = "PC1",
  ylab = "PC2"
)

legend("topright",
       legend = c("tumor","normal"),
       col = c(1,2),
       pch = 16)
```

PCA after downsampling shows a more balanced distribution of tumor and normal samples, with partial separation between the two classes.

## 5.2 Machine Learning

```{r}
# Step 1：train/test split

set.seed(1)

n <- ncol(final_expr)
train_idx <- sample(1:n, size = 0.7*n)

train_data <- t(final_expr[, train_idx])
test_data  <- t(final_expr[, -train_idx])

train_labels <- final_labels[train_idx]
test_labels  <- final_labels[-train_idx]
```

```{r}
# Step 2：Random Forest

rf_model <- randomForest(
  x = train_data,
  y = as.factor(train_labels),
  ntree = 100
)
```

```{r}
pred <- predict(rf_model, test_data)
```

## 5.3 Evaluation

```{r}
# Accuracy

accuracy <- mean(pred == test_labels)
accuracy
```

```{r}
# Precision / Recall / F1

confusionMatrix(pred, as.factor(test_labels))
```

## 6. Feature selection

```{r}
# Step 1：Feature Selection（t-test）
p_values <- apply(final_expr, 1, function(x) {
  t.test(x ~ final_labels)$p.value
})
```

```{r}
# Step 2：top 100 genes
top_genes <- names(sort(p_values))[1:100]

# Step 3：building new data
expr_selected <- final_expr[top_genes, ]
```

```{r}
# Step 4：Train / Test Split
set.seed(1)

n <- ncol(expr_selected)
train_idx <- sample(1:n, 0.7*n)

train_data <- t(expr_selected[, train_idx])
test_data  <- t(expr_selected[, -train_idx])

train_labels <- final_labels[train_idx]
test_labels  <- final_labels[-train_idx]
```

```{r}
# Step 5：Random Forest（Improved Model）
rf_model_fs <- randomForest(
  x = train_data,
  y = as.factor(train_labels),
  ntree = 100
)

pred_fs <- predict(rf_model_fs, test_data)
```

```{r}
# Accuracy
acc_fs <- mean(pred_fs == test_labels)
acc_fs

# Confusion Matrix
confusionMatrix(pred_fs, as.factor(test_labels))

```

## 7. Model Selection

```{r}
# SVM model

svm_model <- svm(
  x = train_data,
  y = as.factor(train_labels),
  kernel = "linear"
)

pred_svm <- predict(svm_model, test_data)

acc_svm <- mean(pred_svm == test_labels)
acc_svm

confusionMatrix(pred_svm, as.factor(test_labels))
```

```{r}
# KNN

pred_knn <- knn(
  train = train_data,
  test = test_data,
  cl = train_labels,
  k = 5
)

# unified levels
pred_knn <- factor(pred_knn, levels = c("normal", "tumor"))
test_labels <- factor(test_labels, levels = c("normal", "tumor"))

# accuracy
acc_knn <- mean(pred_knn == test_labels)
acc_knn

# confusion matrix
confusionMatrix(pred_knn, test_labels)
```

```{r}
model_results <- data.frame(
  Model = c("RF (full)", "RF (feature selection)", "SVM (feature selection)", "KNN (feature selection)"),
  Accuracy = c(accuracy, acc_fs, acc_svm, acc_knn)
)

model_results
```

## 8. ROC curve

-   ROC curves were used to evaluate the performance of different classification models.

```{r}
library(pROC)

# RF
prob_rf <- predict(rf_model_fs, test_data, type = "prob")[, "tumor"]

# SVM
svm_model <- svm(
  x = train_data,
  y = as.factor(train_labels),
  kernel = "linear",
  probability = TRUE
)

pred_svm <- predict(svm_model, test_data, probability = TRUE)

prob_svm <- attr(pred_svm, "probabilities")[, "tumor"]

# KNN
pred_knn <- knn(
  train = train_data,
  test = test_data,
  cl = train_labels,
  k = 5,
  prob = TRUE
)

prob_knn <- ifelse(pred_knn == "tumor",
                   attr(pred_knn, "prob"),
                   1 - attr(pred_knn, "prob"))

# Draw ROC
# convert label
test_binary <- ifelse(test_labels == "tumor", 1, 0)

roc_rf  <- roc(test_binary, prob_rf)
roc_svm <- roc(test_binary, prob_svm)
roc_knn <- roc(test_binary, prob_knn)

# draw plots
plot(roc_rf, col = "red",   lwd = 2, main = "ROC Curve Comparison")
plot(roc_svm,col = "blue", lwd = 2, add = TRUE)
plot(roc_knn,col = "purple", lwd = 2, add = TRUE)

legend("bottomright",
       legend = c("RF", "SVM", "KNN"),
       col = c("red", "blue","purple"),
       lwd = 2)
```

-   RF has the best discriminative performance, SVM performs comparably well but slightly worse than RF, KNN performs worse due to high-dimensional data.

-   The curve is very steep indicates that the models achieve high sensitivity with low false positives. Good performance overall.

```{r}
auc(roc_rf)
auc(roc_svm)
auc(roc_knn)
```

-   RF（0.9674）close to 1 (strong performance)

-   All three models have an AUC \> 0.9

-   Indicating that the data itself is highly separable (strong biological signal)
