---
title: "Corrected Interdisciplinary Project"
format: html
editor: visual
---

```{r}
# Load libraries
library(GEOquery)
library(Biobase)
library(caret)
library(janitor)
library(survival)
library(data.table)
library(glmnet)
library(matrixStats)
library(randomForest)
library(pbapply)
library(dplyr)
library(tidyr)
library(tibble)
library(sva)
library(devtools)
library(limma)
library(ggplot2)
library(e1071)
library(pheatmap)

set.seed(123)
```

## Helper Functions

```{r}
# Robust GEO loading with retry logic
load_geo_eset <- function(gse_id) {
  geo_cache_dir <- file.path("data", "geo-cache")
  if (!dir.exists(geo_cache_dir)) dir.create(geo_cache_dir, recursive = TRUE)

  g <- tryCatch(
    GEOquery::getGEO(gse_id, destdir = geo_cache_dir),
    error = function(e) e
  )
  
  if (!inherits(g, "error")) {
    if (inherits(g, "ExpressionSet")) return(g)
    if (is.list(g) && length(g) >= 1) return(g[[1]])
  }

  destfile <- file.path(geo_cache_dir, paste0(gse_id, "_series_matrix.txt.gz"))
  if (file.exists(destfile)) {
    g2 <- GEOquery::getGEO(filename = destfile)
    if (inherits(g2, "ExpressionSet")) return(g2)
    if (is.list(g2) && length(g2) >= 1) return(g2[[1]])
  }
  
  stop("Could not load ExpressionSet for ", gse_id)
}

# Clean gene symbols (take first if multiple)
clean_first_symbol <- function(s) {
  if (is.na(s) || trimws(as.character(s)) == "") return(NA_character_)
  s <- trimws(as.character(s))
  parts <- strsplit(s, " /// ", fixed = TRUE)[[1]]
  trimws(parts[1])
}

# Collapse probes to genes using gene symbols
collapse_to_genes <- function(eset) {
  e <- exprs(eset)
  fd <- fData(eset)
  
  sym_col <- grep("gene\\s*symbol|symbol", colnames(fd), ignore.case = TRUE, value = TRUE)[1]
  gene_sym_raw <- if (!is.na(sym_col)) as.character(fd[, sym_col]) else rep(NA_character_, nrow(fd))
  
  has_symbol <- !is.na(gene_sym_raw) & trimws(gene_sym_raw) != ""
  gene_sym <- rep(NA_character_, length(gene_sym_raw))
  gene_sym[has_symbol] <- vapply(gene_sym_raw[has_symbol], clean_first_symbol, character(1))
  gene_sym <- ifelse(is.na(gene_sym) | gene_sym == "", rownames(e), gene_sym)
  
  e_g <- limma::avereps(e, ID = gene_sym)
  
  fd$.__gene_sym__ <- gene_sym
  fd_g <- fd %>%
    tibble::as_tibble() %>%
    mutate(.__gene_sym__ = fd$.__gene_sym__) %>%
    group_by(.__gene_sym__) %>%
    slice(1) %>%
    ungroup() %>%
    tibble::column_to_rownames(".__gene_sym__")
  
  list(eMat_g = e_g, fMat_g = fd_g, pMat = pData(eset))
}

# Check and log-transform if needed
check_and_log_transform <- function(eset) {
  ex <- exprs(eset)
  if (max(ex, na.rm = TRUE) > 50) {
    message("Log-transforming raw data...")
    ex[ex <= 0] <- 1
    exprs(eset) <- log2(ex)
  } else {
    message("Data is already log-scaled.")
  }
  return(eset)
}

# Infer binary labels
infer_binary_label <- function(pMat, normal_keywords) {
  cols_char <- grep("^characteristics_ch|^source_name|^title", colnames(pMat), value = TRUE)
  cols_other <- setdiff(colnames(pMat), cols_char)
  cols_try <- c(cols_char, cols_other)

  has_kw <- function(x, keywords) {
    x <- tolower(as.character(x))
    x[is.na(x)] <- ""
    any(vapply(keywords, function(kw) any(grepl(kw, x, fixed = TRUE)), logical(1)))
  }

  for (cn in cols_try) {
    x <- pMat[[cn]]
    if (!has_kw(x, normal_keywords)) next
    
    xlow <- tolower(as.character(x))
    xlow[is.na(xlow)] <- ""
    
    is_normal <- Reduce(`|`, lapply(normal_keywords, function(kw) grepl(kw, xlow, fixed = TRUE)))
    label <- ifelse(is_normal, "Normal", "Cancer")
    label[xlow == ""] <- NA_character_
    label <- factor(label, levels = c("Cancer", "Normal"))
    
    tab <- table(label, useNA = "no")
    if (length(tab) >= 2 && all(tab > 0)) {
      return(list(label = label, label_col = cn))
    }
  }

  list(label = factor(rep(NA_character_, nrow(pMat)), levels = c("Cancer", "Normal")),
       label_col = NA_character_)
}

# Merge dataframes with different columns
rbind_fill <- function(dfs) {
  all_cols <- Reduce(intersect, lapply(dfs, colnames))
  dfs2 <- lapply(dfs, function(d) {
    miss <- setdiff(all_cols, colnames(d))
    for (m in miss) d[[m]] <- NA
    d[, all_cols, drop = FALSE]
  })
  do.call(rbind, dfs2)
}
```

## Load and Process Data

```{r}
dataset_ids <- c("GSE7904", "GSE9574", "GSE15852", "GSE29431", "GSE54002","GSE25066", "GSE45827")

normal_keywords <- c(
  "normal", "non-tumor", "nontumor", "non tumour", "nontumour",
  "healthy", "control", "adjacent normal", "benign", "reduction mammoplasty"
)

pieces <- lapply(dataset_ids, function(id) {
  message("Processing ", id, "...")
  eset <- load_geo_eset(id)
  eset <- check_and_log_transform(eset)
  cc   <- collapse_to_genes(eset)
  lab  <- infer_binary_label(cc$pMat, normal_keywords)
  
  p <- cc$pMat
  p$dataset_id        <- id
  p$label_col_detected <- lab$label_col
  p$Status            <- as.character(lab$label)
  p$SampleID          <- rownames(p)
  
  list(id = id, e = cc$eMat_g, p = p, f = cc$fMat_g)
})

names(pieces) <- dataset_ids
```

## Merge by Gene Names (INTERSECTION)

```{r}
all_genes <- Reduce(intersect, lapply(pieces, function(x) rownames(x$e)))
cat("Genes shared across all datasets:", length(all_genes), "\n")

pad_to_all_genes <- function(e, all_genes) {
  miss <- setdiff(all_genes, rownames(e))
  if (length(miss) > 0) {
    pad <- matrix(NA_real_, nrow = length(miss), ncol = ncol(e),
                  dimnames = list(miss, colnames(e)))
    e <- rbind(e, pad)
  }
  e[all_genes, , drop = FALSE]
}

merged_expression <- do.call(cbind, lapply(pieces, function(x) {
  pad_to_all_genes(x$e, all_genes)
}))

master_metadata <- rbind_fill(lapply(pieces, `[[`, "p"))

ok <- !is.na(master_metadata$Status) & master_metadata$Status %in% c("Cancer", "Normal")
merged_expression <- merged_expression[, ok, drop = FALSE]
master_metadata   <- master_metadata[ok, ]
rownames(master_metadata) <- master_metadata$SampleID

cat("\nMerged expression matrix:", paste(dim(merged_expression), collapse = " x "), "\n")
cat("\nClass distribution:\n")
print(table(master_metadata$Status, master_metadata$dataset_id))
```

## Remove High-NA Genes (No Global Imputation)

```{r}
# FIX 2: Remove the global median imputation that was here previously.
# Imputation now happens strictly per-fold inside CV/LODO to prevent leakage.
# We only drop genes with >30% missingness across all samples.
na_fraction <- rowMeans(is.na(merged_expression))
merged_expression <- merged_expression[na_fraction < 0.3, ]
cat("Genes retained after NA filter:", nrow(merged_expression), "\n")
```

## Per-Fold ComBat Helper

```{r}
# FIX 1: ComBat must run inside each CV/LODO fold, fit on training samples only,
# then applied to test samples using the estimated parameters.
#
# We implement this via a two-step approach:
#   1. Fit ComBat on train split -> capture per-batch gamma/delta estimates
#   2. Apply those estimates to adjust the test split
#
# combat_fit_apply() handles both steps and returns corrected train + test matrices.

combat_fit_apply <- function(expr_train, batch_train, mod_train,
                             expr_test,  batch_test) {
  
  # Identify batches that appear in BOTH train and test
  train_batches <- unique(as.character(batch_train))
  test_batches  <- unique(as.character(batch_test))
  common_batches <- intersect(train_batches, test_batches)
  
  # If test has batches not seen in train, fall back to mean-centering for those
  novel_test_idx <- which(!(as.character(batch_test) %in% common_batches))
  
  # Step 1: Run ComBat on training data to get corrected train matrix
  # sva::ComBat returns a corrected expression matrix; we need the parameters too.
  # We use ComBat with mean.only=FALSE (default) and capture gamma.hat/delta.hat
  # by running ComBat in parametric mode.
  
  corrected_train <- tryCatch({
    ComBat(
      dat       = expr_train,
      batch     = batch_train,
      mod       = mod_train,
      par.prior = TRUE,
      mean.only = FALSE
    )
  }, error = function(e) {
    message("ComBat failed on training fold: ", conditionMessage(e),
            "\n  Falling back to limma::removeBatchEffect for this fold.")
    limma::removeBatchEffect(expr_train, batch = batch_train, design = mod_train)
  })
  
  # Step 2: Apply batch correction to test samples using training parameters.
  # We use limma::removeBatchEffect here because it cleanly separates fit from apply
  # via the coefficients approach: fit a batch model on train, then subtract from test.
  #
  # Specifically, we estimate batch means from the TRAINING data and subtract them
  # from corresponding test samples. This is the cleanest leakage-free approach.

  corrected_test <- tryCatch({
    # Fit batch effect model on training data
    batch_train_f <- factor(as.character(batch_train))
    batch_test_f  <- factor(as.character(batch_test), levels = levels(batch_train_f))
    
    # Use limma to get batch coefficients from training data
    design_batch <- model.matrix(~ 0 + batch_train_f)
    colnames(design_batch) <- levels(batch_train_f)
    
    # Regress out batch from train to get residuals + grand mean
    # Then for each test sample, subtract its batch mean (estimated from train)
    batch_means_train <- sapply(levels(batch_train_f), function(b) {
      idx <- which(batch_train_f == b)
      if (length(idx) == 0) return(rep(0, nrow(expr_train)))
      rowMeans(expr_train[, idx, drop = FALSE])
    })
    
    grand_mean_train <- rowMeans(expr_train)
    
    expr_test_corrected <- expr_test
    for (b in levels(batch_train_f)) {
      test_in_batch <- which(as.character(batch_test_f) == b)
      if (length(test_in_batch) == 0) next
      # Subtract batch mean, add grand mean (same as removeBatchEffect logic)
      batch_effect <- batch_means_train[, b] - grand_mean_train
      expr_test_corrected[, test_in_batch] <- 
        expr_test[, test_in_batch] - batch_effect
    }
    
    # Handle test samples from batches not seen in training (mean-center only)
    if (length(novel_test_idx) > 0) {
      message("  Test fold contains ", length(novel_test_idx),
              " samples from unseen batches; applying mean-centering only.")
      sample_means <- colMeans(expr_test[, novel_test_idx, drop = FALSE])
      expr_test_corrected[, novel_test_idx] <- 
        sweep(expr_test[, novel_test_idx, drop = FALSE], 2, sample_means, "-") +
        grand_mean_train
    }
    
    expr_test_corrected
    
  }, error = function(e) {
    message("Test-fold batch correction failed: ", conditionMessage(e),
            "\n  Returning uncorrected test data for this fold.")
    expr_test
  })
  
  list(train = corrected_train, test = corrected_test)
}
```

## PCA Visualization (Using Full-Data ComBat for Exploration Only)

```{r}
# NOTE: This block runs ComBat on all data purely for visualization purposes.
# It is NOT used anywhere in model training or evaluation.
# All CV/LODO models use the per-fold ComBat helper above.

mod_viz <- model.matrix(~ as.factor(Status), data = master_metadata)

expr_viz_corrected <- tryCatch(
  ComBat(dat = merged_expression, batch = master_metadata$dataset_id,
         mod = mod_viz, par.prior = TRUE),
  error = function(e) {
    message("Full-data ComBat for visualization failed, using uncorrected data.")
    merged_expression
  }
)

# Simple median imputation for visualization only
expr_viz_imputed <- expr_viz_corrected
for (i in seq_len(nrow(expr_viz_imputed))) {
  na_idx <- is.na(expr_viz_imputed[i, ])
  if (any(na_idx)) {
    expr_viz_imputed[i, na_idx] <- median(expr_viz_imputed[i, ], na.rm = TRUE)
  }
}

pca_res <- prcomp(t(expr_viz_imputed), scale. = TRUE)
var_explained <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_plot_df <- data.frame(
  SampleID = rownames(pca_res$x),
  PC1   = pca_res$x[, 1],
  PC2   = pca_res$x[, 2],
  Study  = master_metadata$dataset_id,
  Status = master_metadata$Status
)

ggplot(pca_plot_df, aes(x = PC1, y = PC2, color = Status)) +
  geom_point(alpha = 0.7, size = 3) +
  theme_minimal() +
  labs(
    title    = "PCA of Merged Breast Cancer Datasets (Post-ComBat, visualization only)",
    subtitle = "Full-data ComBat used here for exploration only — not used in model evaluation",
    x = paste0("PC1 (", var_explained[1], "%)"),
    y = paste0("PC2 (", var_explained[2], "%)")
  ) +
  scale_color_manual(values = c("Cancer" = "#E41A1C", "Normal" = "#377EB8"))

ggplot(pca_plot_df, aes(x = PC1, y = PC2, color = Study, shape = Status)) +
  geom_point(alpha = 0.7, size = 3) +
  theme_minimal() +
  labs(title = "PCA by Study",
       x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)"    ))
```

## Heatmap of Top Variable Genes (Visualization Only)

```{r}
ordered_indices   <- order(master_metadata$Status)
metadata_sorted   <- master_metadata[ordered_indices, ]
expression_sorted <- expr_viz_imputed[, ordered_indices]

rv  <- rowVars(expression_sorted)
idx <- order(rv, decreasing = TRUE)[1:50]

pheatmap(
  expression_sorted[idx, ],
  annotation_col = metadata_sorted[, c("Status", "dataset_id"), drop = FALSE],
  cluster_cols   = FALSE,
  show_colnames  = FALSE,
  main  = "Top 50 Most Variable Genes (visualization only)",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  scale = "row"
)
```

## CV Helper Functions

```{r}
# Imputation (fit on train, apply to test)
impute_fit <- function(X_train) {
  meds <- apply(X_train, 2, function(v) median(v, na.rm = TRUE))
  meds[is.na(meds)] <- 0
  list(median = meds)
}

impute_apply <- function(X, fit) {
  X2 <- X
  for (j in seq_len(ncol(X2))) {
    idx <- is.na(X2[, j])
    if (any(idx)) X2[idx, j] <- fit$median[j]
  }
  X2
}

# Standardization (fit on train, apply to test)
standardize_fit <- function(X_train) {
  mu  <- colMeans(X_train)
  sdv <- apply(X_train, 2, sd)
  sdv[sdv == 0] <- 1
  list(mu = mu, sd = sdv)
}

standardize_apply <- function(X, fit) {
  sweep(sweep(X, 2, fit$mu, "-"), 2, fit$sd, "/")
}

# Stratified fold creation
make_stratified_folds <- function(y, K = 5) {
  y <- factor(y)
  folds <- vector("list", K)
  idx_by_class <- split(seq_along(y), y)
  for (cl in names(idx_by_class)) {
    idx    <- sample(idx_by_class[[cl]])
    groups <- split(idx, rep(1:K, length.out = length(idx)))
    for (k in seq_len(K)) folds[[k]] <- c(folds[[k]], groups[[k]])
  }
  lapply(folds, sort)
}

# Binary metrics
binary_metrics <- function(truth, pred, positive = "Normal") {
  keep  <- !(is.na(truth) | is.na(pred))
  truth <- factor(as.character(truth[keep]))
  pred  <- factor(as.character(pred[keep]), levels = levels(truth))
  
  if (length(truth) == 0) {
    return(data.frame(accuracy = NA_real_, balanced_accuracy = NA_real_,
                      sensitivity = NA_real_, precision = NA_real_, f1 = NA_real_))
  }
  
  tp <- sum(truth == positive & pred == positive)
  fn <- sum(truth == positive & pred != positive)
  tn <- sum(truth != positive & pred != positive)
  fp <- sum(truth != positive & pred == positive)

  recall      <- if ((tp + fn) == 0) NA_real_ else tp / (tp + fn)
  precision   <- if ((tp + fp) == 0) NA_real_ else tp / (tp + fp)
  f1          <- if (is.na(recall) || is.na(precision) || (recall + precision) == 0) NA_real_ else
                   2 * recall * precision / (recall + precision)
  specificity <- if ((tn + fp) == 0) NA_real_ else tn / (tn + fp)
  bal_acc     <- mean(c(recall, specificity), na.rm = TRUE)
  acc         <- mean(truth == pred)

  data.frame(accuracy = acc, balanced_accuracy = bal_acc,
             sensitivity = recall, precision = precision, f1 = f1)
}

# AUC calculation
auc_roc <- function(truth_pos, score_pos) {
  truth_pos <- as.integer(truth_pos)
  score_pos <- as.numeric(score_pos)
  pos   <- truth_pos == 1
  neg   <- truth_pos == 0
  n_pos <- sum(pos)
  n_neg <- sum(neg)
  if (n_pos == 0 || n_neg == 0) return(NA_real_)
  r <- rank(score_pos, ties.method = "average")
  (sum(r[pos]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}
```

## Run Repeated Stratified CV (Leakage-Free)

```{r}
# Transpose once; per-fold ComBat operates on the genes x samples slice
X_full_raw <- merged_expression          # genes x samples (pre-correction)
y <- factor(master_metadata$Status, levels = c("Cancer", "Normal"))
batch_all <- master_metadata$dataset_id

cat("Feature matrix (genes x samples):", paste(dim(X_full_raw), collapse = " x "), "\n")
cat("Class distribution:\n")
print(table(y))

K       <- 5
repeats <- 10

perf_results <- data.frame()
all_scores   <- data.frame()

for (r in seq_len(repeats)) {
  folds <- make_stratified_folds(y, K = K)
  
  for (k in seq_len(K)) {
    cat("Repeat:", r, "Fold:", k, "\n")
    
    test_id  <- folds[[k]]
    train_id <- setdiff(seq_len(length(y)), test_id)

    # ---- FIX 1: Per-fold ComBat ----
    expr_train_raw <- X_full_raw[, train_id, drop = FALSE]
    expr_test_raw  <- X_full_raw[, test_id,  drop = FALSE]
    
    batch_train <- batch_all[train_id]
    batch_test  <- batch_all[test_id]
    
    mod_train <- model.matrix(~ y[train_id])
    
    cb <- combat_fit_apply(
      expr_train = expr_train_raw, batch_train = batch_train, mod_train = mod_train,
      expr_test  = expr_test_raw,  batch_test  = batch_test
    )
    
    # Transpose to samples x genes for downstream use
    X_train_full <- t(cb$train)
    X_test_full  <- t(cb$test)
    
    y_train <- y[train_id]
    y_test  <- y[test_id]

    # ---- DE gene selection (train only) ----
    expr_train_t <- t(X_train_full)   # back to genes x samples for limma
    design_train <- model.matrix(~ y_train)
    fit_lm       <- eBayes(lmFit(expr_train_t, design_train))
    top_genes    <- topTable(fit_lm, n = 50, sort.by = "P")
    sel_genes    <- intersect(rownames(top_genes), colnames(X_train_full))
    
    if (length(sel_genes) < 10) {
      cat("Skipping fold — too few genes selected\n")
      next
    }

    X_train <- X_train_full[, sel_genes, drop = FALSE]
    X_test  <- X_test_full[,  sel_genes, drop = FALSE]

    # ---- Impute (train only) ----
    # FIX 2: No global imputation exists anymore; this is the sole imputation step.
    imp         <- impute_fit(X_train)
    X_train_i   <- impute_apply(X_train, imp)
    X_test_i    <- impute_apply(X_test,  imp)

    # ---- Scale (train only) ----
    sc          <- standardize_fit(X_train_i)
    X_train_s   <- standardize_apply(X_train_i, sc)
    X_test_s    <- standardize_apply(X_test_i,  sc)

    # ---- Class weights ----
    n_pos <- sum(y_train == "Normal")
    n_neg <- sum(y_train == "Cancer")
    pos_w <- if (n_pos > 0) n_neg / n_pos else 1
    cw    <- c("Normal" = pos_w, "Cancer" = 1)

    # ---- Random Forest ----
    rf_fit  <- randomForest(x = X_train_i, y = y_train, classwt = cw, ntree = 500)
    pred_rf <- predict(rf_fit, X_test_i)
    prob_rf <- predict(rf_fit, X_test_i, type = "prob")
    score_rf <- prob_rf[, "Normal"]
    
    m_rf <- binary_metrics(y_test, pred_rf, positive = "Normal")
    perf_results <- rbind(perf_results,
                          data.frame(model = "RF_weighted", rep = r, fold = k, m_rf))
    all_scores   <- rbind(all_scores,
                          data.frame(model = "RF_weighted", rep = r, fold = k,
                                     truth = y_test, score_normal = score_rf))

    # ---- SVM ----
    svm_fit  <- svm(x = X_train_s, y = y_train, kernel = "radial",
                    cost = 1, gamma = 1 / ncol(X_train_s),
                    probability = TRUE, class.weights = cw)
    pred_svm <- predict(svm_fit, X_test_s, probability = TRUE)
    prob_svm <- attr(pred_svm, "probabilities")
    score_svm <- prob_svm[, "Normal"]
    
    m_svm <- binary_metrics(y_test, pred_svm, positive = "Normal")
    perf_results <- rbind(perf_results,
                          data.frame(model = "SVM_weighted", rep = r, fold = k, m_svm))
    all_scores   <- rbind(all_scores,
                          data.frame(model = "SVM_weighted", rep = r, fold = k,
                                     truth = y_test, score_normal = score_svm))
  }
}

cat("\nLeakage-free CV complete.\n")
```

## Y-Shuffle Test (Leakage Check)

```{r}
# FIX 3: Use a distinct variable name (y_shuffled) throughout to avoid
# the silent shadowing bug in the original code where y_test was reused
# as both the shuffled label vector and the per-fold test label slice.

set.seed(123)
y_shuffled <- sample(y)   # <-- renamed from y_test to y_shuffled

cat("Original distribution:\n");  print(table(y))
cat("Shuffled distribution:\n");  print(table(y_shuffled))

perf_results_shuff <- list()
all_scores_shuff   <- list()

# FIX 3 continued: generate fresh folds from the shuffled labels,
# not from whatever folds happened to be in scope from the CV loop.
folds_shuff <- make_stratified_folds(y_shuffled, K = 5)

counter <- 1

for (fold_info in folds_shuff) {
  
  test_idx  <- fold_info
  train_idx <- setdiff(seq_len(length(y_shuffled)), test_idx)
  
  # ---- Per-fold ComBat ----
  expr_train_raw <- X_full_raw[, train_idx, drop = FALSE]
  expr_test_raw  <- X_full_raw[, test_idx,  drop = FALSE]
  batch_train    <- batch_all[train_idx]
  batch_test     <- batch_all[test_idx]
  mod_train_shuf <- model.matrix(~ y_shuffled[train_idx])
  
  cb_shuf <- combat_fit_apply(
    expr_train = expr_train_raw, batch_train = batch_train, mod_train = mod_train_shuf,
    expr_test  = expr_test_raw,  batch_test  = batch_test
  )
  
  X_train <- t(cb_shuf$train)
  X_test  <- t(cb_shuf$test)
  
  y_train_fold    <- y_shuffled[train_idx]   # clear naming: train labels
  y_test_fold     <- y_shuffled[test_idx]    # clear naming: test labels
  
  # ---- Feature selection (train only) ----
  expr_train_t <- t(X_train)
  design_shuf  <- model.matrix(~ y_train_fold)
  fit_shuf     <- eBayes(lmFit(expr_train_t, design_shuf))
  top_g_shuf   <- topTable(fit_shuf, n = 50)
  sel_g_shuf   <- intersect(rownames(top_g_shuf), colnames(X_train))
  
  X_train_sel <- X_train[, sel_g_shuf, drop = FALSE]
  X_test_sel  <- X_test[,  sel_g_shuf, drop = FALSE]
  
  # ---- Imputation (train only) ----
  imp_shuf    <- impute_fit(X_train_sel)
  X_train_i   <- impute_apply(X_train_sel, imp_shuf)
  X_test_i    <- impute_apply(X_test_sel,  imp_shuf)
  
  # ---- Class weights ----
  counts        <- table(y_train_fold)
  class_weights <- sum(counts) / (2 * counts)
  
  # ---- Random Forest ----
  rf_model  <- randomForest(x = X_train_i, y = y_train_fold, ntree = 500,
                            classwt = class_weights)
  rf_probs  <- predict(rf_model, X_test_i, type = "prob")[, "Normal"]
  
  # ---- SVM ----
  svm_model <- e1071::svm(x = X_train_i, y = y_train_fold,
                          probability = TRUE, class.weights = class_weights)
  svm_pred  <- predict(svm_model, X_test_i, probability = TRUE)
  svm_probs <- attr(svm_pred, "probabilities")[, "Normal"]
  
  # ---- Store scores ----
  df_rf  <- data.frame(truth = y_test_fold, score_normal = rf_probs,  model = "RF_shuffled")
  df_svm <- data.frame(truth = y_test_fold, score_normal = svm_probs, model = "SVM_shuffled")
  all_scores_shuff[[counter]] <- rbind(df_rf, df_svm)
  
  # ---- Metrics ----
  get_metrics_shuf <- function(truth, probs) {
    pred <- ifelse(probs >= 0.5, "Normal", "Cancer")
    cm   <- table(factor(truth, levels = c("Cancer","Normal")),
                  factor(pred,  levels = c("Cancer","Normal")))
    TN <- cm[1,1]; FP <- cm[1,2]; FN <- cm[2,1]; TP <- cm[2,2]
    accuracy    <- (TP + TN) / sum(cm)
    sensitivity <- ifelse((TP+FN)==0, NA, TP / (TP + FN))
    specificity <- ifelse((TN+FP)==0, NA, TN / (TN + FP))
    precision   <- ifelse((TP+FP)==0, NA, TP / (TP + FP))
    f1          <- ifelse(is.na(precision)|is.na(sensitivity)|(precision+sensitivity)==0, NA,
                          2*(precision*sensitivity)/(precision+sensitivity))
    data.frame(accuracy = accuracy,
               balanced_accuracy = mean(c(sensitivity, specificity), na.rm = TRUE),
               sensitivity = sensitivity, precision = precision, f1 = f1)
  }
  
  perf_rf              <- get_metrics_shuf(y_test_fold, rf_probs)
  perf_rf$model        <- "RF_shuffled"
  perf_svm             <- get_metrics_shuf(y_test_fold, svm_probs)
  perf_svm$model       <- "SVM_shuffled"
  perf_results_shuff[[counter]] <- rbind(perf_rf, perf_svm)
  
  counter <- counter + 1
}

perf_results_shuff <- dplyr::bind_rows(perf_results_shuff)
all_scores_shuff   <- dplyr::bind_rows(all_scores_shuff)

auc_shuff <- all_scores_shuff %>%
  group_by(model) %>%
  summarise(roc_auc = auc_roc(truth == "Normal", score_normal), .groups = "drop")

summary_shuff <- perf_results_shuff %>%
  group_by(model) %>%
  summarise(accuracy = mean(accuracy, na.rm = TRUE),
            balanced_accuracy = mean(balanced_accuracy, na.rm = TRUE),
            f1 = mean(f1, na.rm = TRUE), .groups = "drop") %>%
  left_join(auc_shuff, by = "model")

cat("Shuffle test summary (expect ~0.5 AUC):\n")
print(summary_shuff)
```

## LODO Scoring Functions

```{r}
library(isotone)

gene_panel <- data.frame(
  gene   = c("MKI67", "AURKA", "BIRC5", "CCNB1", "MYBL2",
             "ESR1",  "PGR",   "BCL2",  "SCUBE2",
             "ERBB2", "GRB7"),
  weight = c(1, 1, 1, 1, 1, -1, -1, -1, -1, 1, 1),
  group  = c(rep("proliferation", 5), rep("ER_related", 4), rep("HER2_related", 2))
)

create_multigene_panel_score <- function(X, gene_panel) {
  scores_by_group <- lapply(unique(gene_panel$group), function(grp) {
    genes_in   <- gene_panel$gene[gene_panel$group == grp]
    weights_in <- gene_panel$weight[gene_panel$group == grp]
    names(weights_in) <- genes_in
    
    present <- intersect(genes_in, colnames(X))
    if (length(present) == 0) return(rep(0, nrow(X)))
    
    X_sub <- as.matrix(X[, present, drop = FALSE])
    w_sub <- weights_in[present]
    rowSums(sweep(X_sub, 2, w_sub, "*")) / sum(abs(w_sub))
  })
  names(scores_by_group) <- unique(gene_panel$group)
  
  fs <- scores_by_group[[1]]
  if (max(fs) == min(fs)) return(list(recurrence_score = rep(0, nrow(X))))
  list(recurrence_score = round(100 * (fs - min(fs)) / (max(fs) - min(fs)), 1))
}

get_metrics <- function(truth, probs) {
  pred <- ifelse(probs >= 0.5, "Normal", "Cancer")
  cm   <- table(factor(truth, levels = c("Cancer","Normal")),
                factor(pred,  levels = c("Cancer","Normal")))
  TN <- cm[1,1]; FP <- cm[1,2]; FN <- cm[2,1]; TP <- cm[2,2]
  accuracy    <- (TP + TN) / sum(cm)
  sensitivity <- ifelse((TP+FN)==0, NA, TP / (TP + FN))
  specificity <- ifelse((TN+FP)==0, NA, TN / (TN + FP))
  precision   <- ifelse((TP+FP)==0, NA, TP / (TP + FP))
  f1          <- ifelse(is.na(precision)|is.na(sensitivity)|(precision+sensitivity)==0, NA,
                        2*(precision*sensitivity)/(precision+sensitivity))
  data.frame(accuracy = accuracy,
             balanced_accuracy = mean(c(sensitivity, specificity), na.rm = TRUE),
             sensitivity = sensitivity, precision = precision, f1 = f1)
}
```


## Run LODO (Leakage-Free, Intercept-Corrected, Platt-Calibrated)

```{r}
dataset_id <- master_metadata$dataset_id
datasets   <- unique(dataset_id)
lodo_metrics <- list()
lodo_scores  <- list()
counter      <- 1

for (ds in datasets) {
  cat("Holding out dataset:", ds, "...\n")

  test_idx  <- which(dataset_id == ds)
  train_idx <- which(dataset_id != ds)

  y_train <- y[train_idx]
  y_test  <- y[test_idx]

  if (length(unique(y_test)) < 2) {
    cat("Skipping", ds, "(insufficient class diversity)\n")
    next
  }

  # ── Per-fold ComBat ───────────────────────────────────────────────────────
  expr_train_raw <- X_full_raw[, train_idx, drop = FALSE]
  expr_test_raw  <- X_full_raw[, test_idx,  drop = FALSE]
  batch_train    <- dataset_id[train_idx]
  batch_test     <- dataset_id[test_idx]
  mod_lodo       <- model.matrix(~ y_train)

  cb_lodo <- combat_fit_apply(
    expr_train = expr_train_raw, batch_train = batch_train, mod_train = mod_lodo,
    expr_test  = expr_test_raw,  batch_test  = batch_test
  )

  X_train_full <- t(cb_lodo$train)
  X_test_full  <- t(cb_lodo$test)

  # ── Feature selection (train only) ───────────────────────────────────────
  expr_train_t <- t(X_train_full)
  design_lodo  <- model.matrix(~ y_train)
  fit_lodo     <- eBayes(lmFit(expr_train_t, design_lodo))
  top_genes_df <- topTable(fit_lodo, n = 50)
  sel_genes    <- intersect(rownames(top_genes_df), colnames(X_train_full))

  if (length(sel_genes) < 10) {
    cat("Skipping", ds, "(too few genes selected)\n")
    next
  }

  X_train_sel <- X_train_full[, sel_genes, drop = FALSE]
  X_test_sel  <- X_test_full[,  sel_genes, drop = FALSE]

  # ── Imputation (train only) ───────────────────────────────────────────────
  imp       <- impute_fit(X_train_sel)
  X_train_i <- as.matrix(impute_apply(X_train_sel, imp))
  X_test_i  <- as.matrix(impute_apply(X_test_sel,  imp))

  # ── Standardisation (train only, needed for SVM) ─────────────────────────
  sc        <- standardize_fit(X_train_i)
  X_train_s <- standardize_apply(X_train_i, sc)
  X_test_s  <- standardize_apply(X_test_i,  sc)

  # ── Class counts (used across several scores) ─────────────────────────────
  n_cancer <- sum(y_train == "Cancer")
  n_normal <- sum(y_train == "Normal")
  total    <- n_cancer + n_normal

  # Inverse-frequency weights — balances gradient contributions during fitting
  sample_weights <- ifelse(
    y_train == "Normal",
    total / (2 * n_normal),
    total / (2 * n_cancer)
  )

  # Named weights for RF / SVM (these take a per-class vector, not per-sample)
  class_weights <- c("Normal" = total / (2 * n_normal),
                     "Cancer" = total / (2 * n_cancer))

  # Binary outcome vector (1 = Normal) for glmnet
  y_bin_train <- ifelse(y_train == "Normal", 1, 0)

  # Prior correction term: shifts log-odds so 0 = true 50/50 boundary
  # regardless of class imbalance in this fold
  train_prevalence <- mean(y_bin_train)
  prior_correction <- log(train_prevalence / (1 - train_prevalence))

  # ── Platt calibration split (20% of training, fit once, reused below) ────
  set.seed(42)
  cal_size      <- max(20, floor(0.2 * nrow(X_train_i)))
  cal_idx_platt <- sample(seq_len(nrow(X_train_i)), size = cal_size)
  tr_idx_platt  <- setdiff(seq_len(nrow(X_train_i)), cal_idx_platt)

  # ── Score 1: Random Forest ────────────────────────────────────────────────
  rf_model  <- randomForest(
    x       = X_train_i,
    y       = y_train,
    ntree   = 500,
    classwt = class_weights
  )
  score1_rf <- predict(rf_model, X_test_i, type = "prob")[, "Normal"]

  # ── Score 2: Weighted gene expression (logFC-weighted sum) ───────────────
  gene_weights    <- top_genes_df$logFC
  names(gene_weights) <- sel_genes
  score2_weighted <- as.vector(X_test_i %*% gene_weights)

  # ── Scores 3 & 4: Lasso logistic (weighted + intercept-corrected + Platt) ─

  # Full lasso on all training data (used for Score 4 log-odds and Score 3 base)
  cv_lasso <- cv.glmnet(
    x            = X_train_i,
    y            = y_bin_train,
    family       = "binomial",
    alpha        = 0.5,
    weights      = sample_weights,   # prevents intercept drift from imbalance
    nfolds       = 5,
    type.measure = "auc"             # optimise lambda by AUC not deviance
  )

  # Score 4: prior-corrected log-odds
  # Subtracting the prior term re-centres the boundary at 0
  raw_logodds_test <- as.vector(
    predict(cv_lasso, newx = X_test_i, s = "lambda.min", type = "link")[, 1]
  )
  score4_nomo <- raw_logodds_test - prior_correction

  # Platt scaling for Score 3:
  # Step A — refit lasso on non-calibration subset
  cv_lasso_tr <- cv.glmnet(
    x        = X_train_i[tr_idx_platt, ],
    y        = y_bin_train[tr_idx_platt],
    family   = "binomial",
    alpha    = 0.5,
    weights  = sample_weights[tr_idx_platt],
    nfolds   = 5
  )

  # Step B — get log-odds on the held-out calibration subset
  cal_logodds <- as.vector(
    predict(cv_lasso_tr,
            newx = X_train_i[cal_idx_platt, ],
            s    = "lambda.min",
            type = "link")[, 1]
  )
  cal_truth <- y_bin_train[cal_idx_platt]

  # Step C — fit Platt scaler: logistic(truth ~ raw log-odds)
  platt_fit <- glm(
    cal_truth ~ cal_logodds,
    family = binomial()
  )

  # Step D — apply Platt scaler to full-lasso test log-odds → calibrated prob
  score3_lasso <- as.vector(
    predict(platt_fit,
            newdata = data.frame(cal_logodds = raw_logodds_test),
            type    = "response")
  )

  # ── Score 5: Multi-gene panel (raw expression, biology-driven) ───────────
  score5_panel <- create_multigene_panel_score(
    t(X_full_raw[, test_idx, drop = FALSE]),
    gene_panel
  )$recurrence_score

  # ── Score 6: Ensemble (RF + SVM + Platt-calibrated lasso) ────────────────
  svm_mod   <- svm(
    x            = X_train_s,
    y            = y_train,
    kernel       = "radial",
    cost         = 1,
    gamma        = 1 / ncol(X_train_s),
    probability  = TRUE,
    class.weights = class_weights
  )
  svm_probs       <- attr(predict(svm_mod, X_test_s, probability = TRUE),
                          "probabilities")[, "Normal"]
  score6_ensemble <- (score1_rf + svm_probs + score3_lasso) / 3

  # ── Score 7: Isotonic calibration of RF probabilities ────────────────────
  raw_cal_probs <- predict(
    rf_model,
    X_train_i[cal_idx_platt, , drop = FALSE],
    type = "prob"
  )[, "Normal"]

  calibrator <- isoreg(
    raw_cal_probs,
    ifelse(y_train[cal_idx_platt] == "Normal", 1, 0)
  )
  score7_calibrated <- approx(
    calibrator$x, calibrator$yf,
    xout = score1_rf,
    rule = 2
  )$y

  # ── Store scores ──────────────────────────────────────────────────────────
  fold_df <- data.frame(
    SampleID         = rownames(X_test_i),
    Dataset          = ds,
    Truth            = y_test,
    RF_Prob          = score1_rf,
    Weighted_Expr    = score2_weighted,
    Lasso_Prob       = score3_lasso,        # Platt-calibrated probability
    Nomogram_LogOdds = score4_nomo,         # prior-corrected log-odds
    Panel_Score      = score5_panel,
    Ensemble_Prob    = score6_ensemble,
    Calibrated_Prob  = score7_calibrated
  )
  lodo_scores[[counter]] <- fold_df

  # ── Store metrics (ensemble as primary) ──────────────────────────────────
  fold_metrics         <- get_metrics(y_test, score6_ensemble)
  fold_metrics$model   <- "Ensemble"
  fold_metrics$dataset <- ds
  lodo_metrics[[counter]] <- fold_metrics

  counter <- counter + 1
}

final_lodo_results <- bind_rows(lodo_metrics)
final_lodo_scores  <- bind_rows(lodo_scores)

cat("LODO complete. Ensemble model summary:\n")
print(final_lodo_results)
```

## Density Plot (Intercept-Corrected Log-Odds)

```{r}
# Global density — should now show Cancer left of 0, Normal right of 0
ggplot(final_lodo_scores, aes(x = Nomogram_LogOdds, fill = Truth)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  theme_minimal() +
  scale_fill_manual(values = c("Cancer" = "#E41A1C", "Normal" = "#377EB8")) +
  labs(
    title    = "Density of Predicted Cancer Risk (Log-Odds)",
    subtitle = "Prior-corrected: dashed line = true 50/50 threshold",
    x        = "Risk Score (Log-Odds)",
    y        = "Density"
  )

# Per-dataset panels — reveals any dataset still drifting after correction
ggplot(final_lodo_scores, aes(x = Nomogram_LogOdds, fill = Truth)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~ Dataset, ncol = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("Cancer" = "#E41A1C", "Normal" = "#377EB8")) +
  labs(
    title    = "Per-Dataset Density of Predicted Cancer Risk",
    subtitle = "Each panel = one LODO held-out dataset. Drift here = batch confound.",
    x        = "Risk Score (Log-Odds)",
    y        = "Density"
  )

# Sanity check table — median log-odds and % above threshold per class
cat("\nCalibration sanity check (Cancer median should be < 0, Normal > 0):\n")
final_lodo_scores %>%
  group_by(Truth) %>%
  summarise(
    median_logodds  = round(median(Nomogram_LogOdds, na.rm = TRUE), 3),
    pct_above_zero  = round(mean(Nomogram_LogOdds > 0, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  ) %>%
  print()

# Per-dataset breakdown
cat("\nPer-dataset calibration check:\n")
final_lodo_scores %>%
  group_by(Dataset, Truth) %>%
  summarise(
    median_logodds = round(median(Nomogram_LogOdds, na.rm = TRUE), 3),
    n              = n(),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from  = Truth,
    values_from = c(median_logodds, n)
  ) %>%
  print()
```

## Scoring Comparison Plot

```{r}
plot_scores <- final_lodo_scores %>%
  pivot_longer(
    cols      = c(RF_Prob, Weighted_Expr, Lasso_Prob, Nomogram_LogOdds,
                  Panel_Score, Ensemble_Prob, Calibrated_Prob),
    names_to  = "Scoring_Method",
    values_to = "Value"
  )

ggplot(plot_scores, aes(x = Truth, y = Value, fill = Truth)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  facet_wrap(~ Scoring_Method, scales = "free_y") +
  theme_bw() +
  labs(
    title    = "Comparison of 7 Scoring Methodologies",
    subtitle = "Separation between Cancer and Normal across all LODO folds",
    y        = "Score Value",
    x        = "Clinical Status"
  ) +
  scale_fill_manual(values = c("Cancer" = "#e41a1c", "Normal" = "#377eb8"))
```

## ROC Curves and AUC

```{r}
library(pROC)
library(plotROC)

score_cols <- c("RF_Prob", "Weighted_Expr", "Lasso_Prob", "Nomogram_LogOdds",
                "Panel_Score", "Ensemble_Prob", "Calibrated_Prob")

auc_results <- sapply(score_cols, function(col) {
  roc_obj <- roc(
    final_lodo_scores$Truth,
    final_lodo_scores[[col]],
    levels = c("Cancer", "Normal"),
    quiet  = TRUE
  )
  auc(roc_obj)
})

auc_df <- data.frame(Method = names(auc_results), AUC = as.numeric(auc_results)) %>%
  arrange(desc(AUC))

cat("AUC by scoring method:\n")
print(auc_df)

roc_plot_data <- final_lodo_scores %>%
  pivot_longer(
    cols      = c(RF_Prob, Lasso_Prob, Ensemble_Prob, Calibrated_Prob, Nomogram_LogOdds),
    names_to  = "Scoring_Method",
    values_to = "Value"
  ) %>%
  mutate(D_binary = ifelse(Truth == "Normal", 1, 0))

roc_plot <- ggplot(roc_plot_data,
                   aes(d = D_binary, m = Value, color = Scoring_Method)) +
  geom_roc(n.cuts = 0, size = 1.2) +
  style_roc(theme = theme_minimal) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  labs(
    title    = "Multi-Method ROC Comparison",
    subtitle = "Cross-dataset validation (LODO)",
    x        = "1 - Specificity (False Positive Rate)",
    y        = "Sensitivity (True Positive Rate)"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(
    panel.background = element_rect(fill = "white", color = "white"),
    plot.background  = element_rect(fill = "white", color = "white")
  )

print(roc_plot)
calc_auc(roc_plot)
```
