---
title: "Breast Cancer PAM50 Subtype Risk Calculator (Multiclass)"
format:
  html:
    embed-resources: true
    code-fold: true
    df-print: paged
    toc: true
editor: visual
---

This notebook builds a **PAM50 subtype classifier** from merged GEO microarray datasets. It is structured in three clearly separated stages:

1.  **QC & normalisation** — per-dataset, completely separate from modelling
2.  **Model building** — LODO cross-validation with fold-safe batch correction
3.  **Risk calculator** — subtype probabilities with uncertainty quantification

```{r setup}
#| message: false
#| warning: false
library(GEOquery)
library(limma)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(sva)
library(glmnet)
if (requireNamespace("nnet", quietly = TRUE)) library(nnet)

set.seed(3888)

# ── Dataset registry ──────────────────────────────────────────────────────────
datasets_multiclass <- c("GSE25066", "GSE45827","GSE65194")

# ── PAM50 subtype dictionary ──────────────────────────────────────────────────
subtype_variants <- list(
  "Luminal A"   = c("luminal a","lum a","luma","lum-a","lum_a","luminala"),
  "Luminal B"   = c("luminal b","lum b","lumb","lum-b","lum_b","luminalb"),
  "Basal"       = c("basal","basal-like","basal like","basallike","basal_like"),
  "HER2"        = c("her2","her2+","her-2","her2-enriched","her2enriched","her2e"),
  "Normal-like" = c("normal-like","normal like","normallike","normal_like")
)
```

------------------------------------------------------------------------

# Stage 1 — QC & Normalisation (per dataset)

> **Design principle:** normalisation is applied within each dataset *before* any merging or modelling. This ensures samples within a study are on a common scale. Cross-dataset harmonisation (ComBat) happens later, *inside* each CV fold.

## Helper functions

```{r helpers-qc}
#| message: false
#| warning: false

# ── GEO loader ────────────────────────────────────────────────────────────────
load_geo_eset <- function(gse_id) {
  geo_cache_dir <- file.path("data", "geo-cache")
  if (!dir.exists(geo_cache_dir)) dir.create(geo_cache_dir, recursive = TRUE)

  gse_matrix_url <- function(id, host = "ftp.ncbi.nlm.nih.gov") {
    num    <- sub("^GSE", "", id)
    prefix <- substr(num, 1, max(1, nchar(num) - 3))
    paste0("https://", host, "/geo/series/GSE", prefix, "nnn/", id,
           "/matrix/", id, "_series_matrix.txt.gz")
  }

  g <- tryCatch(GEOquery::getGEO(gse_id, destdir = geo_cache_dir), error = function(e) e)
  if (!inherits(g, "error")) {
    if (inherits(g, "ExpressionSet")) return(g)
    if (is.list(g) && length(g) >= 1) return(g[[1]])
  }

  destfile <- file.path(geo_cache_dir, paste0(gse_id, "_series_matrix.txt.gz"))
  for (host in c("ftp.ncbi.nlm.nih.gov", "download.ncbi.nlm.nih.gov")) {
    for (attempt in 1:3) {
      try({
        h <- curl::new_handle(); curl::handle_setopt(h, connecttimeout=60, timeout=1800)
        curl::curl_download(gse_matrix_url(gse_id, host), destfile, quiet=FALSE, handle=h)
      }, silent = TRUE)
      if (file.exists(destfile) && file.info(destfile)$size > 0) break
      Sys.sleep(3 * attempt)
    }
  }
  g2 <- GEOquery::getGEO(filename = destfile)
  if (inherits(g2, "ExpressionSet")) return(g2)
  if (is.list(g2) && length(g2) >= 1) return(g2[[1]])
  stop("Could not load ExpressionSet for ", gse_id)
}

# ── Log2 transform (GEO2R heuristic) ─────────────────────────────────────────
needs_log2 <- function(e) {
  v <- as.numeric(e); v <- v[is.finite(v)]
  if (length(v) < 10) return(FALSE)
  qx <- quantile(v, c(0, 0.25, 0.5, 0.75, 0.99, 1), na.rm = TRUE)
  (qx[5] > 100) || (qx[6] - qx[1] > 50 && qx[2] > 0) ||
    (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
}

safe_log2 <- function(e) {
  pos    <- e[e > 0 & is.finite(e)]
  pseudo <- min(pos, na.rm = TRUE) / 2
  e2 <- e; e2[!is.finite(e2)] <- NA_real_; e2[e2 <= 0] <- pseudo
  log2(e2)
}

# ── Within-dataset quantile normalisation ────────────────────────────────────
quantile_normalise <- function(e) limma::normalizeBetweenArrays(e, method = "quantile")

# ── Probe → gene collapse (avereps) ──────────────────────────────────────────
collapse_probes <- function(eset) {
  e   <- exprs(eset)
  fd  <- fData(eset)
  sym_col <- grep("gene\\s*symbol|^symbol$", colnames(fd), ignore.case=TRUE, value=TRUE)[1]
  gene_sym_raw <- if (!is.na(sym_col)) as.character(fd[, sym_col]) else rep(NA_character_, nrow(fd))
  gene_sym <- ifelse(is.na(gene_sym_raw) | trimws(gene_sym_raw) == "",
                     rownames(e),
                     vapply(gene_sym_raw, function(s) {
                       s <- trimws(as.character(s))
                       if (is.na(s) || s == "") return(NA_character_)
                       trimws(strsplit(s, " /// ", fixed=TRUE)[[1]][1])
                     }, character(1)))
  gene_sym[is.na(gene_sym) | gene_sym == ""] <- rownames(e)[is.na(gene_sym) | gene_sym == ""]
  limma::avereps(e, ID = gene_sym)
}

# ── PAM50 label inference ─────────────────────────────────────────────────────
norm_label_txt <- function(x) {
  x <- tolower(as.character(x)); x[is.na(x)] <- ""
  trimws(gsub("[^a-z0-9]+", " ", x))
}

map_to_subtype <- function(x_raw) {
  x_chr  <- as.character(x_raw)
  x_norm <- norm_label_txt(x_chr)
  if (x_norm == "") return(NA_character_)
  if (grepl("pam50_class", tolower(x_chr), fixed=TRUE)) {
    m <- regexpr("pam50_class:\\s*", x_chr, ignore.case=TRUE)
    if (m > 0) x_norm <- norm_label_txt(substring(x_chr, m + attr(m,"match.length")))
  }
  for (canon in names(subtype_variants)) {
    if (any(vapply(subtype_variants[[canon]],
                   function(kw) grepl(kw, x_norm, fixed=TRUE), logical(1))))
      return(canon)
  }
  NA_character_
}

infer_multiclass_label <- function(pMat) {
  cols_try <- c(
    grep("pam50",  colnames(pMat), ignore.case=TRUE, value=TRUE),
    grep("^characteristics_ch", colnames(pMat), value=TRUE),
    setdiff(colnames(pMat), c(
      grep("pam50", colnames(pMat), ignore.case=TRUE, value=TRUE),
      grep("^characteristics_ch", colnames(pMat), value=TRUE)))
  )
  for (cn in cols_try) {
    lab <- vapply(as.character(pMat[[cn]]), map_to_subtype, character(1))
    y   <- droplevels(factor(lab[!is.na(lab)]))
    if (nlevels(y) >= 2) return(list(label = factor(lab), label_col = cn))
  }
  list(label = factor(rep(NA_character_, nrow(pMat))), label_col = NA_character_)
}

# ── Full per-dataset QC pipeline ──────────────────────────────────────────────
process_dataset_qc <- function(gse_id) {
  eset <- load_geo_eset(gse_id)
  e    <- exprs(eset)

  log2_applied <- needs_log2(e)
  if (log2_applied) e <- safe_log2(e)

  e_norm <- quantile_normalise(e)
  exprs(eset) <- e_norm

  e_gene <- collapse_probes(eset)
  lab    <- infer_multiclass_label(pData(eset))

  p              <- pData(eset)
  p$dataset_id   <- gse_id
  p$label_mc     <- as.character(lab$label)

  list(
    id           = gse_id,
    e_raw        = e,          # log2-only, pre-normalisation (for QC plot)
    e_norm       = e_gene,     # quantile-normalised + gene-collapsed
    pMat         = p,
    log2_applied = log2_applied,
    label_col    = lab$label_col
  )
}
```

## Load and normalise each dataset

```{r load-datasets}
#| message: false
#| warning: false
qc_pieces <- lapply(datasets_multiclass, function(id) {
  tryCatch(process_dataset_qc(id), error = function(e) {
    message("Failed: ", id, " — ", conditionMessage(e)); NULL
  })
})
names(qc_pieces) <- datasets_multiclass
qc_pieces <- Filter(Negate(is.null), qc_pieces)

for (id in names(qc_pieces)) {
  p <- qc_pieces[[id]]
  cat("\n", id, "— log2 applied:", p$log2_applied,
      "| genes:", nrow(p$e_norm),
      "| samples:", ncol(p$e_norm), "\n")
  print(table(p$pMat$label_mc, useNA = "ifany"))
}
```

## QC: Per-sample expression distributions

```{r qc-boxplots, fig.cap="Per-sample expression distributions after within-dataset quantile normalisation. Boxes should be tightly aligned within each dataset."}
#| message: false
#| warning: false
set.seed(3888)
qc_box_list <- lapply(names(qc_pieces), function(id) {
  p       <- qc_pieces[[id]]
  n_genes <- min(1500L, nrow(p$e_norm))
  gsamp   <- sample(rownames(p$e_norm), n_genes)
  e_sub   <- p$e_norm[gsamp, , drop = FALSE]
  lab_vec <- p$pMat$label_mc

  data.frame(
    expr    = as.vector(e_sub),
    sample  = rep(seq_len(ncol(e_sub)), each = nrow(e_sub)),
    dataset = id,
    subtype = rep(lab_vec, each = nrow(e_sub)),
    stringsAsFactors = FALSE
  )
})

qc_box_df <- bind_rows(qc_box_list)

# Sample median for ordering
sample_med <- qc_box_df %>%
  group_by(dataset, sample) %>%
  summarise(med = median(expr, na.rm=TRUE), .groups="drop") %>%
  arrange(dataset, med) %>%
  group_by(dataset) %>%
  mutate(sample_ord = row_number()) %>%
  ungroup()

qc_box_df <- qc_box_df %>%
  left_join(sample_med %>% select(dataset, sample, sample_ord),
            by = c("dataset","sample"))

for (ds in unique(qc_box_df$dataset)) {
  df_ds <- qc_box_df %>% filter(dataset == ds)
  print(
    ggplot(df_ds, aes(x = factor(sample_ord), y = expr, colour = subtype)) +
      geom_boxplot(outlier.shape = NA, size = 0.2) +
      theme_bw(base_size = 10) +
      labs(x = "Sample (ordered by median)", y = "log2 expression",
           colour = "Subtype",
           title = paste0(ds, " — per-sample expression after quantile normalisation")) +
      theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  )
}

# Summary table
qc_box_df %>%
  group_by(dataset) %>%
  summarise(n_samples  = n_distinct(sample),
            median_expr = round(median(expr, na.rm=TRUE), 3),
            IQR_expr    = round(IQR(expr,   na.rm=TRUE), 3),
            q01         = round(quantile(expr, .01, na.rm=TRUE), 3),
            q99         = round(quantile(expr, .99, na.rm=TRUE), 3),
            .groups = "drop") %>%
  knitr::kable(caption = "Per-dataset expression summary (post normalisation)")
```

## QC: PCA per dataset

```{r qc-pca, fig.cap="PCA of each dataset after normalisation. Points coloured by PAM50 subtype. Subtypes should separate along PCs."}
#| message: false
#| warning: false
for (id in names(qc_pieces)) {
  p   <- qc_pieces[[id]]
  ok  <- !is.na(p$pMat$label_mc) & p$pMat$label_mc != ""
  e_s <- p$e_norm[, ok, drop=FALSE]
  y_s <- p$pMat$label_mc[ok]

  e_v <- e_s[apply(e_s, 1, var, na.rm=TRUE) > 0, ]
  if (nrow(e_v) < 5 || ncol(e_v) < 3) { cat(id, ": skipping PCA\n"); next }

  # impute NAs with row means before PCA
  e_v[is.na(e_v)] <- rowMeans(e_v, na.rm=TRUE)[row(e_v)[is.na(e_v)]]

  pca <- prcomp(t(e_v), scale.=TRUE)
  ve  <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 1)

  df <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2], subtype = y_s)
  print(
    ggplot(df, aes(PC1, PC2, colour = subtype)) +
      geom_point(size = 2, alpha = .8) +
      theme_bw() +
      labs(x = paste0("PC1 (", ve[1], "%)"),
           y = paste0("PC2 (", ve[2], "%)"),
           colour = "Subtype",
           title  = paste0(id, " — PCA after normalisation"))
  )
}
```

------------------------------------------------------------------------

# Stage 2 — Model Building

> **Design principle:** normalised datasets are merged by gene intersection. Batch correction (ComBat) is applied **inside** each LODO fold using training samples only, then transferred to the test fold.

## Merge normalised datasets

```{r merge-datasets}
#| message: false
#| warning: false

# Keep only genes present in all datasets
all_genes_mc <- Reduce(intersect, lapply(qc_pieces, function(p) rownames(p$e_norm)))
cat("Shared genes across all multiclass datasets:", length(all_genes_mc), "\n")

# Filter genes with >30% missing across all samples
e_merged_mc <- do.call(cbind, lapply(qc_pieces, function(p)
  p$e_norm[all_genes_mc, , drop=FALSE]))

na_frac <- rowMeans(is.na(e_merged_mc))
e_merged_mc <- e_merged_mc[na_frac < 0.3, , drop=FALSE]
cat("Genes after NA filter:", nrow(e_merged_mc), "\n")

p_merged_mc <- do.call(rbind, lapply(qc_pieces, function(p) {
  data.frame(sample_id  = rownames(p$pMat),
             dataset_id = p$id,
             label_mc   = p$pMat$label_mc,
             stringsAsFactors = FALSE)
}))

ok_mc <- !is.na(p_merged_mc$label_mc) & p_merged_mc$label_mc != ""
y_mc   <- factor(p_merged_mc$label_mc[ok_mc])
e_mc   <- e_merged_mc[, ok_mc, drop=FALSE]
batch_mc <- p_merged_mc$dataset_id[ok_mc]

cat("\nClass counts (merged):\n"); print(table(y_mc))
cat("\nSamples per dataset:\n");  print(table(batch_mc))
```

## Modelling helper functions

```{r helpers-model}
#| message: false
#| warning: false

# ── Macro F1 ──────────────────────────────────────────────────────────────────
macro_f1 <- function(truth, pred) {
  truth <- factor(truth); pred <- factor(pred, levels=levels(truth))
  lvls  <- levels(truth)
  f1s <- vapply(lvls, function(k) {
    tp <- sum(pred==k & truth==k, na.rm=TRUE)
    fp <- sum(pred==k & truth!=k, na.rm=TRUE)
    fn <- sum(pred!=k & truth==k, na.rm=TRUE)
    p  <- if ((tp+fp)==0) NA_real_ else tp/(tp+fp)
    r  <- if ((tp+fn)==0) NA_real_ else tp/(tp+fn)
    if (is.na(p)||is.na(r)||(p+r)==0) NA_real_ else 2*p*r/(p+r)
  }, numeric(1))
  mean(f1s, na.rm=TRUE)
}

# ── Macro OvR AUC ────────────────────────────────────────────────────────────
auc_roc_vec <- function(truth01, score) {
  truth01 <- as.integer(truth01); score <- as.numeric(score)
  n_pos <- sum(truth01==1); n_neg <- sum(truth01==0)
  if (n_pos==0 || n_neg==0) return(NA_real_)
  r <- rank(score, ties.method="average")
  (sum(r[truth01==1]) - n_pos*(n_pos+1)/2) / (n_pos*n_neg)
}

macro_auc_ovr <- function(truth, prob_mat) {
  lvls <- colnames(prob_mat)
  aucs <- vapply(lvls, function(k) {
    tk <- as.integer(as.character(truth) == k)
    tk[is.na(tk)] <- 0L
    n_pos <- sum(tk, na.rm=TRUE)
    if (n_pos==0L || n_pos==length(tk)) return(NA_real_)
    auc_roc_vec(tk, prob_mat[, k])
  }, numeric(1))
  mean(aucs, na.rm=TRUE)
}

# ── ComBat: fit on train, apply to test ───────────────────────────────────────
combat_fit_apply <- function(expr_train, batch_train, mod_train,
                              expr_test, batch_test) {
  if (length(unique(as.character(batch_train))) < 2) {
    message("Single batch in fold — skipping ComBat.")
    return(list(train=expr_train, test=expr_test))
  }
  corrected_train <- tryCatch(
    sva::ComBat(dat=expr_train, batch=batch_train, mod=mod_train,
                par.prior=TRUE, mean.only=FALSE),
    error = function(e) {
      message("ComBat failed, falling back to removeBatchEffect: ", conditionMessage(e))
      limma::removeBatchEffect(expr_train, batch=batch_train, design=mod_train)
    }
  )
  # Apply training batch means to test
  batch_train_f <- factor(as.character(batch_train))
  batch_test_f  <- factor(as.character(batch_test), levels=levels(batch_train_f))
  grand_mean    <- rowMeans(expr_train)
  batch_means   <- sapply(levels(batch_train_f), function(b) {
    idx <- which(batch_train_f==b)
    if (length(idx)==0) return(rep(0, nrow(expr_train)))
    rowMeans(expr_train[, idx, drop=FALSE])
  })
  corrected_test <- expr_test
  for (b in levels(batch_train_f)) {
    idx <- which(as.character(batch_test_f)==b)
    if (length(idx)==0) next
    corrected_test[, idx] <- expr_test[, idx, drop=FALSE] -
      (batch_means[, b] - grand_mean)
  }
  # Novel test batches: mean-centre only
  novel <- which(!(as.character(batch_test) %in% levels(batch_train_f)))
  if (length(novel) > 0) {
    smeans <- colMeans(expr_test[, novel, drop=FALSE])
    corrected_test[, novel] <- sweep(expr_test[, novel, drop=FALSE], 2, smeans, "-") + grand_mean
  }
  list(train=corrected_train, test=corrected_test)
}

# ── Imputation & standardisation (fit on train, apply to test) ───────────────
impute_fit    <- function(X) list(med=apply(X,2,function(v) {m<-median(v,na.rm=TRUE); if(is.na(m)) 0 else m}))
impute_apply  <- function(X, fit) { X2<-X; for(j in seq_len(ncol(X2))) { idx<-is.na(X2[,j]); if(any(idx)) X2[idx,j]<-fit$med[j] }; X2 }
std_fit       <- function(X) { mu<-colMeans(X); sd<-apply(X,2,sd); sd[sd==0]<-1; list(mu=mu,sd=sd) }
std_apply     <- function(X, fit) sweep(sweep(X,2,fit$mu,"-"),2,fit$sd,"/")

# ── OvR limma gene selection ──────────────────────────────────────────────────
select_genes_ovr <- function(e_mat, y, n_per_class=25) {
  ok <- !is.na(y) & y!=""; e<-e_mat[, ok, drop=FALSE]; y<-factor(y[ok])
  lvls_s <- make.names(levels(y))
  design  <- model.matrix(~0+y); colnames(design)<-lvls_s
  fit     <- lmFit(e, design)
  genes   <- character(0)
  for (i in seq_along(levels(y))) {
    lvl <- lvls_s[i]; others <- lvls_s[-i]
    cont_str <- paste0(lvl," - (",paste(others,collapse="+"),")/",length(others))
    cont <- makeContrasts(contrasts=cont_str, levels=design)
    fit2 <- eBayes(contrasts.fit(fit, cont))
    genes <- union(genes, rownames(topTable(fit2, n=n_per_class, sort.by="P")))
  }
  genes
}

# ── Stratified split helper ───────────────────────────────────────────────────
stratified_split <- function(y, dataset, train_prop=0.75) {
  dev_idx <- c(); test_idx <- c()
  for (ds in unique(as.character(dataset))) {
    for (cl in unique(as.character(y))) {
      idx <- which(as.character(dataset)==ds & as.character(y)==cl)
      if (length(idx)==0) next
      idx <- sample(idx)
      n_dev <- max(1, min(floor(length(idx)*train_prop), length(idx)-1))
      dev_idx  <- c(dev_idx,  idx[seq_len(n_dev)])
      test_idx <- c(test_idx, idx[(n_dev+1):length(idx)])
    }
  }
  list(dev=sort(unique(dev_idx)), test=sort(unique(test_idx)))
}
```

## LODO cross-validation

```{r lodo-cv}
#| message: false
#| warning: false

run_lodo_multiclass <- function(X, y, batch, n_per_class=25, use_weights=TRUE) {
  y     <- factor(y)
  batch <- as.character(batch)
  perf_list  <- list()
  probs_list <- list()

  for (ds in unique(batch)) {
    test_id  <- which(batch==ds)
    train_id <- setdiff(seq_along(batch), test_id)

    y_train <- droplevels(y[train_id])
    y_test  <- factor(as.character(y[test_id]), levels=levels(y_train))
    if (nlevels(y_train)<2) next

    # Batch correction (train only → apply to test)
    cb <- combat_fit_apply(
      expr_train=t(X[train_id,,drop=FALSE]), batch_train=batch[train_id],
      mod_train=model.matrix(~y_train),
      expr_test=t(X[test_id,,drop=FALSE]),  batch_test=batch[test_id]
    )
    X_tr_full <- t(cb$train); X_te_full <- t(cb$test)

    # Gene selection from training data only
    genes_sel <- select_genes_ovr(t(X_tr_full), y=y_train, n_per_class=n_per_class)
    sel_g     <- intersect(genes_sel, colnames(X_tr_full))
    if (length(sel_g)<10) next

    X_tr <- X_tr_full[, sel_g]; X_te <- X_te_full[, sel_g]
    imp  <- impute_fit(X_tr); X_tr_i <- impute_apply(X_tr,imp); X_te_i <- impute_apply(X_te,imp)
    sc   <- std_fit(X_tr_i);  X_tr_s <- std_apply(X_tr_i,sc);  X_te_s <- std_apply(X_te_i,sc)

    lvls <- levels(y_train); n_cl <- length(lvls); n_te <- nrow(X_te_s)
    tab_tr <- table(y_train); n_tot <- sum(tab_tr)
    cw <- if (use_weights)
      setNames(vapply(lvls, function(k) n_tot/(n_cl*max(1,tab_tr[k])), numeric(1)), lvls)
    else setNames(rep(1,n_cl), lvls)
    sw <- cw[as.character(y_train)]

    y_te_chr <- as.character(y_test)

    record <- function(nm, pred_chr, prob_mat) {
      perf_list[[length(perf_list)+1]] <<- tibble::tibble(
        fold=ds, model=nm,
        accuracy  = mean(pred_chr==y_te_chr, na.rm=TRUE),
        macro_f1  = macro_f1(y_te_chr, pred_chr),
        macro_auc = macro_auc_ovr(y_te_chr, prob_mat)
      )
      pd <- as.data.frame(prob_mat); pd$fold<-ds; pd$model<-nm; pd$truth<-y_te_chr
      probs_list[[length(probs_list)+1]] <<- pd
    }

    # kNN
    pk  <- class::knn(X_tr_s, X_te_s, y_train, k=7, prob=TRUE)
    kw  <- attr(pk,"prob"); pkc <- as.character(pk)
    pm_knn <- matrix(0,n_te,n_cl,dimnames=list(NULL,lvls))
    for(i in seq_len(n_te)) {
      pm_knn[i,pkc[i]] <- kw[i]
      oth <- setdiff(lvls,pkc[i])
      if (length(oth)) pm_knn[i,oth] <- (1-kw[i])/length(oth)
    }
    record("kNN", pkc, pm_knn)

    # SVM-RBF
    svm_fit <- tryCatch(
      e1071::svm(X_tr_s, y_train, kernel="radial", cost=1,
                 gamma=1/ncol(X_tr_s), probability=TRUE,
                 class.weights=if(use_weights)cw else NULL),
      error=function(e) NULL
    )
    if (!is.null(svm_fit)) {
      ps  <- predict(svm_fit, X_te_s, probability=TRUE)
      pm  <- attr(ps,"probabilities")[,lvls,drop=FALSE]
      record(if(use_weights)"SVM_RBF_w"else"SVM_RBF", as.character(ps), pm)
    }

    # Random Forest
    rf_fit  <- randomForest::randomForest(X_tr_i, y_train,
                 classwt=if(use_weights)cw else NULL)
    pred_rf <- predict(rf_fit, X_te_i)
    prob_rf <- predict(rf_fit, X_te_i, type="prob")[,lvls,drop=FALSE]
    record(if(use_weights)"RF_w"else"RF", as.character(pred_rf), prob_rf)

    # Multinomial Lasso
    if (requireNamespace("glmnet", quietly=TRUE)) {
      lf <- tryCatch(
        glmnet::cv.glmnet(as.matrix(X_tr_s), y_train, family="multinomial",
                          alpha=1, weights=sw, standardize=FALSE,
                          type.multinomial="grouped"),
        error=function(e) NULL
      )
      if (!is.null(lf)) {
        pr <- predict(lf, as.matrix(X_te_s), type="response", s="lambda.min")[,,1]
        if (is.null(dim(pr))) pr <- t(as.matrix(pr))
        pr <- pr[, intersect(lvls,colnames(pr)), drop=FALSE]
        record(if(use_weights)"LassoMC_w"else"LassoMC", lvls[max.col(pr)], pr)
      }
    }

    # Multinomial logistic (nnet)
    if (requireNamespace("nnet", quietly=TRUE)) {
      mn <- tryCatch(
        nnet::multinom(y_train~., data=data.frame(y_train, X_tr_s),
                       weights=sw, trace=FALSE, MaxNWts=100000),
        error=function(e) NULL
      )
      if (!is.null(mn)) {
        pm_mn <- predict(mn, newdata=data.frame(X_te_s), type="probs")
        if (is.null(dim(pm_mn))) pm_mn <- t(as.matrix(pm_mn))
        pm_mn <- pm_mn[, intersect(lvls,colnames(pm_mn)), drop=FALSE]
        record(if(use_weights)"Multinom_w"else"Multinom", lvls[max.col(pm_mn)], pm_mn)
      }
    }
  }

  list(perf=bind_rows(perf_list), probs=bind_rows(probs_list))
}

X_mc <- t(e_mc)
cat("Running LODO CV...\n")
res_lodo <- run_lodo_multiclass(X_mc, y_mc, batch_mc, n_per_class=25, use_weights=TRUE)
cat("Models evaluated:", paste(unique(res_lodo$perf$model), collapse=", "), "\n")
```

------------------------------------------------------------------------

# Stage 3 — Results

## Model scorecard

```{r scorecard}
#| message: false
#| warning: false
scorecard <- res_lodo$perf %>%
  group_by(model) %>%
  summarise(
    macro_auc_mean = mean(macro_auc, na.rm=TRUE),
    macro_auc_sd   = sd(macro_auc,   na.rm=TRUE),
    macro_f1_mean  = mean(macro_f1,  na.rm=TRUE),
    macro_f1_sd    = sd(macro_f1,    na.rm=TRUE),
    accuracy_mean  = mean(accuracy,  na.rm=TRUE),
    accuracy_sd    = sd(accuracy,    na.rm=TRUE),
    .groups="drop"
  ) %>%
  arrange(desc(macro_auc_mean), desc(macro_f1_mean)) %>%
  mutate(
    rank      = row_number(),
    macro_auc = sprintf("%.3f ± %.3f", macro_auc_mean, macro_auc_sd),
    macro_f1  = sprintf("%.3f ± %.3f", macro_f1_mean,  macro_f1_sd),
    accuracy  = sprintf("%.3f ± %.3f", accuracy_mean,  accuracy_sd)
  ) %>%
  select(rank, model, macro_auc, macro_f1, accuracy)

knitr::kable(scorecard,
  caption="Model scorecard — LODO CV (ranked by macro OvR AUC → macro F1 → accuracy)")

best_model_mc <- res_lodo$perf %>%
  group_by(model) %>%
  summarise(m=mean(macro_auc, na.rm=TRUE), .groups="drop") %>%
  arrange(desc(m)) %>% slice(1) %>% pull(model)
cat("\nBest single model:", best_model_mc, "\n")
```

## Performance plots

```{r perf-plots, fig.cap="LODO CV performance across models. Higher is better for all three metrics."}
#| message: false
#| warning: false
res_lodo$perf %>%
  pivot_longer(cols=c(macro_auc, macro_f1, accuracy),
               names_to="metric", values_to="value") %>%
  mutate(metric = recode(metric,
    macro_auc = "Macro OvR AUC",
    macro_f1  = "Macro F1",
    accuracy  = "Accuracy")) %>%
  ggplot(aes(x=reorder(model, value, median), y=value, fill=model)) +
  geom_boxplot(alpha=0.75, outlier.size=0.8) +
  facet_wrap(~metric, scales="free_y") +
  coord_flip() +
  theme_bw() +
  labs(x=NULL, y=NULL, title="LODO CV: model comparison") +
  theme(legend.position="none")
```

## Confusion matrix (best model)

```{r confusion-best}
#| message: false
#| warning: false
preds_best <- res_lodo$probs %>%
  filter(model==best_model_mc) %>%
  mutate(pred = {
    prob_cols <- setdiff(colnames(.), c("fold","model","truth"))
    lvls_here <- prob_cols[prob_cols %in% levels(y_mc)]
    lvls_here[max.col(.[ , lvls_here, drop=FALSE])]
  })

tab_best <- table(truth=preds_best$truth, pred=preds_best$pred)
as.data.frame(tab_best) %>%
  ggplot(aes(x=pred, y=truth, fill=Freq)) +
  geom_tile(colour="white") +
  geom_text(aes(label=Freq), size=4) +
  scale_fill_gradient(low="white", high="#2C7FB8") +
  theme_bw() +
  labs(x="Predicted", y="True",
       title=paste0("Confusion matrix — ", best_model_mc, " (pooled LODO)")) +
  theme(axis.text.x=element_text(angle=40, hjust=1))
```

------------------------------------------------------------------------

# Single model vs Ensemble — is it worth it?

```{r ensemble-eval}
#| message: false
#| warning: false

# ── Dev / hold-out split ──────────────────────────────────────────────────────
split_mc <- stratified_split(y_mc, batch_mc, train_prop=0.75)
dev_idx  <- split_mc$dev; hold_idx <- split_mc$test

X_dev  <- X_mc[dev_idx,,drop=FALSE]; y_dev  <- y_mc[dev_idx]; batch_dev  <- batch_mc[dev_idx]
X_hold <- X_mc[hold_idx,,drop=FALSE]; y_hold <- y_mc[hold_idx]; batch_hold <- batch_mc[hold_idx]
cat("Dev:", nrow(X_dev), " Hold-out:", nrow(X_hold), "\n")

# LODO on dev to rank models
res_dev <- run_lodo_multiclass(X_dev, y_dev, batch_dev, n_per_class=25)

dev_rank <- res_dev$perf %>%
  group_by(model) %>%
  summarise(mean_auc=mean(macro_auc,na.rm=TRUE),
            mean_f1 =mean(macro_f1, na.rm=TRUE), .groups="drop") %>%
  arrange(desc(mean_auc)) %>% mutate(dev_rank=row_number())

# ── Fit on full dev, predict hold-out ─────────────────────────────────────────
fit_once_mc <- function(X_tr, y_tr, batch_tr, X_te, batch_te,
                        n_per_class=25, use_weights=TRUE) {
  y_tr  <- factor(y_tr); lvls <- levels(y_tr); n_cl <- length(lvls)
  cb    <- combat_fit_apply(t(X_tr), batch_tr, model.matrix(~y_tr), t(X_te), batch_te)
  X_trc <- t(cb$train); X_tec <- t(cb$test)
  gs    <- intersect(select_genes_ovr(t(X_trc), y=y_tr, n_per_class=n_per_class), colnames(X_trc))
  if (length(gs)<10) return(NULL)
  X_t2 <- X_trc[,gs]; X_e2 <- X_tec[,gs]
  imp  <- impute_fit(X_t2); X_ti <- impute_apply(X_t2,imp); X_ei <- impute_apply(X_e2,imp)
  sc   <- std_fit(X_ti);   X_ts <- std_apply(X_ti,sc);    X_es <- std_apply(X_ei,sc)
  tab_tr <- table(y_tr); n_tot <- sum(tab_tr); n_te <- nrow(X_es)
  cw <- if(use_weights) setNames(vapply(lvls,function(k) n_tot/(n_cl*max(1,tab_tr[k])),numeric(1)),lvls) else setNames(rep(1,n_cl),lvls)
  sw <- cw[as.character(y_tr)]
  out <- list()

  pk <- class::knn(X_ts,X_es,y_tr,k=7,prob=TRUE)
  kw <- attr(pk,"prob"); pkc <- as.character(pk)
  pm_k <- matrix(0,n_te,n_cl,dimnames=list(NULL,lvls))
  for(i in seq_len(n_te)){pm_k[i,pkc[i]]<-kw[i]; oth<-setdiff(lvls,pkc[i]); if(length(oth))pm_k[i,oth]<-(1-kw[i])/length(oth)}
  out[["kNN"]] <- pm_k

  rf2 <- randomForest::randomForest(X_ti,y_tr,classwt=if(use_weights)cw else NULL)
  out[[if(use_weights)"RF_w"else"RF"]] <- predict(rf2,X_ei,type="prob")[,lvls,drop=FALSE]

  if(requireNamespace("glmnet",quietly=TRUE)){
    lf <- tryCatch(glmnet::cv.glmnet(as.matrix(X_ts),y_tr,family="multinomial",alpha=1,weights=sw,standardize=FALSE,type.multinomial="grouped"),error=function(e)NULL)
    if(!is.null(lf)){pr<-predict(lf,as.matrix(X_es),type="response",s="lambda.min")[,,1]; if(is.null(dim(pr)))pr<-t(as.matrix(pr)); out[[if(use_weights)"LassoMC_w"else"LassoMC"]]<-pr[,intersect(lvls,colnames(pr)),drop=FALSE]}
  }
  if(requireNamespace("nnet",quietly=TRUE)){
    mn <- tryCatch(nnet::multinom(y_tr~.,data=data.frame(y_tr,X_ts),weights=sw,trace=FALSE,MaxNWts=100000),error=function(e)NULL)
    if(!is.null(mn)){pm_mn<-predict(mn,newdata=data.frame(X_es),type="probs"); if(is.null(dim(pm_mn)))pm_mn<-t(as.matrix(pm_mn)); out[[if(use_weights)"Multinom_w"else"Multinom"]]<-pm_mn[,intersect(lvls,colnames(pm_mn)),drop=FALSE]}
  }
  out
}

hold_probs <- fit_once_mc(X_dev, y_dev, batch_dev, X_hold, batch_hold)

if (!is.null(hold_probs)) {
  lvls_hold <- levels(y_dev)

  # Single model metrics on hold-out
  single_hold <- bind_rows(lapply(names(hold_probs), function(nm) {
    pm       <- hold_probs[[nm]]
    lv       <- intersect(lvls_hold, colnames(pm))
    pred_chr <- lv[max.col(pm[,lv,drop=FALSE])]
    tibble::tibble(
      type      = "Single model",
      id        = nm,
      accuracy  = mean(pred_chr==as.character(y_hold), na.rm=TRUE),
      macro_f1  = macro_f1(as.character(y_hold), pred_chr),
      macro_auc = macro_auc_ovr(as.character(y_hold), pm[,lv,drop=FALSE])
    )
  })) %>% arrange(desc(macro_auc))

  # Ensemble metrics (top-k equal-weight averaging)
  ens_hold <- bind_rows(lapply(c(2,3,4), function(k) {
    sel_mods <- dev_rank %>% slice_head(n=k) %>% pull(model)
    sel_mods <- intersect(sel_mods, names(hold_probs))
    if (length(sel_mods)<2) return(NULL)
    avg_pm   <- Reduce("+", lapply(sel_mods, function(nm) {
      pm <- hold_probs[[nm]]; pm[,intersect(lvls_hold,colnames(pm)),drop=FALSE]
    })) / length(sel_mods)
    pred_chr <- colnames(avg_pm)[max.col(avg_pm)]
    tibble::tibble(
      type      = paste0("Ensemble (top-", k, ")"),
      id        = paste(sel_mods, collapse=" + "),
      accuracy  = mean(pred_chr==as.character(y_hold), na.rm=TRUE),
      macro_f1  = macro_f1(as.character(y_hold), pred_chr),
      macro_auc = macro_auc_ovr(as.character(y_hold), avg_pm)
    )
  })) %>% arrange(desc(macro_auc))

  comparison_tbl <- bind_rows(single_hold, ens_hold) %>%
    mutate(across(c(macro_auc, macro_f1, accuracy), round, 3))

  knitr::kable(comparison_tbl,
    caption="Hold-out: single model vs ensemble (equal-weight probability averaging)")

  # ── Visual comparison ─────────────────────────────────────────────────────
  comparison_tbl %>%
    pivot_longer(cols=c(macro_auc, macro_f1, accuracy),
                 names_to="metric", values_to="value") %>%
    mutate(metric=recode(metric,
      macro_auc="Macro OvR AUC", macro_f1="Macro F1", accuracy="Accuracy")) %>%
    ggplot(aes(x=reorder(id, value), y=value, fill=type)) +
    geom_col(alpha=0.85) +
    geom_text(aes(label=round(value,3)), hjust=-0.1, size=3) +
    facet_wrap(~metric, scales="free_x") +
    coord_flip() +
    theme_bw() +
    labs(x=NULL, y=NULL, fill=NULL,
         title="Single model vs ensemble — hold-out performance",
         subtitle="Is the extra complexity of an ensemble justified?") +
    theme(legend.position="bottom")
}
```

------------------------------------------------------------------------

# Uncertainty Quantification

> **Two complementary uncertainty signals:**
>
> 1.  **Model disagreement** — when ensemble members assign very different probabilities to the same sample, gene expression alone may not be sufficient to make a confident call.
>
> 2.  **Bootstrap confidence intervals** — resampling the hold-out set 500 times gives a distribution over macro AUC and per-sample risk scores, quantifying how stable the estimates are.

## Model disagreement per sample

```{r uncertainty-disagreement, fig.cap="Model disagreement (SD of predicted subtype probability across ensemble members). High disagreement = low confidence in the call."}
#| message: false
#| warning: false
if (!is.null(hold_probs) && length(hold_probs) >= 2) {
  lvls_u <- lvls_hold
  sel_mods_u <- dev_rank %>% slice_head(n=min(4,nrow(dev_rank))) %>%
    pull(model) %>% intersect(names(hold_probs))

  # For each sample × subtype, collect probabilities across models
  prob_stack <- array(NA_real_, dim=c(nrow(X_hold), length(lvls_u), length(sel_mods_u)),
                      dimnames=list(NULL, lvls_u, sel_mods_u))
  for (nm in sel_mods_u) {
    pm <- hold_probs[[nm]]; lv <- intersect(lvls_u, colnames(pm))
    prob_stack[, lv, nm] <- pm[, lv]
  }

  # Predicted class = argmax of ensemble average
  avg_stack <- apply(prob_stack, c(1,2), mean, na.rm=TRUE)
  pred_class <- lvls_u[max.col(avg_stack)]

  # Disagreement = SD of predicted-class probability across models
  pred_class_idx <- match(pred_class, lvls_u)
  disagreement <- vapply(seq_len(nrow(X_hold)), function(i) {
    probs_for_pred <- prob_stack[i, pred_class_idx[i], ]
    sd(probs_for_pred, na.rm=TRUE)
  }, numeric(1))

  # Entropy of ensemble mean (overall uncertainty)
  entropy <- apply(avg_stack, 1, function(p) {
    p <- p[p > 0]; -sum(p * log(p))
  })

  uncertainty_df <- tibble::tibble(
    sample_idx    = seq_len(nrow(X_hold)),
    true_subtype  = as.character(y_hold),
    pred_subtype  = pred_class,
    correct       = pred_class == as.character(y_hold),
    confidence    = apply(avg_stack, 1, max),
    disagreement  = disagreement,
    entropy       = entropy
  )

  # Plot: confidence vs disagreement
  p1 <- ggplot(uncertainty_df, aes(x=confidence, y=disagreement, colour=correct)) +
    geom_point(alpha=0.75, size=2) +
    scale_colour_manual(values=c("TRUE"="#41AB5D","FALSE"="#D7191C"),
                        labels=c("TRUE"="Correct","FALSE"="Incorrect")) +
    theme_bw() +
    labs(x="Ensemble confidence (max P)", y="Model disagreement (SD of P)",
         colour=NULL,
         title="Confidence vs model disagreement",
         subtitle="Top-right = high confidence but models disagree (flag for review)")

  # Plot: disagreement by true subtype
  p2 <- ggplot(uncertainty_df, aes(x=true_subtype, y=disagreement, fill=true_subtype)) +
    geom_boxplot(alpha=0.75, outlier.size=0.8) +
    theme_bw() +
    labs(x="True subtype", y="Model disagreement (SD)",
         title="Which subtypes are hardest to predict consistently?") +
    theme(legend.position="none",
          axis.text.x=element_text(angle=30, hjust=1))

  print(p1); print(p2)

  cat("\nHigh-uncertainty samples (disagreement > 75th percentile):\n")
  threshold_u <- quantile(uncertainty_df$disagreement, 0.75, na.rm=TRUE)
  knitr::kable(
    uncertainty_df %>%
      filter(disagreement > threshold_u) %>%
      arrange(desc(disagreement)) %>%
      mutate(across(c(confidence, disagreement, entropy), round, 3)) %>%
      head(15),
    caption="High-uncertainty samples — models disagree most on these"
  )
}
```

## Per-patient bootstrap CI on subtype probabilities

> For each bootstrap iteration we: **(1)** resample the dev (training) set with replacement, **(2)** refit the best model (Multinomial Lasso) on that resample, **(3)** predict subtype probabilities on the **fixed** hold-out patients. The spread of B predictions per patient is the CI on that patient's risk score.

```{r bootstrap-ci}
#| message: false
#| warning: false
set.seed(3888)
B_boot_mc <- 200

run_bootstrap_mc <- function(X_tr, y_tr, batch_tr, X_te, batch_te,
                              B = 200, n_per_class = 25) {
  y_tr  <- factor(y_tr)
  lvls  <- levels(y_tr)
  n_cl  <- length(lvls)
  n_te  <- nrow(X_te)

  # Batch-correct full training set once
  cb <- combat_fit_apply(t(X_tr), batch_tr, model.matrix(~y_tr), t(X_te), batch_te)
  X_trc <- t(cb$train); X_tec <- t(cb$test)

  # Gene selection on full corrected training set
  gs <- tryCatch({
    sel <- select_genes_ovr(t(X_trc), y=y_tr, n_per_class=n_per_class)
    intersect(sel, colnames(X_trc))
  }, error=function(e) character(0))
  if (length(gs) < 10) { message("Too few genes for bootstrap"); return(NULL) }

  X_t2 <- X_trc[,gs]; X_e2 <- X_tec[,gs]
  imp  <- impute_fit(X_t2)
  X_ti <- impute_apply(X_t2,imp); X_ei <- impute_apply(X_e2,imp)
  sc   <- std_fit(X_ti)
  X_ts <- std_apply(X_ti,sc);    X_es <- std_apply(X_ei,sc)

  n_tr <- nrow(X_ts)
  tab_tr <- table(y_tr); n_tot <- sum(tab_tr)
  cw <- setNames(vapply(lvls,function(k) n_tot/(n_cl*max(1,tab_tr[k])),numeric(1)),lvls)
  sw_full <- cw[as.character(y_tr)]

  # Find best lambda ONCE on full training set — reuse in every bootstrap iteration
  # This avoids 10-fold inner CV per iteration (~10x speedup)
  cv_once <- tryCatch(
    glmnet::cv.glmnet(as.matrix(X_ts), y_tr, family="multinomial",
                      alpha=1, weights=sw_full, standardize=FALSE,
                      type.multinomial="grouped"),
    error=function(e) NULL
  )
  if (is.null(cv_once)) { message("cv.glmnet setup failed"); return(NULL) }
  lambda_fixed_mc <- cv_once$lambda.min
  cat(sprintf("  Fixed lambda: %.5f\n", lambda_fixed_mc))

  # Boot array: samples × subtypes × iterations
  boot_arr <- array(NA_real_, dim=c(n_te, n_cl, B),
                    dimnames=list(NULL, lvls, NULL))

  for (b in seq_len(B)) {
    idx_b  <- sample(n_tr, n_tr, replace=TRUE)
    y_b    <- droplevels(y_tr[idx_b])
    X_b    <- X_ts[idx_b,,drop=FALSE]
    if (nlevels(y_b) < 2) next

    tab_b <- table(y_b); n_tot_b <- sum(tab_b)
    lvls_b <- levels(y_b)
    sw_b   <- setNames(vapply(lvls_b,function(k) n_tot_b/(length(lvls_b)*max(1,tab_b[k])),
                              numeric(1)), lvls_b)[as.character(y_b)]

    # Plain glmnet with fixed lambda — no inner CV per iteration
    lf <- tryCatch(
      glmnet::glmnet(as.matrix(X_b), y_b, family="multinomial",
                     alpha=1, weights=sw_b, standardize=FALSE,
                     lambda=lambda_fixed_mc,
                     type.multinomial="grouped"),
      error=function(e) NULL
    )
    if (is.null(lf)) next

    pr <- predict(lf, as.matrix(X_es), type="response",
                  s=lambda_fixed_mc)[,,1]
    if (is.null(dim(pr))) pr <- t(as.matrix(pr))
    lv_common <- intersect(lvls, colnames(pr))
    boot_arr[, lv_common, b] <- pr[, lv_common]
  }

  boot_arr
}

cat("Running", B_boot_mc, "bootstrap iterations (Multinomial Lasso refit)...
")
boot_arr_mc <- run_bootstrap_mc(
  X_dev, y_dev, batch_dev, X_hold, batch_hold,
  B=B_boot_mc, n_per_class=25
)

if (!is.null(boot_arr_mc)) {
  lvls_b <- dimnames(boot_arr_mc)[[2]]

  # Per-patient, per-subtype: mean and 95% CI across B iterations
  prob_mean <- apply(boot_arr_mc, c(1,2), mean, na.rm=TRUE)
  prob_lo   <- apply(boot_arr_mc, c(1,2), quantile, 0.025, na.rm=TRUE)
  prob_hi   <- apply(boot_arr_mc, c(1,2), quantile, 0.975, na.rm=TRUE)
  ci_width  <- prob_hi - prob_lo

  # Predicted subtype = argmax of mean probability
  pred_subtype_boot <- lvls_b[max.col(prob_mean)]
  confidence_boot   <- apply(prob_mean, 1, max)
  ci_of_pred        <- vapply(seq_len(nrow(prob_mean)), function(i)
    ci_width[i, pred_subtype_boot[i]], numeric(1))

  boot_risk_mc <- tibble::tibble(
    sample_id      = seq_len(nrow(X_hold)),
    true_subtype   = as.character(y_hold),
    pred_subtype   = pred_subtype_boot,
    correct        = pred_subtype_boot == as.character(y_hold),
    confidence     = round(confidence_boot, 3),
    ci_width_pred  = round(ci_of_pred, 3),
    uncertainty_band = case_when(
      ci_of_pred <= 0.10 ~ "Narrow (confident)",
      ci_of_pred <= 0.20 ~ "Moderate",
      TRUE               ~ "Wide (uncertain — review)"
    )
  )

  # Add per-subtype probability + CI columns
  for (k in lvls_b) {
    boot_risk_mc[[paste0("P_", k)]]      <- round(prob_mean[, k], 3)
    boot_risk_mc[[paste0("CI_lo_", k)]]  <- round(prob_lo[,  k], 3)
    boot_risk_mc[[paste0("CI_hi_", k)]]  <- round(prob_hi[,  k], 3)
  }

  cat("
Uncertainty band distribution:
")
  print(table(boot_risk_mc$uncertainty_band))
  cat("
Accuracy (bootstrap mean prediction):",
      round(mean(boot_risk_mc$correct, na.rm=TRUE), 3), "
")

  knitr::kable(
    boot_risk_mc %>%
      arrange(desc(confidence)) %>%
      select(sample_id, true_subtype, pred_subtype, correct,
             confidence, ci_width_pred, uncertainty_band) %>%
      head(20),
    caption=paste0("Per-patient PAM50 predictions with bootstrap 95% CI (B=",
                   B_boot_mc, " Lasso refits). Wide CI = uncertain.")
  )

  # ── Forest plot: top 40 most confident patients ──
  boot_risk_mc %>%
    arrange(desc(confidence)) %>%
    slice_head(n=40) %>%
    mutate(sample_id=factor(sample_id, levels=rev(sample_id))) %>%
    ggplot(aes(x=confidence, y=sample_id, colour=correct)) +
    geom_errorbarh(aes(xmin=confidence - ci_width_pred/2,
                       xmax=confidence + ci_width_pred/2),
                   height=0.4, alpha=0.6) +
    geom_point(size=2) +
    scale_colour_manual(values=c("TRUE"="#41AB5D","FALSE"="#D7191C"),
                        labels=c("TRUE"="Correct","FALSE"="Incorrect")) +
    theme_bw(base_size=9) +
    labs(x="Predicted subtype confidence  [point = mean, bar = 95% CI]",
         y="Hold-out patient",
         colour=NULL,
         title=paste0("Per-patient subtype confidence with bootstrap uncertainty (B=",B_boot_mc,")"),
         subtitle="Wide bars = model uncertain. Dashed = 0.5 threshold.") +
    geom_vline(xintercept=0.5, linetype=2, colour="grey50") +
    theme(axis.text.y=element_blank(), axis.ticks.y=element_blank(),
          legend.position="bottom")

  # ── CI width by true subtype ──
  ggplot(boot_risk_mc, aes(x=true_subtype, y=ci_width_pred, fill=true_subtype)) +
    geom_violin(alpha=0.7, trim=FALSE) +
    geom_boxplot(width=0.15, fill="white", outlier.size=0.8) +
    theme_bw() +
    labs(x="True subtype", y="Bootstrap 95% CI width on predicted subtype",
         title="Which subtypes are hardest to predict with confidence?",
         subtitle="Wider CI = model less certain across training resamples") +
    theme(legend.position="none",
          axis.text.x=element_text(angle=30, hjust=1))

  cat(sprintf("\nMedian CI width: %.3f  |  Mean CI width: %.3f\n",
    median(boot_risk_mc$ci_width_pred, na.rm=TRUE),
    mean(boot_risk_mc$ci_width_pred, na.rm=TRUE)))
} else {
  cat("Bootstrap could not run.\n")
}
```

------------------------------------------------------------------------

# PAM50 Subtype Risk Calculator

```{r risk-calculator}
#| message: false
#| warning: false
if (!is.null(hold_probs) && exists("uncertainty_df")) {
  risk_df <- uncertainty_df %>%
    bind_cols(as.data.frame(avg_stack)) %>%
    mutate(
      risk_band = case_when(
        confidence >= 0.70 & disagreement <= quantile(disagreement, 0.25, na.rm=TRUE) ~ "High confidence",
        confidence >= 0.50 ~ "Moderate confidence",
        TRUE ~ "Low confidence — review recommended"
      )
    ) %>%
    select(sample_idx, true_subtype, pred_subtype, correct,
           confidence, disagreement,
           any_of("boot_ci_width"),  # populated later by bootstrap section
           entropy, risk_band,
           all_of(lvls_hold))

  cat("\nRisk band distribution:\n")
  print(table(risk_df$risk_band))

  knitr::kable(
    risk_df %>%
      arrange(desc(confidence)) %>%
      mutate(across(all_of(lvls_hold), ~paste0(round(.x * 100, 1), "%")),
       confidence    = paste0(round(confidence * 100, 1), "%"),
       across(any_of("boot_ci_width"), ~paste0(round(.x * 100, 1), "%")),
       across(c(disagreement, entropy), round, 3)) %>%
      head(20),
    caption="PAM50 subtype risk calculator — top 20 most confident predictions"
  )

  # Distribution of confidence by predicted subtype
  ggplot(risk_df, aes(x=pred_subtype, y=confidence, fill=pred_subtype)) +
    geom_violin(alpha=0.7, trim=FALSE) +
    geom_boxplot(width=0.15, fill="white", outlier.size=0.8) +
    theme_bw() +
    labs(x="Predicted subtype", y="Confidence (max P)",
         title="Prediction confidence by PAM50 subtype") +
    theme(legend.position="none",
          axis.text.x=element_text(angle=30, hjust=1))

  # Stacked probability bars (top 30 samples sorted by confidence)
  risk_df %>%
    arrange(desc(confidence)) %>%
    slice_head(n=30) %>%
    mutate(sample_idx=factor(sample_idx, levels=sample_idx)) %>%
    pivot_longer(cols=all_of(lvls_hold), names_to="subtype", values_to="prob") %>%
    ggplot(aes(x=sample_idx, y=prob, fill=subtype)) +
    geom_col(width=0.9) +
    theme_bw(base_size=9) +
    labs(x="Sample (top 30 by confidence)", y="P(subtype)", fill="Subtype",
         title="Predicted subtype probability profiles — top 30 most confident") +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
}
```

## Summary for clinical interpretation

```{r summary-table}
#| message: false
#| warning: false
if (exists("risk_df")) {
  cat("=== CLASSIFIER SUMMARY ===\n\n")
  cat("Best single model (LODO CV):", best_model_mc, "\n")

  best_ens_k <- ens_hold %>% arrange(desc(macro_auc)) %>% slice(1)
  cat("Best ensemble:", best_ens_k$id, "\n")
  cat(sprintf("  High-confidence predictions: %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band == "High confidence"),
              nrow(risk_df),
              100 * mean(risk_df$risk_band == "High confidence", na.rm=TRUE)))
  cat(sprintf("  Moderate-confidence:         %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band == "Moderate confidence"),
              nrow(risk_df),
              100 * mean(risk_df$risk_band == "Moderate confidence", na.rm=TRUE)))
  cat(sprintf("  Low-confidence (review):     %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band == "Low confidence — review recommended"),
              nrow(risk_df),
              100 * mean(risk_df$risk_band == "Low confidence — review recommended", na.rm=TRUE)))
  cat(sprintf("\n  Median predicted confidence: %.1f%%\n",
              median(as.numeric(gsub("%","",risk_df$confidence)), na.rm=TRUE)*100))

  if (exists("boot_arr_mc") && !is.null(boot_arr_mc)) {
    cat(sprintf("  Bootstrap 95%% CI (median ci_width): %.1f%%\n",
                median(boot_risk_mc$ci_width_pred, na.rm=TRUE)*100))
  }

  cat(sprintf("\nHigh-confidence predictions (≥0.70, low disagreement): %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band=="High confidence"),
              nrow(risk_df),
              100*mean(risk_df$risk_band=="High confidence", na.rm=TRUE)))
}
```
