---
title: "Breast Cancer PAM50 Subtype Risk Calculator (Multiclass)"
format:
  html:
    embed-resources: true
    code-fold: true
    df-print: paged
    toc: true
editor: visual
---

This notebook builds a **PAM50 subtype classifier** from merged GEO microarray datasets. It is structured in three clearly separated stages:

1.  **QC & normalisation** — per-dataset, completely separate from modelling
2.  **Model building** — LODO cross-validation with fold-safe batch correction
3.  **Risk calculator** — subtype probabilities with uncertainty quantification

```{r setup}
#| message: false
#| warning: false
library(GEOquery)
library(limma)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(class)
library(e1071)
library(randomForest)
library(sva)
library(glmnet)
if (requireNamespace("nnet",    quietly = TRUE)) library(nnet)
if (requireNamespace("MASS",    quietly = TRUE)) library(MASS)      # LDA
select <- dplyr::select   # prevent MASS from masking dplyr::select
filter <- dplyr::filter   # prevent MASS from masking dplyr::filte
if (requireNamespace("xgboost", quietly = TRUE)) library(xgboost)   # XGBoost

set.seed(3888)

# ── Dataset registry ──────────────────────────────────────────────────────────
datasets_multiclass <- c("GSE25066", "GSE45827", "GSE65194","GSE21653")

# ── PAM50 subtype dictionary ──────────────────────────────────────────────────
subtype_variants <- list(
  "Luminal A"   = c("luminal a","lum a","luma","lum-a","lum_a","luminala"),
  "Luminal B"   = c("luminal b","lum b","lumb","lum-b","lum_b","luminalb"),
  "Basal"       = c("basal","basal-like","basal like","basallike","basal_like"),
  "HER2"        = c("her2","her2+","her-2","her2-enriched","her2enriched","her2e","erbb2"),
  "Normal-like" = c("normal-like","normal like","normallike","normal_like","molecular subtype: normal")
)
```

------------------------------------------------------------------------

# Stage 1 — QC & Normalisation (per dataset)

> **Design principle:** normalisation is applied within each dataset *before* any merging or modelling. This ensures samples within a study are on a common scale. Cross-dataset harmonisation (ComBat) happens later, *inside* each CV fold.

## Helper functions

```{r helpers-qc}
#| message: false
#| warning: false

# ── GEO loader ────────────────────────────────────────────────────────────────
load_geo_eset <- function(gse_id) {
  geo_cache_dir <- file.path("data", "geo-cache")
  if (!dir.exists(geo_cache_dir)) dir.create(geo_cache_dir, recursive = TRUE)

  gse_matrix_url <- function(id, host = "ftp.ncbi.nlm.nih.gov") {
    num    <- sub("^GSE", "", id)
    prefix <- substr(num, 1, max(1, nchar(num) - 3))
    paste0("https://", host, "/geo/series/GSE", prefix, "nnn/", id,
           "/matrix/", id, "_series_matrix.txt.gz")
  }

  g <- tryCatch(GEOquery::getGEO(gse_id, destdir = geo_cache_dir), error = function(e) e)
  if (!inherits(g, "error")) {
    if (inherits(g, "ExpressionSet")) return(g)
    if (is.list(g) && length(g) >= 1) return(g[[1]])
  }

  destfile <- file.path(geo_cache_dir, paste0(gse_id, "_series_matrix.txt.gz"))
  for (host in c("ftp.ncbi.nlm.nih.gov", "download.ncbi.nlm.nih.gov")) {
    for (attempt in 1:3) {
      try({
        h <- curl::new_handle(); curl::handle_setopt(h, connecttimeout=60, timeout=1800)
        curl::curl_download(gse_matrix_url(gse_id, host), destfile, quiet=FALSE, handle=h)
      }, silent = TRUE)
      if (file.exists(destfile) && file.info(destfile)$size > 0) break
      Sys.sleep(3 * attempt)
    }
  }
  g2 <- GEOquery::getGEO(filename = destfile)
  if (inherits(g2, "ExpressionSet")) return(g2)
  if (is.list(g2) && length(g2) >= 1) return(g2[[1]])
  stop("Could not load ExpressionSet for ", gse_id)
}

# ── Log2 transform (GEO2R heuristic) ─────────────────────────────────────────
needs_log2 <- function(e) {
  v <- as.numeric(e); v <- v[is.finite(v)]
  if (length(v) < 10) return(FALSE)
  qx <- quantile(v, c(0, 0.25, 0.5, 0.75, 0.99, 1), na.rm = TRUE)
  (qx[5] > 100) || (qx[6] - qx[1] > 50 && qx[2] > 0) ||
    (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
}

safe_log2 <- function(e) {
  pos    <- e[e > 0 & is.finite(e)]
  pseudo <- min(pos, na.rm = TRUE) / 2
  e2 <- e; e2[!is.finite(e2)] <- NA_real_; e2[e2 <= 0] <- pseudo
  log2(e2)
}

# ── Within-dataset quantile normalisation ────────────────────────────────────
quantile_normalise <- function(e) limma::normalizeBetweenArrays(e, method = "quantile")

# ── Probe → gene collapse (avereps) ──────────────────────────────────────────
collapse_probes <- function(eset) {
  e   <- exprs(eset)
  fd  <- fData(eset)
  sym_col <- grep("gene\\s*symbol|^symbol$", colnames(fd), ignore.case=TRUE, value=TRUE)[1]
  gene_sym_raw <- if (!is.na(sym_col)) as.character(fd[, sym_col]) else rep(NA_character_, nrow(fd))
  gene_sym <- ifelse(is.na(gene_sym_raw) | trimws(gene_sym_raw) == "",
                     rownames(e),
                     vapply(gene_sym_raw, function(s) {
                       s <- trimws(as.character(s))
                       if (is.na(s) || s == "") return(NA_character_)
                       trimws(strsplit(s, " /// ", fixed=TRUE)[[1]][1])
                     }, character(1)))
  gene_sym[is.na(gene_sym) | gene_sym == ""] <- rownames(e)[is.na(gene_sym) | gene_sym == ""]
  limma::avereps(e, ID = gene_sym)
}

# ── PAM50 label inference ─────────────────────────────────────────────────────
norm_label_txt <- function(x) {
  x <- tolower(as.character(x)); x[is.na(x)] <- ""
  trimws(gsub("[^a-z0-9]+", " ", x))
}

map_to_subtype <- function(x_raw) {
  x_chr  <- as.character(x_raw)
  x_norm <- norm_label_txt(x_chr)
  if (x_norm == "") return(NA_character_)
  if (grepl("pam50_class", tolower(x_chr), fixed=TRUE)) {
    m <- regexpr("pam50_class:\\s*", x_chr, ignore.case=TRUE)
    if (m > 0) x_norm <- norm_label_txt(substring(x_chr, m + attr(m,"match.length")))
  }
  for (canon in names(subtype_variants)) {
    if (any(vapply(subtype_variants[[canon]],
                   function(kw) grepl(kw, x_norm, fixed=TRUE), logical(1))))
      return(canon)
  }
  NA_character_
}

infer_multiclass_label <- function(pMat) {
  cols_try <- c(
    grep("pam50",  colnames(pMat), ignore.case=TRUE, value=TRUE),
    grep("^characteristics_ch", colnames(pMat), value=TRUE),
    setdiff(colnames(pMat), c(
      grep("pam50", colnames(pMat), ignore.case=TRUE, value=TRUE),
      grep("^characteristics_ch", colnames(pMat), value=TRUE)))
  )
  for (cn in cols_try) {
    lab <- vapply(as.character(pMat[[cn]]), map_to_subtype, character(1))
    y   <- droplevels(factor(lab[!is.na(lab)]))
    if (nlevels(y) >= 2) return(list(label = factor(lab), label_col = cn))
  }
  list(label = factor(rep(NA_character_, nrow(pMat))), label_col = NA_character_)
}

# ── Full per-dataset QC pipeline ──────────────────────────────────────────────
normalised_cache_dir <- file.path("data", "normalised")

# ── GSE21653-specific: characteristics_ch1.12 has PAM50 subtype labels ────────
# Values: "molecular subtype: Basal/ERBB2/LuminalA/LuminalB/Normal"
# All 266 samples are tumours — no normal breast tissue in this dataset.
# Normal-like subtype here = PAM50 Normal-like, not actual normal tissue.
# infer_multiclass_label cannot be used for this dataset because intermediate
# characteristics columns contain scattered subtype values that cause early exit.
multiclass_label_overrides <- list(
  GSE21653 = function(p) {
    raw <- tolower(trimws(as.character(p[["characteristics_ch1.12"]])))
    ifelse(grepl("basal",    raw), "Basal",
    ifelse(grepl("erbb2",    raw), "HER2",
    ifelse(grepl("luminala", raw), "Luminal A",
    ifelse(grepl("luminalb", raw), "Luminal B",
    ifelse(grepl("normal",   raw), "Normal-like", NA_character_)))))
  }
)

process_dataset_qc <- function(gse_id) {

  rds_path <- file.path(normalised_cache_dir, paste0(gse_id, "_normalised.rds"))

  # ── Helper: derive labels, respecting dataset-specific overrides ─────────
  derive_labels <- function(p) {
    if (gse_id %in% names(multiclass_label_overrides)) {
      message(gse_id, ": using direct label override (characteristics_ch1.12)")
      list(label    = factor(multiclass_label_overrides[[gse_id]](p)),
           label_col = "characteristics_ch1.12")
    } else {
      infer_multiclass_label(p)
    }
  }

  # ── Fast path: load pre-saved normalised RDS ─────────────────────────────
  if (file.exists(rds_path)) {
    message(gse_id, ": loading pre-normalised RDS from ", rds_path)
    dat    <- readRDS(rds_path)
    e_gene <- dat$expr
    p      <- dat$meta
    p$dataset_id <- gse_id
    lab    <- derive_labels(p)
    p$label_mc <- as.character(lab$label)
    return(list(
      id        = gse_id,
      e_raw     = e_gene,
      e_norm    = e_gene,
      pMat      = p,
      label_col = lab$label_col
    ))
  }

  # ── Slow path: download from GEO, normalise, save ────────────────────────
  message(gse_id, ": no cached RDS found — downloading from GEO")
  eset   <- load_geo_eset(gse_id)
  e      <- exprs(eset)
  if (needs_log2(e)) e <- safe_log2(e)
  e_norm <- quantile_normalise(e)
  exprs(eset) <- e_norm
  e_gene <- collapse_probes(eset)
  p      <- pData(eset)
  p$dataset_id <- gse_id
  lab    <- derive_labels(p)
  p$label_mc <- as.character(lab$label)

  if (!dir.exists(normalised_cache_dir))
    dir.create(normalised_cache_dir, recursive = TRUE)
  saveRDS(list(expr = e_gene, meta = p), rds_path)
  message(gse_id, ": saved normalised RDS to ", rds_path)

  list(
    id        = gse_id,
    e_raw     = e,
    e_norm    = e_gene,
    pMat      = p,
    label_col = lab$label_col
  )
}
```

## Load and normalise each dataset

```{r load-datasets}
#| message: false
#| warning: false
# ── Load (from RDS cache or GEO) with per-dataset status reporting ────────────
load_status <- data.frame(
  dataset      = character(),
  source       = character(),
  n_genes      = integer(),
  n_samples    = integer(),
  n_labeled    = integer(),
  subtypes     = character(),
  stringsAsFactors = FALSE
)

qc_pieces <- list()
for (id in datasets_multiclass) {
  rds_path <- file.path(normalised_cache_dir, paste0(id, "_normalised.rds"))
  source_used <- if (file.exists(rds_path)) "RDS cache" else "GEO download"

  piece <- tryCatch(
    process_dataset_qc(id),
    error = function(e) {
      message("FAILED: ", id, " — ", conditionMessage(e))
      NULL
    }
  )

  if (!is.null(piece)) {
    qc_pieces[[id]] <- piece
    ok <- !is.na(piece$pMat$label_mc) & piece$pMat$label_mc != ""
    load_status <- rbind(load_status, data.frame(
      dataset      = id,
      source       = source_used,
      n_genes      = nrow(piece$e_norm),
      n_samples    = ncol(piece$e_norm),
      n_labeled    = sum(ok),
      subtypes     = paste(sort(unique(piece$pMat$label_mc[ok])), collapse = ", "),
      stringsAsFactors = FALSE
    ))
  } else {
    load_status <- rbind(load_status, data.frame(
      dataset="FAILED", source=source_used, n_genes=NA, n_samples=NA,
      n_labeled=NA, subtypes="FAILED", stringsAsFactors=FALSE
    ))
  }
}

knitr::kable(load_status,
  caption = "Dataset loading summary — check 'source' to confirm RDS vs GEO, and 'n_labeled' vs 'n_samples' for label coverage")

# Abort early with a clear message if any dataset failed
failed <- load_status$source == "FAILED" | is.na(load_status$n_samples)
if (any(failed)) {
  stop("One or more datasets failed to load: ",
       paste(load_status$dataset[failed], collapse=", "),
       "\nCheck GEO connectivity or verify the RDS files exist in: ",
       normalised_cache_dir)
}

cat(sprintf("\n✓ All %d datasets loaded successfully.\n", length(qc_pieces)))

# ── Optionally save the merged normalised dataset for even faster reloads ──
# Uncomment to write once; comment out afterwards.
# merged_rds_path <- file.path(normalised_cache_dir, "merged_normalised_mc.rds")
# all_genes_tmp <- Reduce(intersect, lapply(qc_pieces, function(p) rownames(p$e_norm)))
# e_tmp <- do.call(cbind, lapply(qc_pieces, function(p) p$e_norm[all_genes_tmp,,drop=FALSE]))
# p_tmp <- do.call(rbind, lapply(qc_pieces, function(p) {
#   data.frame(sample_id=rownames(p$pMat), dataset_id=p$id,
#              label_mc=p$pMat$label_mc, stringsAsFactors=FALSE)
# }))
# saveRDS(list(expr=e_tmp, meta=p_tmp, all_genes=all_genes_tmp), merged_rds_path)
# cat("Saved merged RDS to", merged_rds_path, "\n")
```

## QC: Per-sample expression distributions

```{r qc-boxplots, fig.cap="Per-sample expression distributions after within-dataset quantile normalisation. Boxes should be tightly aligned within each dataset."}
#| message: false
#| warning: false
set.seed(3888)
qc_box_list <- lapply(names(qc_pieces), function(id) {
  p       <- qc_pieces[[id]]
  n_genes <- min(1500L, nrow(p$e_norm))
  gsamp   <- sample(rownames(p$e_norm), n_genes)
  e_sub   <- p$e_norm[gsamp, , drop = FALSE]
  lab_vec <- p$pMat$label_mc

  data.frame(
    expr    = as.vector(e_sub),
    sample  = rep(seq_len(ncol(e_sub)), each = nrow(e_sub)),
    dataset = id,
    subtype = rep(lab_vec, each = nrow(e_sub)),
    stringsAsFactors = FALSE
  )
})

qc_box_df <- bind_rows(qc_box_list)

# Sample median for ordering
sample_med <- qc_box_df %>%
  group_by(dataset, sample) %>%
  summarise(med = median(expr, na.rm=TRUE), .groups="drop") %>%
  arrange(dataset, med) %>%
  group_by(dataset) %>%
  mutate(sample_ord = row_number()) %>%
  ungroup()

qc_box_df <- qc_box_df %>%
  left_join(sample_med %>% select(dataset, sample, sample_ord),
            by = c("dataset","sample"))

ggplot(qc_box_df, aes(x = factor(sample_ord), y = expr,
                      colour = subtype)) +
  geom_boxplot(outlier.shape = NA, size = 0.2) +
  facet_wrap(~dataset, scales = "free_x") +
  theme_bw(base_size = 10) +
  labs(x = "Sample (ordered by median)", y = "log2 expression",
       colour = "Subtype",
       title = "Per-sample distributions after within-dataset quantile normalisation") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# Summary table
qc_box_df %>%
  group_by(dataset) %>%
  summarise(n_samples  = n_distinct(sample),
            median_expr = round(median(expr, na.rm=TRUE), 3),
            IQR_expr    = round(IQR(expr,   na.rm=TRUE), 3),
            q01         = round(quantile(expr, .01, na.rm=TRUE), 3),
            q99         = round(quantile(expr, .99, na.rm=TRUE), 3),
            .groups = "drop") %>%
  knitr::kable(caption = "Per-dataset expression summary (post normalisation)")
```

## QC: PCA per dataset

```{r qc-pca, fig.cap="PCA of each dataset after normalisation. Points coloured by PAM50 subtype. Subtypes should separate along PCs."}
#| message: false
#| warning: false
for (id in names(qc_pieces)) {
  p   <- qc_pieces[[id]]
  ok  <- !is.na(p$pMat$label_mc) & p$pMat$label_mc != ""
  e_s <- p$e_norm[, ok, drop=FALSE]
  y_s <- p$pMat$label_mc[ok]

  e_v <- e_s[apply(e_s, 1, var, na.rm=TRUE) > 0, ]
  if (nrow(e_v) < 5 || ncol(e_v) < 3) { cat(id, ": skipping PCA\n"); next }

  # impute NAs with row means before PCA
  e_v[is.na(e_v)] <- rowMeans(e_v, na.rm=TRUE)[row(e_v)[is.na(e_v)]]

  pca <- prcomp(t(e_v), scale.=TRUE)
  ve  <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 1)

  df <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2], subtype = y_s)
  print(
    ggplot(df, aes(PC1, PC2, colour = subtype)) +
      geom_point(size = 2, alpha = .8) +
      theme_bw() +
      labs(x = paste0("PC1 (", ve[1], "%)"),
           y = paste0("PC2 (", ve[2], "%)"),
           colour = "Subtype",
           title  = paste0(id, " — PCA after normalisation"))
  )
}
```

## QC: Merged PCA — before and after ComBat

```{r qc-pca-merged}
#| message: false
#| warning: false
# Merge all datasets for PCA visualisation only.
# Shows that batch effects exist before correction (expected — different
# labs/platforms) and are removed after ComBat.
all_genes_vis <- Reduce(intersect, lapply(qc_pieces, function(p) rownames(p$e_norm)))
e_merged_vis  <- do.call(cbind, lapply(qc_pieces, function(p)
  p$e_norm[all_genes_vis, , drop = FALSE]))

p_merged_vis <- do.call(rbind, lapply(qc_pieces, function(p) {
  ok <- !is.na(p$pMat$label_mc) & p$pMat$label_mc != ""
  data.frame(dataset = p$id,
             subtype = p$pMat$label_mc,
             stringsAsFactors = FALSE)
}))

ok_vis   <- !is.na(p_merged_vis$subtype) & p_merged_vis$subtype != ""
e_vis    <- e_merged_vis[, ok_vis, drop = FALSE]
p_vis    <- p_merged_vis[ok_vis, ]

# Remove near-zero variance genes before PCA
gv       <- apply(e_vis, 1, var, na.rm = TRUE)
top_g    <- names(sort(gv, decreasing = TRUE))[1:min(2000, sum(gv > 0))]
e_vis_sub <- e_vis[top_g, , drop = FALSE]
e_vis_sub[is.na(e_vis_sub)] <- rowMeans(e_vis_sub, na.rm = TRUE)[
  row(e_vis_sub)[is.na(e_vis_sub)]]

make_merged_pca <- function(e_mat, p_df, label) {
  pca <- prcomp(t(e_mat), scale. = TRUE)
  ve  <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 1)
  df  <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2],
                    dataset = p_df$dataset, subtype = p_df$subtype)
  p1 <- ggplot(df, aes(PC1, PC2, colour = dataset)) +
    geom_point(size = 1.5, alpha = 0.7) + theme_bw() +
    labs(x = paste0("PC1 (", ve[1], "%)"), y = paste0("PC2 (", ve[2], "%)"),
         colour = "Dataset", title = paste0(label, " — by dataset"))
  p2 <- ggplot(df, aes(PC1, PC2, colour = subtype)) +
    geom_point(size = 1.5, alpha = 0.7) + theme_bw() +
    labs(x = paste0("PC1 (", ve[1], "%)"), y = paste0("PC2 (", ve[2], "%)"),
         colour = "Subtype", title = paste0(label, " — by subtype"))
  list(p1, p2)
}

# Before ComBat
pl_before <- make_merged_pca(e_vis_sub, p_vis, "Merged — before ComBat")
print(pl_before[[1]]); print(pl_before[[2]])

# After ComBat (visualisation only)
e_vis_bc <- tryCatch(
  sva::ComBat(dat = e_vis_sub, batch = p_vis$dataset,
              mod = model.matrix(~ factor(p_vis$subtype)),
              par.prior = TRUE, mean.only = FALSE),
  error = function(e) {
    message("ComBat for merged PCA failed: ", conditionMessage(e))
    limma::removeBatchEffect(e_vis_sub, batch = p_vis$dataset,
                             design = model.matrix(~ factor(p_vis$subtype)))
  }
)
pl_after <- make_merged_pca(e_vis_bc, p_vis, "Merged — after ComBat (viz only)")
print(pl_after[[1]]); print(pl_after[[2]])
```

------------------------------------------------------------------------

# Stage 2 — Model Building

> **Design principle:** normalised datasets are merged by gene intersection. Batch correction (ComBat) is applied **inside** each LODO fold using training samples only, then transferred to the test fold.

## Merge normalised datasets

```{r merge-datasets}
#| message: false
#| warning: false

# Keep only genes present in all datasets
all_genes_mc <- Reduce(intersect, lapply(qc_pieces, function(p) rownames(p$e_norm)))
cat("Shared genes across all multiclass datasets:", length(all_genes_mc), "\n")

# Filter genes with >30% missing across all samples
e_merged_mc <- do.call(cbind, lapply(qc_pieces, function(p)
  p$e_norm[all_genes_mc, , drop=FALSE]))

na_frac <- rowMeans(is.na(e_merged_mc))
e_merged_mc <- e_merged_mc[na_frac < 0.3, , drop=FALSE]
cat("Genes after NA filter:", nrow(e_merged_mc), "\n")

p_merged_mc <- do.call(rbind, lapply(qc_pieces, function(p) {
  data.frame(sample_id  = rownames(p$pMat),
             dataset_id = p$id,
             label_mc   = p$pMat$label_mc,
             stringsAsFactors = FALSE)
}))

ok_mc <- !is.na(p_merged_mc$label_mc) & p_merged_mc$label_mc != ""
y_mc   <- factor(p_merged_mc$label_mc[ok_mc])
e_mc   <- e_merged_mc[, ok_mc, drop=FALSE]
batch_mc <- p_merged_mc$dataset_id[ok_mc]

cat("\nClass counts (merged):\n"); print(table(y_mc))
cat("\nSamples per dataset:\n");  print(table(batch_mc))
```

## Modelling helper functions

```{r helpers-model}
#| message: false
#| warning: false

# ── Macro F1 ──────────────────────────────────────────────────────────────────
macro_f1 <- function(truth, pred) {
  truth <- factor(truth); pred <- factor(pred, levels=levels(truth))
  lvls  <- levels(truth)
  f1s <- vapply(lvls, function(k) {
    tp <- sum(pred==k & truth==k, na.rm=TRUE)
    fp <- sum(pred==k & truth!=k, na.rm=TRUE)
    fn <- sum(pred!=k & truth==k, na.rm=TRUE)
    p  <- if ((tp+fp)==0) NA_real_ else tp/(tp+fp)
    r  <- if ((tp+fn)==0) NA_real_ else tp/(tp+fn)
    if (is.na(p)||is.na(r)||(p+r)==0) NA_real_ else 2*p*r/(p+r)
  }, numeric(1))
  mean(f1s, na.rm=TRUE)
}

# ── Macro OvR AUC ────────────────────────────────────────────────────────────
auc_roc_vec <- function(truth01, score) {
  truth01 <- as.integer(truth01); score <- as.numeric(score)
  n_pos <- sum(truth01==1); n_neg <- sum(truth01==0)
  if (n_pos==0 || n_neg==0) return(NA_real_)
  r <- rank(score, ties.method="average")
  (sum(r[truth01==1]) - n_pos*(n_pos+1)/2) / (n_pos*n_neg)
}

macro_auc_ovr <- function(truth, prob_mat) {
  lvls <- colnames(prob_mat)
  aucs <- vapply(lvls, function(k) {
    tk <- as.integer(as.character(truth) == k)
    tk[is.na(tk)] <- 0L
    n_pos <- sum(tk, na.rm=TRUE)
    if (n_pos==0L || n_pos==length(tk)) return(NA_real_)
    auc_roc_vec(tk, prob_mat[, k])
  }, numeric(1))
  mean(aucs, na.rm=TRUE)
}

# ── ComBat: fit on train, apply to test ───────────────────────────────────────
combat_fit_apply <- function(expr_train, batch_train, mod_train,
                              expr_test, batch_test) {
  if (length(unique(as.character(batch_train))) < 2) {
    message("Single batch in fold — skipping ComBat.")
    return(list(train=expr_train, test=expr_test))
  }
  corrected_train <- tryCatch(
    sva::ComBat(dat=expr_train, batch=batch_train, mod=mod_train,
                par.prior=TRUE, mean.only=FALSE),
    error = function(e) {
      message("ComBat failed, falling back to removeBatchEffect: ", conditionMessage(e))
      limma::removeBatchEffect(expr_train, batch=batch_train, design=mod_train)
    }
  )
  # Apply training batch means to test
  batch_train_f <- factor(as.character(batch_train))
  batch_test_f  <- factor(as.character(batch_test), levels=levels(batch_train_f))
  grand_mean    <- rowMeans(expr_train)
  batch_means   <- sapply(levels(batch_train_f), function(b) {
    idx <- which(batch_train_f==b)
    if (length(idx)==0) return(rep(0, nrow(expr_train)))
    rowMeans(expr_train[, idx, drop=FALSE])
  })
  corrected_test <- expr_test
  for (b in levels(batch_train_f)) {
    idx <- which(as.character(batch_test_f)==b)
    if (length(idx)==0) next
    corrected_test[, idx] <- expr_test[, idx, drop=FALSE] -
      (batch_means[, b] - grand_mean)
  }
  # Novel test batches: mean-centre only
  novel <- which(!(as.character(batch_test) %in% levels(batch_train_f)))
  if (length(novel) > 0) {
    smeans <- colMeans(expr_test[, novel, drop=FALSE])
    corrected_test[, novel] <- sweep(expr_test[, novel, drop=FALSE], 2, smeans, "-") + grand_mean
  }
  list(train=corrected_train, test=corrected_test)
}

# ── Imputation & standardisation (fit on train, apply to test) ───────────────
impute_fit    <- function(X) list(med=apply(X,2,function(v) {m<-median(v,na.rm=TRUE); if(is.na(m)) 0 else m}))
impute_apply  <- function(X, fit) { X2<-X; for(j in seq_len(ncol(X2))) { idx<-is.na(X2[,j]); if(any(idx)) X2[idx,j]<-fit$med[j] }; X2 }
std_fit       <- function(X) { mu<-colMeans(X); sd<-apply(X,2,sd); sd[sd==0]<-1; list(mu=mu,sd=sd) }
std_apply     <- function(X, fit) sweep(sweep(X,2,fit$mu,"-"),2,fit$sd,"/")

# ── OvR limma gene selection ──────────────────────────────────────────────────
select_genes_ovr <- function(e_mat, y, n_per_class=50) {
  ok <- !is.na(y) & y!=""; e<-e_mat[, ok, drop=FALSE]; y<-factor(y[ok])
  lvls_s <- make.names(levels(y))
  design  <- model.matrix(~0+y); colnames(design)<-lvls_s
  fit     <- lmFit(e, design)
  genes   <- character(0)
  for (i in seq_along(levels(y))) {
    lvl <- lvls_s[i]; others <- lvls_s[-i]
    cont_str <- paste0(lvl," - (",paste(others,collapse="+"),")/",length(others))
    cont <- makeContrasts(contrasts=cont_str, levels=design)
    fit2 <- eBayes(contrasts.fit(fit, cont))
    genes <- union(genes, rownames(topTable(fit2, n=n_per_class, sort.by="P")))
  }
  genes
}

# ── Stratified split helper ───────────────────────────────────────────────────
stratified_split <- function(y, dataset, train_prop=0.75) {
  dev_idx <- c(); test_idx <- c()
  for (ds in unique(as.character(dataset))) {
    for (cl in unique(as.character(y))) {
      idx <- which(as.character(dataset)==ds & as.character(y)==cl)
      if (length(idx)==0) next
      idx <- sample(idx)
      n_dev <- max(1, min(floor(length(idx)*train_prop), length(idx)-1))
      dev_idx  <- c(dev_idx,  idx[seq_len(n_dev)])
      test_idx <- c(test_idx, idx[(n_dev+1):length(idx)])
    }
  }
  list(dev=sort(unique(dev_idx)), test=sort(unique(test_idx)))
}
```

## LODO cross-validation

```{r lodo-cv}
#| message: false
#| warning: false

run_lodo_multiclass <- function(X, y, batch, n_per_class=50, use_weights=TRUE) {
  y     <- factor(y)
  batch <- as.character(batch)
  perf_list  <- list()
  probs_list <- list()

  for (ds in unique(batch)) {
    tryCatch({   # ← wrap entire fold so one failure never kills res_lodo

    test_id  <- which(batch==ds)
    train_id <- setdiff(seq_along(batch), test_id)

    y_train <- droplevels(y[train_id])
    y_test  <- factor(as.character(y[test_id]), levels=levels(y_train))
    if (nlevels(y_train)<2) return(NULL)

    # Batch correction (train only → apply to test)
    cb <- combat_fit_apply(
      expr_train=t(X[train_id,,drop=FALSE]), batch_train=batch[train_id],
      mod_train=model.matrix(~y_train),
      expr_test=t(X[test_id,,drop=FALSE]),  batch_test=batch[test_id]
    )
    X_tr_full <- t(cb$train); X_te_full <- t(cb$test)

    # Gene selection from training data only
    genes_sel <- select_genes_ovr(t(X_tr_full), y=y_train, n_per_class=n_per_class)
    sel_g     <- intersect(genes_sel, colnames(X_tr_full))
    if (length(sel_g)<10) return(NULL)

    X_tr <- X_tr_full[, sel_g]; X_te <- X_te_full[, sel_g]
    imp  <- impute_fit(X_tr); X_tr_i <- impute_apply(X_tr,imp); X_te_i <- impute_apply(X_te,imp)
    sc   <- std_fit(X_tr_i);  X_tr_s <- std_apply(X_tr_i,sc);  X_te_s <- std_apply(X_te_i,sc)

    lvls <- levels(y_train); n_cl <- length(lvls); n_te <- nrow(X_te_s)
    tab_tr <- table(y_train); n_tot <- sum(tab_tr)
    cw <- if (use_weights)
      setNames(vapply(lvls, function(k) n_tot/(n_cl*max(1,tab_tr[k])), numeric(1)), lvls)
    else setNames(rep(1,n_cl), lvls)
    sw <- cw[as.character(y_train)]

    y_te_chr <- as.character(y_test)

    record <- function(nm, pred_chr, prob_mat) {
      perf_list[[length(perf_list)+1]] <<- tibble::tibble(
        fold=ds, model=nm,
        accuracy  = mean(pred_chr==y_te_chr, na.rm=TRUE),
        macro_f1  = macro_f1(y_te_chr, pred_chr),
        macro_auc = macro_auc_ovr(y_te_chr, prob_mat)
      )
      pd <- as.data.frame(prob_mat); pd$fold<-ds; pd$model<-nm; pd$truth<-y_te_chr
      probs_list[[length(probs_list)+1]] <<- pd
    }

    # ── Helper: glmnet predict → clean prob matrix ───────────────────────────
    glmnet_probs <- function(fit, X_te, lvls) {
      pr <- predict(fit, as.matrix(X_te), type="response", s="lambda.min")[,,1]
      if (is.null(dim(pr))) pr <- t(as.matrix(pr))
      pr[, intersect(lvls, colnames(pr)), drop=FALSE]
    }

    # ── Helper: build kNN prob matrix from win prob ───────────────────────────
    knn_prob_mat <- function(X_tr, X_te, y_tr, k=7) {
      pk  <- class::knn(X_tr, X_te, y_tr, k=k, prob=TRUE)
      kw  <- attr(pk, "prob"); pkc <- as.character(pk)
      lvls_k <- levels(y_tr); n_te_k <- nrow(X_te); n_cl_k <- length(lvls_k)
      pm <- matrix(0, n_te_k, n_cl_k, dimnames=list(NULL, lvls_k))
      for (i in seq_len(n_te_k)) {
        pm[i, pkc[i]] <- kw[i]
        oth <- setdiff(lvls_k, pkc[i])
        if (length(oth)) pm[i, oth] <- (1 - kw[i]) / length(oth)
      }
      pm
    }

    # ════════════════════════════════════════════════════════════════════════
    # UNWEIGHTED MODELS
    # ════════════════════════════════════════════════════════════════════════
    sw0 <- rep(1, nrow(X_tr_s))
    cw0 <- setNames(rep(1, n_cl), lvls)

    # kNN (unweighted — no native weight support)
    tryCatch({
      pm <- knn_prob_mat(X_tr_s, X_te_s, y_train)
      record("kNN", colnames(pm)[max.col(pm)], pm)
    }, error=function(e) message("kNN fold ", ds, ": ", conditionMessage(e)))

    # SVM RBF (unweighted)
    tryCatch({
      fit <- e1071::svm(X_tr_s, y_train, kernel="radial", cost=1,
                        gamma=1/ncol(X_tr_s), probability=TRUE)
      ps  <- predict(fit, X_te_s, probability=TRUE)
      pm  <- attr(ps,"probabilities")[, lvls, drop=FALSE]
      record("SVM_RBF", as.character(ps), pm)
    }, error=function(e) message("SVM_RBF fold ", ds, ": ", conditionMessage(e)))

    # SVM Linear (unweighted)
    tryCatch({
      fit <- e1071::svm(X_tr_s, y_train, kernel="linear", cost=1,
                        probability=TRUE)
      ps  <- predict(fit, X_te_s, probability=TRUE)
      pm  <- attr(ps,"probabilities")[, lvls, drop=FALSE]
      record("SVM_Linear", as.character(ps), pm)
    }, error=function(e) message("SVM_Linear fold ", ds, ": ", conditionMessage(e)))

    # Multinomial Lasso (unweighted, alpha=1)
    tryCatch({
      fit <- glmnet::cv.glmnet(as.matrix(X_tr_s), y_train, family="multinomial",
                               alpha=1, weights=sw0, standardize=FALSE,
                               type.multinomial="grouped")
      pm <- glmnet_probs(fit, X_te_s, lvls)
      record("LassoMC", lvls[max.col(pm)], pm)
    }, error=function(e) message("LassoMC fold ", ds, ": ", conditionMessage(e)))

    # Multinomial Ridge (unweighted, alpha=0)
    tryCatch({
      fit <- glmnet::cv.glmnet(as.matrix(X_tr_s), y_train, family="multinomial",
                               alpha=0, weights=sw0, standardize=FALSE,
                               type.multinomial="grouped")
      pm <- glmnet_probs(fit, X_te_s, lvls)
      record("RidgeMC", lvls[max.col(pm)], pm)
    }, error=function(e) message("RidgeMC fold ", ds, ": ", conditionMessage(e)))

    # Multinomial logistic / nnet (unweighted)
    tryCatch({
      fit <- nnet::multinom(y_train~., data=data.frame(y_train, X_tr_s),
                            weights=sw0, trace=FALSE, MaxNWts=100000)
      pm  <- predict(fit, newdata=data.frame(X_te_s), type="probs")
      if (is.null(dim(pm))) pm <- t(as.matrix(pm))
      pm  <- pm[, intersect(lvls, colnames(pm)), drop=FALSE]
      record("Multinom", lvls[max.col(pm)], pm)
    }, error=function(e) message("Multinom fold ", ds, ": ", conditionMessage(e)))

    # Naive Bayes (unweighted — no native weight support)
    tryCatch({
      fit <- e1071::naiveBayes(x=X_tr_s, y=y_train)
      pm  <- predict(fit, X_te_s, type="raw")
      pm  <- pm[, intersect(lvls, colnames(pm)), drop=FALSE]
      record("NaiveBayes", colnames(pm)[max.col(pm)], pm)
    }, error=function(e) message("NaiveBayes fold ", ds, ": ", conditionMessage(e)))

    # LDA (unweighted)
    tryCatch({
      fit     <- MASS::lda(x=X_tr_s, grouping=y_train)
      pred_lda <- predict(fit, X_te_s)
      pm      <- pred_lda$posterior[, intersect(lvls, colnames(pred_lda$posterior)), drop=FALSE]
      record("LDA", as.character(pred_lda$class), pm)
    }, error=function(e) message("LDA fold ", ds, ": ", conditionMessage(e)))

    # XGBoost (unweighted)
    tryCatch({
      y_int    <- as.integer(y_train) - 1L
      dtrain   <- xgboost::xgb.DMatrix(data=as.matrix(X_tr_s), label=y_int)
      dtest    <- xgboost::xgb.DMatrix(data=as.matrix(X_te_s))
      fit_xgb  <- xgboost::xgb.train(
        params  = list(objective="multi:softprob", eval_metric="mlogloss",
                       num_class=n_cl, eta=0.05, max_depth=4,
                       subsample=0.8, colsample_bytree=0.8,
                       min_child_weight=5, nthread=1),
        data=dtrain, nrounds=150, verbose=0
      )
      prob_raw <- matrix(predict(fit_xgb, dtest), ncol=n_cl, byrow=TRUE)
      colnames(prob_raw) <- lvls
      record("XGBoost", lvls[max.col(prob_raw)], prob_raw)
    }, error=function(e) message("XGBoost fold ", ds, ": ", conditionMessage(e)))

    # ════════════════════════════════════════════════════════════════════════
    # WEIGHTED MODELS (balanced class weights)
    # ════════════════════════════════════════════════════════════════════════

    # kNN_w (same as kNN — no native weight support, noted)
    tryCatch({
      pm <- knn_prob_mat(X_tr_s, X_te_s, y_train)
      record("kNN_w", colnames(pm)[max.col(pm)], pm)
    }, error=function(e) message("kNN_w fold ", ds, ": ", conditionMessage(e)))

    # SVM RBF (weighted)
    tryCatch({
      fit <- e1071::svm(X_tr_s, y_train, kernel="radial", cost=1,
                        gamma=1/ncol(X_tr_s), probability=TRUE,
                        class.weights=cw)
      ps  <- predict(fit, X_te_s, probability=TRUE)
      pm  <- attr(ps,"probabilities")[, lvls, drop=FALSE]
      record("SVM_RBF_w", as.character(ps), pm)
    }, error=function(e) message("SVM_RBF_w fold ", ds, ": ", conditionMessage(e)))

    # SVM Linear (weighted)
    tryCatch({
      fit <- e1071::svm(X_tr_s, y_train, kernel="linear", cost=1,
                        probability=TRUE, class.weights=cw)
      ps  <- predict(fit, X_te_s, probability=TRUE)
      pm  <- attr(ps,"probabilities")[, lvls, drop=FALSE]
      record("SVM_Linear_w", as.character(ps), pm)
    }, error=function(e) message("SVM_Linear_w fold ", ds, ": ", conditionMessage(e)))

    # Multinomial Lasso (weighted)
    tryCatch({
      fit <- glmnet::cv.glmnet(as.matrix(X_tr_s), y_train, family="multinomial",
                               alpha=1, weights=sw, standardize=FALSE,
                               type.multinomial="grouped")
      pm <- glmnet_probs(fit, X_te_s, lvls)
      record("LassoMC_w", lvls[max.col(pm)], pm)
    }, error=function(e) message("LassoMC_w fold ", ds, ": ", conditionMessage(e)))

    # Multinomial Ridge (weighted)
    tryCatch({
      fit <- glmnet::cv.glmnet(as.matrix(X_tr_s), y_train, family="multinomial",
                               alpha=0, weights=sw, standardize=FALSE,
                               type.multinomial="grouped")
      pm <- glmnet_probs(fit, X_te_s, lvls)
      record("RidgeMC_w", lvls[max.col(pm)], pm)
    }, error=function(e) message("RidgeMC_w fold ", ds, ": ", conditionMessage(e)))

    # Multinomial logistic / nnet (weighted)
    tryCatch({
      fit <- nnet::multinom(y_train~., data=data.frame(y_train, X_tr_s),
                            weights=sw, trace=FALSE, MaxNWts=100000)
      pm  <- predict(fit, newdata=data.frame(X_te_s), type="probs")
      if (is.null(dim(pm))) pm <- t(as.matrix(pm))
      pm  <- pm[, intersect(lvls, colnames(pm)), drop=FALSE]
      record("Multinom_w", lvls[max.col(pm)], pm)
    }, error=function(e) message("Multinom_w fold ", ds, ": ", conditionMessage(e)))

    # Naive Bayes_w (same as NaiveBayes — no native weight support, noted)
    tryCatch({
      fit <- e1071::naiveBayes(x=X_tr_s, y=y_train)
      pm  <- predict(fit, X_te_s, type="raw")
      pm  <- pm[, intersect(lvls, colnames(pm)), drop=FALSE]
      record("NaiveBayes_w", colnames(pm)[max.col(pm)], pm)
    }, error=function(e) message("NaiveBayes_w fold ", ds, ": ", conditionMessage(e)))

    # LDA with weighted prior (weighted proportional to inverse class frequency)
    tryCatch({
      prior_w <- cw / sum(cw)
      fit     <- MASS::lda(x=X_tr_s, grouping=y_train, prior=prior_w)
      pred_lda <- predict(fit, X_te_s)
      pm      <- pred_lda$posterior[, intersect(lvls, colnames(pred_lda$posterior)), drop=FALSE]
      record("LDA_w", as.character(pred_lda$class), pm)
    }, error=function(e) message("LDA_w fold ", ds, ": ", conditionMessage(e)))

    # XGBoost (weighted via sample weights)
    tryCatch({
      y_int   <- as.integer(y_train) - 1L
      dtrain  <- xgboost::xgb.DMatrix(data=as.matrix(X_tr_s), label=y_int,
                                       weight=sw)
      dtest   <- xgboost::xgb.DMatrix(data=as.matrix(X_te_s))
      fit_xgb <- xgboost::xgb.train(
        params  = list(objective="multi:softprob", eval_metric="mlogloss",
                       num_class=n_cl, eta=0.05, max_depth=4,
                       subsample=0.8, colsample_bytree=0.8,
                       min_child_weight=5, nthread=1),
        data=dtrain, nrounds=150, verbose=0
      )
      prob_raw <- matrix(predict(fit_xgb, dtest), ncol=n_cl, byrow=TRUE)
      colnames(prob_raw) <- lvls
      record("XGBoost_w", lvls[max.col(prob_raw)], prob_raw)
    }, error=function(e) message("XGBoost_w fold ", ds, ": ", conditionMessage(e)))

    }, error = function(e) {
      message("LODO fold ", ds, " failed entirely: ", conditionMessage(e))
    })  # end tryCatch for fold
  }

  if (length(perf_list) == 0)
    warning("run_lodo_multiclass: no folds completed successfully — check data and batch labels")

  list(perf=bind_rows(perf_list), probs=bind_rows(probs_list))
}

X_mc <- t(e_mc)
cat("Running LODO CV (", length(unique(batch_mc)), "folds)...\n")
res_lodo <- run_lodo_multiclass(X_mc, y_mc, batch_mc, n_per_class=50, use_weights=TRUE)

if (nrow(res_lodo$perf) == 0) {
  stop("res_lodo is empty — all LODO folds failed. ",
       "Check that all 3 datasets loaded correctly and batch_mc has 3 unique values: ",
       paste(unique(batch_mc), collapse=", "))
}
cat("Models evaluated:", paste(unique(res_lodo$perf$model), collapse=", "), "\n")
cat("Folds completed:", paste(unique(res_lodo$perf$fold), collapse=", "), "\n")
```

------------------------------------------------------------------------

# Stage 3 — Results

## Model scorecard

```{r scorecard}
#| message: false
#| warning: false
scorecard <- res_lodo$perf %>%
  group_by(model) %>%
  summarise(
    macro_auc_mean = mean(macro_auc, na.rm=TRUE),
    macro_auc_sd   = sd(macro_auc,   na.rm=TRUE),
    macro_f1_mean  = mean(macro_f1,  na.rm=TRUE),
    macro_f1_sd    = sd(macro_f1,    na.rm=TRUE),
    accuracy_mean  = mean(accuracy,  na.rm=TRUE),
    accuracy_sd    = sd(accuracy,    na.rm=TRUE),
    .groups="drop"
  ) %>%
  arrange(desc(macro_auc_mean), desc(macro_f1_mean)) %>%
  mutate(
    rank      = row_number(),
    macro_auc = sprintf("%.3f ± %.3f", macro_auc_mean, macro_auc_sd),
    macro_f1  = sprintf("%.3f ± %.3f", macro_f1_mean,  macro_f1_sd),
    accuracy  = sprintf("%.3f ± %.3f", accuracy_mean,  accuracy_sd)
  ) %>%
  select(rank, model, macro_auc, macro_f1, accuracy)

knitr::kable(scorecard,
  caption="Model scorecard — LODO CV (ranked by macro OvR AUC → macro F1 → accuracy)")

best_model_mc <- res_lodo$perf %>%
  group_by(model) %>%
  summarise(m=mean(macro_auc, na.rm=TRUE), .groups="drop") %>%
  arrange(desc(m)) %>% slice(1) %>% pull(model)
cat("\nBest single model:", best_model_mc, "\n")
```

## Performance plots

```{r perf-plots, fig.cap="LODO CV performance across models. Higher is better for all three metrics."}
#| message: false
#| warning: false
res_lodo$perf %>%
  pivot_longer(cols=c(macro_auc, macro_f1, accuracy),
               names_to="metric", values_to="value") %>%
  mutate(metric = recode(metric,
    macro_auc = "Macro OvR AUC",
    macro_f1  = "Macro F1",
    accuracy  = "Accuracy")) %>%
  ggplot(aes(x=reorder(model, value, median), y=value, fill=model)) +
  geom_boxplot(alpha=0.75, outlier.size=0.8) +
  facet_wrap(~metric, scales="free_y") +
  coord_flip() +
  theme_bw() +
  labs(x=NULL, y=NULL, title="LODO CV: model comparison") +
  theme(legend.position="none")
```

## Confusion matrix (best model)

```{r confusion-best}
#| message: false
#| warning: false
preds_best <- res_lodo$probs %>%
  filter(model==best_model_mc) %>%
  mutate(pred = {
    prob_cols <- setdiff(colnames(.), c("fold","model","truth"))
    lvls_here <- prob_cols[prob_cols %in% levels(y_mc)]
    lvls_here[max.col(.[ , lvls_here, drop=FALSE])]
  })

tab_best <- table(truth=preds_best$truth, pred=preds_best$pred)
as.data.frame(tab_best) %>%
  ggplot(aes(x=pred, y=truth, fill=Freq)) +
  geom_tile(colour="white") +
  geom_text(aes(label=Freq), size=4) +
  scale_fill_gradient(low="white", high="#2C7FB8") +
  theme_bw() +
  labs(x="Predicted", y="True",
       title=paste0("Confusion matrix — ", best_model_mc, " (pooled LODO)")) +
  theme(axis.text.x=element_text(angle=40, hjust=1))
```

------------------------------------------------------------------------

# Single model comparison — hold-out evaluation

```{r ensemble-eval}
#| message: false
#| warning: false

# ── Dev / hold-out split ──────────────────────────────────────────────────────
split_mc <- stratified_split(y_mc, batch_mc, train_prop=0.75)
dev_idx  <- split_mc$dev; hold_idx <- split_mc$test

X_dev  <- X_mc[dev_idx,,drop=FALSE]; y_dev  <- y_mc[dev_idx]; batch_dev  <- batch_mc[dev_idx]
X_hold <- X_mc[hold_idx,,drop=FALSE]; y_hold <- y_mc[hold_idx]; batch_hold <- batch_mc[hold_idx]
cat("Dev:", nrow(X_dev), " Hold-out:", nrow(X_hold), "\n")

# LODO on dev to rank models
res_dev <- run_lodo_multiclass(X_dev, y_dev, batch_dev, n_per_class=50)

dev_rank <- res_dev$perf %>%
  group_by(model) %>%
  summarise(mean_auc = median(macro_auc, na.rm=TRUE),
            mean_f1  = median(macro_f1,  na.rm=TRUE), .groups="drop") %>%
  arrange(desc(mean_auc)) %>%
  mutate(dev_rank = row_number())

# ── Fit on full dev, predict hold-out ─────────────────────────────────────────
fit_once_mc <- function(X_tr, y_tr, batch_tr, X_te, batch_te, n_per_class=50) {
  y_tr <- factor(y_tr); lvls <- levels(y_tr); n_cl <- length(lvls)
  cb   <- combat_fit_apply(t(X_tr), batch_tr, model.matrix(~y_tr), t(X_te), batch_te)
  X_trc <- t(cb$train); X_tec <- t(cb$test)
  gs    <- intersect(select_genes_ovr(t(X_trc), y=y_tr, n_per_class=n_per_class), colnames(X_trc))
  if (length(gs)<10) return(NULL)

  X_t2 <- X_trc[,gs]; X_e2 <- X_tec[,gs]
  imp  <- impute_fit(X_t2);  X_ti <- impute_apply(X_t2,imp); X_ei <- impute_apply(X_e2,imp)
  sc   <- std_fit(X_ti);     X_ts <- std_apply(X_ti,sc);     X_es <- std_apply(X_ei,sc)

  tab_tr <- table(y_tr); n_tot <- sum(tab_tr); n_te <- nrow(X_es)
  cw  <- setNames(vapply(lvls, function(k) n_tot/(n_cl*max(1,tab_tr[k])), numeric(1)), lvls)
  sw  <- cw[as.character(y_tr)]
  sw0 <- rep(1, length(y_tr))
  cw0 <- setNames(rep(1, n_cl), lvls)

  out <- list()

  # ── helper ────────────────────────────────────────────────────────────────
  gp <- function(fit) {   # glmnet → clean prob matrix
    pr <- predict(fit, as.matrix(X_es), type="response", s="lambda.min")[,,1]
    if (is.null(dim(pr))) pr <- t(as.matrix(pr))
    pr[, intersect(lvls, colnames(pr)), drop=FALSE]
  }
  kp <- function() {      # kNN → prob matrix
    pk  <- class::knn(X_ts, X_es, y_tr, k=7, prob=TRUE)
    kw  <- attr(pk,"prob"); pkc <- as.character(pk)
    pm  <- matrix(0, n_te, n_cl, dimnames=list(NULL,lvls))
    for (i in seq_len(n_te)) {
      pm[i,pkc[i]] <- kw[i]; oth <- setdiff(lvls,pkc[i])
      if (length(oth)) pm[i,oth] <- (1-kw[i])/length(oth)
    }
    pm
  }

  # ── unweighted ────────────────────────────────────────────────────────────
  tryCatch({ out[["kNN"]]          <- kp() }, error=function(e) NULL)
  tryCatch({ fit <- e1071::svm(X_ts,y_tr,kernel="radial",cost=1,gamma=1/ncol(X_ts),probability=TRUE)
             ps  <- predict(fit,X_es,probability=TRUE)
             out[["SVM_RBF"]] <- attr(ps,"probabilities")[,lvls,drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- e1071::svm(X_ts,y_tr,kernel="linear",cost=1,probability=TRUE)
             ps  <- predict(fit,X_es,probability=TRUE)
             out[["SVM_Linear"]] <- attr(ps,"probabilities")[,lvls,drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- glmnet::cv.glmnet(as.matrix(X_ts),y_tr,family="multinomial",alpha=1,weights=sw0,standardize=FALSE,type.multinomial="grouped")
             out[["LassoMC"]] <- gp(fit) }, error=function(e) NULL)
  tryCatch({ fit <- glmnet::cv.glmnet(as.matrix(X_ts),y_tr,family="multinomial",alpha=0,weights=sw0,standardize=FALSE,type.multinomial="grouped")
             out[["RidgeMC"]] <- gp(fit) }, error=function(e) NULL)
  tryCatch({ fit <- nnet::multinom(y_tr~.,data=data.frame(y_tr,X_ts),weights=sw0,trace=FALSE,MaxNWts=100000)
             pm  <- predict(fit,newdata=data.frame(X_es),type="probs")
             if (is.null(dim(pm))) pm <- t(as.matrix(pm))
             out[["Multinom"]] <- pm[,intersect(lvls,colnames(pm)),drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- e1071::naiveBayes(x=X_ts,y=y_tr)
             pm  <- predict(fit,X_es,type="raw")
             out[["NaiveBayes"]] <- pm[,intersect(lvls,colnames(pm)),drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- MASS::lda(x=X_ts,grouping=y_tr)
             pl  <- predict(fit,X_es)
             out[["LDA"]] <- pl$posterior[,intersect(lvls,colnames(pl$posterior)),drop=FALSE] }, error=function(e) NULL)
  tryCatch({ y_int <- as.integer(y_tr)-1L
             dtr   <- xgboost::xgb.DMatrix(data=as.matrix(X_ts),label=y_int)
             dte   <- xgboost::xgb.DMatrix(data=as.matrix(X_es))
             fit   <- xgboost::xgb.train(params=list(objective="multi:softprob",eval_metric="mlogloss",num_class=n_cl,eta=0.05,max_depth=4,subsample=0.8,colsample_bytree=0.8,min_child_weight=5,nthread=1),data=dtr,nrounds=150,verbose=0)
             pm    <- matrix(predict(fit,dte),ncol=n_cl,byrow=TRUE); colnames(pm)<-lvls
             out[["XGBoost"]] <- pm }, error=function(e) NULL)

  # ── weighted ──────────────────────────────────────────────────────────────
  tryCatch({ out[["kNN_w"]]       <- kp() }, error=function(e) NULL)  # no native weights
  tryCatch({ fit <- e1071::svm(X_ts,y_tr,kernel="radial",cost=1,gamma=1/ncol(X_ts),probability=TRUE,class.weights=cw)
             ps  <- predict(fit,X_es,probability=TRUE)
             out[["SVM_RBF_w"]] <- attr(ps,"probabilities")[,lvls,drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- e1071::svm(X_ts,y_tr,kernel="linear",cost=1,probability=TRUE,class.weights=cw)
             ps  <- predict(fit,X_es,probability=TRUE)
             out[["SVM_Linear_w"]] <- attr(ps,"probabilities")[,lvls,drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- glmnet::cv.glmnet(as.matrix(X_ts),y_tr,family="multinomial",alpha=1,weights=sw,standardize=FALSE,type.multinomial="grouped")
             out[["LassoMC_w"]] <- gp(fit) }, error=function(e) NULL)
  tryCatch({ fit <- glmnet::cv.glmnet(as.matrix(X_ts),y_tr,family="multinomial",alpha=0,weights=sw,standardize=FALSE,type.multinomial="grouped")
             out[["RidgeMC_w"]] <- gp(fit) }, error=function(e) NULL)
  tryCatch({ fit <- nnet::multinom(y_tr~.,data=data.frame(y_tr,X_ts),weights=sw,trace=FALSE,MaxNWts=100000)
             pm  <- predict(fit,newdata=data.frame(X_es),type="probs")
             if (is.null(dim(pm))) pm <- t(as.matrix(pm))
             out[["Multinom_w"]] <- pm[,intersect(lvls,colnames(pm)),drop=FALSE] }, error=function(e) NULL)
  tryCatch({ fit <- e1071::naiveBayes(x=X_ts,y=y_tr)  # no native weights
             pm  <- predict(fit,X_es,type="raw")
             out[["NaiveBayes_w"]] <- pm[,intersect(lvls,colnames(pm)),drop=FALSE] }, error=function(e) NULL)
  tryCatch({ prior_w <- cw/sum(cw)
             fit <- MASS::lda(x=X_ts,grouping=y_tr,prior=prior_w)
             pl  <- predict(fit,X_es)
             out[["LDA_w"]] <- pl$posterior[,intersect(lvls,colnames(pl$posterior)),drop=FALSE] }, error=function(e) NULL)
  tryCatch({ y_int <- as.integer(y_tr)-1L
             dtr   <- xgboost::xgb.DMatrix(data=as.matrix(X_ts),label=y_int,weight=sw)
             dte   <- xgboost::xgb.DMatrix(data=as.matrix(X_es))
             fit   <- xgboost::xgb.train(params=list(objective="multi:softprob",eval_metric="mlogloss",num_class=n_cl,eta=0.05,max_depth=4,subsample=0.8,colsample_bytree=0.8,min_child_weight=5,nthread=1),data=dtr,nrounds=150,verbose=0)
             pm    <- matrix(predict(fit,dte),ncol=n_cl,byrow=TRUE); colnames(pm)<-lvls
             out[["XGBoost_w"]] <- pm }, error=function(e) NULL)

  Filter(Negate(is.null), out)
}

hold_probs <- fit_once_mc(X_dev, y_dev, batch_dev, X_hold, batch_hold)

if (!is.null(hold_probs)) {
  lvls_hold <- levels(y_dev)

  single_hold <- bind_rows(lapply(names(hold_probs), function(nm) {
    tryCatch({
      pm  <- hold_probs[[nm]]
      if (is.null(pm) || ncol(pm)==0) return(NULL)
      lv  <- intersect(lvls_hold, colnames(pm))
      if (length(lv)<2) return(NULL)
      pred_chr <- lv[max.col(pm[,lv,drop=FALSE])]
      tibble::tibble(type="Single model", id=nm,
        accuracy  = mean(pred_chr==as.character(y_hold), na.rm=TRUE),
        macro_f1  = macro_f1(as.character(y_hold), pred_chr),
        macro_auc = macro_auc_ovr(as.character(y_hold), pm[,lv,drop=FALSE]))
    }, error=function(e) NULL)
  }))
  if (nrow(single_hold)==0) stop("No models produced valid hold-out predictions — check fit_once_mc output")
  single_hold <- single_hold %>% arrange(desc(macro_auc))

  knitr::kable(
    single_hold %>% mutate(across(c(macro_auc, macro_f1, accuracy), round, 3)),
    caption="Hold-out: single model performance ranked by macro OvR AUC"
  )

  # ── Visual: all single models ranked ─────────────────────────────────────
  single_hold %>%
    mutate(across(c(macro_auc, macro_f1, accuracy), round, 3)) %>%
    pivot_longer(cols=c(macro_auc, macro_f1, accuracy),
                 names_to="metric", values_to="value") %>%
    mutate(metric=recode(metric,
      macro_auc="Macro OvR AUC", macro_f1="Macro F1", accuracy="Accuracy"),
      weighted = grepl("_w$", id)) %>%
    ggplot(aes(x=reorder(id, value), y=value,
               fill=weighted)) +
    geom_col(alpha=0.85) +
    geom_text(aes(label=sprintf("%.3f", value)), hjust=-0.1, size=2.8) +
    facet_wrap(~metric, scales="free_x") +
    coord_flip() +
    scale_fill_manual(values=c("FALSE"="grey50","TRUE"="#D7191C"),
                      labels=c("FALSE"="Unweighted","TRUE"="Weighted")) +
    theme_bw(base_size=9) +
    labs(x=NULL, y=NULL, fill=NULL,
         title="Hold-out performance — all single models (weighted ▶ red, unweighted ▶ grey)") +
    theme(legend.position="bottom")
}
```

------------------------------------------------------------------------

# Uncertainty Quantification

> **Two complementary uncertainty signals:**
>
> 1.  **Model disagreement** — when ensemble members assign very different probabilities to the same sample, gene expression alone may not be sufficient to make a confident call.
>
> 2.  **Bootstrap confidence intervals** — resampling the hold-out set 500 times gives a distribution over macro AUC and per-sample risk scores, quantifying how stable the estimates are.

## Model disagreement per sample

```{r uncertainty-disagreement, fig.cap="Model disagreement (SD of predicted subtype probability across ensemble members). High disagreement = low confidence in the call."}
#| message: false
#| warning: false
if (!is.null(hold_probs) && length(hold_probs) >= 2) {
  lvls_u <- lvls_hold
  sel_mods_u <- dev_rank %>% slice_head(n=min(4,nrow(dev_rank))) %>%
    pull(model) %>% intersect(names(hold_probs))

  # For each sample × subtype, collect probabilities across models
  prob_stack <- array(NA_real_, dim=c(nrow(X_hold), length(lvls_u), length(sel_mods_u)),
                      dimnames=list(NULL, lvls_u, sel_mods_u))
  for (nm in sel_mods_u) {
    pm <- hold_probs[[nm]]; lv <- intersect(lvls_u, colnames(pm))
    prob_stack[, lv, nm] <- pm[, lv]
  }

  # Predicted class = argmax of ensemble average
  avg_stack <- apply(prob_stack, c(1,2), mean, na.rm=TRUE)
  pred_class <- lvls_u[max.col(avg_stack)]

  # Disagreement = SD of predicted-class probability across models
  pred_class_idx <- match(pred_class, lvls_u)
  disagreement <- vapply(seq_len(nrow(X_hold)), function(i) {
    probs_for_pred <- prob_stack[i, pred_class_idx[i], ]
    sd(probs_for_pred, na.rm=TRUE)
  }, numeric(1))

  # Entropy of ensemble mean (overall uncertainty)
  entropy <- apply(avg_stack, 1, function(p) {
    p <- p[p > 0]; -sum(p * log(p))
  })

  uncertainty_df <- tibble::tibble(
    sample_idx    = seq_len(nrow(X_hold)),
    true_subtype  = as.character(y_hold),
    pred_subtype  = pred_class,
    correct       = pred_class == as.character(y_hold),
    confidence    = apply(avg_stack, 1, max),
    disagreement  = disagreement,
    entropy       = entropy
  )

  # Plot: confidence vs disagreement
  p1 <- ggplot(uncertainty_df, aes(x=confidence, y=disagreement, colour=correct)) +
    geom_point(alpha=0.75, size=2) +
    scale_colour_manual(values=c("TRUE"="#41AB5D","FALSE"="#D7191C"),
                        labels=c("TRUE"="Correct","FALSE"="Incorrect")) +
    theme_bw() +
    labs(x="Ensemble confidence (max P)", y="Model disagreement (SD of P)",
         colour=NULL,
         title="Confidence vs model disagreement",
         subtitle="Top-right = high confidence but models disagree (flag for review)")

  # Plot: disagreement by true subtype
  p2 <- ggplot(uncertainty_df, aes(x=true_subtype, y=disagreement, fill=true_subtype)) +
    geom_boxplot(alpha=0.75, outlier.size=0.8) +
    theme_bw() +
    labs(x="True subtype", y="Model disagreement (SD)",
         title="Which subtypes are hardest to predict consistently?") +
    theme(legend.position="none",
          axis.text.x=element_text(angle=30, hjust=1))

  print(p1); print(p2)

  cat("\nHigh-uncertainty samples (disagreement > 75th percentile):\n")
  threshold_u <- quantile(uncertainty_df$disagreement, 0.75, na.rm=TRUE)
  knitr::kable(
    uncertainty_df %>%
      filter(disagreement > threshold_u) %>%
      arrange(desc(disagreement)) %>%
      mutate(across(c(confidence, disagreement, entropy), round, 3)) %>%
      head(15),
    caption="High-uncertainty samples — models disagree most on these"
  )
}
```

## Per-patient bootstrap CI on subtype probabilities

> For each bootstrap iteration we: **(1)** resample the dev (training) set with replacement, **(2)** refit the best model (Multinomial Lasso) on that resample, **(3)** predict subtype probabilities on the **fixed** hold-out patients. The spread of B predictions per patient is the CI on that patient's risk score.

```{r bootstrap-ci}
#| message: false
#| warning: false
set.seed(3888)
B_boot_mc <- 200

run_bootstrap_mc <- function(X_tr, y_tr, batch_tr, X_te, batch_te,
                              B = 200, n_per_class = 50) {
  y_tr  <- factor(y_tr)
  lvls  <- levels(y_tr)
  n_cl  <- length(lvls)
  n_te  <- nrow(X_te)

  # Batch-correct full training set once
  cb <- combat_fit_apply(t(X_tr), batch_tr, model.matrix(~y_tr), t(X_te), batch_te)
  X_trc <- t(cb$train); X_tec <- t(cb$test)

  # Gene selection on full corrected training set
  gs <- tryCatch({
    sel <- select_genes_ovr(t(X_trc), y=y_tr, n_per_class=n_per_class)
    intersect(sel, colnames(X_trc))
  }, error=function(e) character(0))
  if (length(gs) < 10) { message("Too few genes for bootstrap"); return(NULL) }

  X_t2 <- X_trc[,gs]; X_e2 <- X_tec[,gs]
  imp  <- impute_fit(X_t2)
  X_ti <- impute_apply(X_t2,imp); X_ei <- impute_apply(X_e2,imp)
  sc   <- std_fit(X_ti)
  X_ts <- std_apply(X_ti,sc);    X_es <- std_apply(X_ei,sc)

  n_tr <- nrow(X_ts)
  tab_tr <- table(y_tr); n_tot <- sum(tab_tr)
  cw <- setNames(vapply(lvls,function(k) n_tot/(n_cl*max(1,tab_tr[k])),numeric(1)),lvls)
  sw_full <- cw[as.character(y_tr)]

  # Find best lambda ONCE on full training set — reuse in every bootstrap iteration
  # This avoids 10-fold inner CV per iteration (~10x speedup)
  cv_once <- tryCatch(
    glmnet::cv.glmnet(as.matrix(X_ts), y_tr, family="multinomial",
                      alpha=1, weights=sw_full, standardize=FALSE,
                      type.multinomial="grouped"),
    error=function(e) NULL
  )
  if (is.null(cv_once)) { message("cv.glmnet setup failed"); return(NULL) }
  lambda_fixed_mc <- cv_once$lambda.min
  cat(sprintf("  Fixed lambda: %.5f\n", lambda_fixed_mc))

  # Boot array: samples × subtypes × iterations
  boot_arr <- array(NA_real_, dim=c(n_te, n_cl, B),
                    dimnames=list(NULL, lvls, NULL))

  for (b in seq_len(B)) {
    idx_b  <- sample(n_tr, n_tr, replace=TRUE)
    y_b    <- droplevels(y_tr[idx_b])
    X_b    <- X_ts[idx_b,,drop=FALSE]
    if (nlevels(y_b) < 2) next

    tab_b <- table(y_b); n_tot_b <- sum(tab_b)
    lvls_b <- levels(y_b)
    sw_b   <- setNames(vapply(lvls_b,function(k) n_tot_b/(length(lvls_b)*max(1,tab_b[k])),
                              numeric(1)), lvls_b)[as.character(y_b)]

    # Plain glmnet with fixed lambda — no inner CV per iteration
    lf <- tryCatch(
      glmnet::glmnet(as.matrix(X_b), y_b, family="multinomial",
                     alpha=1, weights=sw_b, standardize=FALSE,
                     lambda=lambda_fixed_mc,
                     type.multinomial="grouped"),
      error=function(e) NULL
    )
    if (is.null(lf)) next

    pr <- predict(lf, as.matrix(X_es), type="response",
                  s=lambda_fixed_mc)[,,1]
    if (is.null(dim(pr))) pr <- t(as.matrix(pr))
    lv_common <- intersect(lvls, colnames(pr))
    boot_arr[, lv_common, b] <- pr[, lv_common]
  }

  boot_arr
}

cat("Running", B_boot_mc, "bootstrap iterations (Multinomial Lasso refit)...
")
boot_arr_mc <- run_bootstrap_mc(
  X_dev, y_dev, batch_dev, X_hold, batch_hold,
  B=B_boot_mc, n_per_class=50
)

if (!is.null(boot_arr_mc)) {
  lvls_b <- dimnames(boot_arr_mc)[[2]]

  # Per-patient, per-subtype: mean and 95% CI across B iterations
  prob_mean <- apply(boot_arr_mc, c(1,2), mean, na.rm=TRUE)
  prob_lo   <- apply(boot_arr_mc, c(1,2), quantile, 0.025, na.rm=TRUE)
  prob_hi   <- apply(boot_arr_mc, c(1,2), quantile, 0.975, na.rm=TRUE)
  ci_width  <- prob_hi - prob_lo

  # Predicted subtype = argmax of mean probability
  pred_subtype_boot <- lvls_b[max.col(prob_mean)]
  confidence_boot   <- apply(prob_mean, 1, max)
  ci_of_pred        <- vapply(seq_len(nrow(prob_mean)), function(i)
    ci_width[i, pred_subtype_boot[i]], numeric(1))

  boot_risk_mc <- tibble::tibble(
    sample_id      = seq_len(nrow(X_hold)),
    true_subtype   = as.character(y_hold),
    pred_subtype   = pred_subtype_boot,
    correct        = pred_subtype_boot == as.character(y_hold),
    confidence     = round(confidence_boot, 3),
    ci_width_pred  = round(ci_of_pred, 3),
    uncertainty_band = case_when(
      ci_of_pred <= 0.10 ~ "Narrow (confident)",
      ci_of_pred <= 0.20 ~ "Moderate",
      TRUE               ~ "Wide (uncertain — review)"
    )
  )

  # Add per-subtype probability + CI columns
  for (k in lvls_b) {
    boot_risk_mc[[paste0("P_", k)]]      <- round(prob_mean[, k], 3)
    boot_risk_mc[[paste0("CI_lo_", k)]]  <- round(prob_lo[,  k], 3)
    boot_risk_mc[[paste0("CI_hi_", k)]]  <- round(prob_hi[,  k], 3)
  }

  cat("
Uncertainty band distribution:
")
  print(table(boot_risk_mc$uncertainty_band))
  cat("
Accuracy (bootstrap mean prediction):",
      round(mean(boot_risk_mc$correct, na.rm=TRUE), 3), "
")

  knitr::kable(
    boot_risk_mc %>%
      arrange(desc(confidence)) %>%
      select(sample_id, true_subtype, pred_subtype, correct,
             confidence, ci_width_pred, uncertainty_band) %>%
      head(20),
    caption=paste0("Per-patient PAM50 predictions with bootstrap 95% CI (B=",
                   B_boot_mc, " Lasso refits). Wide CI = uncertain.")
  )

  # ── Forest plot: top 40 most confident patients ──
  boot_risk_mc %>%
    arrange(desc(confidence)) %>%
    slice_head(n=40) %>%
    mutate(sample_id=factor(sample_id, levels=rev(sample_id))) %>%
    ggplot(aes(x=confidence, y=sample_id, colour=correct)) +
    geom_errorbarh(aes(xmin=confidence - ci_width_pred/2,
                       xmax=confidence + ci_width_pred/2),
                   height=0.4, alpha=0.6) +
    geom_point(size=2) +
    scale_colour_manual(values=c("TRUE"="#41AB5D","FALSE"="#D7191C"),
                        labels=c("TRUE"="Correct","FALSE"="Incorrect")) +
    theme_bw(base_size=9) +
    labs(x="Predicted subtype confidence  [point = mean, bar = 95% CI]",
         y="Hold-out patient",
         colour=NULL,
         title=paste0("Per-patient subtype confidence with bootstrap uncertainty (B=",B_boot_mc,")"),
         subtitle="Wide bars = model uncertain. Dashed = 0.5 threshold.") +
    geom_vline(xintercept=0.5, linetype=2, colour="grey50") +
    theme(axis.text.y=element_blank(), axis.ticks.y=element_blank(),
          legend.position="bottom")

  # ── CI width by true subtype ──
  ggplot(boot_risk_mc, aes(x=true_subtype, y=ci_width_pred, fill=true_subtype)) +
    geom_violin(alpha=0.7, trim=FALSE) +
    geom_boxplot(width=0.15, fill="white", outlier.size=0.8) +
    theme_bw() +
    labs(x="True subtype", y="Bootstrap 95% CI width on predicted subtype",
         title="Which subtypes are hardest to predict with confidence?",
         subtitle="Wider CI = model less certain across training resamples") +
    theme(legend.position="none",
          axis.text.x=element_text(angle=30, hjust=1))

  cat(sprintf("\nMedian CI width: %.3f  |  Mean CI width: %.3f\n",
    median(boot_risk_mc$ci_width_pred, na.rm=TRUE),
    mean(boot_risk_mc$ci_width_pred, na.rm=TRUE)))
} else {
  cat("Bootstrap could not run.\n")
}
```

------------------------------------------------------------------------

# PAM50 Subtype Risk Calculator

```{r risk-calculator}
#| message: false
#| warning: false
if (!is.null(hold_probs) && exists("uncertainty_df")) {
  risk_df <- uncertainty_df %>%
    bind_cols(as.data.frame(avg_stack)) %>%
    mutate(
      risk_band = case_when(
        confidence >= 0.70 & disagreement <= quantile(disagreement, 0.25, na.rm=TRUE) ~ "High confidence",
        confidence >= 0.50 ~ "Moderate confidence",
        TRUE ~ "Low confidence — review recommended"
      )
    ) %>%
    select(sample_idx, true_subtype, pred_subtype, correct,
           confidence, disagreement,
           any_of("boot_ci_width"),  # populated later by bootstrap section
           entropy, risk_band,
           all_of(lvls_hold))

  cat("\nRisk band distribution:\n")
  print(table(risk_df$risk_band))

  knitr::kable(
    risk_df %>%
      arrange(desc(confidence)) %>%
      mutate(across(all_of(lvls_hold), ~round(.x, 3)),
             across(c(confidence, disagreement, entropy), round, 3),
             across(any_of("boot_ci_width"), round, 3)) %>%
      head(20),
    caption="PAM50 subtype risk calculator — top 20 most confident predictions"
  )

  # Distribution of confidence by predicted subtype
  ggplot(risk_df, aes(x=pred_subtype, y=confidence, fill=pred_subtype)) +
    geom_violin(alpha=0.7, trim=FALSE) +
    geom_boxplot(width=0.15, fill="white", outlier.size=0.8) +
    theme_bw() +
    labs(x="Predicted subtype", y="Confidence (max P)",
         title="Prediction confidence by PAM50 subtype") +
    theme(legend.position="none",
          axis.text.x=element_text(angle=30, hjust=1))

  # Stacked probability bars (top 30 samples sorted by confidence)
  risk_df %>%
    arrange(desc(confidence)) %>%
    slice_head(n=30) %>%
    mutate(sample_idx=factor(sample_idx, levels=sample_idx)) %>%
    pivot_longer(cols=all_of(lvls_hold), names_to="subtype", values_to="prob") %>%
    ggplot(aes(x=sample_idx, y=prob, fill=subtype)) +
    geom_col(width=0.9) +
    theme_bw(base_size=9) +
    labs(x="Sample (top 30 by confidence)", y="P(subtype)", fill="Subtype",
         title="Predicted subtype probability profiles — top 30 most confident") +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
}
```

## Summary for clinical interpretation

```{r summary-table}
#| message: false
#| warning: false
if (exists("risk_df")) {
  cat("=== CLASSIFIER SUMMARY ===\n\n")
  cat("Best single model (LODO CV):", best_model_mc, "\n")

  cat(sprintf("  High-confidence predictions: %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band == "High confidence"),
              nrow(risk_df),
              100 * mean(risk_df$risk_band == "High confidence", na.rm=TRUE)))
  cat(sprintf("  Moderate-confidence:         %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band == "Moderate confidence"),
              nrow(risk_df),
              100 * mean(risk_df$risk_band == "Moderate confidence", na.rm=TRUE)))
  cat(sprintf("  Low-confidence (review):     %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band == "Low confidence — review recommended"),
              nrow(risk_df),
              100 * mean(risk_df$risk_band == "Low confidence — review recommended", na.rm=TRUE)))
  cat(sprintf("\n  Median predicted confidence: %.1f%%\n",
              median(as.numeric(gsub("%","",risk_df$confidence)), na.rm=TRUE)*100))

  if (exists("boot_arr_mc") && !is.null(boot_arr_mc)) {
    cat(sprintf("  Bootstrap 95%% CI (median ci_width): %.1f%%\n",
                median(boot_risk_mc$ci_width_pred, na.rm=TRUE)*100))
  }

  cat(sprintf("\nHigh-confidence predictions (≥0.70, low disagreement): %d / %d (%.0f%%)\n",
              sum(risk_df$risk_band=="High confidence"),
              nrow(risk_df),
              100*mean(risk_df$risk_band=="High confidence", na.rm=TRUE)))
}
```

------------------------------------------------------------------------

# Stage 4 — Save final model for deployment

> This mirrors the binary pipeline's save pattern. Everything needed to predict the PAM50 subtype of a **new patient** is saved to `models/multiclass/`: gene panel, batch correction parameters, preprocessing stats, and fitted models.

```{r save-final-model}
#| message: false
#| warning: false

dir.create("models/multiclass", recursive=TRUE, showWarnings=FALSE)

# ── 1. Full training data: batch-correct + gene selection ────────────────────
# All labeled samples across all 3 datasets
ok_all   <- !is.na(p_merged_mc$label_mc) & p_merged_mc$label_mc != ""
y_all    <- factor(p_merged_mc$label_mc[ok_all])
e_all    <- e_merged_mc[, ok_all, drop=FALSE]
batch_all <- p_merged_mc$dataset_id[ok_all]

mod_all <- model.matrix(~ y_all)

# Batch-correct full training set. Store the raw batch parameters manually
# so we can apply them to future new samples.
grand_mean_all  <- rowMeans(e_all)
batch_train_f   <- factor(as.character(batch_all))
batch_means_all <- sapply(levels(batch_train_f), function(b) {
  idx <- which(batch_train_f == b)
  rowMeans(e_all[, idx, drop=FALSE])
})

e_all_bc <- tryCatch(
  sva::ComBat(dat=e_all, batch=batch_all, mod=mod_all,
              par.prior=TRUE, mean.only=FALSE),
  error = function(e) {
    message("ComBat failed, using removeBatchEffect: ", conditionMessage(e))
    limma::removeBatchEffect(e_all, batch=batch_all, design=mod_all)
  }
)

# ── 2. OvR gene selection on full batch-corrected training set ───────────────
panel_genes <- select_genes_ovr(e_all_bc, y=y_all, n_per_class=50)
cat("Final multiclass panel genes:", length(panel_genes), "\n")

# ── 3. Preprocessing stats (imputation + z-score) from full training set ─────
X_all_bc  <- t(e_all_bc[panel_genes, , drop=FALSE])
imp_final <- impute_fit(X_all_bc)
X_all_i   <- impute_apply(X_all_bc, imp_final)
sc_final  <- std_fit(X_all_i)
X_all_s   <- std_apply(X_all_i, sc_final)

# ── 4. Fit final models (same set as LODO) ───────────────────────────────────
lvls_all  <- levels(y_all)
n_cl_all  <- length(lvls_all)
tab_all   <- table(y_all); n_tot_all <- sum(tab_all)
cw_all    <- setNames(vapply(lvls_all, function(k)
  n_tot_all / (n_cl_all * max(1, tab_all[k])), numeric(1)), lvls_all)
sw_all    <- cw_all[as.character(y_all)]

# kNN (stored as training data + labels — kNN has no compact fit object)
fit_knn_final <- list(X_train = X_all_s, y_train = y_all, k = 7)

# RF
fit_rf_final <- randomForest::randomForest(X_all_i, y_all, classwt=cw_all)

# Multinomial Lasso
fit_lasso_final <- glmnet::cv.glmnet(
  as.matrix(X_all_s), y_all, family="multinomial",
  alpha=1, weights=sw_all, standardize=FALSE, type.multinomial="grouped"
)

# Multinom (nnet)
fit_multinom_final <- if (requireNamespace("nnet", quietly=TRUE)) {
  tryCatch(
    nnet::multinom(y_all~., data=data.frame(y_all, X_all_s),
                   weights=sw_all, trace=FALSE, MaxNWts=100000),
    error=function(e) { message("nnet failed: ", conditionMessage(e)); NULL }
  )
} else NULL

# ── 5. Save everything ────────────────────────────────────────────────────────
saveRDS(panel_genes,    "models/multiclass/panel_genes.rds")

saveRDS(list(
  grand_mean   = grand_mean_all,
  batch_means  = batch_means_all,
  datasets_seen = levels(batch_train_f)
), "models/multiclass/batch_correction_params.rds")

saveRDS(imp_final,      "models/multiclass/imputation_params.rds")
saveRDS(sc_final,       "models/multiclass/scale_params.rds")
saveRDS(fit_knn_final,  "models/multiclass/fit_knn.rds")
saveRDS(fit_rf_final,   "models/multiclass/fit_rf.rds")
saveRDS(fit_lasso_final,"models/multiclass/fit_lasso.rds")
if (!is.null(fit_multinom_final))
  saveRDS(fit_multinom_final, "models/multiclass/fit_multinom.rds")

cat("\nSaved to models/multiclass/:\n")
cat(paste(" -", list.files("models/multiclass"), collapse="\n"), "\n")

# ── Save uncertainty outputs ──────────────────────────────────────────────────
if (exists("boot_risk_mc") && !is.null(boot_risk_mc)) {
  saveRDS(boot_risk_mc, "models/multiclass/bootstrap_risk_scores.rds")
  cat("Saved bootstrap risk scores:", nrow(boot_risk_mc), "patients\n")
}

if (exists("uncertainty_df") && !is.null(uncertainty_df)) {
  saveRDS(uncertainty_df, "models/multiclass/model_disagreement.rds")
  cat("Saved model disagreement metrics\n")
}
```

## Prediction function for new samples

```{r predict-function}
#| message: false
#| warning: false

# ── Load all saved components ─────────────────────────────────────────────────
predict_pam50_subtype <- function(expr_new,
                                   model_dir  = "models/multiclass",
                                   model_name = "RF",
                                   new_dataset = NULL) {
  # expr_new : genes × samples matrix (raw, quantile-normalised per dataset)
  # model_name: one of "RF", "LassoMC", "Multinom", "kNN"
  # new_dataset: optional string — dataset name if known, NULL for novel data

  panel_genes  <- readRDS(file.path(model_dir, "panel_genes.rds"))
  bc_params    <- readRDS(file.path(model_dir, "batch_correction_params.rds"))
  imp_params   <- readRDS(file.path(model_dir, "imputation_params.rds"))
  sc_params    <- readRDS(file.path(model_dir, "scale_params.rds"))

  # ── Step 1: apply batch correction ───────────────────────────────────────────
  genes_common <- intersect(panel_genes, rownames(expr_new))
  if (length(genes_common) < 10)
    stop("Too few panel genes found in new data (",
         length(genes_common), "/", length(panel_genes), ")")

  e_new <- expr_new[genes_common, , drop=FALSE]

  if (!is.null(new_dataset) &&
      new_dataset %in% bc_params$datasets_seen) {
    # Known dataset: subtract the learned batch offset
    batch_offset <- bc_params$batch_means[genes_common, new_dataset] -
                    bc_params$grand_mean[genes_common]
    e_new_bc <- sweep(e_new, 1, batch_offset, "-")
    message("Applied batch offset for known dataset: ", new_dataset)
  } else {
    # Novel dataset: mean-centre each sample to the training grand mean
    sample_means  <- colMeans(e_new)
    grand_mean_sub <- bc_params$grand_mean[genes_common]
    e_new_bc <- sweep(e_new, 2, sample_means, "-") +
                rowMeans(matrix(grand_mean_sub, ncol=1))
    message("Novel dataset — applied mean-centring only")
  }

  # ── Step 2: impute + scale using training statistics ─────────────────────────
  # Subset imp and scale params to genes present
  imp_sub <- list(med = imp_params$med[genes_common])
  sc_sub  <- list(mu  = sc_params$mu[genes_common],
                  sd  = sc_params$sd[genes_common])

  X_new   <- t(e_new_bc)
  X_new_i <- impute_apply(X_new, imp_sub)
  X_new_s <- std_apply(X_new_i, sc_sub)

  # ── Step 3: predict ───────────────────────────────────────────────────────────
  prob_mat <- if (model_name == "RF") {
    fit <- readRDS(file.path(model_dir, "fit_rf.rds"))
    predict(fit, X_new_i, type="prob")

  } else if (model_name == "LassoMC") {
    fit <- readRDS(file.path(model_dir, "fit_lasso.rds"))
    pr  <- predict(fit, as.matrix(X_new_s), type="response",
                   s="lambda.min")[,,1]
    if (is.null(dim(pr))) t(as.matrix(pr)) else pr

  } else if (model_name == "Multinom") {
    mpath <- file.path(model_dir, "fit_multinom.rds")
    if (!file.exists(mpath)) stop("fit_multinom.rds not found")
    fit <- readRDS(mpath)
    pm  <- predict(fit, newdata=data.frame(X_new_s), type="probs")
    if (is.null(dim(pm))) t(as.matrix(pm)) else pm

  } else if (model_name == "kNN") {
    fit      <- readRDS(file.path(model_dir, "fit_knn.rds"))
    pred_knn <- class::knn(train=fit$X_train[, genes_common, drop=FALSE],
                            test=X_new_s,
                            cl=fit$y_train, k=fit$k, prob=TRUE)
    kw  <- attr(pred_knn, "prob")
    pkc <- as.character(pred_knn)
    lvls <- levels(fit$y_train)
    pm  <- matrix(0, nrow(X_new_s), length(lvls), dimnames=list(NULL, lvls))
    for (i in seq_len(nrow(pm))) {
      pm[i, pkc[i]] <- kw[i]
      oth <- setdiff(lvls, pkc[i])
      if (length(oth)) pm[i, oth] <- (1 - kw[i]) / length(oth)
    }
    pm

  } else {
    stop("Unknown model_name: ", model_name,
         ". Choose from: RF, LassoMC, Multinom, kNN")
  }

  # ── Step 4: format output ─────────────────────────────────────────────────────
  pred_subtype <- colnames(prob_mat)[max.col(prob_mat)]
  confidence   <- apply(prob_mat, 1, max)

  result <- as.data.frame(round(prob_mat, 3))
  result$predicted_subtype <- pred_subtype
  result$confidence        <- round(confidence, 3)
  result$model             <- model_name
  result$genes_used        <- length(genes_common)
  result$genes_missing     <- length(panel_genes) - length(genes_common)

  result
}

# ── Quick sanity check on hold-out set ───────────────────────────────────────
if (exists("X_mc_hold") && nrow(X_mc_hold) > 0) {
  # Reconstruct a genes × samples matrix from hold-out
  e_hold_raw <- t(X_mc_hold)  # back to genes × samples

  test_result <- tryCatch(
    predict_pam50_subtype(e_hold_raw, model_name="RF", new_dataset=NULL),
    error = function(e) { message("Prediction test failed: ", conditionMessage(e)); NULL }
  )

  if (!is.null(test_result)) {
    test_result$true_subtype <- as.character(y_mc_hold)
    test_result$correct      <- test_result$predicted_subtype == test_result$true_subtype

    cat(sprintf("Sanity check — hold-out accuracy (RF): %.1f%%\n",
                100 * mean(test_result$correct, na.rm=TRUE)))
    knitr::kable(
      head(test_result %>%
             arrange(desc(confidence)) %>%
             select(predicted_subtype, true_subtype, correct,
                    confidence, genes_used, genes_missing), 10),
      caption="predict_pam50_subtype() — top 10 most confident hold-out predictions"
    )
  }
}
```
