---
title: "Breast Cancer Risk Report"
author: "Zerek Lee"
date: "`r Sys.Date()`"
output: pdf_document
params:
  patient_id: NULL
  risk_score: NULL
  risk_category: NULL
  ci_low: NULL
  ci_high: NULL
  rf: NULL
  ridge: NULL
  nb: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Patient Information

**Patient ID:** `r params$patient_id`

---

# Main Risk Result

## Risk Score

`r params$risk_score`

## Risk Category

`r params$risk_category`

## 95% Confidence Interval

`r params$ci_low` — `r params$ci_high`

---

# Base Model Predictions

| Model | Output |
|---|---|
| Random Forest | `r params$rf` |
| Ridge Regression | `r params$ridge` |
| Naive Bayes | `r params$nb` |

---

# Clinical Interpretation

```{r clinical_interpretation}
if (params$risk_category == "High Risk") {
  cat("The gene expression profile indicates a high predicted risk.")
} else if (params$risk_category == "Medium Risk") {
  cat("The gene expression profile indicates an intermediate predicted risk.")
} else {
  cat("The gene expression profile indicates a low predicted risk.")
}
```

---

# Important Notice

This report is generated from a prototype clinical decision-support dashboard.

Clinical interpretation should always be combined with professional medical judgement.