---
title: "Breast Cancer Risk Report"
author: "Clinician"
date: "`r Sys.Date()`"
output: pdf_document
params:
  patient_id: NULL
  risk_score: NULL
  risk_category: NULL
  ci_low: NULL
  ci_high: NULL
  gene_table: NULL
  high_genes: NULL
  low_genes: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
```

# Patient Information

**Patient ID:** `r params$patient_id`

---

# Main Risk Result

## Risk Score

`r params$risk_score`

## Risk Category

`r params$risk_category`

## 95% Confidence Interval

`r params$ci_low` — `r params$ci_high`

---

# Clinical Interpretation

`r if (params$risk_category == "High Risk") {
  "The gene expression profile indicates a high predicted risk. Further clinical assessment and molecular subtype assessment is recommended."
} else if (params$risk_category == "Medium Risk") {
  "The gene expression profile indicates an intermediate predicted risk. Additional clinical review may be useful before making decisions."
} else {
  "The gene expression profile indicates a low predicted risk. Routine monitoring may be appropriate depending on clinical context."
}`

---

# Top Abnormal Gene Summary

## High Expression Genes

```{r high_genes_table}
if (!is.null(params$high_genes) && nrow(params$high_genes) > 0) {
  kable(
    params$high_genes,
    row.names = FALSE,
    align = "c"
  )
} else {
  cat("No high-expression genes detected.")
}
```

## Low Expression Genes

```{r low_genes_table}
if (!is.null(params$low_genes) && nrow(params$low_genes) > 0) {
  kable(
    params$low_genes,
    row.names = FALSE,
    align = "c"
  )
} else {
  cat("No low-expression genes detected.")
}
```

---

# Selected Patient Gene Values

```{r gene_table}
if (!is.null(params$gene_table)) {
  kable(
    params$gene_table,
    row.names = FALSE,
    align = "c"
  )
}
```

---

# Selected Sample Gene Expression

The dashboard visualization compares the selected patient against the cohort-level distribution for each gene.

- Horizontal boxplots represent cohort gene-expression distributions.
- Red triangle markers indicate the selected patient.
- Percentile rankings are used to identify unusually high or low expression patterns.

---

# Important Notice

This report is generated from a prototype clinical decision-support dashboard.

Current risk predictions are demonstration outputs and will be replaced with fully trained machine-learning models in the final version.

Clinical interpretation should always be combined with professional medical judgement.