---
title: "Model building"
format:
  html:
    embed-resources: true
    code-fold: true
    df-print: paged
    toc: true
editor: visual
---
```{r}
library(dplyr)
library(tibble)

seed_base <- 3888L
set.seed(seed_base)

seed_from_label <- function(label, offset = 0L) {
  label_int <- utf8ToInt(as.character(label))
  label_hash <- sum(seq_along(label_int) * label_int)
  as.integer((seed_base + label_hash + offset) %% (.Machine$integer.max - 1L))
}

make_cv_fold_spec <- function(y, label, max_folds = 5L) {
  y_chr <- as.character(y)
  min_class_n <- min(as.integer(table(y_chr)))
  nfolds <- min(as.integer(max_folds), min_class_n)
  nfolds <- max(2L, nfolds)
  shift <- seed_from_label(label, 0L) %% nfolds
  
  foldid <- integer(length(y_chr))
  
  for (cls in sort(unique(y_chr))) {
    idx <- which(y_chr == cls)
    idx <- idx[order(idx)]
    cls_folds <- rep(seq_len(nfolds), length.out = length(idx))
    cls_folds <- ((cls_folds - 1L + shift) %% nfolds) + 1L
    foldid[idx] <- cls_folds
  }
  
  list(
    foldid = foldid,
    nfolds = nfolds
  )
}

#load data
dat <- readRDS("data/normalised/merged_normalised.rds")

expr <- dat$expr          # genes x samples expression matrix
meta <- dat$meta          # sample metadata
all_genes <- dat$all_genes
dataset_ids <- dat$dataset_ids

# Exclude datasets that should not be used in this model-building workflow
excluded_datasets <- c("GSE15852", "GSE25066", "GSE9574")
external_test_ds <- "GSE29044"
expected_train_datasets <- c(
  "GSE10810", "GSE29431", "GSE42568", "GSE45827",
  "GSE54002", "GSE65194", "GSE7904"
)
expected_retained_datasets <- c(expected_train_datasets, external_test_ds)

keep_idx <- !(meta$dataset %in% excluded_datasets)
expr <- expr[, keep_idx, drop = FALSE]
meta <- meta[keep_idx, , drop = FALSE]

# Keep dataset_ids aligned if it stores one id per sample
if (length(dataset_ids) == length(keep_idx)) {
  dataset_ids <- dataset_ids[keep_idx]
}

dataset <- meta$dataset
y <- factor(meta$Status, levels = c("Normal", "Cancer"))
dataset_list <- sort(unique(as.character(dataset)))

stopifnot(
  setequal(unique(as.character(dataset)), expected_retained_datasets)
)

# Basic checks
dim(expr)
head(meta)
table(meta$dataset, meta$Status)
# =========================================================
# Define external test dataset
# =========================================================

external_idx <- dataset == external_test_ds
train_idx_all <- dataset != external_test_ds

expr_external <- expr[, external_idx, drop = FALSE]
meta_external <- meta[external_idx, , drop = FALSE]
expr <- expr[, train_idx_all, drop = FALSE]
meta <- meta[train_idx_all, , drop = FALSE]

# dataset and y
dataset <- meta$dataset
y <- factor(meta$Status, levels = c("Normal", "Cancer"))

dataset_list <- sort(unique(as.character(dataset)))

stopifnot(
  !any(dataset %in% c(excluded_datasets, external_test_ds)),
  !any(meta_external$dataset %in% excluded_datasets),
  ncol(expr_external) > 0,
  nrow(meta_external) > 0,
  all(meta_external$dataset == external_test_ds),
  setequal(unique(as.character(dataset)), expected_train_datasets)
)

cat("Excluded datasets:", paste(excluded_datasets, collapse = ", "), "\n")
cat("External test dataset:", external_test_ds, "\n")
cat("Training datasets:", paste(sort(unique(as.character(dataset))), collapse = ", "), "\n")
cat("External dataset distribution:\n")
print(table(meta_external$dataset, meta_external$Status))

all(colnames(expr) == rownames(meta))
```
```{r}
# Ensure sample order matches
meta <- meta[colnames(expr), ]
# Final safety check
sum(is.na(meta$Status))
# =========================================================
# Step 2: Align samples and check labels
# =========================================================

# Make sure meta rows match expression columns
meta <- meta[colnames(expr), ]

# Check whether sample order is correct
all(rownames(meta) == colnames(expr))

# Check label distribution
table(meta$Status)
dataset <- meta$dataset
# Check dataset distribution
table(meta$dataset, meta$Status)

# Convert label to factor
meta$Status <- factor(meta$Status, levels = c("Normal", "Cancer"))
y <- factor(meta$Status, levels = c("Normal", "Cancer"))



table(y)
```

```{r}
# =========================================================
# Step 3: LODO with fold-specific limma feature selection
# =========================================================
library(sva)
library(limma)

top_n <- 50
dataset_list <- sort(unique(as.character(dataset)))

lodo_gene_list <- list()

for (test_ds in dataset_list) {
  
  cat("\nLODO fold:", test_ds, "\n")
  
  train_idx <- dataset != test_ds
  
  expr_train <- expr[, train_idx, drop = FALSE]
  meta_train <- meta[train_idx, ]
  # -----------------------------
  # Batch correction on TRAIN fold only
  # -----------------------------
  batch_train <- factor(meta_train$dataset)
  
group <- factor(y[train_idx], levels = c("Normal", "Cancer"))


  mod <- model.matrix(~ group)
  
  expr_train <- ComBat(
    dat = as.matrix(expr_train),
    batch = batch_train,
    mod = mod,
    par.prior = TRUE
  )
  
  
  
  design <- model.matrix(~ group)
  
  fit <- lmFit(expr_train, design)
  fit <- eBayes(fit)
  
  pvals <- fit$p.value[, "groupCancer"]
  top_genes_fold <- names(sort(pvals))[1:top_n]
  
  lodo_gene_list[[test_ds]] <- top_genes_fold
  
  cat("Selected genes:", length(top_genes_fold), "\n")
}
```
```{r}
class(y)
levels(y)
table(y)
```

```{r}
# =========================================================
# Step 4: LODO predictions with fold-specific top genes
# =========================================================

library(glmnet)
library(randomForest)
library(e1071)
library(class)
library(sva)

model_cols <- c(
  "Ridge", "RF", "SVM_LINEAR", "Logistic", "ElasticNet",
  "LASSO", "SVM_RBF", "kNN", "NaiveBayes"
)

lodo_pred <- data.frame(
  sample_id = rownames(meta),
  dataset = dataset,
  truth = y
)

for (m in model_cols) {
  lodo_pred[[m]] <- NA_real_
}

for (test_ds in dataset_list) {
  
  cat("\nLODO test dataset:", test_ds, "\n")
  fold_seed <- seed_from_label(test_ds, 1000L)
  
  train_idx <- dataset != test_ds
  test_idx  <- dataset == test_ds
  
  # use fold-specific genes
  genes_fold <- lodo_gene_list[[test_ds]]
  
  X_train <- t(expr[genes_fold, train_idx, drop = FALSE])
  X_test  <- t(expr[genes_fold, test_idx, drop = FALSE])
  
  y_train <- y[train_idx]
  y_train_num <- ifelse(y_train == "Cancer", 1, 0)
  cv_spec <- make_cv_fold_spec(y_train, paste0("lodo_", test_ds))
  
  # -----------------------------
  # Ridge
  # -----------------------------
  set.seed(fold_seed + 1L)
  fit_ridge <- cv.glmnet(
    x = as.matrix(X_train),
    y = y_train_num,
    family = "binomial",
    alpha = 0,
    type.measure = "auc",
    foldid = cv_spec$foldid,
    nfolds = cv_spec$nfolds
  )
  
  lodo_pred$Ridge[test_idx] <- as.numeric(
    predict(fit_ridge, newx = as.matrix(X_test),
            s = "lambda.min", type = "response")
  )
  
  # -----------------------------
  # Random Forest
  # -----------------------------
  set.seed(fold_seed + 2L)
  fit_rf <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )
  
  lodo_pred$RF[test_idx] <- predict(
    fit_rf,
    newdata = X_test,
    type = "prob"
  )[,"Cancer"]
  
  # -----------------------------
  # SVM Linear
  # -----------------------------
  set.seed(fold_seed + 3L)
  fit_svm_linear <- svm(
    x = X_train,
    y = y_train,
    kernel = "linear",
    probability = TRUE,
    scale = TRUE
  )
  
  pred_linear <- predict(fit_svm_linear, X_test, probability = TRUE)
  lodo_pred$SVM_LINEAR[test_idx] <- attr(pred_linear, "probabilities")[,"Cancer"]
  
  # -----------------------------
  # Logistic Regression
  # -----------------------------
  df_train <- data.frame(y_train = y_train, X_train)
  df_test  <- data.frame(X_test)
  
  fit_logistic <- glm(
    y_train ~ .,
    data = df_train,
    family = binomial()
  )
  
  lodo_pred$Logistic[test_idx] <- predict(
    fit_logistic,
    newdata = df_test,
    type = "response"
  )
  
  # -----------------------------
  # Elastic Net
  # -----------------------------
  set.seed(fold_seed + 4L)
  fit_enet <- cv.glmnet(
    x = as.matrix(X_train),
    y = y_train_num,
    family = "binomial",
    alpha = 0.5,
    type.measure = "auc",
    foldid = cv_spec$foldid,
    nfolds = cv_spec$nfolds
  )
  
  lodo_pred$ElasticNet[test_idx] <- as.numeric(
    predict(fit_enet, newx = as.matrix(X_test),
            s = "lambda.min", type = "response")
  )
  
  # -----------------------------
  # LASSO
  # -----------------------------
  set.seed(fold_seed + 5L)
  fit_lasso <- cv.glmnet(
    x = as.matrix(X_train),
    y = y_train_num,
    family = "binomial",
    alpha = 1,
    type.measure = "auc",
    foldid = cv_spec$foldid,
    nfolds = cv_spec$nfolds
  )
  
  lodo_pred$LASSO[test_idx] <- as.numeric(
    predict(fit_lasso, newx = as.matrix(X_test),
            s = "lambda.min", type = "response")
  )
  
  # -----------------------------
  # SVM RBF
  # -----------------------------
  set.seed(fold_seed + 6L)
  fit_svm_rbf <- svm(
    x = X_train,
    y = y_train,
    kernel = "radial",
    probability = TRUE,
    scale = TRUE
  )
  
  pred_rbf <- predict(fit_svm_rbf, X_test, probability = TRUE)
  lodo_pred$SVM_RBF[test_idx] <- attr(pred_rbf, "probabilities")[,"Cancer"]
  
  # -----------------------------
  # kNN
  # -----------------------------
  pred_knn <- knn(
    train = X_train,
    test = X_test,
    cl = y_train,
    k = 5,
    prob = TRUE
  )
  
  prob_knn <- attr(pred_knn, "prob")
  prob_cancer <- ifelse(pred_knn == "Cancer", prob_knn, 1 - prob_knn)
  
  lodo_pred$kNN[test_idx] <- prob_cancer
  
  # -----------------------------
  # Naive Bayes
  # -----------------------------
  fit_nb <- naiveBayes(
    x = X_train,
    y = y_train
  )
  
  lodo_pred$NaiveBayes[test_idx] <- predict(
    fit_nb,
    newdata = X_test,
    type = "raw"
  )[,"Cancer"]
  
  cat("Finished:", test_ds, "\n")
}
```
```{r}
# =========================================================
# Step 6: Performance vs Redundancy plot
# =========================================================

library(pROC)
library(ggplot2)
library(ggrepel)
library(dplyr)

model_cols <- c(
  "Ridge", "RF", "SVM_LINEAR", "Logistic", "ElasticNet",
  "LASSO", "SVM_RBF", "kNN", "NaiveBayes"
)

# -----------------------------
# 1. Calculate LODO AUC
# -----------------------------
auc_tbl <- data.frame(
  model = model_cols,
  auc = NA_real_
)

for (m in model_cols) {
  
  roc_obj <- roc(
    response = lodo_pred$truth,
    predictor = lodo_pred[[m]],
    levels = c("Normal", "Cancer"),
    direction = "<"
  )
  
  auc_tbl$auc[auc_tbl$model == m] <- as.numeric(auc(roc_obj))
}

auc_tbl
# -----------------------------
# 2. Calculate mean absolute correlation
# -----------------------------

pred_mat <- lodo_pred[, model_cols]

cor_mat <- cor(pred_mat, use = "pairwise.complete.obs")

mean_abs_cor <- sapply(model_cols, function(m) {
  other_models <- setdiff(model_cols, m)
  mean(abs(cor_mat[m, other_models]), na.rm = TRUE)
})

cor_tbl <- data.frame(
  model = names(mean_abs_cor),
  mean_abs_cor = as.numeric(mean_abs_cor)
)

cor_tbl
# -----------------------------
# 3. Combine performance + redundancy
# -----------------------------

plot_tbl <- auc_tbl %>%
  left_join(cor_tbl, by = "model")

plot_tbl
# -----------------------------
# 4. Plot
# -----------------------------

ggplot(plot_tbl, aes(x = mean_abs_cor, y = auc, label = model)) +
  geom_point(size = 3, color = "grey20") +
  ggrepel::geom_text_repel(
    size = 4.5,
    max.overlaps = Inf,
    box.padding = 0.45,
    point.padding = 0.3,
    min.segment.length = 0,
    segment.color = "grey60",
    seed = 3888
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 16) +
  theme(
    plot.margin = margin(10, 80, 10, 10)
  ) +
  labs(
    title = "Performance vs Redundancy",
    subtitle = "Better candidates are high AUC and relatively low correlation",
    x = "Mean Absolute Correlation with Other Models",
    y = "Mean FULL-DATA LODO AUC"
  )
```

```{r}
# =========================================================
# Step 7: Prepare meta-model training data
# =========================================================

selected_models <- c("NaiveBayes", "ElasticNet", "SVM_RBF")

meta_X <- lodo_pred[, selected_models]
meta_y <- factor(lodo_pred$truth, levels = c("Normal", "Cancer"))

head(meta_X)
table(meta_y)
```
```{r}
# =========================================================
# Attention-based meta model using optim()
# No torch version
# =========================================================

# -----------------------------
# 1. Select final base models
# -----------------------------

selected_models <- c("NaiveBayes", "ElasticNet", "SVM_RBF")

meta_train_data <- lodo_pred[, c("sample_id", "dataset", "truth", selected_models)]

head(meta_train_data)
# -----------------------------
# 2. Helper function: clip probability
# -----------------------------

clip_prob <- function(p, eps = 1e-6) {
  pmin(pmax(as.numeric(p), eps), 1 - eps)
}
# -----------------------------
# 3. Prepare X and y
# -----------------------------

# Original probabilities for final weighted risk score
X_meta_train_prob <- as.matrix(meta_train_data[, selected_models])
X_meta_train_prob <- apply(X_meta_train_prob, 2, clip_prob)

# Logit-transformed probabilities for attention weight calculation
X_meta_train_att <- qlogis(X_meta_train_prob)

# Outcome: Cancer = 1, Normal = 0
y_meta_train <- as.numeric(meta_train_data$truth == "Cancer")

cat("\nX_meta_train_prob dimension:\n")
print(dim(X_meta_train_prob))

cat("\nX_meta_train_att dimension:\n")
print(dim(X_meta_train_att))

cat("\ny_meta_train distribution:\n")
print(table(y_meta_train))
# -----------------------------
# 4. Smoothed attention model
# -----------------------------

softmax_temp <- function(z, temperature = 1.5) {
  z <- z / temperature
  z <- z - max(z)
  exp_z <- exp(z)
  exp_z / sum(exp_z)
}

attention_predict <- function(params, X_att, X_prob, temperature = 1.5) {
  
  n_models <- ncol(X_att)
  
  W <- matrix(
    params[1:(n_models * n_models)],
    nrow = n_models,
    ncol = n_models
  )
  
  b <- params[
    (n_models * n_models + 1):(n_models * n_models + n_models)
  ]
  
  weights <- t(apply(X_att, 1, function(x) {
    logits <- as.numeric(x %*% W + b)
    softmax_temp(logits, temperature = temperature)
  }))
  
  final_score <- rowSums(weights * X_prob)
  
  list(
    weights = weights,
    final_score = clip_prob(final_score)
  )
}
# -----------------------------
# 5. Loss function
# -----------------------------

entropy_loss <- function(weights) {
  eps <- 1e-8
  entropy <- -rowSums(weights * log(weights + eps))
  -mean(entropy)
}

bce_loss_smoothed <- function(
    params,
    X_att,
    X_prob,
    y,
    temperature = 1.5,
    l2_lambda = 0.0005,
    entropy_lambda = 0.01
) {
  
  pred_obj <- attention_predict(
    params = params,
    X_att = X_att,
    X_prob = X_prob,
    temperature = temperature
  )
  
  pred <- clip_prob(pred_obj$final_score)
  weights <- pred_obj$weights
  
  bce <- -mean(
    y * log(pred) +
      (1 - y) * log(1 - pred)
  )
  
  l2_penalty <- l2_lambda * sum(params^2)
  entropy_penalty <- entropy_lambda * entropy_loss(weights)
  
  total_loss <- bce + l2_penalty + entropy_penalty
  
  total_loss
}
# -----------------------------
# 6. Train attention meta model
# -----------------------------

set.seed(seed_from_label("attention_meta_model", 4000L))

n_models <- ncol(X_meta_train_att)

temperature_use <- 1.5
l2_lambda_use <- 0.0005
entropy_lambda_use <- 0.01

init_params <- c(
  rnorm(n_models * n_models, mean = 0, sd = 0.001),
  rep(0, n_models)
)

fit_attention <- optim(
  par = init_params,
  fn = bce_loss_smoothed,
  X_att = X_meta_train_att,
  X_prob = X_meta_train_prob,
  y = y_meta_train,
  temperature = temperature_use,
  l2_lambda = l2_lambda_use,
  entropy_lambda = entropy_lambda_use,
  method = "BFGS",
  control = list(maxit = 1000)
)

fit_attention$value
fit_attention$convergence
# -----------------------------
# 7. Save attention meta model
# -----------------------------



attention_model <- list(
  params = fit_attention$par,
  selected_models = selected_models,
  temperature = temperature_use,
  l2_lambda = l2_lambda_use,
  entropy_lambda = entropy_lambda_use
)
dir.create("models", showWarnings = FALSE)

saveRDS(attention_model, "models/metamodel(attention).rds")

# -----------------------------
# 8. Get training predictions and attention weights
# -----------------------------

att_train_pred <- attention_predict(
  params = attention_model$params,
  X_att = X_meta_train_att,
  X_prob = X_meta_train_prob,
  temperature = attention_model$temperature
)

meta_pred <- att_train_pred$final_score
meta_weights <- att_train_pred$weights

colnames(meta_weights) <- selected_models

head(meta_pred)
head(meta_weights)

colMeans(meta_weights)
# -----------------------------
# 9. Plot average attention weights
# -----------------------------

library(ggplot2)

att_df <- data.frame(
  model = names(colMeans(meta_weights)),
  attention = as.numeric(colMeans(meta_weights))
)

ggplot(att_df, aes(x = reorder(model, attention), y = attention)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Average Attention Weight of Selected Base Models",
    x = "Base model",
    y = "Average attention weight"
  )
```
```{r}
# =========================================================
# Plot: Smoothed dynamic attention weights by truth
# =========================================================

library(dplyr)
library(tidyr)
library(ggplot2)

# meta_weights: attention weights matrix
# columns should be selected_models
# rows should match meta_train_data / lodo_pred samples

att_plot_df <- as.data.frame(meta_weights)

att_plot_df$sample_id <- meta_train_data$sample_id
att_plot_df$truth <- meta_train_data$truth

att_long <- att_plot_df %>%
  pivot_longer(
    cols = all_of(selected_models),
    names_to = "Base_model",
    values_to = "Attention_weight"
  )

ggplot(att_long, aes(
  x = Base_model,
  y = Attention_weight,
  fill = truth
)) +
  geom_boxplot(
    alpha = 0.75,
    outlier.size = 0.8,
    position = position_dodge(width = 0.8)
  ) +
  theme_minimal(base_size = 16) +
  labs(
    title = "Smoothed dynamic attention weights",
    x = "Base model",
    y = "Attention weight",
    fill = "truth"
  )
```
```{r}
library(limma)
library(sva)


# batch variable: 
batch <- factor(meta$dataset)

# biological label
group <- factor(meta$Status, levels = c("Normal", "Cancer"))

# model matrix
mod <- model.matrix(~ group)

# batch correction on TRAINING DATA ONLY
expr_precombat <- expr

expr <- ComBat(
  dat = as.matrix(expr),
  batch = batch,
  mod = mod,
  par.prior = TRUE
)

# FINAL GENE SELECTION
top_n <- 50

group <- factor(meta$Status, levels = c("Normal", "Cancer"))
design <- model.matrix(~ group)

fit <- lmFit(expr, design)
fit <- eBayes(fit)

top_table <- topTable(
  fit,
  coef = "groupCancer",
  number = Inf,
  adjust.method = "BH"
)
final_top_genes <- rownames(top_table)[1:top_n]

length(final_top_genes)
head(final_top_genes, 20)

dir.create("models", showWarnings = FALSE)

saveRDS(final_top_genes, "models/final_top_genes.rds")

# Load final genes
final_top_genes <- readRDS("models/final_top_genes.rds")

# Expression matrix for final model
X_full <- t(expr[final_top_genes, , drop = FALSE])

# Label
y_full <- factor(meta$Status, levels = c("Normal", "Cancer"))
y_full_num <- ifelse(y_full == "Cancer", 1, 0)

# Check
dim(X_full)
table(y_full)

if (exists("expr_precombat")) {
  gene_var_precombat <- apply(expr_precombat, 1, var, na.rm = TRUE)
  keep_genes_precombat <- gene_var_precombat > 0
  
  if (sum(keep_genes_precombat) >= 5 && ncol(expr_precombat) >= 2) {
    top_pca_genes_precombat <- names(sort(
      gene_var_precombat[keep_genes_precombat],
      decreasing = TRUE
    ))[
      1:min(2000, sum(keep_genes_precombat))
    ]
    
    expr_pca_precombat <- expr_precombat[top_pca_genes_precombat, , drop = FALSE]
    
    if (anyNA(expr_pca_precombat)) {
      row_means_precombat <- rowMeans(expr_pca_precombat, na.rm = TRUE)
      expr_pca_precombat[is.na(expr_pca_precombat)] <-
        row_means_precombat[row(expr_pca_precombat)[is.na(expr_pca_precombat)]]
    }
    
    pca_precombat <- prcomp(t(expr_pca_precombat), scale. = FALSE)
    var_exp_precombat <- pca_precombat$sdev^2 / sum(pca_precombat$sdev^2)
    
    pca_df_precombat <- data.frame(
      PC1 = pca_precombat$x[, 1],
      PC2 = pca_precombat$x[, 2],
      dataset = meta$dataset,
      Status = meta$Status,
      sample_id = rownames(meta)
    )
    
    ggplot2::ggplot(
      pca_df_precombat,
      ggplot2::aes(x = PC1, y = PC2, color = dataset, shape = Status)
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.85) +
      ggplot2::theme_bw(base_size = 13) +
      ggplot2::labs(
        title = "PCA of merged training set before ComBat",
        subtitle = paste0("External test dataset excluded: ", external_test_ds),
        x = paste0("PC1 (", round(var_exp_precombat[1] * 100, 1), "%)"),
        y = paste0("PC2 (", round(var_exp_precombat[2] * 100, 1), "%)"),
        color = "Dataset",
        shape = "Status"
      )
  } else {
    message("Skipping pre-ComBat PCA plot: insufficient genes or samples.")
  }
}

# PCA of merged training set after ComBat
gene_var <- apply(expr, 1, var, na.rm = TRUE)
keep_genes <- gene_var > 0

if (sum(keep_genes) >= 5 && ncol(expr) >= 2) {
  top_pca_genes <- names(sort(gene_var[keep_genes], decreasing = TRUE))[
    1:min(2000, sum(keep_genes))
  ]
  
  expr_pca <- expr[top_pca_genes, , drop = FALSE]
  
  if (anyNA(expr_pca)) {
    row_means <- rowMeans(expr_pca, na.rm = TRUE)
    expr_pca[is.na(expr_pca)] <- row_means[row(expr_pca)[is.na(expr_pca)]]
  }
  
  pca <- prcomp(t(expr_pca), scale. = FALSE)
  var_exp <- pca$sdev^2 / sum(pca$sdev^2)
  
  pca_df <- data.frame(
    PC1 = pca$x[, 1],
    PC2 = pca$x[, 2],
    dataset = meta$dataset,
    Status = meta$Status,
    sample_id = rownames(meta)
  )
  
  ggplot2::ggplot(
    pca_df,
    ggplot2::aes(x = PC1, y = PC2, color = dataset, shape = Status)
  ) +
    ggplot2::geom_point(size = 3, alpha = 0.85) +
    ggplot2::theme_bw(base_size = 13) +
    ggplot2::labs(
      title = "PCA of merged training set after ComBat",
      subtitle = paste0("External test dataset excluded: ", external_test_ds),
      x = paste0("PC1 (", round(var_exp[1] * 100, 1), "%)"),
      y = paste0("PC2 (", round(var_exp[2] * 100, 1), "%)"),
      color = "Dataset",
      shape = "Status"
    )
} else {
  message("Skipping PCA plot: insufficient genes or samples.")
}
```
# Ensemble train(base models)
```{r}
library(e1071)
library(randomForest)
library(glmnet)
#train base model 
# Naive Bayes
fit_nb_full <- naiveBayes(
  x = X_full,
  y = y_full
)

# Elastic Net
set.seed(seed_from_label("full_elasticnet", 5000L))
full_cv_spec <- make_cv_fold_spec(y_full, "full_elasticnet")
fit_enet_full <- cv.glmnet(
  x = as.matrix(X_full),
  y = y_full_num,
  family = "binomial",
  alpha = 0.5,
  type.measure = "auc",
  foldid = full_cv_spec$foldid,
  nfolds = full_cv_spec$nfolds
)

# SVM RBF
set.seed(seed_from_label("full_svm_rbf", 5001L))
fit_svm_rbf_full <- svm(
  x = X_full,
  y = y_full,
  kernel = "radial",
  probability = TRUE,
  scale = TRUE
)

full_pred <- data.frame(
  NaiveBayes = predict(
    fit_nb_full,
    newdata = X_full,
    type = "raw"
  )[,"Cancer"],
  
  ElasticNet = as.numeric(
    predict(
      fit_enet_full,
      newx = as.matrix(X_full),
      s = "lambda.min",
      type = "response"
    )
  ),
  
  SVM_RBF = attr(
    predict(fit_svm_rbf_full, X_full, probability = TRUE),
    "probabilities"
  )[,"Cancer"]
)
head(full_pred)
summary(full_pred)
saveRDS(fit_nb_full, "models/nb_full.rds")
saveRDS(fit_enet_full, "models/elasticnet_full.rds")
saveRDS(fit_svm_rbf_full, "models/svm_rbf_full.rds")
   
```
# Ensemble Testing(main model)
```{r}
# =========================================================
# External test on the held-out dataset
# =========================================================

library(pROC)
library(caret)
library(ggplot2)

# 1. Prepare external test matrix
X_external <- t(expr_external[final_top_genes, , drop = FALSE])

y_external <- factor(meta_external$Status, levels = c("Normal", "Cancer"))

dim(X_external)
table(y_external)
# 2. Base model predictions on the held-out dataset

pred_svm_rbf_external <- predict(
  fit_svm_rbf_full,
  X_external,
  probability = TRUE
)

external_pred <- data.frame(
  NaiveBayes = predict(
    fit_nb_full,
    newdata = X_external,
    type = "raw"
  )[,"Cancer"],
  
  ElasticNet = as.numeric(
    predict(
      fit_enet_full,
      newx = as.matrix(X_external),
      s = "lambda.min",
      type = "response"
    )
  ),
  
  SVM_RBF = attr(pred_svm_rbf_external, "probabilities")[,"Cancer"]
)

head(external_pred)
summary(external_pred)
# 3. Apply LODO-trained attention meta model

attention_model <- readRDS("models/metamodel(attention).rds")

X_external_prob <- as.matrix(external_pred[, attention_model$selected_models])
X_external_prob <- apply(X_external_prob, 2, clip_prob)

X_external_att <- qlogis(X_external_prob)

att_external <- attention_predict(
  params = attention_model$params,
  X_att = X_external_att,
  X_prob = X_external_prob,
  temperature = attention_model$temperature
)

external_risk_score <- att_external$final_score
external_weights <- att_external$weights

colnames(external_weights) <- attention_model$selected_models

head(external_risk_score)
colMeans(external_weights)
# 4. Evaluate performance
eps <- 1e-6

external_logit_score <- qlogis(
  pmin(pmax(external_risk_score, eps), 1 - eps)
)

summary(external_logit_score)

roc_obj <- roc(
  response = y_external,
  predictor = external_risk_score,
  levels = c("Normal", "Cancer"),
  direction = "<"
)

external_auc <- as.numeric(auc(roc_obj))
external_auc

# threshold = 0.5
external_pred_class <- ifelse(external_risk_score >= 0.5, "Cancer", "Normal")
external_pred_class <- factor(external_pred_class, levels = c("Normal", "Cancer"))

cm <- confusionMatrix(
  data = external_pred_class,
  reference = y_external,
  positive = "Cancer"
)

cm

# Extract key metrics
external_results <- data.frame(
  AUC = external_auc,
  Accuracy = cm$overall["Accuracy"],
  BalancedAccuracy = cm$byClass["Balanced Accuracy"],
  Sensitivity = cm$byClass["Sensitivity"],
  Specificity = cm$byClass["Specificity"],
  Precision = cm$byClass["Precision"],
  F1 = cm$byClass["F1"]
)

external_results

# 5. Plot ROC curve

plot(
  roc_obj,
  main = paste0("External Test ROC: ", external_test_ds, ", AUC = ", round(external_auc, 3))
)

# 6. Risk score distribution plot

plot_df <- data.frame(
  risk_score = external_logit_score,
  truth = y_external
)

ggplot(plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.45) +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("External Test Risk Score Distribution: ", external_test_ds),
    x = "Predicted cancer risk score",
    y = "Density",
    fill = "True label"
  )
ggplot(plot_df, aes(x = risk_score, fill = truth)) +
  geom_histogram(
    bins = 30,
    alpha = 0.5,
    position = "identity"
  ) +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("External Test Risk Score Distribution: ", external_test_ds),
    x = "Predicted risk score",
    y = "Count",
    fill = "True label"
  )
```
# Single best testing
```{r}
X_external <- t(expr_external[final_top_genes, , drop = FALSE])

y_external <- factor(meta_external$Status, levels = c("Normal", "Cancer"))

dim(X_external)
table(y_external)
enet_pred <- as.numeric(
  predict(
    fit_enet_full,
    newx = as.matrix(X_external),
    s = "lambda.min",
    type = "response"
  )
)

head(enet_pred)
summary(enet_pred)
eps <- 1e-6

enet_logit_score <- qlogis(
  pmin(pmax(enet_pred, eps), 1 - eps)
)

summary(enet_logit_score)
library(pROC)

roc_obj <- roc(
  response = y_external,
  predictor = enet_pred,
  levels = c("Normal", "Cancer"),
  direction = "<"
)

enet_auc <- as.numeric(auc(roc_obj))
enet_auc
library(caret)

enet_class <- ifelse(enet_pred >= 0.5, "Cancer", "Normal")
enet_class <- factor(enet_class, levels = c("Normal", "Cancer"))

cm <- confusionMatrix(
  data = enet_class,
  reference = y_external,
  positive = "Cancer"
)

cm
enet_results <- data.frame(
  AUC = enet_auc,
  Accuracy = cm$overall["Accuracy"],
  BalancedAccuracy = cm$byClass["Balanced Accuracy"],
  Sensitivity = cm$byClass["Sensitivity"],
  Specificity = cm$byClass["Specificity"],
  Precision = cm$byClass["Precision"],
  F1 = cm$byClass["F1"]
)

enet_results
library(ggplot2)

plot_df <- data.frame(
  risk_score = enet_logit_score,
  truth = y_external
)

ggplot(plot_df, aes(x = risk_score, fill = truth)) +
  geom_density(alpha = 0.45) +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("Elastic Net Risk Score Distribution (External Test ", external_test_ds, ")"),
    x = "Predicted risk score",
    y = "Density",
    fill = "True label"
  )
ggplot(plot_df, aes(x = risk_score, fill = truth)) +
  geom_histogram(
    bins = 30,
    alpha = 0.5,
    position = "identity"
  ) +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("Elastic Net Risk Score Distribution (External Test ", external_test_ds, ")"),
    x = "Predicted risk score",
    y = "Count",
    fill = "True label"
  )
```

# BOOTSTRAP
```{r}
allocate_bootstrap_counts <- function(weights, total_n) {
  weight_names <- names(weights)
  weights_num <- as.numeric(weights)
  weights_num <- weights_num / sum(weights_num)
  
  raw_counts <- weights_num * total_n
  counts <- floor(raw_counts)
  remainder <- total_n - sum(counts)
  
  if (remainder > 0L) {
    add_idx <- order(raw_counts - counts, decreasing = TRUE)[seq_len(remainder)]
    counts[add_idx] <- counts[add_idx] + 1L
  }
  
  stats::setNames(as.integer(counts), weight_names)
}

make_bootstrap_indices <- function(dataset, y) {
  dataset_chr <- as.character(dataset)
  y_chr <- as.character(y)
  dataset_counts <- table(dataset_chr)
  
  total_boot_n <- 2L * floor(length(y_chr) / 2L)
  n_per_class <- total_boot_n / 2L
  dataset_target_per_class <- allocate_bootstrap_counts(dataset_counts, n_per_class)
  
  boot_idx <- unlist(lapply(names(dataset_target_per_class), function(ds) {
    ds_target_n <- dataset_target_per_class[[ds]]
    cancer_idx <- which(dataset_chr == ds & y_chr == "Cancer")
    normal_idx <- which(dataset_chr == ds & y_chr == "Normal")
    
    if (length(cancer_idx) == 0L || length(normal_idx) == 0L) {
      stop("Balanced bootstrap requires both classes in dataset: ", ds)
    }
    
    c(
      sample(cancer_idx, ds_target_n, replace = TRUE),
      sample(normal_idx, ds_target_n, replace = TRUE)
    )
  }), use.names = FALSE)
  
  boot_idx
}
```

```{r}
set.seed(seed_from_label("bootstrap_preview", 6000L))
boot_idx <- make_bootstrap_indices(dataset, y)

table(dataset, y)

table(dataset[boot_idx], y[boot_idx])
```
```{r}
B <- 100

X_external <- t(expr_external[final_top_genes, , drop = FALSE])

bootstrap_scores <- matrix(
  NA_real_,
  nrow = ncol(expr_external),
  ncol = B
)

rownames(bootstrap_scores) <- colnames(expr_external)

for (b in 1:B) {
  
  cat("Bootstrap:", b, "\n")
  boot_seed <- seed_from_label(paste0("bootstrap_", b), 7000L)
  
  set.seed(boot_seed)
  boot_idx <- make_bootstrap_indices(dataset, y)
  
  X_boot <- t(expr[final_top_genes, boot_idx, drop = FALSE])
  y_boot <- y[boot_idx]
  y_boot_num <- ifelse(y_boot == "Cancer", 1, 0)
  
  fit_nb_boot <- naiveBayes(
    x = X_boot,
    y = y_boot
  )
  
  set.seed(boot_seed + 1L)
  boot_cv_spec <- make_cv_fold_spec(y_boot, paste0("bootstrap_", b, "_elasticnet"))
  fit_enet_boot <- cv.glmnet(
    x = as.matrix(X_boot),
    y = y_boot_num,
    family = "binomial",
    alpha = 0.5,
    type.measure = "auc",
    foldid = boot_cv_spec$foldid,
    nfolds = boot_cv_spec$nfolds
  )
  
  set.seed(boot_seed + 2L)
  fit_svm_rbf_boot <- svm(
    x = X_boot,
    y = y_boot,
    kernel = "radial",
    probability = TRUE,
    scale = TRUE
  )
  
  # save models
  boot_dir <- file.path("models", sprintf("bootstrap_%03d", b))
  dir.create(boot_dir, recursive = TRUE, showWarnings = FALSE)
  
  saveRDS(fit_nb_boot, file.path(boot_dir, "nb.rds"))
  saveRDS(fit_enet_boot, file.path(boot_dir, "elasticnet.rds"))
  saveRDS(fit_svm_rbf_boot, file.path(boot_dir, "svm_rbf.rds"))
  
  # predict external test
  pred_svm_rbf_boot <- predict(
    fit_svm_rbf_boot,
    X_external,
    probability = TRUE
  )
  
  pred_ext_boot <- data.frame(
    NaiveBayes = predict(fit_nb_boot, X_external, type = "raw")[,"Cancer"],
    ElasticNet = as.numeric(
      predict(
        fit_enet_boot,
        newx = as.matrix(X_external),
        s = "lambda.min",
        type = "response"
      )
    ),
    SVM_RBF = attr(pred_svm_rbf_boot, "probabilities")[,"Cancer"]
  )
  
  X_prob_boot <- as.matrix(pred_ext_boot[, attention_model$selected_models])
  X_prob_boot <- apply(X_prob_boot, 2, clip_prob)
  X_att_boot <- qlogis(X_prob_boot)
  
  att_boot <- attention_predict(
    params = attention_model$params,
    X_att = X_att_boot,
    X_prob = X_prob_boot,
    temperature = attention_model$temperature
  )
  
  bootstrap_scores[, b] <- att_boot$final_score
}
```

```{r}
bootstrap_ci_level <- 0.80
bootstrap_alpha <- 1 - bootstrap_ci_level
bootstrap_ci_label <- paste0(round(bootstrap_ci_level * 100), "% bootstrap interval")

bootstrap_summary <- data.frame(
  sample_id = rownames(bootstrap_scores),
  truth = y_external,
  risk_mean = rowMeans(bootstrap_scores, na.rm = TRUE),
  risk_lower = apply(
    bootstrap_scores, 1, quantile,
    probs = bootstrap_alpha / 2, na.rm = TRUE
  ),
  risk_upper = apply(
    bootstrap_scores, 1, quantile,
    probs = 1 - bootstrap_alpha / 2, na.rm = TRUE
  )
)

head(bootstrap_summary)
```
```{r}
library(ggplot2)
library(dplyr)

plot_boot <- bootstrap_summary %>%
  mutate(
    sample_order = reorder(sample_id, risk_mean),
    truth = factor(truth, levels = c("Normal", "Cancer"))
  )

ggplot(plot_boot, aes(x = sample_order, y = risk_mean, color = truth)) +
  geom_point(size = 2.2) +
  geom_errorbar(
    aes(ymin = risk_lower, ymax = risk_upper),
    width = 0.25,
    alpha = 0.7
  ) +
  geom_hline(
    yintercept = 0.5,
    linetype = "dashed"
  ) +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = paste0("Bootstrap Risk Score with ", bootstrap_ci_label, " on ", external_test_ds),
    x = "External test samples",
    y = "Predicted cancer risk score",
    color = "True label"
  )
ggplot(plot_boot, aes(x = risk_mean, fill = truth)) +
  geom_density(alpha = 0.45) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("Bootstrap Mean Risk Score Distribution on ", external_test_ds),
    x = "Bootstrap mean cancer risk score",
    y = "Density",
    fill = "True label"
  )

```
```{r}
# example for
uncertain_samples <- bootstrap_summary %>%
  mutate(distance_to_05 = abs(risk_mean - 0.5)) %>%
  arrange(distance_to_05)

head(uncertain_samples, 10)
sample_id_use <- uncertain_samples$sample_id[1]
sample_idx <- which(rownames(bootstrap_scores) == sample_id_use)

main_risk <- external_risk_score[sample_idx]

boot_risks <- bootstrap_scores[sample_idx, ]

ci_low <- quantile(boot_risks, bootstrap_alpha / 2, na.rm = TRUE)
ci_high <- quantile(boot_risks, 1 - bootstrap_alpha / 2, na.rm = TRUE)

main_risk
ci_low
ci_high
library(ggplot2)

single_df <- data.frame(
  risk = as.numeric(boot_risks)
)

ggplot(single_df, aes(x = risk)) +
  geom_histogram(bins = 25, alpha = 0.7) +
  geom_vline(xintercept = main_risk, linewidth = 1.2) +
  geom_vline(xintercept = ci_low, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = ci_high, linetype = "dashed", linewidth = 1) +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("Prediction uncertainty for sample: ", sample_id_use),
    subtitle = paste0(
      "Main risk = ", round(main_risk, 3),
      " | ", bootstrap_ci_label, " = [", round(ci_low, 3), ", ", round(ci_high, 3), "]"
    ),
    x = "Predicted cancer risk score",
    y = "Bootstrap model count"
  )
```

# Elastic Net only bootstrap uncertainty
```{r}
elastic_bootstrap_scores <- matrix(
  NA_real_,
  nrow = nrow(X_external),
  ncol = B
)

rownames(elastic_bootstrap_scores) <- rownames(X_external)

for (b in 1:B) {
  enet_file <- file.path(
    "models",
    sprintf("bootstrap_%03d", b),
    "elasticnet.rds"
  )
  
  stopifnot(file.exists(enet_file))
  
  fit_enet_boot <- readRDS(enet_file)
  
  elastic_bootstrap_scores[, b] <- as.numeric(
    predict(
      fit_enet_boot,
      newx = as.matrix(X_external),
      s = "lambda.min",
      type = "response"
    )
  )
}

dim(elastic_bootstrap_scores)
head(elastic_bootstrap_scores[, 1:min(5, ncol(elastic_bootstrap_scores)), drop = FALSE])
```

```{r}
elastic_bootstrap_summary <- data.frame(
  sample_id = rownames(elastic_bootstrap_scores),
  truth = y_external,
  risk_mean = rowMeans(elastic_bootstrap_scores, na.rm = TRUE),
  risk_lower = apply(
    elastic_bootstrap_scores, 1, quantile,
    probs = bootstrap_alpha / 2, na.rm = TRUE
  ),
  risk_upper = apply(
    elastic_bootstrap_scores, 1, quantile,
    probs = 1 - bootstrap_alpha / 2, na.rm = TRUE
  )
)

head(elastic_bootstrap_summary)
```

```{r}
plot_elastic_boot <- elastic_bootstrap_summary %>%
  mutate(
    sample_order = reorder(sample_id, risk_mean),
    truth = factor(truth, levels = c("Normal", "Cancer"))
  )

ggplot(plot_elastic_boot, aes(x = sample_order, y = risk_mean, color = truth)) +
  geom_point(size = 2.2) +
  geom_errorbar(
    aes(ymin = risk_lower, ymax = risk_upper),
    width = 0.25,
    alpha = 0.7
  ) +
  geom_hline(
    yintercept = 0.5,
    linetype = "dashed"
  ) +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = paste0(
      "Elastic Net Bootstrap Risk Score with ",
      bootstrap_ci_label,
      " on ",
      external_test_ds
    ),
    x = "External test samples",
    y = "Predicted cancer risk score",
    color = "True label"
  )

ggplot(plot_elastic_boot, aes(x = risk_mean, fill = truth)) +
  geom_density(alpha = 0.45) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("Elastic Net Bootstrap Mean Risk Score Distribution on ", external_test_ds),
    x = "Bootstrap mean cancer risk score",
    y = "Density",
    fill = "True label"
  )
```

```{r}
elastic_uncertain_samples <- elastic_bootstrap_summary %>%
  mutate(distance_to_05 = abs(risk_mean - 0.5)) %>%
  arrange(distance_to_05)

head(elastic_uncertain_samples, 10)
elastic_sample_id_use <- elastic_uncertain_samples$sample_id[1]
elastic_sample_idx <- which(rownames(elastic_bootstrap_scores) == elastic_sample_id_use)

elastic_main_risk <- enet_pred[elastic_sample_idx]
elastic_boot_risks <- elastic_bootstrap_scores[elastic_sample_idx, ]

elastic_ci_low <- quantile(elastic_boot_risks, bootstrap_alpha / 2, na.rm = TRUE)
elastic_ci_high <- quantile(elastic_boot_risks, 1 - bootstrap_alpha / 2, na.rm = TRUE)

elastic_single_df <- data.frame(
  risk = as.numeric(elastic_boot_risks)
)

ggplot(elastic_single_df, aes(x = risk)) +
  geom_histogram(bins = 25, alpha = 0.7) +
  geom_vline(xintercept = elastic_main_risk, linewidth = 1.2) +
  geom_vline(xintercept = elastic_ci_low, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = elastic_ci_high, linetype = "dashed", linewidth = 1) +
  theme_minimal(base_size = 15) +
  labs(
    title = paste0("Elastic Net prediction uncertainty for sample: ", elastic_sample_id_use),
    subtitle = paste0(
      "Main risk = ", round(elastic_main_risk, 3),
      " | ", bootstrap_ci_label, " = [",
      round(elastic_ci_low, 3), ", ", round(elastic_ci_high, 3), "]"
    ),
    x = "Predicted cancer risk score",
    y = "Bootstrap model count"
  )
```

# Uncertainty for Metrics

The section below keeps the final Elastic Net model fixed and quantifies how uncertain its external-test performance metrics are under 100 class-stratified bootstrap resamples of the held-out dataset.

```{r}
make_external_metric_boot_idx <- function(y) {
  y_chr <- as.character(y)
  
  unlist(lapply(levels(y), function(cls) {
    cls_idx <- which(y_chr == cls)
    sample(cls_idx, length(cls_idx), replace = TRUE)
  }), use.names = FALSE)
}

safe_divide <- function(num, den) {
  if (den == 0) {
    return(NA_real_)
  }
  
  num / den
}

calc_binary_metric_set <- function(truth, score, threshold = 0.5, eps = 1e-6) {
  truth <- factor(truth, levels = c("Normal", "Cancer"))
  truth_chr <- as.character(truth)
  truth_num <- ifelse(truth_chr == "Cancer", 1, 0)
  score_clip <- pmin(pmax(score, eps), 1 - eps)
  
  pred_chr <- ifelse(score >= threshold, "Cancer", "Normal")
  
  tp <- sum(pred_chr == "Cancer" & truth_chr == "Cancer")
  tn <- sum(pred_chr == "Normal" & truth_chr == "Normal")
  fp <- sum(pred_chr == "Cancer" & truth_chr == "Normal")
  fn <- sum(pred_chr == "Normal" & truth_chr == "Cancer")
  
  sensitivity <- safe_divide(tp, tp + fn)
  specificity <- safe_divide(tn, tn + fp)
  precision <- safe_divide(tp, tp + fp)
  npv <- safe_divide(tn, tn + fn)
  
  f1 <- if (is.na(precision) || is.na(sensitivity) || (precision + sensitivity) == 0) {
    NA_real_
  } else {
    2 * precision * sensitivity / (precision + sensitivity)
  }
  
  mcc_denom <- sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
  
  auc_value <- tryCatch(
    as.numeric(
      pROC::auc(
        pROC::roc(
          response = truth,
          predictor = score,
          levels = c("Normal", "Cancer"),
          direction = "<",
          quiet = TRUE
        )
      )
    ),
    error = function(e) NA_real_
  )
  
  c(
    AUC = auc_value,
    Accuracy = (tp + tn) / length(truth_chr),
    BalancedAccuracy = mean(c(sensitivity, specificity), na.rm = TRUE),
    Sensitivity = sensitivity,
    Specificity = specificity,
    Precision = precision,
    NPV = npv,
    F1 = f1,
    MCC = safe_divide((tp * tn) - (fp * fn), mcc_denom),
    BrierScore = mean((score - truth_num) ^ 2)
  )
}
```

```{r}
metric_boot_ci_level <- 0.95
metric_boot_alpha <- 1 - metric_boot_ci_level
metric_boot_ci_label <- paste0(round(metric_boot_ci_level * 100), "% percentile interval")
metric_display_order <- c(
  "AUC", "Accuracy", "BalancedAccuracy", "Sensitivity", "Specificity",
  "Precision", "F1"
)
metric_labels <- c(
  AUC = "AUC",
  Accuracy = "Accuracy",
  BalancedAccuracy = "Balanced accuracy",
  Sensitivity = "Sensitivity",
  Specificity = "Specificity",
  Precision = "Precision",
  F1 = "F1 score"
)
metric_direction <- c(
  AUC = "Higher is better",
  Accuracy = "Higher is better",
  BalancedAccuracy = "Higher is better",
  Sensitivity = "Higher is better",
  Specificity = "Higher is better",
  Precision = "Higher is better",
  F1 = "Higher is better"
)

elastic_metric_point <- calc_binary_metric_set(
  truth = y_external,
  score = enet_pred,
  threshold = 0.5
)

elastic_metric_boot <- do.call(
  rbind,
  lapply(seq_len(B), function(b) {
    set.seed(seed_from_label(paste0("external_metric_bootstrap_", b), 9000L))
    boot_idx <- make_external_metric_boot_idx(y_external)
    
    calc_binary_metric_set(
      truth = y_external[boot_idx],
      score = enet_pred[boot_idx],
      threshold = 0.5
    )
  })
)

elastic_metric_boot <- as.data.frame(elastic_metric_boot)

elastic_metric_summary <- bind_rows(lapply(metric_display_order, function(metric_name) {
  metric_values <- elastic_metric_boot[[metric_name]]
  
  data.frame(
    metric = metric_name,
    metric_label = unname(metric_labels[metric_name]),
    direction = unname(metric_direction[metric_name]),
    point_estimate = as.numeric(elastic_metric_point[metric_name]),
    bootstrap_mean = mean(metric_values, na.rm = TRUE),
    bootstrap_sd = stats::sd(metric_values, na.rm = TRUE),
    ci_lower = as.numeric(
      stats::quantile(metric_values, probs = metric_boot_alpha / 2, na.rm = TRUE)
    ),
    ci_upper = as.numeric(
      stats::quantile(metric_values, probs = 1 - metric_boot_alpha / 2, na.rm = TRUE)
    )
  )
})) %>%
  mutate(
    metric_label = factor(
      metric_label,
      levels = unname(metric_labels[metric_display_order])
    )
  )

elastic_metric_summary

dir.create("results", showWarnings = FALSE)
elastic_metric_summary_rds <- file.path(
  "results",
  paste0("elastic_metric_summary_", external_test_ds, ".rds")
)

saveRDS(
  elastic_metric_summary,
  file = elastic_metric_summary_rds
)

message("Saved Elastic Net metric summary to: ", elastic_metric_summary_rds)
```

```{r}
elastic_metric_boot_long <- stack(elastic_metric_boot[, metric_display_order, drop = FALSE])
colnames(elastic_metric_boot_long) <- c("value", "metric")

elastic_metric_boot_long <- elastic_metric_boot_long %>%
  mutate(
    metric = as.character(metric),
    metric_label = factor(
      unname(metric_labels[metric]),
      levels = unname(metric_labels[metric_display_order])
    )
  ) %>%
  left_join(
    elastic_metric_summary %>%
      mutate(metric = as.character(metric)) %>%
      select(metric, point_estimate),
    by = "metric"
  )

ggplot(elastic_metric_summary, aes(x = metric_label, y = point_estimate)) +
  geom_linerange(
    aes(ymin = ci_lower, ymax = ci_upper),
    linewidth = 1.1,
    color = "#4C78A8"
  ) +
  geom_point(size = 2.8, color = "#E45756") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = paste0("Elastic Net Metric Uncertainty on ", external_test_ds),
    subtitle = paste0(
      B, " class-stratified bootstrap resamples of the external test dataset (",
      metric_boot_ci_label, ")"
    ),
    x = NULL,
    y = "Metric value"
  )

ggplot(elastic_metric_boot_long, aes(x = value)) +
  geom_histogram(
    bins = 18,
    fill = "#72B7B2",
    color = "white",
    alpha = 0.9
  ) +
  geom_vline(
    aes(xintercept = point_estimate),
    color = "#E45756",
    linewidth = 0.9
  ) +
  facet_wrap(~ metric_label, scales = "free", ncol = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = paste0("Bootstrap Distributions of Elastic Net Metrics on ", external_test_ds),
    subtitle = "Red vertical line shows the observed metric on the full external dataset",
    x = "Metric value",
    y = "Bootstrap count"
  )
```

```{r}
# Additional visualisation: top 5 models by mean LODO AUC
top5_auc_lodo <- auc_tbl %>%
  arrange(desc(auc)) %>%
  slice_head(n = 5) %>%
  mutate(
    model = factor(model, levels = rev(model)),
    group = ifelse(model == "ElasticNet", "Elastic Net", "Other models"),
    auc_label = sprintf("%.3f", auc)
  )

ggplot(top5_auc_lodo, aes(x = model, y = auc, fill = group)) +
  geom_col(width = 0.72, alpha = 0.95) +
  geom_text(
    aes(label = auc_label),
    hjust = -0.05,
    size = 4.1,
    fontface = "bold",
    color = "grey15"
  ) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c("Elastic Net" = "#D55E00", "Other models" = "#4C78A8")) +
  scale_y_continuous(
    limits = c(0, min(1.05, max(top5_auc_lodo$auc, na.rm = TRUE) + 0.10)),
    expand = expansion(mult = c(0, 0.02))
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 28, 5.5, 5.5),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30")
  ) +
  labs(
    title = "Top 5 Models by Mean LODO AUC",
    subtitle = "This barplot uses the same previously computed LODO AUC results as the scatter plot above",
    x = NULL,
    y = "Mean FULL-DATA LODO AUC"
  )
```

```{r}
# Additional visualisation: focused bootstrap metrics
metric_focus <- c(
  F1 = "F1 score",
  Sensitivity = "Sensitivity",
  Specificity = "Specificity",
  BalancedAccuracy = "Balanced accuracy",
  AUC = "AUC value"
)
metric_focus_pal <- c(
  "F1 score" = "#E76F51",
  "Sensitivity" = "#F4A261",
  "Specificity" = "#2A9D8F",
  "Balanced accuracy" = "#4C78A8",
  "AUC value" = "#264653"
)

metric_focus_summary <- elastic_metric_summary %>%
  filter(metric %in% names(metric_focus)) %>%
  transmute(
    metric = as.character(metric),
    metric_label = unname(metric_focus[metric]),
    point_estimate = point_estimate,
    ci_lower = ci_lower,
    ci_upper = ci_upper
  ) %>%
  mutate(
    metric_label = factor(metric_label, levels = rev(unname(metric_focus))),
    point_label = sprintf("%.3f", point_estimate)
  )

metric_focus_long <- elastic_metric_boot_long %>%
  filter(metric %in% names(metric_focus)) %>%
  mutate(
    metric = as.character(metric),
    metric_label = factor(
      unname(metric_focus[metric]),
      levels = rev(unname(metric_focus))
    )
  )

metric_x_min <- max(0, min(metric_focus_summary$ci_lower, na.rm = TRUE) - 0.05)
metric_x_max <- min(1.05, max(metric_focus_summary$ci_upper, na.rm = TRUE) + 0.10)

ggplot(metric_focus_summary, aes(x = point_estimate, y = metric_label, color = metric_label)) +
  geom_segment(
    aes(x = ci_lower, xend = ci_upper, yend = metric_label),
    linewidth = 4,
    alpha = 0.28,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(size = 4, show.legend = FALSE) +
  geom_text(
    aes(label = point_label),
    nudge_x = 0.02,
    hjust = 0,
    size = 4,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = metric_focus_pal) +
  coord_cartesian(xlim = c(metric_x_min, metric_x_max)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey30")
  ) +
  labs(
    title = paste0("Focused Elastic Net Metric Summary on ", external_test_ds),
    subtitle = paste0("Only key metrics are shown (", metric_boot_ci_label, ", B = ", B, ")"),
    x = "Metric value",
    y = NULL
  )

ggplot(metric_focus_long, aes(x = value, y = metric_label, fill = metric_label)) +
  geom_violin(alpha = 0.88, trim = FALSE, color = NA) +
  geom_boxplot(
    width = 0.14,
    outlier.shape = NA,
    fill = "white",
    color = "grey20",
    linewidth = 0.5
  ) +
  geom_point(
    data = metric_focus_summary,
    aes(x = point_estimate, y = metric_label),
    inherit.aes = FALSE,
    shape = 21,
    size = 3.2,
    stroke = 0.9,
    fill = "white",
    color = "grey10"
  ) +
  scale_fill_manual(values = metric_focus_pal) +
  coord_cartesian(xlim = c(metric_x_min, metric_x_max)) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30")
  ) +
  labs(
    title = paste0("Bootstrap Distributions of Key Elastic Net Metrics on ", external_test_ds),
    subtitle = "White dots show the observed metric on the full external dataset",
    x = "Metric value",
    y = NULL
  )
```

```{r}
# Additional visualisation: forest-style AUC summary across held-out datasets
# This block only summarises the existing LODO predictions in lodo_pred.
# It does not refit or rerun any classifier.
auc_lodo_by_dataset <- bind_rows(lapply(sort(unique(as.character(lodo_pred$dataset))), function(test_ds) {
  fold_df <- lodo_pred %>% filter(dataset == test_ds)
  
  bind_rows(lapply(model_cols, function(m) {
    data.frame(
      dataset = test_ds,
      model = m,
      auc = as.numeric(
        pROC::auc(
          pROC::roc(
            response = fold_df$truth,
            predictor = fold_df[[m]],
            levels = c("Normal", "Cancer"),
            direction = "<",
            quiet = TRUE
          )
        )
      )
    )
  }))
}))

auc_lodo_forest <- auc_lodo_by_dataset %>%
  group_by(model) %>%
  summarise(
    median_auc = median(auc, na.rm = TRUE),
    q1_auc = as.numeric(stats::quantile(auc, 0.25, na.rm = TRUE)),
    q3_auc = as.numeric(stats::quantile(auc, 0.75, na.rm = TRUE)),
    min_auc = min(auc, na.rm = TRUE),
    max_auc = max(auc, na.rm = TRUE),
    .groups = "drop"
  )

auc_lodo_forest <- bind_rows(
  auc_lodo_forest %>% filter(model == "ElasticNet"),
  auc_lodo_forest %>% filter(model != "ElasticNet") %>% arrange(desc(median_auc))
) %>%
  mutate(
    display_label = ifelse(model == "ElasticNet", "ElasticNet (selected)", model),
    display_label = factor(display_label, levels = rev(display_label)),
    highlight = ifelse(model == "ElasticNet", "Selected classifier", "Alternative"),
    selected_label = ifelse(model == "ElasticNet", sprintf("%.3f", median_auc), NA_character_)
  )

forest_x_min <- max(0, min(auc_lodo_forest$min_auc, na.rm = TRUE) - 0.03)
forest_x_max <- min(1.06, max(auc_lodo_forest$max_auc, na.rm = TRUE) + 0.08)

auc_lodo_forest <- auc_lodo_forest %>%
  mutate(
    label_x = ifelse(
      model == "ElasticNet",
      pmin(max_auc + 0.006, forest_x_max - 0.018),
      NA_real_
    )
  )

ggplot(auc_lodo_forest, aes(x = median_auc, y = display_label)) +
  geom_segment(
    data = auc_lodo_forest %>% filter(model != "ElasticNet"),
    aes(x = min_auc, xend = max_auc, yend = display_label),
    linewidth = 1.2,
    color = "#C9C9C9"
  ) +
  geom_segment(
    data = auc_lodo_forest %>% filter(model == "ElasticNet"),
    aes(x = min_auc, xend = max_auc, yend = display_label),
    linewidth = 1.2,
    color = "#F28E2B",
    alpha = 0.65
  ) +
  geom_segment(
    aes(x = q1_auc, xend = q3_auc, yend = display_label, color = highlight),
    linewidth = 3.2,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(
    aes(fill = highlight, color = highlight),
    shape = 21,
    size = 4.3,
    stroke = 1.1,
    show.legend = FALSE
  ) +
  geom_text(
    data = auc_lodo_forest %>% filter(model == "ElasticNet"),
    inherit.aes = FALSE,
    aes(x = label_x, y = display_label, label = selected_label, color = highlight),
    hjust = 0,
    size = 2.8,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = c("Selected classifier" = "#F28E2B", "Alternative" = "#9E9E9E")) +
  scale_fill_manual(values = c("Selected classifier" = "#F28E2B", "Alternative" = "#9E9E9E")) +
  scale_x_continuous(
    limits = c(forest_x_min, forest_x_max),
    breaks = seq(0.50, 1.00, by = 0.10)
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 18, 8, 5.5),
    plot.title = element_text(face = "bold", size = 15, lineheight = 0.95),
    plot.subtitle = element_text(color = "grey30", size = 9.5, lineheight = 1.0)
  ) +
  labs(
    title = "Classifier Performance\nAcross Held-out Datasets",
    subtitle = "Existing LODO predictions only\nPoint = median AUC; thick bar = IQR; thin bar = range",
    x = "Discrimination (AUC) - higher is better",
    y = NULL
  )
```

```{r}
# Additional experiment: single Elastic Net with different LIMMA top-gene counts
library(glmnet)
library(pROC)
library(caret)
library(dplyr)
library(tidyr)
library(ggplot2)

top_gene_grid <- c(5, 10, 20, 50, 100, 200)

top_gene_elastic_results <- bind_rows(lapply(top_gene_grid, function(top_n_use) {
  top_genes_use <- rownames(top_table)[seq_len(min(top_n_use, nrow(top_table)))]
  
  X_train_top <- t(expr[top_genes_use, , drop = FALSE])
  X_test_top <- t(expr_external[top_genes_use, , drop = FALSE])
  
  set.seed(seed_from_label("full_elasticnet", 5000L))
  fit_top_elastic <- cv.glmnet(
    x = as.matrix(X_train_top),
    y = y_full_num,
    family = "binomial",
    alpha = 0.5,
    type.measure = "auc",
    foldid = full_cv_spec$foldid,
    nfolds = full_cv_spec$nfolds
  )
  
  top_elastic_pred <- as.numeric(
    predict(
      fit_top_elastic,
      newx = as.matrix(X_test_top),
      s = "lambda.min",
      type = "response"
    )
  )
  
  top_elastic_roc <- pROC::roc(
    response = y_external,
    predictor = top_elastic_pred,
    levels = c("Normal", "Cancer"),
    direction = "<",
    quiet = TRUE
  )
  
  top_elastic_auc <- as.numeric(pROC::auc(top_elastic_roc))
  
  top_elastic_best <- pROC::coords(
    top_elastic_roc,
    x = "best",
    best.method = "youden",
    ret = c("threshold", "sensitivity", "specificity"),
    transpose = FALSE
  )
  
  top_elastic_best <- as.data.frame(top_elastic_best)
  top_elastic_best <- top_elastic_best[1, , drop = FALSE]
  
  top_best_threshold <- as.numeric(top_elastic_best$threshold)
  top_best_sensitivity <- as.numeric(top_elastic_best$sensitivity)
  top_best_specificity <- as.numeric(top_elastic_best$specificity)
  
  top_elastic_class_best <- factor(
    ifelse(top_elastic_pred >= top_best_threshold, "Cancer", "Normal"),
    levels = c("Normal", "Cancer")
  )
  
  top_elastic_cm_best <- caret::confusionMatrix(
    data = top_elastic_class_best,
    reference = y_external,
    positive = "Cancer"
  )
  
  top_elastic_class_05 <- factor(
    ifelse(top_elastic_pred >= 0.5, "Cancer", "Normal"),
    levels = c("Normal", "Cancer")
  )
  
  top_elastic_cm_05 <- caret::confusionMatrix(
    data = top_elastic_class_05,
    reference = y_external,
    positive = "Cancer"
  )
  
  data.frame(
    TopGenes = top_n_use,
    AUC = top_elastic_auc,
    Threshold05 = 0.5,
    BalancedAccuracy = as.numeric(top_elastic_cm_05$byClass["Balanced Accuracy"]),
    Sensitivity05 = as.numeric(top_elastic_cm_05$byClass["Sensitivity"]),
    Specificity05 = as.numeric(top_elastic_cm_05$byClass["Specificity"]),
    BestThreshold = top_best_threshold,
    BalancedAccuracyBestThreshold = as.numeric(
      top_elastic_cm_best$byClass["Balanced Accuracy"]
    ),
    SensitivityBestThreshold = top_best_sensitivity,
    SpecificityBestThreshold = top_best_specificity
  )
}))

top_gene_elastic_results %>%
  select(
    TopGenes,
    AUC,
    BalancedAccuracy,
    Sensitivity05,
    Specificity05,
    BestThreshold,
    BalancedAccuracyBestThreshold
  )

top_gene_threshold_diagnostic <- top_gene_elastic_results %>%
  mutate(
    BalancedAccuracyGap = BalancedAccuracyBestThreshold - BalancedAccuracy
  ) %>%
  select(
    TopGenes,
    AUC,
    BalancedAccuracy05 = BalancedAccuracy,
    BestThreshold,
    BalancedAccuracyBestThreshold,
    BalancedAccuracyGap
  )

top_gene_threshold_diagnostic

top50_check <- top_gene_elastic_results %>%
  filter(TopGenes == length(final_top_genes)) %>%
  transmute(
    TopGenes,
    OriginalAUC = as.numeric(enet_results$AUC),
    RefitAUC = AUC,
    OriginalBalancedAccuracy = as.numeric(enet_results$BalancedAccuracy),
    RefitBalancedAccuracy = BalancedAccuracy
  )

top50_check

top50_refit_genes <- rownames(top_table)[seq_len(length(final_top_genes))]
X_train_top50_refit <- t(expr[top50_refit_genes, , drop = FALSE])
X_test_top50_refit <- t(expr_external[top50_refit_genes, , drop = FALSE])

set.seed(seed_from_label("full_elasticnet", 5000L))
fit_top50_refit <- cv.glmnet(
  x = as.matrix(X_train_top50_refit),
  y = y_full_num,
  family = "binomial",
  alpha = 0.5,
  type.measure = "auc",
  foldid = full_cv_spec$foldid,
  nfolds = full_cv_spec$nfolds
)

top50_refit_pred <- as.numeric(
  predict(
    fit_top50_refit,
    newx = as.matrix(X_test_top50_refit),
    s = "lambda.min",
    type = "response"
  )
)

top50_refit_class <- factor(
  ifelse(top50_refit_pred >= 0.5, "Cancer", "Normal"),
  levels = c("Normal", "Cancer")
)

top50_refit_cm <- caret::confusionMatrix(
  data = top50_refit_class,
  reference = y_external,
  positive = "Cancer"
)

top50_refit_auc <- as.numeric(
  pROC::auc(
    pROC::roc(
      response = y_external,
      predictor = top50_refit_pred,
      levels = c("Normal", "Cancer"),
      direction = "<",
      quiet = TRUE
    )
  )
)

top50_refit_prediction_process_check <- data.frame(
  SameGenesAsOriginal = identical(
    as.character(top50_refit_genes),
    as.character(final_top_genes)
  ),
  MaxAbsPredictionDifference = max(abs(top50_refit_pred - enet_pred)),
  MeanAbsPredictionDifference = mean(abs(top50_refit_pred - enet_pred)),
  OriginalAUC = as.numeric(enet_results$AUC),
  RefitAUC = top50_refit_auc,
  OriginalBalancedAccuracy = as.numeric(enet_results$BalancedAccuracy),
  RefitBalancedAccuracy = as.numeric(top50_refit_cm$byClass["Balanced Accuracy"])
)

top50_refit_prediction_process_check

top_gene_elastic_long <- top_gene_elastic_results %>%
  pivot_longer(
    cols = c(AUC, BalancedAccuracy),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    TopGenesLabel = factor(
      paste0("Top ", TopGenes),
      levels = paste0("Top ", top_gene_grid)
    ),
    Metric = recode(
      Metric,
      AUC = "AUC",
      BalancedAccuracy = "Balanced accuracy (cutoff = 0.5)"
    )
  )

ggplot(
  top_gene_elastic_long,
  aes(x = TopGenesLabel, y = Value, color = Metric, group = Metric)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.2) +
  geom_text(
    aes(label = sprintf("%.3f", Value)),
    vjust = -0.8,
    size = 3.4,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_y_continuous(
    limits = c(
      max(0, min(top_gene_elastic_long$Value, na.rm = TRUE) - 0.06),
      min(1.04, max(top_gene_elastic_long$Value, na.rm = TRUE) + 0.06)
    )
  ) +
  scale_color_manual(
    values = c(
      "AUC" = "#D55E00",
      "Balanced accuracy (cutoff = 0.5)" = "#4C78A8"
    )
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30", size = 11)
  ) +
  labs(
    title = "Single Elastic Net Performance by LIMMA Top Genes",
    subtitle = paste0("External test dataset: ", external_test_ds),
    x = "Number of LIMMA-selected top genes",
    y = "Metric value"
  )

top_gene_threshold_long <- top_gene_elastic_results %>%
  transmute(
    TopGenesLabel = factor(
      paste0("Top ", TopGenes),
      levels = paste0("Top ", top_gene_grid)
    ),
    `0.5 cutoff` = BalancedAccuracy,
    `ROC/Youden cutoff` = BalancedAccuracyBestThreshold
  ) %>%
  pivot_longer(
    cols = c(`0.5 cutoff`, `ROC/Youden cutoff`),
    names_to = "ThresholdMethod",
    values_to = "BalancedAccuracy"
  )

ggplot(
  top_gene_threshold_long,
  aes(
    x = TopGenesLabel,
    y = BalancedAccuracy,
    color = ThresholdMethod,
    group = ThresholdMethod
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.2) +
  geom_text(
    aes(label = sprintf("%.3f", BalancedAccuracy)),
    vjust = -0.8,
    size = 3.2,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_y_continuous(limits = c(0.45, 1.03)) +
  scale_color_manual(
    values = c("0.5 cutoff" = "#4C78A8", "ROC/Youden cutoff" = "#F28E2B")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey30", size = 10.5)
  ) +
  labs(
    title = "Balanced Accuracy Is Sensitive to the Classification Cutoff",
    subtitle = "ROC/Youden cutoff is shown as a diagnostic; fixed 0.5 cutoff matches the original reporting rule",
    x = "Number of LIMMA-selected top genes",
    y = "Balanced accuracy"
  )
```

# Bootstrap metric uncertainty across LIMMA top-gene sets
```{r}
#| fig-width: 9
#| fig-height: 5.8
# Additional bootstrap metric uncertainty plots for Top 10/20/50/100/200 genes.
# Each Elastic Net model is fitted once, then the external test set is
# class-stratified bootstrap-resampled to estimate metric uncertainty.
library(glmnet)
library(pROC)
library(caret)
library(dplyr)
library(ggplot2)

top_gene_boot_grid <- c(10, 20, 50, 100, 200)
top_gene_boot_B <- 100
top_gene_boot_ci_level <- 0.95
top_gene_boot_alpha <- 1 - top_gene_boot_ci_level
top_gene_boot_ci_label <- paste0(
  round(top_gene_boot_ci_level * 100),
  "% percentile interval"
)

top_gene_boot_predictions <- lapply(top_gene_boot_grid, function(top_n_use) {
  top_genes_use <- rownames(top_table)[seq_len(min(top_n_use, nrow(top_table)))]
  
  X_train_top <- t(expr[top_genes_use, , drop = FALSE])
  X_test_top <- t(expr_external[top_genes_use, , drop = FALSE])
  
  set.seed(seed_from_label("full_elasticnet", 5000L))
  fit_top_elastic <- cv.glmnet(
    x = as.matrix(X_train_top),
    y = y_full_num,
    family = "binomial",
    alpha = 0.5,
    type.measure = "auc",
    foldid = full_cv_spec$foldid,
    nfolds = full_cv_spec$nfolds
  )
  
  top_elastic_pred <- as.numeric(
    predict(
      fit_top_elastic,
      newx = as.matrix(X_test_top),
      s = "lambda.min",
      type = "response"
    )
  )
  
  data.frame(
    TopGenes = top_n_use,
    sample_id = rownames(X_test_top),
    truth = y_external,
    score = top_elastic_pred
  )
})

top_gene_boot_predictions <- bind_rows(top_gene_boot_predictions)

top_gene_metric_boot <- bind_rows(lapply(top_gene_boot_grid, function(top_n_use) {
  pred_df <- top_gene_boot_predictions %>%
    filter(TopGenes == top_n_use)
  
  boot_metrics <- do.call(
    rbind,
    lapply(seq_len(top_gene_boot_B), function(b) {
      set.seed(seed_from_label(
        paste0("top_gene_external_metric_bootstrap_", top_n_use, "_", b),
        12000L
      ))
      
      boot_idx <- make_external_metric_boot_idx(pred_df$truth)
      
      calc_binary_metric_set(
        truth = pred_df$truth[boot_idx],
        score = pred_df$score[boot_idx],
        threshold = 0.5
      )
    })
  )
  
  boot_metrics <- as.data.frame(boot_metrics)
  boot_metrics$TopGenes <- top_n_use
  boot_metrics
}))

top_gene_metric_points <- bind_rows(lapply(top_gene_boot_grid, function(top_n_use) {
  pred_df <- top_gene_boot_predictions %>%
    filter(TopGenes == top_n_use)
  
  point_metrics <- calc_binary_metric_set(
    truth = pred_df$truth,
    score = pred_df$score,
    threshold = 0.5
  )
  
  data.frame(
    TopGenes = top_n_use,
    metric = names(point_metrics),
    point_estimate = as.numeric(point_metrics)
  )
}))

top_gene_metric_summary <- bind_rows(lapply(top_gene_boot_grid, function(top_n_use) {
  bind_rows(lapply(metric_display_order, function(metric_name) {
    metric_values <- top_gene_metric_boot %>%
      filter(TopGenes == top_n_use) %>%
      pull(metric_name)
    
    point_value <- top_gene_metric_points %>%
      filter(TopGenes == top_n_use, metric == metric_name) %>%
      pull(point_estimate)
    
    data.frame(
      TopGenes = top_n_use,
      TopGenesLabel = paste0("Top ", top_n_use),
      metric = metric_name,
      metric_label = unname(metric_labels[metric_name]),
      point_estimate = point_value,
      bootstrap_mean = mean(metric_values, na.rm = TRUE),
      bootstrap_sd = stats::sd(metric_values, na.rm = TRUE),
      ci_lower = as.numeric(stats::quantile(
        metric_values,
        probs = top_gene_boot_alpha / 2,
        na.rm = TRUE
      )),
      ci_upper = as.numeric(stats::quantile(
        metric_values,
        probs = 1 - top_gene_boot_alpha / 2,
        na.rm = TRUE
      ))
    )
  }))
})) %>%
  mutate(
    TopGenesLabel = factor(
      TopGenesLabel,
      levels = paste0("Top ", top_gene_boot_grid)
    ),
    metric_label = factor(
      metric_label,
      levels = unname(metric_labels[metric_display_order])
    )
  )

top_gene_metric_summary

for (top_label_use in levels(top_gene_metric_summary$TopGenesLabel)) {
  plot_df <- top_gene_metric_summary %>%
    filter(TopGenesLabel == top_label_use)
  
  p <- ggplot(plot_df, aes(x = metric_label, y = point_estimate)) +
    geom_linerange(
      aes(ymin = ci_lower, ymax = ci_upper),
      linewidth = 1.05,
      color = "#4C78A8"
    ) +
    geom_point(size = 2.7, color = "#E45756") +
    coord_flip() +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.margin = margin(8, 18, 8, 8),
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey30", size = 10)
    ) +
    labs(
      title = paste0(
        "Elastic Net Metric Uncertainty on ",
        external_test_ds,
        " - ",
        top_label_use
      ),
      x = NULL,
      y = "Metric value"
    )
  
  print(p)
}
```


## Confusion matrix (single best model: Elastic Net)

```{r confusion-matrix-elastic-external}
#| message: false
#| warning: false
elastic_external_pred <- factor(
  ifelse(enet_pred >= 0.5, "Cancer", "Normal"),
  levels = c("Normal", "Cancer")
)

elastic_external_cm <- caret::confusionMatrix(
  data = elastic_external_pred,
  reference = y_external,
  positive = "Cancer"
)

elastic_external_cm

elastic_external_cm_df <- as.data.frame(
  table(
    Truth = y_external,
    Predicted = elastic_external_pred
  )
) %>%
  dplyr::group_by(Truth) %>%
  dplyr::mutate(RowPct = Freq / sum(Freq)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Label = paste0(Freq, "\n", sprintf("%.1f%%", 100 * RowPct))
  )

ggplot(elastic_external_cm_df, aes(x = Predicted, y = Truth, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = Label), size = 4.2, fontface = "bold") +
  scale_fill_gradient(low = "#F7FBFF", high = "#2C7FB8") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30")
  ) +
  labs(
    title = paste0("Confusion Matrix: Elastic Net on ", external_test_ds),
    x = "Predicted label",
    y = "True label",
    fill = "Count"
  )
```

## LODO model summary with AUC, balanced accuracy, and sensitivity

```{r lodo-model-metric-summary}
#| message: false
#| warning: false
library(pROC)
library(caret)
library(dplyr)

lodo_truth_eval <- factor(lodo_pred$truth, levels = c("Normal", "Cancer"))

lodo_metric_tbl <- bind_rows(lapply(model_cols, function(m) {
  pred_prob <- lodo_pred[[m]]
  pred_class <- factor(
    ifelse(pred_prob >= 0.5, "Cancer", "Normal"),
    levels = c("Normal", "Cancer")
  )
  
  roc_obj <- pROC::roc(
    response = lodo_truth_eval,
    predictor = pred_prob,
    levels = c("Normal", "Cancer"),
    direction = "<",
    quiet = TRUE
  )
  
  cm_obj <- caret::confusionMatrix(
    data = pred_class,
    reference = lodo_truth_eval,
    positive = "Cancer"
  )
  
  data.frame(
    model = m,
    AUC = as.numeric(pROC::auc(roc_obj)),
    BalancedAccuracy = as.numeric(cm_obj$byClass["Balanced Accuracy"]),
    Sensitivity = as.numeric(cm_obj$byClass["Sensitivity"])
  )
}))

lodo_metric_tbl %>%
  arrange(desc(AUC))
```

## Coefficient formulas for Ridge, LASSO, Logistic Regression, and Elastic Net

```{r coefficient-formulas-four-models}
#| message: false
#| warning: false
library(glmnet)
library(dplyr)

# Refit the four gene-level linear models on the same final gene set used above.
set.seed(seed_from_label("full_ridge_formula", 15001L))
fit_ridge_formula <- cv.glmnet(
  x = as.matrix(X_full),
  y = y_full_num,
  family = "binomial",
  alpha = 0,
  type.measure = "auc",
  foldid = full_cv_spec$foldid,
  nfolds = full_cv_spec$nfolds
)

set.seed(seed_from_label("full_lasso_formula", 15002L))
fit_lasso_formula <- cv.glmnet(
  x = as.matrix(X_full),
  y = y_full_num,
  family = "binomial",
  alpha = 1,
  type.measure = "auc",
  foldid = full_cv_spec$foldid,
  nfolds = full_cv_spec$nfolds
)

fit_elastic_formula <- fit_enet_full

logistic_formula_df <- data.frame(
  y_full = y_full,
  X_full,
  check.names = FALSE
)

fit_logistic_formula <- glm(
  y_full ~ .,
  data = logistic_formula_df,
  family = binomial()
)

extract_coef_tbl <- function(model_name, coef_obj) {
  coef_vec <- if (is.matrix(coef_obj)) {
    stats::setNames(as.numeric(coef_obj[, 1]), rownames(coef_obj))
  } else {
    coef_obj
  }
  
  data.frame(
    model = model_name,
    term = names(coef_vec),
    coefficient = as.numeric(coef_vec),
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      term = ifelse(term == "(Intercept)", "Intercept", term),
      abs_coefficient = abs(coefficient)
    )
}

coef_formula_tbl <- bind_rows(
  extract_coef_tbl(
    "Ridge",
    as.matrix(coef(fit_ridge_formula, s = "lambda.min"))
  ),
  extract_coef_tbl(
    "LASSO",
    as.matrix(coef(fit_lasso_formula, s = "lambda.min"))
  ),
  extract_coef_tbl(
    "ElasticNet",
    as.matrix(coef(fit_elastic_formula, s = "lambda.min"))
  ),
  extract_coef_tbl(
    "LogisticRegression",
    stats::coef(fit_logistic_formula)
  )
) %>%
  filter(term == "Intercept" | coefficient != 0)

coef_gene_tbl <- coef_formula_tbl %>%
  filter(term != "Intercept") %>%
  arrange(model, desc(abs_coefficient), term) %>%
  select(model, gene = term, coefficient)

coef_gene_tbl

build_formula_string <- function(one_model_tbl) {
  intercept_value <- one_model_tbl %>%
    filter(term == "Intercept") %>%
    pull(coefficient)
  
  intercept_value <- if (length(intercept_value) == 0) 0 else intercept_value[1]
  
  term_tbl <- one_model_tbl %>%
    filter(term != "Intercept") %>%
    arrange(desc(abs_coefficient), term)
  
  if (nrow(term_tbl) == 0) {
    return(paste0("Y = ", sprintf("%.6f", intercept_value)))
  }
  
  term_strings <- paste0(
    ifelse(term_tbl$coefficient >= 0, " + ", " - "),
    sprintf("%.6f", abs(term_tbl$coefficient)),
    "*",
    term_tbl$term
  )
  
  paste0(
    "Y = ",
    sprintf("%.6f", intercept_value),
    paste(term_strings, collapse = "")
  )
}

formula_tbl <- coef_formula_tbl %>%
  group_by(model) %>%
  group_split() %>%
  lapply(function(one_model_tbl) {
    data.frame(
      model = one_model_tbl$model[1],
      formula = build_formula_string(one_model_tbl),
      stringsAsFactors = FALSE
    )
  }) %>%
  bind_rows() %>%
  mutate(
    model = factor(
      model,
      levels = c("Ridge", "LASSO", "ElasticNet", "LogisticRegression")
    )
  ) %>%
  arrange(model)

formula_tbl

cat("\nThese are binomial models, so the linear predictor is:\n")
for (i in seq_len(nrow(formula_tbl))) {
  cat("\n", as.character(formula_tbl$model[i]), ":\n", sep = "")
  cat(
    paste(strwrap(formula_tbl$formula[i], width = 110), collapse = "\n"),
    "\n",
    sep = ""
  )
}
cat("\nP(Cancer) = 1 / (1 + exp(-Y))\n")
```

## Coefficient visualisation for the four linear models

```{r coefficient-visualisation-four-models}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)

coef_plot_tbl <- coef_formula_tbl %>%
  filter(term != "Intercept") %>%
  mutate(
    direction = ifelse(coefficient >= 0, "Positive", "Negative")
  )

# Heatmap: show the genes with the largest coefficient magnitude in any model.
top_heatmap_genes <- coef_plot_tbl %>%
  group_by(term) %>%
  summarise(max_abs_coef = max(abs(coefficient), na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(max_abs_coef)) %>%
  slice_head(n = 25) %>%
  pull(term)

coef_heatmap_tbl <- coef_plot_tbl %>%
  filter(term %in% top_heatmap_genes) %>%
  group_by(term) %>%
  mutate(gene_order_value = max(abs(coefficient), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    term = factor(
      term,
      levels = rev(
        coef_plot_tbl %>%
          filter(term %in% top_heatmap_genes) %>%
          group_by(term) %>%
          summarise(max_abs_coef = max(abs(coefficient), na.rm = TRUE), .groups = "drop") %>%
          arrange(max_abs_coef) %>%
          pull(term)
      )
    ),
    model = factor(
      model,
      levels = c("Ridge", "LASSO", "ElasticNet", "LogisticRegression")
    )
  )

ggplot(coef_heatmap_tbl, aes(x = model, y = term, fill = coefficient)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "white",
    high = "#D7191C",
    midpoint = 0
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30")
  ) +
  labs(
    title = "Coefficient Heatmap Across Four Models",
    subtitle = "Top 25 genes by maximum absolute coefficient across Ridge, LASSO, Elastic Net, and Logistic Regression",
    fill = "Coefficient"
  )

# Faceted bar plot: top genes within each model.
coef_top_by_model_tbl <- coef_plot_tbl %>%
  group_by(model) %>%
  arrange(desc(abs(coefficient)), .by_group = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(
    gene_model_label = paste(model, term, sep = "___")
  ) %>%
  ungroup() %>%
  arrange(model, coefficient) %>%
  mutate(
    gene_model_label = factor(gene_model_label, levels = unique(gene_model_label))
  )

ggplot(
  coef_top_by_model_tbl,
  aes(x = coefficient, y = gene_model_label, fill = direction)
) +
  geom_col(width = 0.72) +
  facet_wrap(~ model, scales = "free_y", ncol = 2) +
  scale_y_discrete(labels = function(x) sub("^.*___", "", x)) +
  scale_fill_manual(values = c("Positive" = "#D55E00", "Negative" = "#0072B2")) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey30")
  ) +
  labs(
    title = "Top Gene Coefficients Within Each Model",
    subtitle = "Largest absolute coefficients; warm colors increase cancer score, cool colors decrease it",
    x = "Coefficient",
    y = "Gene"
  )
```

## Faceted boxplot for Top 10/20/50/100/200 metric uncertainty

```{r top-gene-metric-facet-boxplot}
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(ggplot2)

top_gene_metric_focus <- c(
  AUC = "AUC",
  BalancedAccuracy = "Balanced accuracy",
  Sensitivity = "Sensitivity",
  Specificity = "Specificity"
)

top_gene_metric_boot_long <- top_gene_metric_boot %>%
  select(TopGenes, all_of(names(top_gene_metric_focus))) %>%
  pivot_longer(
    cols = all_of(names(top_gene_metric_focus)),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    TopGenesLabel = factor(
      paste0("Top ", TopGenes),
      levels = paste0("Top ", c(10, 20, 50, 100, 200))
    ),
    metric_label = factor(
      unname(top_gene_metric_focus[metric]),
      levels = c("Sensitivity", "Specificity", "Balanced accuracy", "AUC")
    )
  )

top_gene_metric_point_focus <- top_gene_metric_summary %>%
  filter(metric %in% names(top_gene_metric_focus)) %>%
  mutate(
    metric_label = factor(
      unname(top_gene_metric_focus[metric]),
      levels = c("Sensitivity", "Specificity", "Balanced accuracy", "AUC")
    ),
    TopGenesLabel = factor(
      TopGenesLabel,
      levels = paste0("Top ", c(10, 20, 50, 100, 200))
    )
  )

ggplot(
  top_gene_metric_boot_long,
  aes(x = TopGenesLabel, y = value, fill = TopGenesLabel)
) +
  geom_jitter(
    width = 0.10,
    height = 0,
    alpha = 0.12,
    size = 0.8,
    color = "grey35",
    show.legend = FALSE
  ) +
  geom_boxplot(
    width = 0.62,
    alpha = 0.88,
    outlier.alpha = 0.35,
    outlier.size = 1.2,
    linewidth = 0.65,
    color = "grey25"
  ) +
  geom_point(
    data = top_gene_metric_point_focus,
    aes(x = TopGenesLabel, y = point_estimate),
    inherit.aes = FALSE,
    color = "#E45756",
    size = 2.8
  ) +
  facet_wrap(~ metric_label, ncol = 2, scales = "free_y") +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.12))) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 15)
  ) +
  labs(
    title = paste0("Uncertainty Across Top-Gene Sets on ", external_test_ds),
    x = "Number of top genes",
    y = "Metric value"
  )
```











